### 2019-12-04 README.Rmd
set up repo structure with rrtools

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# QRFpaper

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/KevinSee/QRFpaper/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: <https://github.com/KevinSee/QRFpaper/archive/master.zip>

Or you can install this compendium as an R package, QRFpaper, from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("KevinSee/QRFpaper")
```

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2019-12-09 analysis/paper/paper.Rmd
added text to manuscript from previous Word version.

---
title: "Estimating Life-Stage Specific Habitat Capacity for Anadromous Salmonids using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@biomark.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@biomark.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richie.Carmichael@biomark.com
      institute: [biomark]
      correspondence: false
  - Sarah Hoffman:
      email: Sarah.Hoffman@biomark.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@biomark.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
# bibliography: references.bib
bibliography:
- /Users/kevin/Dropbox/Bibliography/Research.bib
- /Users/kevin/Dropbox/Bibliography/SoftwareCitations.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r packages}
library(tidyverse)
library(QRFcapacity) # for the moment, until we know exactly what data to copy over to this repo
library(janitor)
library(knitr)

# setwd('analysis/paper')
```

```{r package-bibtex}
knitr::write_bib(c("FSA", "Rcapture", "minerva", "quantregForest", "sf", "knitr", "rmarkdown"),
                 file = 'packages.bib')
```

```{r read-data}
# fish/habitat data for fitting QRF models
# data("fh_sum_champ_2017")
# fh_data = fh_sum_champ_2017 %>%
#   filter(Species == 'Chinook') %>%
#   mutate_at(vars(Watershed, Species, FishCrew),
#             list(fct_drop))

data("fh_sum_champ_2014")
fh_data = fh_sum_champ_2014 %>%
  filter(Species == 'Chinook') %>%
  mutate_at(vars(Watershed, Species, FishCrew),
            list(fct_drop))

# habitat data used for predicting capacity
data("champ_site_2011_17")
hab_data = champ_site_2011_17 %>%
  filter(!Watershed %in% c('Big-Navarro-Garcia (CA)',
                           'CHaMP Training',
                           'Region 17',
                           'Walla Walla',
                           'Umatilla'))

data("champ_site_2011_17_avg")
hab_avg = champ_site_2011_17_avg

data("hab_dict_2017")
hab_dict = hab_dict_2017

```

```{r load-qrf-model}
load('../data/derived_data/qrf_juv_summer.rda')
qrf_mod = qrf_mods[[1]]
qrf_data = qrf_mod_df %>%
  filter(Species == 'Chinook')
```

# Introduction

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes might most cost-effectively and sustainably reduce mortality. Recently released “de-listing” criteria (NOAA 2016 ***need reference details***) identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required. These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured. Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. For perhaps the first time, the necessity of tributary habitat restoration can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”

Pacific salmon (*Oncorhynchus spp.*) species have experienced large declines in abundance throughout much of their range [@Good2005]. Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*Oncorhynchus mykiss*) in the Pacific Northwest, USA. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity, both historic and contemporary, for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge. Reliable methods to better understand fish-habitat relationships as well as to estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity, and further, to better direct tributary rehabilitation efforts.

@Fausch1988 conducted a thorough review of attempts to predict the standing crop of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for some salmonid species. @Nickelson1992 found that juvenile coho were found in higher densities within pool habitat on the Oregon coast. @Sharma2001 determined that pool and pond densities were good predictors of coho smolt densities in western Washington. @Pess2002 demonstrated that densities of adult spawning coho were higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed. @Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. @Bryant2009 found a positive relationship between juvenile coho abundance and large wood at the reach scale, and a negative one between coho and the number of pools. @Braun2011 found positive associations between spawner densities of sockeye in the Fraser River and percent undercuts, large wood and percent pools. However, all of these studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship, which often results in weak predictive power.

There have been a number of studies that utilized other modeling approaches to elicit fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density. @Eastwood2003 also utilized a quantile regression model to map the potential extent of sole in the English Channel and southern North Sea. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. The results from these studies highlight the importance of using techniques that can accommodate non-linear fish habitat relationships.

Most studies that have investigated fish habitat relationships have focused on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity. Also, many of these studies focus on only one or two measures of habitat. @Sweka2010 did estimate carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves. However, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009], and it often fails to account for habitat covariates. This manuscript incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.

In fisheries it has long been recognized that that biotic and abiotic factors limit productivity within and across life-stages. For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile regression forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

The fish abundance/density and habitat data used to fit the QRF model presented here were available from nine watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA. Within the interior CRB two major runs of Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may only spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia Spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River Spring/Summer-run is listed as threatened (***need citation***).

In this study, we developed a QRF model to 1) elicit fish-habitat relationships and 2) predict habitat rearing capacity at the reach scale (200 – 500 m) for juvenile (summer parr) stream-type Chinook salmon based on paired fish abundance/density and habitat data. Importantly, the QRF model places no constraints on possible fish-habitat relationships; instead, relationships are estimated from the data regardless of being positive, negative, linear, non-linear, etc. Based on observed fish-habitat relationships we can then predict capacity at the site scale using measurements of the habitat characteristics that were used to populate the model. We then extrapolated to larger spatial scales and validated our estimates of capacity at the watershed scale with those from spawner recruit curves from a number of watersheds around the interior Columbia basin. Moreover, the model provides managers a framework to guide the identification, prioritization, and development of habitat enhancement actions to recover salmon populations.

# Methods

## Study Site

Fish and habitat data used in our study were collected from nine watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (Figure 1). The CRB covers greater than 668,000 km$^2$ and drains large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (@CHaMP2016; @Volk2017] and were downloaded from https://www.champmonitoring.org. Data from the following `r n_distinct(hab_data$Watershed)` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`. Juvenile density and abundance data collected at CHaMP survey reaches were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].

## Data

CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin (CRB) selected based on a spatially balanced Generalized Random Tesselation Stratified (GRTS) sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites are collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, size (depth, width, discharge), substrate, temperature, and water quality.

Juvenile fish surveys were conducted for sp/sum Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, and single-pass electrofishing, as well as snorkeling. These data were used to estimate sp/sum Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013. Two-pass removal estimates used the estimator described by @Seber2002. Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964. These estimates were made using the removal function from the FSA package [@R-FSA] or the closedp.bc function from the Rcapture package [@R-Rcapture] in R software [@Rsoftware2018]. Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew / watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

Abundance estimates at all sites were then translated into linear (parr/m) fish densities and density estimates were paired with the associated CHaMP habitat data. For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity. After removing duplicate samples, our initial dataset contained `r n_distinct(fh_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fh-samp-table)).

## Habitat Covariate Selection

A key step in developing a QRF model to predict fish capacities is selecting the habitat covariates to include in the model. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another. However, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).

The CHaMP protocol produces more than 100 metrics describing the quantity and quality of fish habitat for each survey site. To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities. We calculated the maximal information coefficient (MIC), using the R package minerva [@Albanese2013], to measure the strength of the linear or non-linear association between two variables [@Reshef2011]. The MIC value between each of the measured habitat characteristics and the response variable, parr density (fish/m), was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.

Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring volume and area were scaled to the wetted area of each site. Within each category, metrics were ranked according to their MIC value (Figure 1). Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.). Additionally, we attempted to avoid covariates that were highly correlated (Figure 2) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.

## QRF Model Fit

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months. QRF models combine the flexibility of random forest models (Breiman 2001) with the ability of quantile regression to extract relationships between quantiles of the data other than the mean (Cade and Noon 2003). Random forests also account for non-linear relationships between the response and predictor variables, and naturally incorporate interactions between the predictor variables, two common features of ecological datasets (Liaw and Wiener 2002). After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response. Therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. We fit the QRF models using the quantregForest function from the quantregForest package (Meinshausen 2016) in R software (R Core Team 2015). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of Sweka and Mackey (2010). One reason for the 90th quantile, instead of something higher, is to avoid using predictions that are aimed at the very upper tails of observed fish density, where the variability of predictions may be influenced by sample size issues.

Summer parr density and habitat data were paired up by site and year. There were some missing values in the habitat dataset. Any site visit with more than three missing covariates was removed; the remaining missing values were imputed using the missForest R package (Stekhoven and Buehlmann 2012; Stekhoven 2013). Ultimately, the spring/summer Chinook parr capacity QRF model was fit using 186 records (paired fish-habitat data) and 14 habitat covariates (13.3 data points per covariate) (Table \@ref(tab:select-mets-table)).

After model fitting, the QRF model was then be used to predict capacity using measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that have been sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated 485 predictions of spring/summer Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, Secesh, Tucannon, Yankee Fork and Wenatchee.

## Extrapolating to Other Sites

Using the QRF model, we made predictions of habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin. However, fisheries biologists are more often interested in capacity at larger scales (e.g., watershed, population). To accomplish this, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) to scale our CHaMP site predictions to a series of larger spatial scales; often corresponding to the entirety of tributary habitat utilized by a given population. The GAA data used here was available from the list of GRTS master sample sites that the CHaMP sites were originally selected from. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed.  The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the svyglm function from the survey package (Lumley 2004, 2016) in R software (R Core Team 2015). Models with all possible combination of GAAs were fit and all the models with cumulative Akaike weights less than 0.95 were model averaged to make predictions of capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart. For sites outside of CHaMP watersheds, only extrapolation models that did not include watershed as a covariate were model averaged to make predictions. Summaries of extrapolation model fit are shown in Table 4.

To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook, as defined by StreamNet http://www.streamnet.org or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale. However, for visualization, predictions of areal capacity (fish/m2) were used.

## Model Validation

Spawner-recruit data from several watersheds within the interior CRB were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts from rotary screw traps. Using estimates of over-winter survival, estimates of parr were calculated from those watersheds (Table 5). A series of spawner-recruit functions were then fit to this data including Beverton-Hold, Ricker, and hockey stick. Estimates of capacity were made from each of these spawner-recruit curves and compared with QRF estimates of capacity (Table 5, Figure 4).

# Results

## Habitat Covariate Selection

We categorized 145 habitat measurements collected using the CHaMP habitat protocol (CHaMP 2016) into 11 habitat groups, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following 14 CHaMP habitat covariates to fit the QRF model: channel unit frequency, thalweg depth CV, average wetted width  to depth ratio, total fish cover, disturbance index, riparian ground cover, average thalweg depth, percent gravel, average substrate size (D50), percent substrate < 6mm, average hourly temperatures in the summer, maximum weekly average temperature, conductivity and frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).

## QRF Model

A QRF model was fit using 14 habitat covariates (Table \@ref(tab:select-mets-table)) and the quantregForest package (Meinshausen 2016) in R (R Core Team 2015) and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure 3). Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots (PDP). These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure 4). However, given that many habitat metrics are somewhat correlated, these marginal effects are not independent of one another making the interpretation of partial dependence plots less straightforward. 
After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

## Extrapolating to Other Sites

## Model Validation

# Discussion

Using a quantile regression forest model, we predicted the summer rearing capacity for spring/summer Chinook parr at the reach scale. Quantile regression forests combine the benefits of random forest analyses (e.g. non-linear responses, correlated covariates) with quantile regression, which allows us to estimate how fish capacity, and not just observed fish density, changes with stream habitat characteristics. We then extrapolated those reach-scale estimates to watershed-scale estimates of capacity, which are more in line with the spatial scale at which management actions take place.

## Biological Expectations from QRF Model

The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations.

<!-- Discussion of biological expectations and PDP plots… -->

## Model Assumptions

<!-- Somewhere in here, mention how data can be collected over a short time span, if enough spatial variation is captured. Or dataset can be built over time, sampling different locations (but be vary of possible year effects, maybe). -->

Model assumptions should always be considered when interpreting results; the following assumptions were made when modeling habitat capacity using QRF methods. First, we assume that sites with higher carrying capacity (i.e., ‘better’ habitat) contain more parr during fish surveys. Second, we assume that carrying capacity is a function of habitat. Third, we assume that estimates of parr densities from fish surveys are unbiased. Fourth, we assume that at least some sites surveys in our dataset are at or close to carrying capacity at the site level. However, if this fourth assumption is not met, we will obtain a conservative estimate of carrying capacity (but the framework of the model is not wrong). Moreover, we assume it is other unmeasured factors (not habitat) that are preventing a site from displaying fish densities near carrying capacity (e.g., insufficient adult spawners, competition, predation, etc.). Finally, we assume that the 90th quantile of predictions at each CHaMP site is a reasonable proxy for carrying capacity. With insufficient data, it is difficult to get a well-defined estimate of higher quantiles (i.e., lots of data is necessary to define the tails of a distribution).

<!-- This paragraph needs to be hashed out.	 -->
Assumptions we don’t need to make. We make no assumptions regarding the shape of fish-habitat relationships. Rather fish-habitat relationships are estimated from the data. Relationships may be positive or negative, linear or non-linear; regardless, no constraints are put on the relationships. Moreover, we make no assumptions regarding correlations among habitat covariates; the random forest framework accounts for possible interactions between habitat metrics, so we don’t need to explicitly place them in the model.

## Extrapolation

<!-- Start with benefits of extrapolation model. For example, allows us to make estimates at larger scales. However, transition to negatives of extrapolation model. -->

<!-- Then open it up to next steps of extrapolation model. i.e., ideally one should collect spatially continuous habitat data. If done, we could eliminate the extrapolation model altogether. Transition to improving habitat data. -->

## Improving Habitat Data

<!-- Discussion of improving habitat data to possibly avoid extrapolation. Collect cheaper/better spatially continuous habitat data. -->

## Conclusions and Next Steps

# Acknowledgements

<!-- The following line inserts a page break  -->
\newpage

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

\newpage

# Tables

```{r fh-samp-table}
# fh_data %>%
qrf_data %>%
  # filter(Year <= 2014) %>%
  tabyl(Watershed) %>%
  adorn_totals() %>%
  adorn_pct_formatting() %>%
  # select(-percent) %>%
  kable(caption = 'The number of unique sites in the initial dataset, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity model')
```

```{r select-mets-table}
qrf_mod$importance %>%
  as_tibble(rownames = 'ShortName') %>%
  rename(imp = IncNodePurity) %>%
  arrange(desc(imp)) %>%
  mutate(Rank = 1:n()) %>%
  select(-imp) %>%
  left_join(hab_dict) %>%
  select(Rank,
         Metric = Name,
         `Metric Category` = MetricCategory,
         Description = DescriptiveText) %>%
  kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months). Metrics are ranked in order of relative importance.",
        format.args = list(width = c(2, 10, 10, 50)))
  # pander(split.cells = c(2, 10, 10, 50))
```

\newpage

# Figures

<!-- We need to make a map of the Columbia River basin that includes polygons for each of the CHaMP watersheds. -->

```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile regression forest (QRF) model to predict habitat capacity, during summer months, for spring/summer Chinook salmon parr"}
qrf_mod$importance %>%
  as_tibble(rownames = 'ShortName') %>%
  rename(imp = IncNodePurity) %>%
  arrange(desc(imp)) %>%
  left_join(hab_dict %>%
              filter(is.na(MetricGroupName) | MetricGroupName != 'Tier 1 Summary')) %>%
  mutate_at(vars(ShortName, Name),
            list(~ fct_reorder(., .x = imp))) %>%
  ggplot(aes(x = Name,
             y = imp)) +
  # geom_col(aes(fill = MetricCategory)) %>%
  geom_col(fill = 'gray40') +
  coord_flip() +
  labs(x = 'Metric',
       y = 'Relative Importance')

```

```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer Chinook salmon parr capacity quantile regression forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values. Tick marks along the X-axis depict observed values, and the sub-basin they came from."}
plot_partial_dependence(qrf_mod,
                        qrf_data,
                        hab_dict,
                        scales = 'free')
```

<!-- The following line inserts a page break  -->
\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2019-12-10 analysis/R_scripts/select_covariates.Rmd
added script to select QRF covariates and make figures for manuscript.

---
title: "QRF Covariate Selection"
author: "Kevin See"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

# load packages for analysis
library(QRFcapacity)
library(maptools)
library(tidyverse)
library(janitor)
library(magrittr)
library(minerva)
library(sf)
library(quantregForest)
library(survey)
library(MuMIn)

# set default theme for ggplot
theme_set(theme_bw())
# setwd('analysis')
```

We loaded the data that had been prepped and saved as part of the `QRFcapacity` package. We used the data from all CHaMP sites, 2011-2017.

```{r load-data}
# determine which set of fish/habitat data to use
data("fh_sum_champ_2017")
fish_hab = fh_sum_champ_2017 %>%
  mutate_at(vars(Watershed, Year),
            list(as.factor))

# and the appropriate habitat dictionrary to go with it
data("hab_dict_2017")
hab_dict = hab_dict_2017

# all the related habitat data
data("champ_site_2011_17")
hab_data = champ_site_2011_17

data("champ_site_2011_17_avg")
hab_avg = champ_site_2011_17_avg

# add temperature metrics
data("champ_temps")

# add average aug temperature to average CHaMP metrics
hab_avg %<>%
  left_join(champ_temps %>%
              as_tibble() %>%
              select(Site, avg_aug_temp = S2_02_11) %>%
              distinct())

# add specific years' temperature, and average for each site
hab_data %<>%
  left_join(hab_data %>%
              select(VisitID, Year = VisitYear) %>%
              distinct() %>%
              left_join(champ_temps %>%
                          as_tibble() %>%
                          select(VisitID, avg_aug_temp = S2_02_11))) %>%
  left_join(hab_data %>%
              select(VisitID, Year = VisitYear) %>%
              distinct() %>%
              left_join(champ_temps %>%
                          as_tibble() %>%
                          select(Site:VisitID, S1_93_11:S36_2015) %>%
                          gather(scenario, aug_temp, S1_93_11:S36_2015) %>%
                          mutate(Year = str_sub(scenario, -4)) %>%
                          mutate_at(vars(Year),
                                    list(as.numeric)) %>%
                          filter(!is.na(Year)) %>%
                          select(Site:VisitID, Year, aug_temp))) %>%
  select(-Year)

```

All sites fell within the range of steelhead, but many of them were outside the range of spring/summer Chinook salmon. However, defining the range of sp/su Chinook is tricky, and it is unclear if the reported data for Chinook salmon includes only sites considered with the range of Chinook, or if we should exclude data from some sites based on our best understanding of the species' range. Currently, we are excluding sites outside the Chinook domain. We created our own list of sites in the John Day that were within a Chinook domain, based on which sites Chinook were found at some point (and downstream of those).

```{r chnk-domain-filter}
data("chnk_domain")

# which sites were sampled for Chinook? 
chnk_samps = fish_hab %>%
  filter(Species == 'Chinook') %>%
  select(Site:Lon, N) %>%
  distinct() %>%
  st_as_sf(coords = c('Lon', 'Lat'),
           crs = 4326) %>%
  st_transform(st_crs(chnk_domain))

# set snap distance (in meters)
st_crs(chnk_samps)
snap_dist = 1000

# which of those sites are in Chinook domain?
chnk_sites = chnk_samps %>%
  as_Spatial() %>%
  maptools::snapPointsToLines(chnk_domain %>%
                                mutate(id = 1:n()) %>%
                                select(id, MPG) %>%
                                as_Spatial(),
                              maxDist = snap_dist,
                              withAttrs = T,
                              idField = 'id') %>%
  as('sf') %>%
  select(-nearest_line_id, -snap_dist) %>%
  # include some sites in the John Day where Chinook were found (or seemed to be close to sites where Chinook were found)
  rbind(st_read('../data/derived_data/Chnk_JohnDay_TrueObs.shp',
                quiet = T) %>%
          st_transform(st_crs(chnk_domain)) %>%
          select(-in_range))

fish_hab %<>%
  filter(Species == 'Steelhead' |
           (Species == 'Chinook' & Site %in% chnk_sites$Site))
```

Next we generated MINE stats for all possible habitat metrics, matching them up against fish densities.

```{r MINE_stats}

# # what are some possible habitat covariates?
poss_hab_mets = hab_dict %>%
  filter(MetricCategory != 'Categorical') %>%
  filter(ShortName %in% names(fish_hab)) %>%
  pull(ShortName) %>%
  unique()

poss_hab_mets = c(poss_hab_mets, 'aug_temp', 'avg_aug_temp') %>%
  unique()

mine_res = fish_hab %>%
  split(list(.$Species)) %>%
  map_df(.id = 'Species',
         .f = function(x) {
           x %>%
             mutate(fish_dens = log(fish_dens + 0.005)) %>%
             estimate_MIC(covars = poss_hab_mets,
                          response = 'fish_dens')
         }) %>%
  left_join(hab_dict %>%
              filter(MetricGroupName == 'Visit Metric') %>%
              select(Metric = ShortName,
                     MetricCategory,
                     Name),
            by = 'Metric') %>%
  mutate_at(vars(MetricCategory),
            list(fct_explicit_na),
            na_level = 'Other') %>%
  mutate_at(vars(Name),
            list(as.character)) %>%
  mutate(Name = if_else(is.na(Name),
                        as.character(Metric),
                        Name)) %>%
  # put the metric names in descending order by MIC
  mutate_at(vars(Metric, Name),
            list(~ fct_reorder(., .x = MIC))) %>%
  select(Species, MetricCategory, Metric, everything()) %>%
  arrange(Species, MetricCategory, desc(MIC))

mine_plot_df = mine_res #%>%
  # filter out some metrics with too many NAs or 0s
  # filter((perc_NA < 0.2 & non_0 > 100) | MetricCategory == 'Temperature') #%>%
  # # filter out metrics with very low variance
  # filter(var < 0.1) %>% select(1:11)
  # # filter out area and volume metrics
  # filter(!grepl('Area', Metric),
  #        !grepl('Vol', Metric),
  #        Metric != 'Lgth_Wet')
```

### 2019-12-10 analysis/R_scripts/fit_qrf_model.Rmd
renamed script to fit QRF models.

---
title: "QRF Covariate Selection"
author: "Kevin See"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

# load packages for analysis
library(QRFcapacity)
library(maptools)
library(tidyverse)
library(janitor)
library(magrittr)
library(minerva)
library(sf)
library(quantregForest)
library(survey)
library(MuMIn)

# set default theme for ggplot
theme_set(theme_bw())
# setwd('analysis/R_scripts')
```

# Select QRF Habitat Covariates

We loaded the data that had been prepped and saved as part of the `QRFcapacity` package. We used the data from all CHaMP sites, 2011-2017.

```{r load-data}
# determine which set of fish/habitat data to use
data("fh_sum_champ_2017")
fish_hab = fh_sum_champ_2017 %>%
  filter(Species == 'Chinook') %>%
  mutate_at(vars(Watershed, Year),
            list(as.factor))

# and the appropriate habitat dictionrary to go with it
data("hab_dict_2017")
hab_dict = hab_dict_2017

# all the related habitat data
data("champ_site_2011_17")
hab_data = champ_site_2011_17 %>%
  filter(!Watershed %in% c('Big-Navarro-Garcia (CA)',
                           'CHaMP Training',
                           'Region 17',
                           'Walla Walla',
                           'Umatilla'))

data("champ_site_2011_17_avg")
hab_avg = champ_site_2011_17_avg %>%
  filter(!Watershed %in% c('Big-Navarro-Garcia (CA)',
                           'CHaMP Training',
                           'Region 17',
                           'Walla Walla',
                           'Umatilla'))

# add temperature metrics
data("champ_temps")

# add average aug temperature to average CHaMP metrics
hab_avg %<>%
  left_join(champ_temps %>%
              as_tibble() %>%
              select(Site, avg_aug_temp = S2_02_11) %>%
              distinct())

# add specific years' temperature, and average for each site
hab_data %<>%
  left_join(hab_data %>%
              select(VisitID, Year = VisitYear) %>%
              distinct() %>%
              left_join(champ_temps %>%
                          as_tibble() %>%
                          select(VisitID, avg_aug_temp = S2_02_11))) %>%
  left_join(hab_data %>%
              select(VisitID, Year = VisitYear) %>%
              distinct() %>%
              left_join(champ_temps %>%
                          as_tibble() %>%
                          select(Site:VisitID, S1_93_11:S36_2015) %>%
                          gather(scenario, aug_temp, S1_93_11:S36_2015) %>%
                          mutate(Year = str_sub(scenario, -4)) %>%
                          mutate_at(vars(Year),
                                    list(as.numeric)) %>%
                          filter(!is.na(Year)) %>%
                          select(Site:VisitID, Year, aug_temp))) %>%
  select(-Year)

```

All sites fell within the range of steelhead, but many of them were outside the range of spring/summer Chinook salmon. However, defining the range of sp/su Chinook is tricky, and it is unclear if the reported data for Chinook salmon includes only sites considered with the range of Chinook, or if we should exclude data from some sites based on our best understanding of the species' range. Currently, we are excluding sites outside the Chinook domain. We created our own list of sites in the John Day that were within a Chinook domain, based on which sites Chinook were found at some point (and downstream of those).

```{r chnk-domain-filter}
data("chnk_domain")

# which sites were sampled for Chinook? 
chnk_samps = fish_hab %>%
  filter(Species == 'Chinook') %>%
  select(Site:Lon, N) %>%
  distinct() %>%
  st_as_sf(coords = c('Lon', 'Lat'),
           crs = 4326) %>%
  st_transform(st_crs(chnk_domain))

# set snap distance (in meters)
st_crs(chnk_samps)
snap_dist = 1000

# which of those sites are in Chinook domain?
chnk_sites = chnk_samps %>%
  as_Spatial() %>%
  maptools::snapPointsToLines(chnk_domain %>%
                                mutate(id = 1:n()) %>%
                                select(id, MPG) %>%
                                as_Spatial(),
                              maxDist = snap_dist,
                              withAttrs = T,
                              idField = 'id') %>%
  as('sf') %>%
  select(-nearest_line_id, -snap_dist) %>%
  # include some sites in the John Day where Chinook were found (or seemed to be close to sites where Chinook were found)
  rbind(st_read('../data/derived_data/Chnk_JohnDay_TrueObs.shp',
                quiet = T) %>%
          st_transform(st_crs(chnk_domain)) %>%
          select(-in_range))

fish_hab %<>%
  filter(Site %in% chnk_sites$Site)
  # filter(Species == 'Steelhead' |
  #          (Species == 'Chinook' & Site %in% chnk_sites$Site))

```

Next we generated MINE stats for all possible habitat metrics, matching them up against the log of fish densities (plus a small value to deal with 0s).

```{r MINE_stats}

# # what are some possible habitat covariates?
poss_hab_mets = hab_dict %>%
  filter(MetricCategory != 'Categorical') %>%
  filter(ShortName %in% names(fish_hab)) %>%
  pull(ShortName) %>%
  unique()

# poss_hab_mets = c(poss_hab_mets, 'aug_temp', 'avg_aug_temp') %>%
#   unique()

mine_res = fish_hab %>%
  mutate_at(vars(starts_with('LWVol')),
            list(~ . / Lgth_Wet)) %>%
  mutate(fish_dens = log(fish_dens + 0.005)) %>%
  estimate_MIC(covars = poss_hab_mets,
               response = 'fish_dens') %>%
  mutate(Species = 'Chinook') %>%
  left_join(hab_dict %>%
              filter(MetricGroupName == 'Visit Metric') %>%
              select(Metric = ShortName,
                     MetricCategory,
                     Name),
            by = 'Metric') %>%
  mutate_at(vars(MetricCategory),
            list(fct_explicit_na),
            na_level = 'Other') %>%
  mutate_at(vars(Name),
            list(as.character)) %>%
  mutate(Name = if_else(is.na(Name),
                        as.character(Metric),
                        Name)) %>%
  # put the metric names in descending order by MIC
  mutate_at(vars(Metric, Name),
            list(~ fct_reorder(., .x = MIC))) %>%
  select(Species, MetricCategory, Metric, everything()) %>%
  arrange(Species, MetricCategory, desc(MIC))

```

Plot the MINE results and save figures for manuscript.

```{r MINE-figures}
mine_plot_df = mine_res %>%
  # filter out some metrics with too many NAs or 0s
  filter((perc_NA < 0.2 & non_0 > 100) | MetricCategory == 'Temperature') %>%
  # filter out metrics with very low variance
  # filter(var < 0.1) %>%
  # filter(obsCV < 0.1)
  # janitor::tabyl(MetricCategory)
  # select(1:11)
  # filter out area and volume metrics
  filter(!grepl('Area$', Metric),
         !grepl('Vol$', Metric),
         !Metric %in% c('Lgth_Wet', 
                        'Lgth_BfChnl',
                        'Lgth_WetChnl',
                        'Area_Wet', 
                        'Area_Bf', 
                        'WetVol', 
                        'BfVol'))

mine_p = mine_plot_df %>%
  ggplot(aes(x = Name,
             y = MIC)) +
  geom_col(position = position_dodge(1),
           fill = 'gray45') +
  coord_flip() +
  facet_wrap(~ MetricCategory,
             scales = 'free_y',
             ncol = 3) +
  scale_fill_brewer(palette = 'Set1',
                    guide = guide_legend(nrow = 1)) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 5)) +
  labs(y = 'Habitat Covariate')

mine_p2 = mine_plot_df %>%
  ggplot(aes(x = Name,
             y = MIC,
             fill = MetricCategory)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = 'Set1',
                    guide = guide_legend(nrow = 1)) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 5)) +
  labs(y = 'Habitat Covariate')

```

```{r, eval = F}
library(corrr)
#----------------------------------------------
# Look at correlations between habitat metrics
#----------------------------------------------
# top metrics
sel_mets = poss_hab_mets
sel_mets = mine_plot_df %>%
  group_by(MetricCategory) %>%
  slice(1:5) %>%
  ungroup() %>%
  pull(Metric) %>%
  unique() %>%
  as.character()

corr_mat = hab_avg %>%
  select(one_of(sel_mets)) %>%
  corrr::correlate()

corr_mat %>%
  corrr::rearrange(absolute = F) %>%
  corrr::shave(upper = T) %>% 
  corrr::stretch() %>%
  filter(!is.na(r)) %>%
  filter(abs(r) > 0.5) %>%
  kable()

corr_p1 = corr_mat %>%
  # rearrange(absolute = F) %>%
  shave(upper = T) %>% 
  rplot(legend = T,
        print_cor = T)

corr_p2 = network_plot(corr_mat)
```

```{r correlation-plots}
#----------------------------------------------
# Look at correlations between habitat metrics
#----------------------------------------------

library(ggcorrplot)

corr_list = hab_dict %>%
  filter(MetricCategory != 'Categorical') %>%
  split(list(.$MetricCategory)) %>%
  map(.f = function(x) {
    catg_mets = x %>%
      pull(ShortName)
    hab_data %>%
      select(one_of(catg_mets)) %>%
      cor(use = "pairwise.complete.obs")
      # corrr::correlate(use = "pairwise.complete.obs",
      #                  method = "pearson",
      #                  quiet = T)
  })

# make correlation plots by metric category
corr_p_list = corr_list %>%
  map(.f = ggcorrplot,
      method = "circle",
      show.diag = F,
      type = 'lower')

# add a title to each plot
for(i in 1:length(corr_p_list)) {
  corr_p_list[[i]] = corr_p_list[[i]] +
    labs(title = names(corr_list)[i])
}

# save as a PDF
pdf("../figures/correlation_plots_all.pdf",
    width = 8,
    height = 8)
for(i in 1:length(corr_p_list)) {
  print(corr_p_list[[i]])
}
dev.off()

```

Based on the MIC results, and trying to stay away from pairs of metrics that were too closely correlated, we came up with the following list of QRF covariates.

```{r habitat-covariates}
sel_hab_mets = tibble(Species = c('Chinook'),
                      Metric = c('UcutArea_Pct',
                                 'FishCovNone',
                                 'SubEstGrvl',
                                 'FstTurb_Freq',
                                 'FstNT_Freq',
                                 'CU_Freq',
                                 'SlowWater_Pct',
                                 'NatPrin1',
                                 'DistPrin1',
                                 'avg_aug_temp',
                                 # 'Sin_CL',
                                 'Sin',
                                 'WetWdth_CV',
                                 'WetBraid',
                                 'WetSC_Pct',
                                 'Q',
                                 'WetWdth_Int',
                                 'LWFreq_Wet',
                                 'LWVol_WetFstTurb'))

# sum(!sel_hab_mets$Metric %in% names(fish_hab))
```

```{r}
sel_hab_mets %>%
  select(ShortName = Metric) %>%
  distinct() %>%
  left_join(hab_dict) %>%
  select(Catg = MetricCategory, 
         Metric = Name, 
         Description = DescriptiveText) %>%
  kable()
```

# Fit QRF Model

```{r}
#-----------------------------------------------------------------
# Fit QRF model
#-----------------------------------------------------------------
covars = sel_hab_mets$Metric
# impute missing data in fish / habitat dataset
set.seed(7)
qrf_data = fish_hab %>%
  impute_missing_data(covars = covars,
                      impute_vars = c('Watershed', 'Elev_M', 'Sin', 'Year', 'CUMDRAINAG'),
                      method = 'missForest') %>%
  select(Site, Watershed, Year, LON_DD, LAT_DD, fish_dens, VisitID, one_of(covars))

# fit the QRF model
# set the density offset (to accommodate 0z)
dens_offset = 0.005

# fit random forest models
set.seed(4)
qrf_mod = quantregForest(x = qrf_data %>%
                               select(one_of(covars)) %>%
                               as.matrix,
                             y = qrf_data %>%
                               mutate_at(vars(fish_dens),
                                         list(~ log(. + dens_offset))) %>%
                               select(fish_dens) %>%
                               as.matrix(),
                             keep.inbag = T,
                             ntree = 1000)

# save some results
use_data(fish_hab, 
         hab_dict,
         hab_data,
         hab_avg,
         sel_hab_mets,
         qrf_data,
         dens_offset,
         qrf_mod,
         version = 2,
         overwrite = T)

```

## Predict at all CHaMP sites

```{r champ_predictions}
# what quantile is a proxy for capacity?
pred_quant = 0.9

covars = unique(sel_hab_mets$Metric)

# impute missing values in habitat data
hab_impute = hab_avg %>%
  mutate_at(vars(Watershed, Channel_Type),
            list(fct_drop)) %>%
  impute_missing_data(data = .,
                      covars = covars,
                      impute_vars = c('Watershed', 
                                      'Elev_M', 
                                      'Channel_Type', 
                                      'CUMDRAINAG'),
                      method = 'missForest') %>%
  select(Site, Watershed, LON_DD, LAT_DD, VisitYear, Lgth_Wet, Area_Wet, one_of(covars))

# predict at all CHaMP sites
pred_hab_sites = hab_impute %>%
  mutate(chnk_per_m = predict(qrf_mod,
                              newdata = select(., one_of(covars)),
                              what = pred_quant),
         chnk_per_m = exp(chnk_per_m) - dens_offset,
         chnk_per_m2 = chnk_per_m * Lgth_Wet / Area_Wet)

# filter out sites outside the Chinook domain
data("chnk_domain")
snap_dist = 1000

pred_hab_sites_chnk = pred_hab_sites %>%
  filter(!is.na(LON_DD)) %>%
  st_as_sf(coords = c('LON_DD', 'LAT_DD'),
           crs = 4326) %>%
  st_transform(crs = st_crs(chnk_domain)) %>%
  as_Spatial() %>%
  maptools::snapPointsToLines(chnk_domain %>%
                                mutate(id = 1:n()) %>%
                                select(id, MPG) %>%
                                as_Spatial(),
                              maxDist = snap_dist,
                              withAttrs = T,
                              idField = 'id') %>%
  as('sf') %>%
  as_tibble() %>%
  select(-nearest_line_id, -snap_dist, -geometry) %>%
  # add sites in the John Day
  bind_rows(st_read('../data/derived_data/Chnk_JohnDay_TrueObs.shp',
                    quiet = T) %>%
              as_tibble() %>%
              select(Site) %>%
              inner_join(pred_hab_sites))

```

# Extrapolation

### 2019-12-10 analysis/paper/paper.Rmd
try using data objects created in QRFpaper repo.

@@
 library(QRFcapacity) # for the moment, until we know exactly what data to copy over to this repo
+library(QRFpaper)
 library(janitor)
@@
                  file = 'packages.bib')
+```
+
+```{r load-data}
+data(fish_hab, 
+     hab_dict, 
+     hab_data, 
+     hab_avg, 
+     qrf_data,
+     dens_offset,
+     qrf_mod,
+     sel_hab_mets)
 ```

### 2019-12-24 analysis/R_scripts/fit_qrf_model.Rmd
re-compiled files on new Merck laptop. Incorporated snapPointsToLines_v2 function from QRFcapacity package.

@@
 st_crs(chnk_samps)
-snap_dist = 1000
+max_snap_dist = 1000
 
@@
   as_Spatial() %>%
-  maptools::snapPointsToLines(chnk_domain %>%
+  snapPointsToLines_v2(points = .,
+  # maptools::snapPointsToLines(points = .,
+                              lines = chnk_domain %>%
                                 mutate(id = 1:n()) %>%
@@
                                 as_Spatial(),
-                              maxDist = snap_dist,
-                              withAttrs = T,
+                              maxDist = max_snap_dist,
+                              withAttrs = TRUE,
                               idField = 'id') %>%
@@
 # save some results
-use_data(fish_hab, 
+usethis::use_data(fish_hab, 
                   hab_dict,
@@
   as_Spatial() %>%
-  maptools::snapPointsToLines(chnk_domain %>%
+  snapPointsToLines_v2(chnk_domain %>%
+  # maptools::snapPointsToLines(chnk_domain %>%
                                 mutate(id = 1:n()) %>%

### 2019-12-24 analysis/paper/paper.Rmd
re-compiled files on new Merck laptop. Incorporated snapPointsToLines_v2 function from QRFcapacity package.

@@
 library(tidyverse)
-library(QRFcapacity) # for the moment, until we know exactly what data to copy over to this repo
+# library(QRFcapacity) # for the moment, until we know exactly what data to copy over to this repo
 library(QRFpaper)
@@
 
-```{r read-data}
+```{r read-data, eval = F}
 # fish/habitat data for fitting QRF models
-# data("fh_sum_champ_2017")
-# fh_data = fh_sum_champ_2017 %>%
+data("fh_sum_champ_2017")
+fh_data = fh_sum_champ_2017 %>%
+  filter(Species == 'Chinook') %>%
+  mutate_at(vars(Watershed, Species, FishCrew),
+            list(fct_drop))
+
+# data("fh_sum_champ_2014")
+# fh_data = fh_sum_champ_2014 %>%
 #   filter(Species == 'Chinook') %>%
@@
 #             list(fct_drop))
-
-data("fh_sum_champ_2014")
-fh_data = fh_sum_champ_2014 %>%
-  filter(Species == 'Chinook') %>%
-  mutate_at(vars(Watershed, Species, FishCrew),
-            list(fct_drop))
 
@@
 
-```{r load-qrf-model}
+```{r load-qrf-model, eval = F}
 load('../data/derived_data/qrf_juv_summer.rda')
@@
 ```{r fh-samp-table}
-# fh_data %>%
 qrf_data %>%
@@
   # pander(split.cells = c(2, 10, 10, 50))
+```
+
+```{r sr-table}
+
 ```

### 2019-12-24 tests/testthat.R

@@ -1,0 +1,4 @@
+library(testthat)
+library(QRFpaper)
+
+test_check("QRFpaper")

### 2019-12-24 analysis/paper/paper.Rmd
added section to re-create extrapolation model fit table with r-squared values

@@
 
+```{r extrap-table}
+mod_fit_df = model_svy_df %>%
+  select(response, data) %>%
+  unnest(cols = data) %>%
+  select(response, obs = qrf_cap) %>%
+  group_by(response) %>%
+  mutate(id = 1:n()) %>%
+  left_join(model_svy_df %>%
+              mutate(pred_full = map(mod_full,
+                                     .f = function(x) {
+                                       predict(x,
+                                               type = 'response') %>%
+                                         as_tibble() %>%
+                                         rename(pred_full = response,
+                                                pred_full_se = SE)
+                                     }),
+                     pred_no_champ = map(mod_no_champ,
+                                         .f = function(x) {
+                                           predict(x,
+                                                   type = 'response') %>%
+                                             as_tibble() %>%
+                                             rename(pred_no_champ = response,
+                                                    pred_no_champ_se = SE)
+                                         })) %>%
+              select(response, starts_with("pred")) %>%
+              unnest(cols = starts_with("pred")) %>%
+              group_by(response) %>%
+              mutate(id = 1:n())) %>%
+  ungroup()
+
+# mod_fit_df %>%
+#   group_by(response) %>%
+#   summarise(r2_full = cor(obs, pred_full)^2,
+#             r2_no_champ = cor(obs, pred_no_champ)^2) %>%
+#   gather(type, r2, -response) %>%
+#   mutate(model = if_else(grepl('no_champ', type),
+#                          'non-CHaMP',
+#                          'CHaMP'),
+#          type = if_else(grepl('adj', type),
+#                         'Adjusted R2',
+#                         'R2')) %>%
+#   spread(type, r2) %>%
+#   mutate(response = recode(response,
+#                            'chnk_per_m' = 'fish/m',
+#                            'chnk_per_m2' = 'fish/m^2')) %>%
+#   select(Model = model,
+#          Response = response,
+#          R2)
+
+mod_fit_df %>%
+  group_by(response) %>%
+  nest() %>%
+  mutate(comp_full = map(data,
+                         .f = function(x) {
+                           lm(pred_full ~ obs,
+                              data = x)
+                         }),
+         comp_no_champ = map(data,
+                             .f = function(x) {
+                               lm(pred_no_champ ~ obs,
+                                  data = x)
+                             })) %>%
+  mutate(r2_full = map_dbl(comp_full,
+                      .f = function(x) summary(x)$r.squared),
+         r2_adj_full = map_dbl(comp_full,
+                           .f = function(x) summary(x)$adj.r.squared),
+         r2_no_champ = map_dbl(comp_no_champ,
+                           .f = function(x) summary(x)$r.squared),
+         r2_adj_no_champ = map_dbl(comp_no_champ,
+                               .f = function(x) summary(x)$adj.r.squared)) %>%
+  ungroup() %>%
+  select(response, starts_with("r2")) %>%
+  gather(type, r2, -response) %>%
+  mutate(model = if_else(grepl('no_champ', type),
+                         'non-CHaMP',
+                         'CHaMP'),
+         type = if_else(grepl('adj', type),
+                        'Adjusted R2',
+                        'R2')) %>%
+  spread(type, r2) %>%
+  mutate(response = recode(response,
+                           'chnk_per_m' = 'fish/m',
+                           'chnk_per_m2' = 'fish/m^2')) %>%
+  select(Model = model,
+         Response = response,
+         R2, `Adjusted R2`) %>%
+  kable(caption = "Summary of model fit for each of the extrapolation models to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to larger scales (e.g., watershed, population).",
+        digits = c(NA, NA, 3, 3))
+
+```
+
 ```{r sr-table}

### 2019-12-27 analysis/R_scripts/fit_qrf_model.Rmd

@@
 # Extrapolation
+
+Because most of our habitat data came from CHaMP sites, which were selected through a generalized random tesselation stratified (GRTS) sample design [@Stevens2003], we wanted to incorporate the design weights for each CHaMP site with a predicted carrying capacity.
+
+```{r survey_design_strata}
+#----------------------------------------
+# pull in survey design related data
+#----------------------------------------
+# Calculate GRTS design weights.
+
+# pull in info about what strata each CHaMP site was assigned to (using 2014 as reference year)
+site_strata = pred_hab_sites_chnk %>%
+  select(Site, Watershed) %>%
+  distinct() %>%
+  left_join(gaa %>%
+              select(Site,
+                     # CHaMPsheds,
+                     strata = AStrat2014)) %>%
+  mutate(site_num = str_split(Site, '-', simplify = T)[,2]) %>%
+  mutate(strata = if_else(Watershed == 'Asotin',
+                          site_num,
+                          if_else(Watershed == 'Entiat' & grepl('ENT00001', Site),
+                                  paste('EntiatIMW', site_num, sep = '_'),
+                                  strata))) %>%
+  mutate(strata = if_else(grepl('EntiatIMW', strata),
+                          str_remove(strata, '[[:digit:]]$'),
+                          strata),
+         strata = if_else(grepl('EntiatIMW', strata),
+                          str_remove(strata, '[[:digit:]]$'),
+                          strata)) %>%
+  filter(!is.na(strata)) %>%
+  mutate(strata = paste(Watershed, strata, sep = '_')) %>%
+  select(-site_num)
+
+# read in data from the CHaMP frame
+champ_frame_df = read_csv('../data/derived_data/champ_frame_data.csv') %>%
+  mutate(Target2014 = ifelse(is.na(AStrat2014), 'Non-Target', Target2014)) %>%
+  mutate(AStrat2014 = ifelse(AStrat2014 == 'Entiat IMW', paste('EntiatIMW', GeoRchIMW, sep = '_'), AStrat2014)) %>%
+  filter(Target2014 == 'Target') %>%
+  rename(Watershed = CHaMPshed)
+
+# what strata do we have?
+frame_strata = champ_frame_df %>%
+  mutate(strata = paste(Watershed, AStrat2014, sep='_')) %>%
+  select(Watershed, 
+         strata) %>%
+  distinct()
+
+# how long is each strata?
+chnk_strata_length = champ_frame_df %>%
+  filter(!is.na(UseTypCHSP)) %>%
+  mutate(strata = paste(Watershed, AStrat2014, sep='_')) %>%
+  select(Watershed, matches("Strat"), FrameLeng) %>%
+  group_by(Watershed, strata) %>%
+  summarise(tot_length_km = sum(FrameLeng) / 1000) %>%
+  ungroup() %>%
+  mutate_at(vars(Watershed, strata), 
+            list(as.factor)) %>%
+  arrange(Watershed, strata)
+
+sthd_strata_length = champ_frame_df %>%
+  filter(!is.na(UseTypSTSU)) %>%
+  mutate(strata = paste(Watershed, AStrat2014, sep='_')) %>%
+  select(Watershed, matches("Strat"), FrameLeng) %>%
+  group_by(Watershed, strata) %>%
+  summarise(tot_length_km = sum(FrameLeng) / 1000) %>%
+  bind_rows(tibble(Watershed = 'Asotin',
+                   strata = paste('Asotin', c('CC', 'NF', 'SF'), sep = '_'),
+                   tot_length_km = 12)) %>%
+  ungroup() %>%
+  mutate_at(vars(Watershed, strata), 
+            list(as.factor)) %>%
+  arrange(Watershed, strata)
+
+# how many sites in each strata? and what is the length of each strata?
+strata_tab = pred_hab_sites_chnk %>%
+  select(Site, Watershed, matches('per_m')) %>%
+  left_join(site_strata) %>%
+  filter(strata != 'Entiat_Entiat IMW') %>%
+  mutate_at(vars(Watershed),
+            list(fct_drop)) %>%
+  group_by(Watershed, strata) %>%
+  summarise(n_sites = n_distinct(Site)) %>%
+  ungroup() %>%
+  full_join(chnk_strata_length) %>%
+  mutate(n_sites = if_else(is.na(n_sites),
+                           as.integer(0),
+                           n_sites)) %>%
+  # calculate the weight of each site in each strata
+  mutate(site_weight = if_else(n_sites > 0,
+                               tot_length_km / n_sites,
+                               as.numeric(NA)))
+
+strata_test = frame_strata %>%
+  full_join(strata_tab) %>%
+  mutate_at(vars(Watershed),
+            list(fct_drop)) %>%
+  mutate(n_sites = if_else(is.na(n_sites),
+                           as.integer(0),
+                           n_sites))
+
+# # what frame strata don't have any sites in them?
+# strata_test %>%
+#   filter(n_sites == 0,
+#          !is.na(tot_length_km)) %>%
+#   arrange(Watershed, strata) %>%
+#   as.data.frame()
+
+# # what strata that we have sites for are not in the frame strata?
+# strata_test %>%
+#   filter(n_sites > 0,
+#          (is.na(tot_length_km) |
+#             tot_length_km == 0)) %>%
+#   as.data.frame()
+
+# how much of each watershed is not accounted for with current sites / strata?
+strata_test %>%
+  group_by(Watershed) %>%
+  summarise_at(vars(tot_length_km),
+               list(sum),
+               na.rm = T) %>%
+  left_join(strata_test %>%
+              filter(n_sites == 0) %>%
+              group_by(Watershed) %>%
+              summarise_at(vars(missing_length = tot_length_km),
+                           list(sum),
+                           na.rm = T)) %>%
+  mutate_at(vars(missing_length),
+            list(~ if_else(is.na(.), 0, .))) %>%
+  mutate(perc_missing = missing_length / tot_length_km) %>%
+  mutate_at(vars(perc_missing),
+            list(~ if_else(is.na(.), 0, .))) %>%
+  arrange(desc(perc_missing)) %>%
+  pander()
+
+```
+
+We then selected a subset of the globally available attributes (GAAs) to use as extrapolation covariates.
+This was based on minimizing the correlations between the covariates.
+We also acknowlegdged which prediction sites had GAAs outside the range of values for the CHaMP sites.
+
+We use the log of predicted capacity wherever we could make those predictions as the response variable, and this group of selected GAAs as the covariates.
+We then fit a linear model that incorporated the survey design weights by using the `survey` pacakge (-@Lumley2014) in R.
+We fit two main models, one that included CHaMP watershed as a covariate, and one that did not.
+The former was used to make predictions within all CHaMP watersheds where we had QRF predictions of carrying capacity.
+The latter was used at all other master sample points.
+
+```{r}
+data(gaa)
+
+gaa_all = gaa %>%
+  rename(Channel_Type = ChanlType)
+
+# possible covariates from among the GAAs
+gaa_covars = c('TRange', 'GDD', 'Precip', 'Elev_M', 'CHaMPsheds', 'NatPrin1', 'NatPrin2', 'DistPrin1', 'BFW_M', 'SrtCumDrn', 'StrmPwr', 'Slp_NHD_v1', 'Channel_Type', 'MAVELV', 'WIDE_BF', 'S2_02_11')
+
+# what type of covariate is each GAA?
+gaa_class = gaa_all %>%
+  select(one_of(gaa_covars)) %>%
+  as.list() %>%
+  map_chr(.f = class)
+
+# which ones are numeric?
+gaa_num = names(gaa_class)[gaa_class %in% c('integer', 'numeric')]
+# which ones are categorical?
+gaa_catg = names(gaa_class)[gaa_class %in% c('factor', 'character')]
+
+# compare range of covariates from model dataset and prediction dataset
+range_comp = bind_rows(gaa_all %>%
+                         filter(!Site %in% unique(pred_hab_sites$Site)) %>%
+                         select(Site,
+                                one_of(gaa_num)) %>%
+                         gather(Metric, value, -Site) %>%
+                         mutate(Source = 'non-DASH Sites'),
+                       gaa_all %>%
+                         filter(Site %in% unique(pred_hab_sites$Site)) %>%
+                         select(Site,
+                                one_of(gaa_num)) %>%
+                         distinct() %>%
+                         gather(Metric, value, -Site) %>%
+                         mutate(Source = 'DASH Sites')) %>%
+  mutate_at(vars(Source, Metric),
+            list(as.factor))
+
+range_max = range_comp %>%
+  group_by(Metric, Source) %>%
+  summarise_at(vars(value),
+               tibble::lst(min, max),
+               na.rm = T) %>%
+  filter(Source == 'DASH Sites') %>%
+  ungroup() %>%
+  gather(type, value, -Metric, -Source)
+
+covar_range_p = range_comp %>%
+  ggplot(aes(x = Source,
+             y = value,
+             fill = Source)) +
+  geom_boxplot() +
+  facet_wrap(~ Metric,
+             scales = 'free') +
+  geom_hline(data = range_comp %>%
+               group_by(Metric, Source) %>%
+               summarise_at(vars(value),
+                            tibble::lst(min, max),
+                            na.rm = T) %>%
+               filter(Source == 'DASH Sites') %>%
+               ungroup() %>%
+               gather(type, value, min, max),
+             aes(yintercept = value),
+             lty = 2,
+             color = 'darkgray') +
+  theme_minimal()
+
+covar_range_p
+
+# look at correlations among GAAs
+library(corrr)
+corr_df = gaa_all %>%
+  select(one_of(gaa_num)) %>%
+  correlate()
+
+corr_df %>%
+  shave(upper = T) %>%
+  stretch() %>%
+  arrange(desc(abs(r))) %>%
+  filter(abs(r) >= 0.5)
+
+gaa_corr_p1 = corr_df %>%
+  rearrange(absolute = T) %>%
+  shave(upper = T) %>%
+  rplot(legend = T,
+        colors = c('blue', 'white', 'indianred2'),
+        print_cor = T)
+
+# correlation matrix
+corr_mat = gaa_all %>%
+  select(one_of(gaa_num)) %>%
+  cor(use = 'pairwise.complete.obs')
+
+library(ggcorrplot)
+gaa_corr_p2 = ggcorrplot(corr_mat,
+           lab = T,
+           lab_size = 2,
+           hc.order = T,
+           colors = c('#2166AC', 'white', '#B2182B'),
+           tl.cex = 8,
+           type = 'full')
+```
+
+```{r}
+# reduce number of covariates bases on correlations
+gaa_covars = c('TRange', 
+               # 'GDD', 
+               # 'Precip',
+               'Elev_M', 
+               'CHaMPsheds', 
+               'NatPrin1', 
+               # 'NatPrin2', 
+               'DistPrin1', 
+               # 'BFW_M', 
+               'SrtCumDrn', 
+               'StrmPwr',
+               'Slp_NHD_v1', 
+               'Channel_Type', 
+               # 'MAVELV', 
+               'WIDE_BF',
+               'S2_02_11')
+
+# what type of covariate is each GAA?
+gaa_class = gaa_all %>%
+  select(one_of(gaa_covars)) %>%
+  as.list() %>%
+  map_chr(.f = class)
+
+# which ones are numeric?
+gaa_num = names(gaa_class)[gaa_class %in% c('integer', 'numeric')]
+# which ones are categorical?
+gaa_catg = names(gaa_class)[gaa_class %in% c('factor', 'character')]
+
+# Center the covariates
+# filter out sites with covariates outside range of covariates used to fit extrapolation model
+out_range_sites = gaa_all %>%
+  # filter out a few areas
+  filter(!HUC6NmNRCS %in% c('Upper Sacramento', 'Southern Oregon Coastal', 'Puget Sound', 'Northern California Coastal', 'Oregon Closed Basins')) %>%
+  # don't use AEM sites in model
+  filter(!grepl('^AEM', Site)) %>%
+  select(one_of(gaa_num), Site) %>%
+  gather(Metric, value, -Site) %>%
+  left_join(select(range_max, -Source) %>%
+              spread(type, value)) %>%
+  group_by(Metric) %>%
+  filter(value > max |
+           value < min) %>%
+  ungroup() %>%
+  pull(Site) %>%
+  unique()
+
+# center covariates
+gaa_summ = inner_join(pred_hab_sites %>%
+                          select(Site) %>%
+                          distinct(),
+                        gaa_all %>%
+                          select(Site, one_of(gaa_num))) %>%
+  gather(GAA, value, -Site) %>%
+  group_by(GAA) %>%
+  summarise(metric_mean = mean(value, na.rm=T),
+            metric_sd = sd(value, na.rm=T)) %>%
+  ungroup()
+
+# extrapolation model data set, with normalized covariates
+mod_data = inner_join(pred_hab_sites_chnk %>%
+                        select(Site:avg_aug_temp,
+                               matches('per_m'),
+                               -one_of(gaa_num)),
+                      gaa_all %>%
+                        select(Site, one_of(gaa_num))) %>%
+  gather(GAA, value, one_of(gaa_num)) %>%
+  left_join(gaa_summ) %>%
+  mutate(norm_value = (value - metric_mean) / metric_sd) %>%
+  select(-(value:metric_sd)) %>%
+  spread(GAA, norm_value) %>%
+  left_join(gaa_all %>%
+              select(Site, one_of(gaa_catg)))
+
+mod_data %<>%
+  bind_cols(mod_data %>%
+              is.na() %>%
+              as_tibble() %>%
+              select(one_of(gaa_covars)) %>%
+              transmute(n_NA = rowSums(.))) %>%
+  filter(n_NA == 0) %>%
+  mutate_at(vars(Watershed, CHaMPsheds, Channel_Type),
+            list(fct_drop))
+
+# where do we want to make extrapolation predictions?
+gaa_pred = gaa_all %>%
+  # filter out a few areas
+  filter(!HUC6NmNRCS %in% c('Upper Sacramento', 'Southern Oregon Coastal', 'Puget Sound', 'Northern California Coastal', 'Oregon Closed Basins')) %>%
+  # don't use AEM sites in model
+  filter(!grepl('^AEM', Site)) %>%
+  # don't use non-GRTS sites
+  filter(SiteID_alt != 'NonGRTSSite' | is.na(SiteID_alt)) %>%
+  filter(!grepl('mega', Site, ignore.case = T)) %>%
+  # note which sites have GAAs outside range of CHaMP sites GAAs
+  mutate(inCovarRange = ifelse(Site %in% out_range_sites, F, T)) %>%
+  select(Site, one_of(gaa_covars), Lon, Lat, inCovarRange, HUC6NmNRCS, HUC8NmNRCS, HUC10NmNRC, HUC12NmNRC, chnk) %>%
+  gather(GAA, value, one_of(gaa_num)) %>%
+  left_join(gaa_summ) %>%
+  mutate(norm_value = (value - metric_mean) / metric_sd) %>%
+  select(-(value:metric_sd)) %>%
+  spread(GAA, norm_value)
+
+```
+
+```{r with_model_weights, eval = T}
+# calculate adjusted weights for all predicted QRF capacity sites
+mod_data_weights = mod_data %>%
+  left_join(site_strata) %>%
+  left_join(strata_tab) %>%
+  filter(!is.na(site_weight)) %>%
+  group_by(Watershed) %>%
+  mutate(sum_weights = sum(site_weight)) %>%
+  ungroup() %>%
+  mutate(adj_weight = site_weight / sum_weights)
+
+#-------------------------------------------------------------
+# Set up the survey design.
+
+# getOption('survey.lonely.psu')
+# this will prevent strata with only 1 site from contributing to the variance
+# options(survey.lonely.psu = 'certainty')
+# this centers strata with only 1 site to the sample grand mean; this is conservative
+options(survey.lonely.psu = 'adjust')
+
+# extrapolation model formula
+full_form = as.formula(paste('log(qrf_cap) ~ -1 + (', paste(gaa_covars, collapse = ' + '), ')'))
+
+# fit various models
+model_svy_df = mod_data_weights %>%
+  gather(response, qrf_cap, matches('per_m')) %>%
+  select(-(n_sites:sum_weights)) %>%
+  group_by(response) %>%
+  nest() %>%
+  mutate(design = map(data,
+                      .f = function(x) {
+                        svydesign(id = ~ 1,
+                                  data = x,
+                                  strata = ~ Watershed,
+                                  # strata = ~ strata,
+                                  weights = ~ adj_weight)
+                      })) %>%
+  mutate(mod_full = map(design,
+                        .f = function(x) {
+                          svyglm(full_form,
+                                 design = x)
+                        }),
+         mod_no_champ = map(design,
+                            .f = function(x) {
+                              svyglm(update(full_form, .~ . -CHaMPsheds),
+                                     design = x)
+                            }))
+```
+
+```{r make-predictions}
+# predictions within CHaMP watersheds, using the model that includes CHaMPsheds as a covariate
+y = gaa_pred %>%
+  filter(CHaMPsheds %in% unique(model_svy_df$data[[1]]$Watershed)) %>%
+  select(Site, CHaMPsheds, one_of(gaa_covars)) %>%
+  na.omit() %>%
+  left_join(gaa_pred) %>%
+  mutate_at(vars(Channel_Type),
+            list(as.factor))
+
+per_m_preds = predict(model_svy_df$mod_full[[1]],
+                      newdata = y,
+                      se = T,
+                      type = 'response')
+
+per_m2_preds = predict(model_svy_df$mod_full[[2]],
+                      newdata = y,
+                      se = T,
+                      type = 'response')
+
+y %<>%
+  bind_cols(per_m_preds %>%
+              as_tibble() %>%
+              rename(chnk_per_m = response,
+                     chnk_per_m_se = SE)) %>%
+  bind_cols(per_m2_preds %>%
+              as_tibble() %>%
+              rename(chnk_per_m2 = response,
+                     chnk_per_m2_se = SE)) %>%
+  mutate_at(vars(matches('per_m')),
+            list(exp))
+
+rm(per_m_preds, per_m2_preds)
+
+# predictions at all points, using the model without CHaMPsheds as a covariate
+z = gaa_pred %>%
+  select(Site, one_of(gaa_covars), -CHaMPsheds) %>%
+  na.omit() %>%
+  left_join(gaa_pred) %>%
+  mutate_at(vars(Channel_Type),
+            list(as.factor))
+
+per_m_preds = predict(model_svy_df$mod_no_champ[[1]],
+                      newdata = z,
+                      se = T,
+                      type = 'response')
+
+per_m2_preds = predict(model_svy_df$mod_no_champ[[2]],
+                      newdata = z,
+                      se = T,
+                      type = 'response')
+
+z %<>%
+  bind_cols(per_m_preds %>%
+              as_tibble() %>%
+              rename(chnk_per_m = response,
+                     chnk_per_m_se = SE)) %>%
+  bind_cols(per_m2_preds %>%
+              as_tibble() %>%
+              rename(chnk_per_m2 = response,
+                     chnk_per_m2_se = SE)) %>%
+  mutate_at(vars(matches('per_m')),
+            list(exp))
+
+rm(per_m_preds, per_m2_preds)
+
+all_preds = y %>%
+  mutate(model = 'CHaMP') %>%
+  bind_rows(z %>%
+              mutate(model = 'non-CHaMP'))
+
+# quick comparison of capacity predicitons with both models
+comp_pred_p = all_preds %>%
+  filter(Site %in% Site[duplicated(Site)]) %>%
+  select(Site, CHaMPsheds, Channel_Type, model, chnk_per_m, chnk_per_m2) %>%
+  arrange(CHaMPsheds, Site, model) %>%
+  gather(dens_type, cap, starts_with('chnk_per_m')) %>%
+  spread(model, cap) %>%
+  ggplot(aes(x = CHaMP,
+             y = `non-CHaMP`)) +
+  geom_point() +
+  geom_abline(linetype = 2,
+              color = 'red') +
+  facet_wrap(~ CHaMPsheds + dens_type,
+             scales = 'free')
+
+# for sites in CHaMP watersheds, use predicions from CHaMP extrapolation model
+all_preds %<>%
+  filter((Site %in% Site[duplicated(Site)] & model == 'CHaMP') |
+           !Site %in% Site[duplicated(Site)])
+
+```
+
+```{r save-extrapolation}
+
+usethis::use_data(gaa_covars,
+                  mod_data_weights,
+                  model_svy_df,
+                  all_preds,
+                  version = 2,
+                  overwrite = T)
+
+```

### 2019-12-27 analysis/paper/paper.Rmd
changed name of chunk for extrapolation r2 table

@@
 
-```{r extrap-table}
+```{r extrap-r2-table}
 mod_fit_df = model_svy_df %>%

### 2020-01-25 analysis/R_scripts/fit_qrf_model.Rmd
added to script to include capacity estimates where RST spawner-recruit data exists. Saved all objects in analysis/data/derived_data as .rds objects, rather than in R package data.

@@
 # save some results
-usethis::use_data(fish_hab, 
-                  hab_dict,
-                  hab_data,
-                  hab_avg,
-                  sel_hab_mets,
-                  qrf_data,
-                  dens_offset,
-                  qrf_mod,
-                  version = 2,
-                  overwrite = T)
+# usethis::use_data(fish_hab, 
+#                   hab_dict,
+#                   hab_data,
+#                   hab_avg,
+#                   sel_hab_mets,
+#                   qrf_data,
+#                   dens_offset,
+#                   qrf_mod,
+#                   version = 2,
+#                   overwrite = T)
+
+write_rds(fish_hab,
+          "../data/derived_data/fish_hab.rds")
+write_rds(hab_dict,
+          "../data/derived_data/hab_dict.rds")
+write_rds(hab_data,
+          "../data/derived_data/hab_data.rds")
+write_rds(sel_hab_mets,
+          "../data/derived_data/sel_hab_mets.rds")
+write_rds(qrf_data,
+          "../data/derived_data/qrf_data.rds")
+write_rds(dens_offset,
+          "../data/derived_data/dens_offset.rds")
+write_rds(qrf_mod,
+          "../data/derived_data/qrf_mod.rds")
 
@@
 
+# for CHaMP sites, use direct QRF esimates, not extrapolation ones
+all_preds %<>%
+  left_join(mod_data %>%
+              select(Site, 
+                     qrf_chnk_per_m = chnk_per_m,
+                     qrf_chnk_per_m2 = chnk_per_m2)) %>%
+  mutate(chnk_per_m = if_else(!is.na(qrf_chnk_per_m),
+                              qrf_chnk_per_m,
+                              chnk_per_m),
+         chnk_per_m_se = if_else(!is.na(qrf_chnk_per_m),
+                                 as.numeric(NA),
+                                 chnk_per_m_se),
+         chnk_per_m2 = if_else(!is.na(qrf_chnk_per_m2),
+                               qrf_chnk_per_m2,
+                               chnk_per_m2),
+         chnk_per_m2_se = if_else(!is.na(qrf_chnk_per_m2),
+                                  as.numeric(NA),
+                                  chnk_per_m2_se)) %>%
+  select(-starts_with('qrf'))
+
 ```
@@
 ```{r save-extrapolation}
-
-usethis::use_data(gaa_covars,
-                  mod_data_weights,
-                  model_svy_df,
-                  all_preds,
-                  version = 2,
-                  overwrite = T)
-
-```
+# usethis::use_data(gaa_covars,
+#                   mod_data_weights,
+#                   # model_svy_df,
+#                   all_preds,
+#                   version = 2,
+#                   overwrite = T)
+
+write_rds(gaa_covars,
+          "../data/derived_data/gaa_covars.rds")
+write_rds(mod_data_weights,
+          "../data/derived_data/mod_data_weights.rds")
+write_rds(model_svy_df,
+          "../data/derived_data/model_svy_df.rds")
+write_rds(all_preds,
+          "../data/derived_data/all_preds.rds")
+
+```
+
+# Validation
+
+We validated our estimates by comparing them with selected areas that had spawner-recruit data available.
+
+```{r}
+library(sf)
+data("chnk_domain")
+# read in trap polygons
+trap_poly = read_sf('../data/derived_data/trap_regions.shp') %>%
+  # st_transform(crs = 5070)
+  st_transform(st_crs(chnk_domain)) %>%
+  select(-Id)
+
+capacity_pts = all_preds %>%
+  st_as_sf(coords = c('Lon', 'Lat'),
+           crs = 4326) %>%
+  st_transform(st_crs(chnk_domain))
+
+qrf_caps = trap_poly %>%
+  split(list(.$trap_name)) %>%
+  map_df(.id = 'Trap',
+         .f = calc_watershed_cap,
+         spp_range = chnk_domain,
+         capacity_pts = capacity_pts,
+         capacity_name = "chnk_per_m",
+         capacity_se_name = 'chnk_per_m_se') %>%
+  mutate(tot_cap_cv = tot_cap_se / tot_cap) %>%
+  mutate(Trap = recode(Trap,
+                       'Minam R.' = 'Minam River',
+                       'Entiat R.' = 'Entiat River',
+                       'Crooked Fork Creek' = 'Crooked Fork',
+                       'Upper Grande Ronde R.' ='Upper Grande Ronde River',
+                       'Toucannon R.' = 'Tucannon River',
+                       'Pahsimeroi' = 'Pahsimeroi River',
+                       'Hood R.' = 'Hood River',
+                       'John Day Middle Fork' = 'Middle Fork John Day',
+                       'Upper Salmon River' = 'Upper Salmon River Chinook',
+                       'John Day Upper Mainstem' = 'Upper Mainstem John Day',
+                       'East Fork Salmon' = 'East Fork Salmon River',
+                       'Lostine R.' = 'Lostine River'))
+
+library(FSA)
+bh1 <- srFuns('BevertonHolt',
+              param = 3)
+
+data("spawn_rec_data")
+# fit Beverton-Holt curve with capacity fixed at QRF capacity prediction
+qrf_params = spawn_recr_data %>%
+  group_by(Population) %>%
+  nest() %>%
+  left_join(qrf_caps,
+            by = c('Population' = 'Trap')) %>%
+  filter(!is.na(tot_cap)) %>%
+  mutate(form = 'QRF',
+         inits = map2(.x = data,
+                      .y = tot_cap,
+                     .f = function(x, y) {
+                       inits = srStarts(Parr ~ Spawners,
+                                        data = x,
+                                        type = 'BevertonHolt',
+                                        param = 3)
+
+                       inits$a = if_else(inits$a < 0, 
+                                         2e-3, 
+                                         inits$a)
+                       inits$b = 1 / y
+                       return(inits)
+                       }),
+         mod_fit = map2(.x = data,
+                        .y = inits,
+                        .f = function(x, y) {
+                          fit = try(nls(log(Parr) ~ log(bh1(Spawners, a, b)),
+                                        data = x,
+                                        start = y,
+                                        algorithm = 'port',
+                                        lower = c(0, y['b']),
+                                        upper = c(Inf, y['b'])))
+                        }),
+         coefs = map(mod_fit,
+                     .f = function(x) {
+                       if(class(x) == 'try-error') {
+                         return(as.numeric(NA))
+                       } else coef(x)
+                       }),
+         preds = map2(.x = data,
+                      .y = mod_fit,
+                      .f = function(x, y) {
+                        if(class(y) != 'try-error') {
+                          tibble(Spawners = 1:max(x$Spawners)) %>%
+                            mutate(Parr = predict(y,
+                                                  newdata = .),
+                                   Parr = exp(Parr))
+                        } else return(tibble(Spawners = NA,
+                                             Parr = NA))
+                      })) %>%
+  rename(est = tot_cap,
+         se = tot_cap_se) %>%
+  select(-c(area:tot_length, tot_cap_cv)) %>%
+  mutate(cv = se / est)
+
+# combine with other spawner recruit estimates
+data("spawn_rec_params")
+
+spawn_rec_params %<>%
+  bind_rows(qrf_params)
+
+# save complete table
+write_rds(spawn_rec_params,
+          path = "../data/derived_data/spawn_rec_params.rds")
+
+```

### 2020-01-25 analysis/paper/paper.Rmd
added several tables and figures. re-knit the word version, and knit an html version as well.

--- 2019-12-24 analysis/paper/paper.Rmd
@@
 output: 
+    bookdown::html_document2:
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::pdf_document2:
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
       - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
       - --lua-filter=../templates/pagebreak.lua
@@
 bibliography:
-- /Users/kevin/Dropbox/Bibliography/Research.bib
-- /Users/kevin/Dropbox/Bibliography/SoftwareCitations.bib
+- /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/Research.bib
+- /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/SoftwareCitations.bib
 - packages.bib
@@
 ```{r packages}
+library(survey)
 library(tidyverse)
@@
 library(knitr)
+library(magrittr)
+
+theme_set(theme_bw())
 
@@
 
-```{r package-bibtex}
-knitr::write_bib(c("FSA", "Rcapture", "minerva", "quantregForest", "sf", "knitr", "rmarkdown"),
+```{r package-bibtex, eval = F}
+knitr::write_bib(c("FSA", 
+                   "Rcapture", 
+                   "minerva", 
+                   "quantregForest", 
+                   "missForest",
+                   "survey",
+                   "sf", 
+                   "knitr", 
+                   "rmarkdown"),
                  file = 'packages.bib')
@@
 ```{r load-data}
-data(fish_hab, 
-     hab_dict, 
-     hab_data, 
-     hab_avg, 
-     qrf_data,
-     dens_offset,
-     qrf_mod,
-     sel_hab_mets)
+fish_hab = read_rds("../data/derived_data/fish_hab.rds")
+hab_dict = read_rds("../data/derived_data/hab_dict.rds")
+qrf_data = read_rds("../data/derived_data/qrf_data.rds")
+dens_offset = read_rds("../data/derived_data/dens_offset.rds")
+qrf_mod = read_rds("../data/derived_data/qrf_mod.rds")
+sel_hab_mets = read_rds("../data/derived_data/sel_hab_mets.rds")
+gaa_covars = read_rds("../data/derived_data/gaa_covars.rds")
+mod_data_weights = read_rds("../data/derived_data/mod_data_weights.rds")
+all_preds = read_rds("../data/derived_data/all_preds.rds")
+spawn_rec_params = read_rds("../data/derived_data/spawn_rec_params.rds")
+model_svy_df = read_rds("../data/derived_data/model_svy_df.rds")
 ```
 
-```{r read-data, eval = F}
-# fish/habitat data for fitting QRF models
-data("fh_sum_champ_2017")
-fh_data = fh_sum_champ_2017 %>%
-  filter(Species == 'Chinook') %>%
-  mutate_at(vars(Watershed, Species, FishCrew),
-            list(fct_drop))
+```{r sp-rec-comp}
+# pull out estimates of capacity
+cap_tbl = spawn_rec_params %>%
+  select(Population, form, est, se, cv) %>%
+  arrange(Population, form)
 
-# data("fh_sum_champ_2014")
-# fh_data = fh_sum_champ_2014 %>%
-#   filter(Species == 'Chinook') %>%
-#   mutate_at(vars(Watershed, Species, FishCrew),
-#             list(fct_drop))
+# make a few estimates NA (too big)
+cap_tbl %<>%
+  mutate_at(vars(est, se, cv),
+            list(~ if_else(Population %in% c('Entiat River',
+                                             'Marsh Creek',
+                                             'Minam River',
+                                             'Tucannon River') &
+                             form == 'BevertonHolt',
+                           as.numeric(NA), .)))
 
-# habitat data used for predicting capacity
-data("champ_site_2011_17")
-hab_data = champ_site_2011_17 %>%
-  filter(!Watershed %in% c('Big-Navarro-Garcia (CA)',
-                           'CHaMP Training',
-                           'Region 17',
-                           'Walla Walla',
-                           'Umatilla'))
+# pull out fitted spawner recruit curves
+curve_fits = spawn_rec_params %>%
+  select(Population, form, preds) %>%
+  unnest(cols = "preds") %>%
+  arrange(Population, form)
 
-data("champ_site_2011_17_avg")
-hab_avg = champ_site_2011_17_avg
+# define polygons for uncertainty shading
+plot_max = Inf
+plot_max = 5e5
+cap_poly = cap_tbl %>%
+  mutate(lwrCI = est + se * qnorm(0.025),
+         uprCI = est + se * qnorm(0.975)) %>%
+  mutate(lwrCI = if_else(lwrCI < 0, 0, lwrCI),
+         uprCI = if_else(uprCI > plot_max, plot_max, uprCI)) %>%
+  left_join(spawn_rec_params %>%
+              select(Population, data) %>%
+              unnest(cols = "data") %>%
+              group_by(Population) %>%
+              summarise_at(vars(Spawners, Parr),
+                           list(min = min,
+                                max = max),
+                           na.rm = T)) %>%
+  # select(Population, form, ends_with('min'), ends_with('max'), ends_with('CI'))
+  group_by(Population, form) %>%
+  summarise(coords = list(tibble(x = c(0, 0, Spawners_max, Spawners_max),
+                                 y = c(lwrCI, uprCI, uprCI, lwrCI)))) %>%
+  ungroup() %>%
+  unnest(cols = c(coords))
 
-data("hab_dict_2017")
-hab_dict = hab_dict_2017
+# which watersheds to plot?
+# keep_wtsds = spawn_rec_params %>%
+#   filter(form == 'QRF',
+#          !is.na(est)) %>%
+#   filter(!Population %in% c('Methow R.')) %>%
+#   pull(Trap)
 
-```
+keep_wtsds = c('Catherine Creek',
+               'Chiwawa R.',
+               'Upper Grande Ronde River',
+               'Hayden Creek',
+               'Minam River',
+               'Tucannon River',
+               'Upper Lemhi')
 
-```{r load-qrf-model, eval = F}
-load('../data/derived_data/qrf_juv_summer.rda')
-qrf_mod = qrf_mods[[1]]
-qrf_data = qrf_mod_df %>%
-  filter(Species == 'Chinook')
+spawn_rec_comp_p = spawn_rec_params %>%
+  # filter out a couple populations with super poor fits
+  filter(Population %in% keep_wtsds) %>%
+  select(Population, data) %>%
+  unnest(cols = "data") %>%
+  distinct() %>%
+  ggplot(aes(x = Spawners,
+             y = Parr)) +
+  geom_polygon(data = cap_poly %>%
+                 # filter out a couple populations with super poor fits
+                 filter(Population %in% keep_wtsds),
+              aes(x = x,
+                  y = y,
+                  fill = form),
+              alpha = 0.2) +
+  geom_line(data = curve_fits %>%
+              # filter out a couple populations with super poor fits
+              filter(Population %in% keep_wtsds),
+            aes(color = form),
+            lwd = 1.5) +
+  geom_hline(data = cap_tbl %>%
+               # filter out a couple populations with super poor fits
+               filter(Population %in% keep_wtsds),
+             aes(yintercept = est,
+                 color = form),
+             linetype = 2) +
+  geom_point(size = 3) +
+  facet_wrap(~ Population,
+             scales = 'free') +
+  scale_color_brewer(palette = 'Set1',
+                     breaks = c('QRF',
+                                'BevertonHolt',
+                                'Ricker',
+                                'Hockey'),
+                     labels = c('QRF', 
+                                'Beverton Holt',
+                                'Ricker',
+                                'Hockey Stick'),
+                     name = 'Model') +
+  scale_fill_brewer(palette = 'Set1',
+                    breaks = c('QRF',
+                               'BevertonHolt',
+                               'Ricker',
+                               'Hockey'),
+                    labels = c('QRF', 
+                               'Beverton Holt',
+                               'Ricker',
+                               'Hockey Stick'),
+                    name = 'Model') +
+  labs(shape = 'Spawner\nType') +
+  theme(legend.position = 'bottom')
+
 ```
@@
 
-Fish and habitat data used in our study were collected from nine watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (Figure 1).
+Fish and habitat data used in our study were collected from nine watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (**Figure ??**).
 The CRB covers greater than 668,000 km$^2$ and drains large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
-The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (@CHaMP2016; @Volk2017] and were downloaded from https://www.champmonitoring.org.
+The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org.
 Data from the following `r n_distinct(hab_data$Watershed)` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`.
@@
 For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity.
-After removing duplicate samples, our initial dataset contained `r n_distinct(fh_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fh-samp-table)).
+After removing duplicate samples, our initial dataset contained `r n_distinct(fish_hab$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fh-samp-table)).
 
@@
 To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities.
-We calculated the maximal information coefficient (MIC), using the R package minerva [@Albanese2013], to measure the strength of the linear or non-linear association between two variables [@Reshef2011].
+We calculated the maximal information coefficient (MIC), using the R package minerva [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011].
 The MIC value between each of the measured habitat characteristics and the response variable, parr density (fish/m), was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
@@
 Habitat metrics measuring volume and area were scaled to the wetted area of each site.
-Within each category, metrics were ranked according to their MIC value (Figure 1).
+Within each category, metrics were ranked according to their MIC value (**Figure ??**).
 Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.).
-Additionally, we attempted to avoid covariates that were highly correlated (Figure 2) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
+Additionally, we attempted to avoid covariates that were highly correlated (***Figure ??***) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
 
@@
 Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months.
-QRF models combine the flexibility of random forest models (Breiman 2001) with the ability of quantile regression to extract relationships between quantiles of the data other than the mean (Cade and Noon 2003).
-Random forests also account for non-linear relationships between the response and predictor variables, and naturally incorporate interactions between the predictor variables, two common features of ecological datasets (Liaw and Wiener 2002).
+QRF models combine the flexibility of random forest models [@Breiman2001] with the ability of quantile regression to extract relationships between quantiles of the data other than the mean [@Cade2003].
+Random forests also account for non-linear relationships between the response and predictor variables, and naturally incorporate interactions between the predictor variables, two common features of ecological datasets [@Liaw2002].
 After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model.
@@
 Therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response.
-We fit the QRF models using the quantregForest function from the quantregForest package (Meinshausen 2016) in R software (R Core Team 2015).
-The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of Sweka and Mackey (2010).
+We fit the QRF models using the quantregForest function from the quantregForest package [@R-quantregForest] in R software [@Rsoftware2018].
+The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010.
 One reason for the 90th quantile, instead of something higher, is to avoid using predictions that are aimed at the very upper tails of observed fish density, where the variability of predictions may be influenced by sample size issues.
@@
 There were some missing values in the habitat dataset.
-Any site visit with more than three missing covariates was removed; the remaining missing values were imputed using the missForest R package (Stekhoven and Buehlmann 2012; Stekhoven 2013).
-Ultimately, the spring/summer Chinook parr capacity QRF model was fit using 186 records (paired fish-habitat data) and 14 habitat covariates (13.3 data points per covariate) (Table \@ref(tab:select-mets-table)).
+Any site visit with more than three missing covariates was removed; the remaining missing values were imputed using the missForest R package [@R-missForest; @Stekhoven2012; @Stekhoven2013].
+Ultimately, the spring/summer Chinook parr capacity QRF model was fit using `r nrow(qrf_data)` records (paired fish-habitat data) and `r nrow(sel_hab_mets)` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
 
@@
 For CHaMP sites that have been sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions.
-In total, we generated 485 predictions of spring/summer Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, Secesh, Tucannon, Yankee Fork and Wenatchee.
+In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of spring/summer Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Yankee Fork and Wenatchee.
 
@@
 To accomplish this, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) to scale our CHaMP site predictions to a series of larger spatial scales; often corresponding to the entirety of tributary habitat utilized by a given population.
-The GAA data used here was available from the list of GRTS master sample sites that the CHaMP sites were originally selected from.
+The GAA data used here was available from the list of GRTS master sample sites that the CHaMP sites were originally selected from (***need citation***).
 Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed.
-The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the svyglm function from the survey package (Lumley 2004, 2016) in R software (R Core Team 2015).
+The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the svyglm function from the survey package [@R-survey] in R software [@Rsoftware2018].
 Models with all possible combination of GAAs were fit and all the models with cumulative Akaike weights less than 0.95 were model averaged to make predictions of capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart.
 For sites outside of CHaMP watersheds, only extrapolation models that did not include watershed as a covariate were model averaged to make predictions.
-Summaries of extrapolation model fit are shown in Table 4.
+Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).
 
 To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
-We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook, as defined by StreamNet http://www.streamnet.org or using expert opinion from local biologists.
+We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists.
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
-However, for visualization, predictions of areal capacity (fish/m2) were used.
+However, for visualization, predictions of areal capacity (fish/m$^2$) were used.
 
@@
 Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts from rotary screw traps.
-Using estimates of over-winter survival, estimates of parr were calculated from those watersheds (Table 5).
+For the latter, estimates of parr were calculated using estimates of over-winter survival.
 A series of spawner-recruit functions were then fit to this data including Beverton-Hold, Ricker, and hockey stick.
-Estimates of capacity were made from each of these spawner-recruit curves and compared with QRF estimates of capacity (Table 5, Figure 4).
+Estimates of capacity were made from each of these spawner-recruit curves and compared with QRF estimates of capacity (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)).
 
@@
 
-We categorized 145 habitat measurements collected using the CHaMP habitat protocol (CHaMP 2016) into 11 habitat groups, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
-We chose the following 14 CHaMP habitat covariates to fit the QRF model: channel unit frequency, thalweg depth CV, average wetted width  to depth ratio, total fish cover, disturbance index, riparian ground cover, average thalweg depth, percent gravel, average substrate size (D50), percent substrate < 6mm, average hourly temperatures in the summer, maximum weekly average temperature, conductivity and frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).
+We categorized 145 habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)])` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
+We chose the following `r nrow(sel_hab_mets)` CHaMP habitat covariates to fit the QRF model: channel unit frequency, thalweg depth CV, average wetted width to depth ratio, total fish cover, disturbance index, riparian ground cover, average thalweg depth, percent gravel, average substrate size (D50), percent substrate < 6mm, average hourly temperatures in the summer, maximum weekly average temperature, conductivity and frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).
 
@@
 
-A QRF model was fit using 14 habitat covariates (Table \@ref(tab:select-mets-table)) and the quantregForest package (Meinshausen 2016) in R (R Core Team 2015) and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
-After model fit, we examined the relative importance of each habitat covariate included in the model (Figure 3).
+A QRF model was fit using 14 habitat covariates (Table \@ref(tab:select-mets-table)) and the quantregForest package [@Meinshausen2006] in R [@Rsoftware2018] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
+After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)).
 Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots (PDP).
-These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure 4).
+These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)).
 However, given that many habitat metrics are somewhat correlated, these marginal effects are not independent of one another making the interpretation of partial dependence plots less straightforward.
@@
   left_join(hab_dict) %>%
+  filter(MetricGroupName != 'Tier 1 Summary') %>%
   select(Rank,
@@
         format.args = list(width = c(2, 10, 10, 50)))
-  # pander(split.cells = c(2, 10, 10, 50))
+```
+
+```{r gaa-table}
+```
+
+```{r extrap-r2-table}
+mod_fit_df = model_svy_df %>%
+  select(response, data) %>%
+  unnest(cols = data) %>%
+  select(response, obs = qrf_cap) %>%
+  group_by(response) %>%
+  mutate(id = 1:n()) %>%
+  left_join(model_svy_df %>%
+              mutate(pred_full = map(mod_full,
+                                     .f = function(x) {
+                                       predict(x,
+                                               type = 'response') %>%
+                                         as_tibble() %>%
+                                         rename(pred_full = response,
+                                                pred_full_se = SE)
+                                     }),
+                     pred_no_champ = map(mod_no_champ,
+                                         .f = function(x) {
+                                           predict(x,
+                                                   type = 'response') %>%
+                                             as_tibble() %>%
+                                             rename(pred_no_champ = response,
+                                                    pred_no_champ_se = SE)
+                                         })) %>%
+              select(response, starts_with("pred")) %>%
+              unnest(cols = starts_with("pred")) %>%
+              group_by(response) %>%
+              mutate(id = 1:n())) %>%
+  ungroup()
+
+# mod_fit_df %>%
+#   group_by(response) %>%
+#   summarise(r2_full = cor(obs, pred_full)^2,
+#             r2_no_champ = cor(obs, pred_no_champ)^2) %>%
+#   gather(type, r2, -response) %>%
+#   mutate(model = if_else(grepl('no_champ', type),
+#                          'non-CHaMP',
+#                          'CHaMP'),
+#          type = if_else(grepl('adj', type),
+#                         'Adjusted R2',
+#                         'R2')) %>%
+#   spread(type, r2) %>%
+#   mutate(response = recode(response,
+#                            'chnk_per_m' = 'fish/m',
+#                            'chnk_per_m2' = 'fish/m^2')) %>%
+#   select(Model = model,
+#          Response = response,
+#          R2)
+
+mod_fit_tab = mod_fit_df %>%
+  group_by(response) %>%
+  nest() %>%
+  mutate(comp_full = map(data,
+                         .f = function(x) {
+                           lm(pred_full ~ obs,
+                              data = x)
+                         }),
+         comp_no_champ = map(data,
+                             .f = function(x) {
+                               lm(pred_no_champ ~ obs,
+                                  data = x)
+                             })) %>%
+  mutate(r2_full = map_dbl(comp_full,
+                      .f = function(x) summary(x)$r.squared),
+         r2_adj_full = map_dbl(comp_full,
+                           .f = function(x) summary(x)$adj.r.squared),
+         r2_no_champ = map_dbl(comp_no_champ,
+                           .f = function(x) summary(x)$r.squared),
+         r2_adj_no_champ = map_dbl(comp_no_champ,
+                               .f = function(x) summary(x)$adj.r.squared)) %>%
+  ungroup() %>%
+  select(response, starts_with("r2")) %>%
+  gather(type, r2, -response) %>%
+  mutate(model = if_else(grepl('no_champ', type),
+                         'non-CHaMP',
+                         'CHaMP'),
+         type = if_else(grepl('adj', type),
+                        'Adjusted R2',
+                        'R2')) %>%
+  spread(type, r2) %>%
+  mutate(response = recode(response,
+                           'chnk_per_m' = 'fish/m',
+                           'chnk_per_m2' = 'fish/m$^2$')) %>%
+  select(Model = model,
+         Response = response,
+         R2, `Adjusted R2`)
+
+names(mod_fit_tab)[c(3:4)] = c('$r^2$', 'Adjusted $r^2$')
+
+kable(mod_fit_tab,
+      escape = F,
+      caption = "Summary of model fit for each of the extrapolation models to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to larger scales (e.g., watershed, population).",
+      digits = c(NA, NA, 3, 3))
+
 ```
@@
 ```{r sr-table}
+spawn_rec_params %>%
+  select(Population, data) %>%
+  unnest(data) %>%
+  group_by(Population) %>%
+  summarise(`n Yrs` = n_distinct(BroodYear),
+            Spawner_type = unique(Spawner_type)) %>%
+  select(Population, `n Yrs`, Spawner_type) %>%
+  mutate(Spawner_type = recode(Spawner_type,
+                               'total spawners' = 'Spawners',
+                               'redd counts' = 'Redds')) %>%
+  rename(`Adult Data` = Spawner_type) %>%
+  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
+                               'Parr Surveys',
+                               'RST')) %>%
+  full_join(spawn_rec_params %>%
+              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
+              select(Population, form, prnt_val) %>%
+              mutate(form = recode(form,
+                                   'BevertonHolt' = 'Beverton Holt',
+                                   'Hockey' = 'Hockey Stick')) %>%
+              mutate(form = factor(form,
+                                   levels = c('Beverton Holt',
+                                              'Ricker',
+                                              'Hockey Stick',
+                                              'QRF'))) %>%
+              spread(form, prnt_val)) %>%
+  filter(Population %in% keep_wtsds) %>%
+  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
+Numbers in parentheses are coefficients of variation.")
 
@@
 ```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer Chinook salmon parr capacity quantile regression forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
-Tick marks along the X-axis depict observed values, and the sub-basin they came from."}
-plot_partial_dependence(qrf_mod,
-                        qrf_data,
-                        hab_dict,
-                        scales = 'free')
+Tick marks along the X-axis depict observed values."}
+plot_partial_dependence(rf_mod = qrf_mod,
+                        data = qrf_data,
+                        data_dict = hab_dict,
+                        type = 'quantile',
+                        pred_quantile = 0.9,
+                        log_transform = T,
+                        log_offset = dens_offset,
+                        scales = 'free') +
+  labs(y = 'Capacity Prediction (per m)') +
+  scale_color_grey(start = 0.5, end = 0.5) +
+  theme(legend.position = 'none')
+```
+
+```{r sr-figure, fig.cap = "Spawner-recruit data from seven watersheds.
+Solid lines are the spawner-recruit curve, dashed lines are the estimated capacity, and shaded polygons depict the 95% confidence intervals of capacity.
+Blue corresponds to Beverton-Holt models, green to Ricker models, yellow to hockey stick models, and red to QRF estimates.
+The QRF solid curve is a Beverton-Holt model with the capacity parameter fixed to the QRF estimate of capacity."}
+
+spawn_rec_comp_p
 ```

### 2020-01-25 analysis/R_scripts/fit_qrf_model.Rmd
added code to read in data about CHaMP frame from csv, but then save as .rds, so that it's smaller and can be pushed to GitHub.

@@
 # read in data from the CHaMP frame
-champ_frame_df = read_csv('../data/derived_data/champ_frame_data.csv') %>%
-  mutate(Target2014 = ifelse(is.na(AStrat2014), 'Non-Target', Target2014)) %>%
-  mutate(AStrat2014 = ifelse(AStrat2014 == 'Entiat IMW', paste('EntiatIMW', GeoRchIMW, sep = '_'), AStrat2014)) %>%
-  filter(Target2014 == 'Target') %>%
-  rename(Watershed = CHaMPshed)
+
+# read it in from .csv once, then save as .rds object (smaller in size)
+# champ_frame_df = read_csv('../data/derived_data/champ_frame_data.csv') %>%
+#   mutate(Target2014 = ifelse(is.na(AStrat2014), 'Non-Target', Target2014)) %>%
+#   mutate(AStrat2014 = ifelse(AStrat2014 == 'Entiat IMW', paste('EntiatIMW', GeoRchIMW, sep = '_'), AStrat2014)) %>%
+#   filter(Target2014 == 'Target') %>%
+#   rename(Watershed = CHaMPshed)
+# 
+# write_rds(champ_frame_df,
+#           "../data/derived_data/champ_frame_data.rds")
+
+champ_frame_df = read_rds("../data/derived_data/champ_frame_data.rds")
 

### 2020-01-27 analysis/paper/paper.Rmd
improved several tables.

@@
   # select(-percent) %>%
+  rename(`n Sites` = n,
+         Percent = percent) %>%
   kable(caption = 'The number of unique sites in the initial dataset, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity model')
@@
 ```{r gaa-table}
+tibble(Covariate = gaa_covars) %>%
+  mutate(Scale = if_else(Covariate %in% c('NatPrin1', 
+                                          'DistPrin1'),
+                         'Watershed (HUC12)',
+                         if_else(Covariate %in% c('WIDE_BF',
+                                                  'Channel_Type',
+                                                  'Elev_M'),
+                                 'Site (300 m)',
+                                 'Reach (2 km)'))) %>%
+  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population).")
 ```
 
-```{r extrap-r2-table}
+```{r, eval=F}
 mod_fit_df = model_svy_df %>%
@@
                          .f = function(x) {
-                           lm(pred_full ~ obs,
+                           lm(pred_full ~ obs - 1,
                               data = x)
@@
                              .f = function(x) {
-                               lm(pred_no_champ ~ obs,
+                               lm(pred_no_champ ~ obs - 1,
                                   data = x)
@@
       caption = "Summary of model fit for each of the extrapolation models to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to larger scales (e.g., watershed, population).",
+      digits = c(NA, NA, 3, 3))
+
+```
+
+```{r extrap-r2-table}
+model_svy_df %>%
+  ungroup() %>%
+  mutate(n_full = map_dbl(data,
+                          .f = nrow),
+         k_full = map_dbl(mod_full,
+                          .f = function(x) {
+                            names(x$coefficients)[!grepl('Intercept', names(x$coefficients))] %>%
+                              length
+                          }),
+         r_full = map_dbl(mod_full,
+                          .f = function(x) {
+                            cor(x$y, x$fitted.values)
+                          }),
+         r2_full = r_full^2,
+         r2_adj_full = 1 - ( ( (1 - r2_full) * (n_full - 1) ) / (n_full - k_full - 1) )) %>%
+  mutate(n_no_champ = map_dbl(data,
+                              .f = nrow),
+         k_no_champ = map_dbl(mod_no_champ,
+                              .f = function(x) {
+                                names(x$coefficients)[!grepl('Intercept', names(x$coefficients))] %>%
+                                  length
+                              }),
+         r_no_champ = map_dbl(mod_no_champ,
+                              .f = function(x) {
+                                cor(x$y, x$fitted.values)
+                              }),
+         r2_no_champ = r_no_champ^2,
+         r2_adj_no_champ = 1 - ( ( (1 - r2_no_champ) * (n_no_champ - 1) ) / (n_no_champ - k_no_champ - 1) )) %>%
+  select(response, starts_with("r2")) %>%
+  gather(type, r2, -response) %>%
+  mutate(model = if_else(grepl('no_champ', type),
+                         'non-CHaMP',
+                         'CHaMP'),
+         type = if_else(grepl('adj', type),
+                        'Adjusted R2',
+                        'R2')) %>%
+  spread(type, r2) %>%
+  mutate(response = recode(response,
+                           'chnk_per_m' = 'fish/m',
+                           'chnk_per_m2' = 'fish/m$^2$')) %>%
+  select(Model = model,
+         Response = response,
+         `$r^2$` = R2, 
+         `Adjusted $r^2$` = `Adjusted R2`) %>%
+  kable(caption = "Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.",
+        escape = F,
         digits = c(NA, NA, 3, 3))

### 2020-01-27 analysis/paper/paper.Rmd
incorporated most of Sarah Hoffman's suggestions into the manuscript. A few of them weren't relevant, because they referred to an older version of the manuscript.

@@
 abstract: |
-  Text of abstract
+  The QRF model presented here provided estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale.
+Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models.
+Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
 keywords: |
@@
 These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation.
-Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes might most cost-effectively and sustainably reduce mortality.
+Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality.
 Recently released “de-listing” criteria (NOAA 2016 ***need reference details***) identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required.
@@
 For perhaps the first time, the necessity of tributary habitat restoration can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”
+
+<!-- Need to define carrying capacity? -->
 
@@
 @Nickelson1992 found that juvenile coho were found in higher densities within pool habitat on the Oregon coast.
-@Sharma2001 determined that pool and pond densities were good predictors of coho smolt densities in western Washington.
-@Pess2002 demonstrated that densities of adult spawning coho were higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed.
+Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001].
+@Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale, however their results demonstrated a negative relationship between abundance and the number of pools.
+@Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools.
+Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002].
 @Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient.
-@Bryant2009 found a positive relationship between juvenile coho abundance and large wood at the reach scale, and a negative one between coho and the number of pools.
-@Braun2011 found positive associations between spawner densities of sockeye in the Fraser River and percent undercuts, large wood and percent pools.
-However, all of these studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed.
+These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed.
 In addition, they all assumed some form of linear fish-habitat relationship, which often results in weak predictive power.
 
-There have been a number of studies that utilized other modeling approaches to elicit fish habitat relationships.
+A number of studies have utilized other modeling approaches to elicit non-salmonid fish habitat relationships.
 @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density.
-@Eastwood2003 also utilized a quantile regression model to map the potential extent of sole in the English Channel and southern North Sea.
+The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003].
 Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
-The results from these studies highlight the importance of using techniques that can accommodate non-linear fish habitat relationships.
+The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships.
 
-Most studies that have investigated fish habitat relationships have focused on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity.
-Also, many of these studies focus on only one or two measures of habitat.
-@Sweka2010 did estimate carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
+Most studies that have investigated fish habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity.
+Further, many of these studies focus on only one or two measures of habitat.
+@Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
 Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves.
-However, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009], and it often fails to account for habitat covariates.
+However, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].
+
+<!-- and it often fails to account for habitat covariates.
 This manuscript incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity.
 Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.
+-->
 
@@
 Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006].
-Quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].
+Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].
 
-The fish abundance/density and habitat data used to fit the QRF model presented here were available from nine watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA.
+The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA.
 Within the interior CRB two major runs of Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics.
@@
 
-In this study, we developed a QRF model to 1) elicit fish-habitat relationships and 2) predict habitat rearing capacity at the reach scale (200 – 500 m) for juvenile (summer parr) stream-type Chinook salmon based on paired fish abundance/density and habitat data.
-Importantly, the QRF model places no constraints on possible fish-habitat relationships; instead, relationships are estimated from the data regardless of being positive, negative, linear, non-linear, etc.
-Based on observed fish-habitat relationships we can then predict capacity at the site scale using measurements of the habitat characteristics that were used to populate the model.
-We then extrapolated to larger spatial scales and validated our estimates of capacity at the watershed scale with those from spawner recruit curves from a number of watersheds around the interior Columbia basin.
-Moreover, the model provides managers a framework to guide the identification, prioritization, and development of habitat enhancement actions to recover salmon populations.
+In this study, we developed a QRF model to
+
+1. Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
+1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
+1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measures (i.e., CHaMP sites within the Columbia River Basin),
+1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
+1. Validate estimates of carrying capacity from QRF across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
+
+This manuscript incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity.
+Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.
 
@@
 Fish and habitat data used in our study were collected from nine watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (**Figure ??**).
-The CRB covers greater than 668,000 km$^2$ and drains large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
+The CRB covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
 The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org.
 Data from the following `r n_distinct(hab_data$Watershed)` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`.
-Juvenile density and abundance data collected at CHaMP survey reaches were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
+Juvenile density and abundance data were collected in a subset of `r n_distinct(fish_hab$Watershed)` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
 
@@
 CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin (CRB) selected based on a spatially balanced Generalized Random Tesselation Stratified (GRTS) sample selection algorithm [@Stevens1999; @Stevens2004].
-Habitat data within CHaMP sites are collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October.
+Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October.
 CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, size (depth, width, discharge), substrate, temperature, and water quality.
@@
 Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964.
-These estimates were made using the removal function from the FSA package [@R-FSA] or the closedp.bc function from the Rcapture package [@R-Rcapture] in R software [@Rsoftware2018].
+These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2018].
 Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts.
@@
 
-Abundance estimates at all sites were then translated into linear (parr/m) fish densities and density estimates were paired with the associated CHaMP habitat data.
+Abundance estimates at all sites were then translated into linear (parr/m) fish densities which were paired with the associated CHaMP habitat data.
 For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity.
@@
 
-A key step in developing a QRF model to predict fish capacities is selecting the habitat covariates to include in the model.
-Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another.
-However, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat).
+A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model.
+Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat).
 Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).
 
-The CHaMP protocol produces more than 100 metrics describing the quantity and quality of fish habitat for each survey site.
+To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site.
 To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities.
-We calculated the maximal information coefficient (MIC), using the R package minerva [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011].
-The MIC value between each of the measured habitat characteristics and the response variable, parr density (fish/m), was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
+We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011].
+The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
 
@@
 Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months.
-QRF models combine the flexibility of random forest models [@Breiman2001] with the ability of quantile regression to extract relationships between quantiles of the data other than the mean [@Cade2003].
-Random forests also account for non-linear relationships between the response and predictor variables, and naturally incorporate interactions between the predictor variables, two common features of ecological datasets [@Liaw2002].
 After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model.
-However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response.
-Therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response.
-We fit the QRF models using the quantregForest function from the quantregForest package [@R-quantregForest] in R software [@Rsoftware2018].
+However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response.
+There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012; @Stekhoven2013].
+We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2018], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r nrow(sel_hab_mets)` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
 The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010.
-One reason for the 90th quantile, instead of something higher, is to avoid using predictions that are aimed at the very upper tails of observed fish density, where the variability of predictions may be influenced by sample size issues.
-
-Summer parr density and habitat data were paired up by site and year.
-There were some missing values in the habitat dataset.
-Any site visit with more than three missing covariates was removed; the remaining missing values were imputed using the missForest R package [@R-missForest; @Stekhoven2012; @Stekhoven2013].
-Ultimately, the spring/summer Chinook parr capacity QRF model was fit using `r nrow(qrf_data)` records (paired fish-habitat data) and `r nrow(sel_hab_mets)` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
+We used the 90th quantile, rather than a higher quantile, to avoid using predictions that are aimed at the very upper tails of observed fish density, where the variability of predictions may be influenced by sample size issues.
 
@@
 For CHaMP sites that have been sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions.
-In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of spring/summer Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Yankee Fork and Wenatchee.
+In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of spring/summer Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.
 
@@
 
-Using the QRF model, we made predictions of habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin.
-However, fisheries biologists are more often interested in capacity at larger scales (e.g., watershed, population).
-To accomplish this, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) to scale our CHaMP site predictions to a series of larger spatial scales; often corresponding to the entirety of tributary habitat utilized by a given population.
-The GAA data used here was available from the list of GRTS master sample sites that the CHaMP sites were originally selected from (***need citation***).
-Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed.
-The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the svyglm function from the survey package [@R-survey] in R software [@Rsoftware2018].
-Models with all possible combination of GAAs were fit and all the models with cumulative Akaike weights less than 0.95 were model averaged to make predictions of capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart.
-For sites outside of CHaMP watersheds, only extrapolation models that did not include watershed as a covariate were model averaged to make predictions.
+Using the QRF model, we predicted habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin.
+To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
+The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from (***need citation***).
+Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)).
+The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2018].
+We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else.
+We then made predictions of linear and areal capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart.
 Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).
@@
 
-A QRF model was fit using 14 habitat covariates (Table \@ref(tab:select-mets-table)) and the quantregForest package [@Meinshausen2006] in R [@Rsoftware2018] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
+A QRF model was fit using `r nrow(sel_hab_mets)` habitat covariates (Table \@ref(tab:select-mets-table)) and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2018] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
 After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)).
@@
 
+Estimates of Chinook salmon parr capacity from the QRF and extrapolation models were comparable to independent estimates from spawner-recruit data (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)).
+QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the seven locations where comparisons were possible (Figure \@ref(fig:sr-figure)).
+Possible additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates.
+Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.905 (Beverton-Holt) to 0.965 (hockey stick).
+This favorable comparison provides strong validation, as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.
+
 # Discussion
 
-Using a quantile regression forest model, we predicted the summer rearing capacity for spring/summer Chinook parr at the reach scale.
-Quantile regression forests combine the benefits of random forest analyses (e.g.
-non-linear responses, correlated covariates) with quantile regression, which allows us to estimate how fish capacity, and not just observed fish density, changes with stream habitat characteristics.
-We then extrapolated those reach-scale estimates to watershed-scale estimates of capacity, which are more in line with the spatial scale at which management actions take place.
+## A Tool To Estimate Habitat Capacity
+
+In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams.
+Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act (ESA) delisting.
+The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat restoration actions to recover depleted salmon populations.
+Moreover, the models presented here can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult).
+Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life stage specific habitat is limiting.
+For example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required for ESA delisting, but that juvenile capacity may not be sufficient to support those levels of adult spawning.
+In such a case, habitat restoration actions may be most cost-effectively and sustainably directed towards juvenile rearing habitat.
+Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.
+
+## What Models Are Currently Available?
+
+Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity.
+The QRF approach employed here allows for the analysis of multiple correlated habitat metrics, often with non-linear relationships to fish density, in order to assess fish habitat relationships for higher quantiles that are assumed to represent carrying capacity.
+
+<!-- Talk about how the gold standard of carrying capacity estimates are spawner-recruit curves, and our validation exercise shows that QRF is getting similar answers.
+Then we can contrast the data needs for both approaches, and how spawner-recruit analyses don't tell you how capacity is related to habitat.
+-->
 
@@
 Model assumptions should always be considered when interpreting results; the following assumptions were made when modeling habitat capacity using QRF methods.
-First, we assume that sites with higher carrying capacity (i.e., ‘better’ habitat) contain more parr during fish surveys.
-Second, we assume that carrying capacity is a function of habitat.
-Third, we assume that estimates of parr densities from fish surveys are unbiased.
-Fourth, we assume that at least some sites surveys in our dataset are at or close to carrying capacity at the site level.
+First, we assumed that sites with higher carrying capacity (i.e., ‘better’ habitat) contain more parr during fish surveys.
+Second, we assumed that carrying capacity is a function of habitat.
+Third, we assumed that estimates of parr densities from fish surveys are unbiased.
+Fourth, we assumed that at least some sites in our dataset were at or close to local carrying capacity at the site level.
 However, if this fourth assumption is not met, we will obtain a conservative estimate of carrying capacity (but the framework of the model is not wrong).
-Moreover, we assume it is other unmeasured factors (not habitat) that are preventing a site from displaying fish densities near carrying capacity (e.g., insufficient adult spawners, competition, predation, etc.).
-Finally, we assume that the 90th quantile of predictions at each CHaMP site is a reasonable proxy for carrying capacity.
-With insufficient data, it is difficult to get a well-defined estimate of higher quantiles (i.e., lots of data is necessary to define the tails of a distribution).
+Finally, we assumed that the 90th quantile of predictions at each CHaMP site is a reasonable proxy for carrying capacity.
+With insufficient data, it is difficult to get a well-defined estimate of higher quantiles because substantial amounts of data are necessary to define the tails of a distribution.
 
@@
 Assumptions we don’t need to make.
-We make no assumptions regarding the shape of fish-habitat relationships.
-Rather fish-habitat relationships are estimated from the data.
-Relationships may be positive or negative, linear or non-linear; regardless, no constraints are put on the relationships.
+Because QRF naturally incorporates non-linear relationships between predictor and response variables, we make no assumptions regarding the shape of fish-habitat relationships.
+Rather, fish-habitat relationships are estimated from the data with no *a priori* constraints.
 Moreover, we make no assumptions regarding correlations among habitat covariates; the random forest framework accounts for possible interactions between habitat metrics, so we don’t need to explicitly place them in the model.
 
-## Extrapolation
+## Extrapolation Model
 
@@
 
-## Improving Habitat Data
+In addition to site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for analyses at larger extents, such as watershed and population level scales.
+This is an efficient technique to leverage existing relationships for meaningful management decisions.
+
+While the Columbia River Basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging.
+On-the-ground habitat data collection is costly, time-consuming, and often not reproducible; thereby limiting the effectiveness of this type of extrapolation.
+Additionally, the extrapolation model relies on Global Available Attribute (GAA) data which ….
+
+<!-- Add here why GAA are messy. And then anything else which may be a consideration for using the extrapolation model moving forward.  -->
+
+<!-- The following sentence doesn't make sense -->
+
+Despite these considerations, we were able to leverage extensive datasets to produce carrying capacity estimates throughout the CRB that significantly increase or habitat restoration prioritization in a cost-effective manner.
+
+## The Future: Improving Habitat Data
 
 <!-- Discussion of improving habitat data to possibly avoid extrapolation. Collect cheaper/better spatially continuous habitat data. -->
+
+<!-- I have no idea if this is what you are looking for so feel free to delete if I am way off base.  -->
+
+Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the CRB, there is a pressing need to develop new tools for habitat analyses.
+Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015].
+UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling.
+Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2017].
+
+Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly improve the broadscale application of QRF.
+Rapid advances drone technology further improves upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing.
+All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts.
+The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat restoration prioritization and planning.
 
@@
 # Acknowledgements
+
+Model development efforts have been funded by the Bonneville Power Administration through projects 2003-017-00 and 2011-006-00 and by the Bureau of Reclamation and Idaho Office of Species Conservation through contract BOR002 16.
+Fish sampling work in the Lemhi River was also funded through the Idaho Office of Species Conservation through the Pacific Coast Salmon Recovery Fund.
+Special thanks to staff from the following agencies for providing data to calculate Chinook salmon parr abundance and density estimates: Columbia River Inter-Tribal Fish Commission, Oregon Department of Fish and Wildlife, and U.S.
+Fish and Wildlife Service.
+The models in this study were improved by conversations with Eric Buhle.
 
@@
          Percent = percent) %>%
-  kable(caption = 'The number of unique sites in the initial dataset, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity model')
+  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model')
 ```
@@
 
+<!-- Need to flesh out this table with better names, and add units as a column -->
+
 ```{r gaa-table}
@@
   kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population).")
-```
-
-```{r, eval=F}
-mod_fit_df = model_svy_df %>%
-  select(response, data) %>%
-  unnest(cols = data) %>%
-  select(response, obs = qrf_cap) %>%
-  group_by(response) %>%
-  mutate(id = 1:n()) %>%
-  left_join(model_svy_df %>%
-              mutate(pred_full = map(mod_full,
-                                     .f = function(x) {
-                                       predict(x,
-                                               type = 'response') %>%
-                                         as_tibble() %>%
-                                         rename(pred_full = response,
-                                                pred_full_se = SE)
-                                     }),
-                     pred_no_champ = map(mod_no_champ,
-                                         .f = function(x) {
-                                           predict(x,
-                                                   type = 'response') %>%
-                                             as_tibble() %>%
-                                             rename(pred_no_champ = response,
-                                                    pred_no_champ_se = SE)
-                                         })) %>%
-              select(response, starts_with("pred")) %>%
-              unnest(cols = starts_with("pred")) %>%
-              group_by(response) %>%
-              mutate(id = 1:n())) %>%
-  ungroup()
-
-# mod_fit_df %>%
-#   group_by(response) %>%
-#   summarise(r2_full = cor(obs, pred_full)^2,
-#             r2_no_champ = cor(obs, pred_no_champ)^2) %>%
-#   gather(type, r2, -response) %>%
-#   mutate(model = if_else(grepl('no_champ', type),
-#                          'non-CHaMP',
-#                          'CHaMP'),
-#          type = if_else(grepl('adj', type),
-#                         'Adjusted R2',
-#                         'R2')) %>%
-#   spread(type, r2) %>%
-#   mutate(response = recode(response,
-#                            'chnk_per_m' = 'fish/m',
-#                            'chnk_per_m2' = 'fish/m^2')) %>%
-#   select(Model = model,
-#          Response = response,
-#          R2)
-
-mod_fit_tab = mod_fit_df %>%
-  group_by(response) %>%
-  nest() %>%
-  mutate(comp_full = map(data,
-                         .f = function(x) {
-                           lm(pred_full ~ obs - 1,
-                              data = x)
-                         }),
-         comp_no_champ = map(data,
-                             .f = function(x) {
-                               lm(pred_no_champ ~ obs - 1,
-                                  data = x)
-                             })) %>%
-  mutate(r2_full = map_dbl(comp_full,
-                      .f = function(x) summary(x)$r.squared),
-         r2_adj_full = map_dbl(comp_full,
-                           .f = function(x) summary(x)$adj.r.squared),
-         r2_no_champ = map_dbl(comp_no_champ,
-                           .f = function(x) summary(x)$r.squared),
-         r2_adj_no_champ = map_dbl(comp_no_champ,
-                               .f = function(x) summary(x)$adj.r.squared)) %>%
-  ungroup() %>%
-  select(response, starts_with("r2")) %>%
-  gather(type, r2, -response) %>%
-  mutate(model = if_else(grepl('no_champ', type),
-                         'non-CHaMP',
-                         'CHaMP'),
-         type = if_else(grepl('adj', type),
-                        'Adjusted R2',
-                        'R2')) %>%
-  spread(type, r2) %>%
-  mutate(response = recode(response,
-                           'chnk_per_m' = 'fish/m',
-                           'chnk_per_m2' = 'fish/m$^2$')) %>%
-  select(Model = model,
-         Response = response,
-         R2, `Adjusted R2`)
-
-names(mod_fit_tab)[c(3:4)] = c('$r^2$', 'Adjusted $r^2$')
-
-kable(mod_fit_tab,
-      escape = F,
-      caption = "Summary of model fit for each of the extrapolation models to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to larger scales (e.g., watershed, population).",
-      digits = c(NA, NA, 3, 3))
-
 ```

### 2020-01-28 analysis/R_scripts/compile_wtsd_bndry.Rmd
created R notebook to read in CHaMP watershed boundaries and save them as .rds object.

---
title: "Create CHaMP map"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

# load packages for analysis
library(tidyverse)
library(janitor)
library(magrittr)
library(sf)
library(ggrepel)

# set default theme for ggplot
theme_set(theme_bw())
# setwd('analysis/R_scripts')
```

Start by looking at which watersheds we need polygon shapes for.

```{r}
fish_hab = read_rds("../data/derived_data/fish_hab.rds")
model_svy_df = read_rds("../data/derived_data/model_svy_df.rds")
```

Watersheds with paired fish-habitat data:

```{r}
tabyl(fish_hab,
      Watershed) %>%
  adorn_pct_formatting()
```

Watersheds with CHaMP habitat data

```{r}
model_svy_df$data[[1]] %>%
  tabyl(Watershed) %>%
  adorn_pct_formatting()
```

Read in some HUC boundaries.

```{r}
huc3 = read_sf('~/OneDrive - Merck Sharp & Dohme, Corp/Data/WatershedBoundaries/WBDHU6.shp') %>%
  select(STATES:NAME) %>%
  st_transform(crs = 5070)
huc4 = read_sf('~/OneDrive - Merck Sharp & Dohme, Corp/Data/WatershedBoundaries/WBDHU8.shp') %>%
  select(STATES:NAME) %>%
  st_transform(st_crs(huc3))
huc5 = read_sf('~/OneDrive - Merck Sharp & Dohme, Corp/Data/WatershedBoundaries/WBDHU10.shp') %>%
  select(STATES:NAME) %>%
  st_transform(st_crs(huc3))

# filter(huc4,
#        grepl('Methow', NAME)) %>%
#   select(NAME) %>% plot()

champ_wtsds = huc3 %>%
  filter(NAME %in% c('John Day')) %>%
  select(NAME) %>%
  rbind(huc4 %>%
          filter(NAME %in% c('Lemhi',
                             'Methow',
                             'Wenatchee',
                             'South Fork Salmon',
                             'Upper Grande Ronde')) %>%
          select(NAME)) %>%
  rbind(huc5 %>%
          filter(NAME %in% c('Minam River',
                             'Secesh River',
                             'Yankee Fork')) %>%
          select(NAME) %>%
          mutate(NAME = str_remove(NAME, ' River$'))) %>%
  rbind(huc5 %>%
          filter(NAME %in% c('Entiat River',
                             'Mad River')) %>%
          st_union() %>%
          st_sf() %>%
          mutate(NAME = 'Entiat')) %>%
  rbind(huc5 %>%
          filter(grepl('Tucannon', NAME)) %>%
          st_union() %>%
          st_sf() %>%
          mutate(NAME = 'Tucannon')) %>%
  arrange(NAME) %>%
  mutate(id = 1:n(),
         hab_data = T,
         fish_data = if_else(NAME %in% c('Secesh', as.character(unique(fish_hab$Watershed))),
                             T, F),
         fish_data = if_else(NAME == 'South Fork Salmon', F, fish_data))

```

Test a plot

```{r}
ggplot(data = champ_wtsds) +
  geom_sf() +
  geom_sf(data = champ_wtsds %>%
            filter(fish_data),
          fill = 'black') +
  geom_sf_label(aes(label = id),
                position = position_dodge(width = 200)) +
                # position = position_nudge(x = 5000,
                #                           y = -5000)) +
  theme(axis.title = element_blank())

```

Looks good. Let's save the polygon sf to use later.

```{r}
write_rds(champ_wtsds,
          path = '../data/derived_data/champ_wtds.rds')
```

### 2020-01-28 analysis/paper/paper.Rmd
copied all references into a local .bib file. Built a map for figure 1.

@@
       - --lua-filter=../templates/pagebreak.lua
-# bibliography: references.bib
 bibliography:
-- /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/Research.bib
-- /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/SoftwareCitations.bib
+# - /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/Research.bib
+# - /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/SoftwareCitations.bib
+- references.bib
 - packages.bib
@@
 library(magrittr)
+library(sf)
 
@@
 model_svy_df = read_rds("../data/derived_data/model_svy_df.rds")
+champ_wtsds = read_rds("../data/derived_data/champ_wtds.rds")
 ```
@@
 
-Fish and habitat data used in our study were collected from nine watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (**Figure ??**).
+Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)).
 The CRB covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
@@
 Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964.
-These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2018].
+These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2019].
 Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts.
@@
 Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.).
-Additionally, we attempted to avoid covariates that were highly correlated (***Figure ??***) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
+Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
 
@@
 However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response.
-There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012; @Stekhoven2013].
-We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2018], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r nrow(sel_hab_mets)` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
+There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012].
+We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r nrow(sel_hab_mets)` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
 The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010.
@@
 Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)).
-The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2018].
+The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019].
 We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else.
@@
 
-A QRF model was fit using `r nrow(sel_hab_mets)` habitat covariates (Table \@ref(tab:select-mets-table)) and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2018] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
+A QRF model was fit using `r nrow(sel_hab_mets)` habitat covariates (Table \@ref(tab:select-mets-table)) and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
 After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)).
@@
 UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling.
-Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2017].
+Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].
 
@@
 # Tables
+
+<div id="tables"></div>
 
@@
 
-```{r gaa-table}
+```{r gaa-covars}
 tibble(Covariate = gaa_covars) %>%
@@
 
-<!-- We need to make a map of the Columbia River basin that includes polygons for each of the CHaMP watersheds. -->
+<div id="figures"></div>
+
+```{r wtsd-map, fig.cap = "Watersheds with CHaMP habitat data.
+Watersheds in black also contain paired fish data.
+Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
+ggplot(data = champ_wtsds) +
+  geom_sf() +
+  geom_sf(data = champ_wtsds %>%
+            filter(fish_data),
+          fill = 'black') +
+  geom_sf_label(aes(label = id),
+                position = position_dodge(width = 200)) +
+                # position = position_nudge(x = 5000,
+                #                           y = -5000)) +
+  theme(axis.title = element_blank())
+```
 

### 2020-01-28 README.Rmd
changed the README to include author names, tenative title, publication year, and journal we're aiming for.

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Estimating Life-Stage Specific Habitat Capacity for Anadromous Salmonids using Quantile Random Forest Models"
+Authors <- "See, Kevin E., Michael W. Ackerman, Richard A. Carmichael, Sarah Hoffman, and Chris Beasley"
+yr <- 2020
+Journal <- "Ecological Modelling"
 ```
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> `r Authors` (`r yr`). _`r Title`_. `r Journal` <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> `r Authors` (`r yr`). _`r Title`_. `r Journal`, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> `r Authors`, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 

### 2020-01-28 analysis/paper/paper.Rmd
added a note to remember to add more to the abstract, and a comment to think about in the conclusions.

@@
 abstract: |
+  Needs more...
+
   The QRF model presented here provided estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale.
@@
 
+<!-- Consider pulling back and thinking about whether these kinds of techniques could be applied to other systems (terrestial, ocean, etc.).
+If not, we may want to target a more fisheries-centric journal -->
+
 # Acknowledgements

### 2020-01-29 analysis/paper/paper.Rmd
re-knit as word doc. Fixed the file that the bibliography points to.

@@
 output: 
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
       - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
       - --lua-filter=../templates/pagebreak.lua
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
+    bookdown::html_document2:
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
       - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
 bibliography:
-# - /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/Research.bib
-# - /Users/seek/OneDrive - Merck Sharp & Dohme, Corp/Bibliography/SoftwareCitations.bib
 - references.bib

### 2020-01-30 analysis/paper/paper.Rmd
Kevin added something to the discussion.
I undid the random comment I made.
deleted one section in discussion.

@@
 
-## What Models Are Currently Available?
+## What Watershed-scale Capacity Models Are Currently Available?
 
@@
 
+<!-- Somewhere in here, mention how data can be collected over a short time span, if enough spatial variation is captured.
+Or dataset can be built over time, sampling different locations (but be vary of possible year effects, maybe).
+-->
+
 ## Biological Expectations from QRF Model
@@
 
-## Model Assumptions
-
-<!-- Somewhere in here, mention how data can be collected over a short time span, if enough spatial variation is captured.
-Or dataset can be built over time, sampling different locations (but be vary of possible year effects, maybe).
--->
-
-Model assumptions should always be considered when interpreting results; the following assumptions were made when modeling habitat capacity using QRF methods.
-First, we assumed that sites with higher carrying capacity (i.e., ‘better’ habitat) contain more parr during fish surveys.
-Second, we assumed that carrying capacity is a function of habitat.
-Third, we assumed that estimates of parr densities from fish surveys are unbiased.
-Fourth, we assumed that at least some sites in our dataset were at or close to local carrying capacity at the site level.
-However, if this fourth assumption is not met, we will obtain a conservative estimate of carrying capacity (but the framework of the model is not wrong).
-Finally, we assumed that the 90th quantile of predictions at each CHaMP site is a reasonable proxy for carrying capacity.
-With insufficient data, it is difficult to get a well-defined estimate of higher quantiles because substantial amounts of data are necessary to define the tails of a distribution.
-
-<!-- This paragraph needs to be hashed out.	 -->
-Assumptions we don’t need to make.
-Because QRF naturally incorporates non-linear relationships between predictor and response variables, we make no assumptions regarding the shape of fish-habitat relationships.
-Rather, fish-habitat relationships are estimated from the data with no *a priori* constraints.
-Moreover, we make no assumptions regarding correlations among habitat covariates; the random forest framework accounts for possible interactions between habitat metrics, so we don’t need to explicitly place them in the model.
+<!-- Try to include images of high capacity / low capacity sites  -->
 

### 2020-01-30 analysis/paper/paper.Rmd
test commit from Sarah's machine. Added comment at start of discussion.
one more test, the first push required credentials
Tried using pander instead of kable to make tables, had to figure out how to get the cross-referenced labels into the pander function (doesn't do it by default, like kable)

@@
 output: 
-    bookdown::word_document2:
+    bookdown::pdf_document2:
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::pdf_document2:
+    bookdown::html_document2:
       fig_caption: yes
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
@@
 library(knitr)
+library(pander)
 library(magrittr)
@@
 theme_set(theme_bw())
+
+panderOptions("big.mark", ',')
+
+options(knitr.kable.NA = '-')
+options(knitr.table.format = "pandoc")
 
@@
 
-```{r fh-samp-table}
+```{r }
 qrf_data %>%
-  # filter(Year <= 2014) %>%
   tabyl(Watershed) %>%
@@
   adorn_pct_formatting() %>%
-  # select(-percent) %>%
   rename(`n Sites` = n,
          Percent = percent) %>%
-  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model')
+  pander(caption = "(\\#tab:fh-samp-table)The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model.")
+  # kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model',
+  #       booktabs = T)
+
 ```
@@
          Description = DescriptiveText) %>%
-  kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
+  pander(caption = "(\\#tab:select-mets-table) Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
+Metrics are ranked in order of relative importance.")
+  # kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
 Metrics are ranked in order of relative importance.",
-        format.args = list(width = c(2, 10, 10, 50)))
+  #       booktabs = T,
+  #       format.args = list(width = c(2, 10, 10, 50)))
+
 ```
@@
                                  'Reach (2 km)'))) %>%
-  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population).")
+  pander(caption = "(\\#tab:gaa-covars)Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population).")
+  # kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population).",
+  #       booktabs = T)
 ```
@@
          `Adjusted $r^2$` = `Adjusted R2`) %>%
-  kable(caption = "Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.",
-        escape = F,
-        digits = c(NA, NA, 3, 3))
+  pander(caption = "(\\#tab:extrap-r2-table)Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.")
+  # kable(caption = "Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.",
+  #       escape = F,
+  #       booktabs = T,
+  #       digits = c(NA, NA, 3, 3))
 
@@
   filter(Population %in% keep_wtsds) %>%
-  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
+  pander(caption = "(\\#tab:sr-table)Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
+Numbers in parentheses are coefficients of variation.")
+  # kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
 Numbers in parentheses are coefficients of variation.")

### 2020-01-31 analysis/paper/paper.Rmd
added some river lines to the watershed map plot

@@
 
+<!-- A test commit and push from Sarah's machine -->
+<!-- need to do one more test -->
+
 ## A Tool To Estimate Habitat Capacity
@@
 Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
+river_lines = QRFcapacity::get_flowlines(stream_order = 6,
+                                         champ_wtsds %>%
+                                           st_union() %>%
+                                           st_make_grid(n = 1)) %>%
+            filter(!grepl('^1701', reachcode),
+                   !grepl('^171', reachcode))
+
+# river_lines %>%
+#   filter(!gnis_name %in% c('Deschutes River',
+#                            'Crooked River',
+#                            'Payette River',
+#                            'North Fork Payette River')) %>%
+#   ggplot() +
+#   geom_sf(aes(color = gnis_name))
+
 ggplot(data = champ_wtsds) +
+  geom_sf(data = river_lines %>%
+            filter(!gnis_name %in% c('Deschutes River',
+                                     'Crooked River',
+                                     'Payette River',
+                                     'North Fork Payette River')),
+          color = 'darkblue') +
   geom_sf() +
@@
                 position = position_dodge(width = 200)) +
-                # position = position_nudge(x = 5000,
-                #                           y = -5000)) +
   theme(axis.title = element_blank())
+
 ```

### 2020-01-31 analysis/paper/paper.Rmd
Fixed Sarah's name

@@
       correspondence: false
-  - Sarah Hoffman:
+  - Sarah L. Hoffmann:
       email: Sarah.Hoffman@biomark.com

### 2020-01-31 analysis/paper/paper.Rmd
.... and email

@@
   - Sarah L. Hoffmann:
-      email: Sarah.Hoffman@biomark.com
+      email: Sarah.Hoffmann@biomark.com
       institute: [biomark]

### 2020-02-04 analysis/R_scripts/fit_qrf_model.Rmd
revised the fit_qrf_model script, including choosing a different subset of metrics. Re-saved all the .rds objects.

@@
 Currently, we are excluding sites outside the Chinook domain.
-We created our own list of sites in the John Day that were within a Chinook domain, based on which sites Chinook were found at some point (and downstream of those).
 
@@
   as_Spatial() %>%
-  snapPointsToLines_v2(points = .,
-  # maptools::snapPointsToLines(points = .,
+  maptools::snapPointsToLines(points = .,
                               lines = chnk_domain %>%
@@
   as('sf') %>%
-  select(-nearest_line_id, -snap_dist) %>%
-  # include some sites in the John Day where Chinook were found (or seemed to be close to sites where Chinook were found)
-  rbind(st_read('../data/derived_data/Chnk_JohnDay_TrueObs.shp',
-                quiet = T) %>%
-          st_transform(st_crs(chnk_domain)) %>%
-          select(-in_range))
+  select(-nearest_line_id, -snap_dist)
 
@@
   filter(Site %in% chnk_sites$Site)
-  # filter(Species == 'Steelhead' |
-  #          (Species == 'Chinook' & Site %in% chnk_sites$Site))
 
@@
 sel_hab_mets = tibble(Species = c('Chinook'),
-                      Metric = c('UcutArea_Pct',
-                                 'FishCovNone',
-                                 'SubEstGrvl',
-                                 'FstTurb_Freq',
-                                 'FstNT_Freq',
-                                 'CU_Freq',
-                                 'SlowWater_Pct',
-                                 'NatPrin1',
-                                 'DistPrin1',
-                                 'avg_aug_temp',
-                                 # 'Sin_CL',
-                                 'Sin',
-                                 'WetWdth_CV',
-                                 'WetBraid',
-                                 'WetSC_Pct',
+                      Metric = c("WetWDRat_Avg",
+                                 "CU_Freq",
+                                 "DpthWet_SD",
+                                 "WetBraid",
+                                 "FishCovTotal",
+                                 "RipCovUstory",
+                                 'WetWdth_Int',
                                  'Q',
-                                 'WetWdth_Int',
-                                 'LWFreq_Wet',
-                                 'LWVol_WetFstTurb'))
+                                 "SubLT6",
+                                 "SubD16",
+                                 "avg_aug_temp",
+                                 "LWFreq_Wet"))
+                      # 
+                      # Metric = c('UcutArea_Pct',
+                      #            'FishCovNone',
+                      #            'SubEstGrvl',
+                      #            'FstTurb_Freq',
+                      #            'FstNT_Freq',
+                      #            'CU_Freq',
+                      #            'SlowWater_Pct',
+                      #            'NatPrin1',
+                      #            'DistPrin1',
+                      #            'avg_aug_temp',
+                      #            # 'Sin_CL',
+                      #            'Sin',
+                      #            'WetWdth_CV',
+                      #            'WetBraid',
+                      #            'WetSC_Pct',
+                      #            'Q',
+                      #            'WetWdth_Int',
+                      #            'LWFreq_Wet',
+                      #            'LWVol_WetFstTurb'))
 
@@
   as_Spatial() %>%
-  snapPointsToLines_v2(chnk_domain %>%
-  # maptools::snapPointsToLines(chnk_domain %>%
+  maptools::snapPointsToLines(chnk_domain %>%
                                 mutate(id = 1:n()) %>%
@@
   as_tibble() %>%
-  select(-nearest_line_id, -snap_dist, -geometry) %>%
-  # add sites in the John Day
-  bind_rows(st_read('../data/derived_data/Chnk_JohnDay_TrueObs.shp',
-                    quiet = T) %>%
-              as_tibble() %>%
-              select(Site) %>%
-              inner_join(pred_hab_sites))
+  select(-nearest_line_id, -snap_dist, -geometry)
 
@@
                          gather(Metric, value, -Site) %>%
-                         mutate(Source = 'non-DASH Sites'),
+                         mutate(Source = 'non-CHaMP Sites'),
                        gaa_all %>%
@@
                          gather(Metric, value, -Site) %>%
-                         mutate(Source = 'DASH Sites')) %>%
+                         mutate(Source = 'CHaMP Sites')) %>%
   mutate_at(vars(Source, Metric),
@@
                na.rm = T) %>%
-  filter(Source == 'DASH Sites') %>%
+  filter(Source == 'CHaMP Sites') %>%
   ungroup() %>%
@@
                             na.rm = T) %>%
-               filter(Source == 'DASH Sites') %>%
+               filter(Source == 'CHaMP Sites') %>%
                ungroup() %>%

### 2020-02-04 analysis/paper/paper.Rmd
added a few more watersheds to the spawner/recruit validation, saved the associated figure as an .svg file.
Used "english" package to translate some numbers into written words.

Re-knit file (overwrote saved figures)

--- 2020-01-31 analysis/paper/paper.Rmd
@@
 output: 
+    bookdown::html_document2:
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
@@
 library(sf)
+library(english)
 
@@
 
+# run this if running chunks directly
 # setwd('analysis/paper')
@@
 # which watersheds to plot?
-# keep_wtsds = spawn_rec_params %>%
-#   filter(form == 'QRF',
-#          !is.na(est)) %>%
-#   filter(!Population %in% c('Methow R.')) %>%
-#   pull(Trap)
+keep_wtsds = spawn_rec_params %>%
+  filter(form == 'QRF',
+         !is.na(est)) %>%
+  filter(!Population %in% c('Methow R.')) %>%
+  pull(Population)
 
@@
                'Tucannon River',
-               'Upper Lemhi')
+               'Upper Lemhi',
+               'Lostine River',
+               'South Fork Salmon River')
 
@@
              y = Parr)) +
-  geom_polygon(data = cap_poly %>%
-                 # filter out a couple populations with super poor fits
-                 filter(Population %in% keep_wtsds),
-              aes(x = x,
-                  y = y,
-                  fill = form),
-              alpha = 0.2) +
+  # geom_polygon(data = cap_poly %>%
+  #                # filter out a couple populations with super poor fits
+  #                filter(Population %in% keep_wtsds),
+  #             aes(x = x,
+  #                 y = y,
+  #                 fill = form),
+  #             alpha = 0.2) +
   geom_line(data = curve_fits %>%
@@
   theme(legend.position = 'bottom')
+
+# # to save a different version
+# ggsave('../figures/sr-test.svg',
+#        spawn_rec_comp_p,
+#        width = 8,
+#        height = 6)
 
@@
 The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org.
-Data from the following `r n_distinct(hab_data$Watershed)` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`.
-Juvenile density and abundance data were collected in a subset of `r n_distinct(fish_hab$Watershed)` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
+Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`.
+Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
 
@@
 There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012].
-We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r nrow(sel_hab_mets)` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
+We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
 The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010.
@@
 
-We categorized 145 habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)])` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
-We chose the following `r nrow(sel_hab_mets)` CHaMP habitat covariates to fit the QRF model: channel unit frequency, thalweg depth CV, average wetted width to depth ratio, total fish cover, disturbance index, riparian ground cover, average thalweg depth, percent gravel, average substrate size (D50), percent substrate < 6mm, average hourly temperatures in the summer, maximum weekly average temperature, conductivity and frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).
+We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
+We chose the following `r nrow(sel_hab_mets)` CHaMP habitat covariates to fit the QRF model: wetted width:depth ratio, channel unit frequency, standard deviation of the wetted depth, braidedness, total fish cover, percent riparian understory, wetted width, observed discharge, percent fines less than 6mm, lower quantile of substrate size (D16), average august temperature and the frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).
 
@@
 
-A QRF model was fit using `r nrow(sel_hab_mets)` habitat covariates (Table \@ref(tab:select-mets-table)) and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
+A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
 After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)).

### 2020-02-04 analysis/paper/paper.Rmd

@@
       correspondence: false
-  - Sarah Hoffman:
-      email: Sarah.Hoffman@biomark.com
+  - Sarah L. Hoffmann:
+      email: Sarah.Hoffmann@biomark.com
       institute: [biomark]

### 2020-02-05 analysis/R_scripts/compile_wtsd_bndry.Rmd
added chunk to make a map that can be saved and imported into the manuscript.

@@
 ```
+
+Now let's make a map that we can import into the manuscript
+
+```{r}
+champ_wtsds = read_rds('../data/derived_data/champ_wtds.rds')
+river_lines = QRFcapacity::get_flowlines(stream_order = 6,
+                                         champ_wtsds %>%
+                                           st_union() %>%
+                                           st_make_grid(n = 1)) %>%
+            filter(!grepl('^1701', reachcode),
+                   !grepl('^171', reachcode))
+
+# river_lines %>%
+#   filter(!gnis_name %in% c('Deschutes River',
+#                            'Crooked River',
+#                            'Payette River',
+#                            'North Fork Payette River')) %>%
+#   ggplot() +
+#   geom_sf(aes(color = gnis_name))
+
+wtsd_map = ggplot(data = champ_wtsds) +
+  geom_sf(data = river_lines %>%
+            filter(!gnis_name %in% c('Deschutes River',
+                                     'Crooked River',
+                                     'Payette River',
+                                     'North Fork Payette River')),
+          color = 'darkblue') +
+  geom_sf() +
+  geom_sf(data = champ_wtsds %>%
+            filter(fish_data),
+          fill = 'black') +
+  geom_sf_label(aes(label = id),
+                position = position_dodge(width = 200)) +
+  theme(axis.title = element_blank())
+
+# save a couple versions
+ggsave('../figures/ChampWtds.svg',
+       wtsd_map,
+       height = 6,
+       width = 6)
+ggsave('../figures/ChampWtds.png',
+       wtsd_map,
+       height = 6,
+       width = 6)
+```

### 2020-02-05 analysis/R_scripts/fit_qrf_model.Rmd
traded riparian cover: understory for riparian cover: big trees in the QRF model. Re-ran whole script.

@@
                                  "FishCovTotal",
-                                 "RipCovUstory",
+                                 # "RipCovUstory",
+                                 "RipCovBigTree",
                                  'WetWdth_Int',

### 2020-02-05 analysis/paper/paper.Rmd
changed title. Added some references to NREI studies. Revised discussion section on other capacity models. Wrote discussion section on biological expectations, and extrapolation models.

@@
 ---
-title: "Estimating Life-Stage Specific Habitat Capacity for Anadromous Salmonids using Quantile Random Forest Models"
+title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
 author:
@@
 Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
+@Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead.
 The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships.
@@
 We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
-We chose the following `r nrow(sel_hab_mets)` CHaMP habitat covariates to fit the QRF model: wetted width:depth ratio, channel unit frequency, standard deviation of the wetted depth, braidedness, total fish cover, percent riparian understory, wetted width, observed discharge, percent fines less than 6mm, lower quantile of substrate size (D16), average august temperature and the frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).
+We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width:depth ratio, channel unit frequency, standard deviation of the wetted depth, braidedness, total fish cover, percent riparian canopy that was big trees, wetted width, observed discharge, percent fines less than 6mm, lower quantile of substrate size (D16), average august temperature and the frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).
 
@@
 These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)).
-However, given that many habitat metrics are somewhat correlated, these marginal effects are not independent of one another making the interpretation of partial dependence plots less straightforward.
+However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another, so care should be taken when interpretting them.
 After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.
@@
 ## Extrapolating to Other Sites
+
+The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars).
+
+<!-- Probably need something more here -->
 
@@
 # Discussion
-
-<!-- A test commit and push from Sarah's machine -->
-<!-- need to do one more test -->
 
@@
 
-## What Watershed-scale Capacity Models Are Currently Available?
+## What Carrying Capacity Models Are Currently Available?
 
-Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity.
+When it comes to estimating carrying capacity, spawner-recruit models are the gold standard (**citation?**).
+However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults.
+Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale.
+QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single season.
+Or the dataset could be built over time, sampling different locations in different years.
+If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.
+
+The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds goes a long way to justifying this approach.
+Although built from completely different data, when these mulitple lines of evidence converge it lends credence to the QRF results.
+
+Bioenergics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the same reach scale that our QRF model operates on [@Wall2016].
+However, there are some potential issues with how the food supply (i.e.
+invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016].
+
+<!-- Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity.
 The QRF approach employed here allows for the analysis of multiple correlated habitat metrics, often with non-linear relationships to fish density, in order to assess fish habitat relationships for higher quantiles that are assumed to represent carrying capacity.
-
-<!-- Talk about how the gold standard of carrying capacity estimates are spawner-recruit curves, and our validation exercise shows that QRF is getting similar answers.
-Then we can contrast the data needs for both approaches, and how spawner-recruit analyses don't tell you how capacity is related to habitat.
--->
-
-<!-- Somewhere in here, mention how data can be collected over a short time span, if enough spatial variation is captured.
-Or dataset can be built over time, sampling different locations (but be vary of possible year effects, maybe).
 -->
@@
 The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations.
+Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases.
+These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations?**).
 
-<!-- Discussion of biological expectations and PDP plots… -->
+The biggest driver of capacity is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response.
+In many ways, these metrics define habitat quantity.
+However, other metrics help define habitat quality, such as cooler temperatures in August, less pool-tail fine sendiment, and higher channel unit frequencies (another way of saying more pool-riffle interfaces).
 
@@
 ## Extrapolation Model
+
+Fish are mobile creatures, and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important.
+In the summer, for Chinook parr, there is clearly movement between multiple channel units (e.g.
+pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit.
+However, it is unlikely that they are moving up and down the entire watershed.
+Therefore, we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity.
+At the same time, we acknowledge that managers, life-cycle modelers and others are often interested in capacity estimates at larger spatial scales.
+While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population level scales.
+This is an efficient technique to leverage existing relationships for meaningful management decisions.
 
@@
 
-In addition to site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for analyses at larger extents, such as watershed and population level scales.
-This is an efficient technique to leverage existing relationships for meaningful management decisions.
-
 While the Columbia River Basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging.
@@
 <!-- Add here why GAA are messy. And then anything else which may be a consideration for using the extrapolation model moving forward.  -->
-
-<!-- The following sentence doesn't make sense -->
-
-Despite these considerations, we were able to leverage extensive datasets to produce carrying capacity estimates throughout the CRB that significantly increase or habitat restoration prioritization in a cost-effective manner.
 
@@
          Percent = percent) %>%
-  pander(caption = "(\\#tab:fh-samp-table)The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model.")
-  # kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model',
-  #       booktabs = T)
+  # pander(caption = "(\\#tab:fh-samp-table)The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model.")
+  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model',
+        booktabs = T)
 
@@
          Description = DescriptiveText) %>%
-  pander(caption = "(\\#tab:select-mets-table) Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
+  # pander(caption = "(\\#tab:select-mets-table) Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
 Metrics are ranked in order of relative importance.")
-  # kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
+  kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
 Metrics are ranked in order of relative importance.",
-  #       booktabs = T,
-  #       format.args = list(width = c(2, 10, 10, 50)))
+        booktabs = T,
+        format.args = list(width = c(2, 10, 10, 50)))
 
@@
 tibble(Covariate = gaa_covars) %>%
+  left_join(readxl::read_excel('../../../QRFcapacity/data/raw/master_sample_pts/GAA_Metadata_20150506.xlsx') %>%
+              mutate(FieldName = recode(FieldName,
+                                        'ChanlType' = 'Channel_Type')) %>%
+              select(Covariate = FieldName,
+                     Name = DisplayName) %>%
+              bind_rows(tibble(Covariate = "S2_02_11",
+                               Name = "NorWeST Avg. Aug. Temp"))) %>%
   mutate(Scale = if_else(Covariate %in% c('NatPrin1', 
@@
                                  'Reach (2 km)'))) %>%
-  pander(caption = "(\\#tab:gaa-covars)Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population).")
-  # kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population).",
-  #       booktabs = T)
+  left_join(model_svy_df$mod_full[[1]] %>%
+              summary() %>%
+              coefficients() %>%
+              as_tibble(rownames = "Covariate") %>%
+              select(1:3)) %>%
+  # pander(caption = "(\\#tab:gaa-covars)Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.")
+  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
+        booktabs = T, 
+        digits = 3)
+
 ```
@@
          `Adjusted $r^2$` = `Adjusted R2`) %>%
-  pander(caption = "(\\#tab:extrap-r2-table)Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.")
-  # kable(caption = "Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.",
-  #       escape = F,
-  #       booktabs = T,
-  #       digits = c(NA, NA, 3, 3))
+  # pander(caption = "(\\#tab:extrap-r2-table)Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.")
+  kable(caption = "Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.",
+        escape = F,
+        booktabs = T,
+        digits = 3)
 
@@
   filter(Population %in% keep_wtsds) %>%
-  pander(caption = "(\\#tab:sr-table)Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
+  # pander(caption = "(\\#tab:sr-table)Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
 Numbers in parentheses are coefficients of variation.")
-  # kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
-Numbers in parentheses are coefficients of variation.")
+  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
+Numbers in parentheses are coefficients of variation.",
+        booktabs = T)
 
@@
 Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
-river_lines = QRFcapacity::get_flowlines(stream_order = 6,
-                                         champ_wtsds %>%
-                                           st_union() %>%
-                                           st_make_grid(n = 1)) %>%
-            filter(!grepl('^1701', reachcode),
-                   !grepl('^171', reachcode))
+knitr::include_graphics('../figures/ChampWtds.svg')
 
-# river_lines %>%
-#   filter(!gnis_name %in% c('Deschutes River',
-#                            'Crooked River',
-#                            'Payette River',
-#                            'North Fork Payette River')) %>%
-#   ggplot() +
-#   geom_sf(aes(color = gnis_name))
-
-ggplot(data = champ_wtsds) +
-  geom_sf(data = river_lines %>%
-            filter(!gnis_name %in% c('Deschutes River',
-                                     'Crooked River',
-                                     'Payette River',
-                                     'North Fork Payette River')),
-          color = 'darkblue') +
-  geom_sf() +
-  geom_sf(data = champ_wtsds %>%
-            filter(fish_data),
-          fill = 'black') +
-  geom_sf_label(aes(label = id),
-                position = position_dodge(width = 200)) +
-  theme(axis.title = element_blank())
-
+# for pdf
+# knitr::include_graphics('../figures/ChampWtds.png')
 ```
@@
   scale_color_grey(start = 0.5, end = 0.5) +
-  theme(legend.position = 'none')
+  theme(legend.position = 'none',
+        strip.text = element_text(size = 6))
 ```
@@
 
-spawn_rec_comp_p
+spawn_rec_comp_p +
+  theme(strip.text = element_text(size = 7))
 ```

### 2020-02-05 analysis/R_scripts/fit_qrf_model.Rmd
changed a few selected habitat metrics, added things to the habitat dictionary, re-ran fit_qrf_model.Rmd script.

--- 2020-02-04 analysis/R_scripts/fit_qrf_model.Rmd
@@
 
-# Select QRF Habitat Covariates
+# Load Data
 
@@
   mutate_at(vars(Watershed, Year),
-            list(as.factor))
+            list(as.factor)) %>%
+  # scale some metrics by site length
+  mutate_at(vars(starts_with('LWVol')),
+            list(~ . / Lgth_Wet * 100)) %>%
+  mutate(RipCovCanSome = 100 - RipCovCanNone)
 
@@
 hab_dict = hab_dict_2017
+
+# change some of the descriptions for large wood volume
+hab_dict %<>%
+  mutate(DescriptiveText = if_else(grepl("^LWVol", ShortName),
+                                   paste0(str_remove(DescriptiveText, ".$"),
+                                         ", scaled by site length."),
+                                   DescriptiveText),
+         UnitOfMeasure = if_else(grepl("^LWVol", ShortName),
+                                   paste0(UnitOfMeasure,
+                                         " per 100 meters"),
+                                   UnitOfMeasure),
+         UnitOfMeasureAbbrv = if_else(grepl("^LWVol", ShortName),
+                                   paste0(UnitOfMeasureAbbrv,
+                                         "/100m"),
+                                   UnitOfMeasureAbbrv)) %>%
+  bind_rows(hab_dict_2017 %>%
+              filter(ShortName == "RipCovCanNone") %>%
+              mutate(ShortName = "RipCovCanSome",
+                     Name = "Riparian Cover: Some Canopy",
+                     DescriptiveText = "Percent of riparian canopy with some vegetation."))
 
@@
                            'Walla Walla',
-                           'Umatilla'))
+                           'Umatilla')) %>%
+  mutate(RipCovCanSome = 100 - RipCovCanNone)
 
@@
 
+# Select QRF Habitat Covariates
+
+## MINE Statistics
+
 Next we generated MINE stats for all possible habitat metrics, matching them up against the log of fish densities (plus a small value to deal with 0s).
@@
 mine_res = fish_hab %>%
-  mutate_at(vars(starts_with('LWVol')),
-            list(~ . / Lgth_Wet)) %>%
   mutate(fish_dens = log(fish_dens + 0.005)) %>%
@@
                                  "FishCovTotal",
-                                 "RipCovUstory",
+                                 # "RipCovUstory",
+                                 # "RipCovBigTree",
+                                 "RipCovCanSome",
                                  'WetWdth_Int',
@@
                                  "avg_aug_temp",
-                                 "LWFreq_Wet"))
+                                 "LWVol_WetSlow"))
+                                 # "LWFreq_Wet"))
                       # 

### 2020-02-05 analysis/paper/paper.Rmd
re-knit paper after selecting new habitat metrics for QRF. Added some text about the extrapolation model to the discussion.

@@
 
-<!-- Start with benefits of extrapolation model. For example, allows us to make estimates at larger scales. However, transition to negatives of extrapolation model. -->
+Our extrapolation model was focused on extrapolating to other master sample points, because that is the dataset available to us, but the methodology could be improved.
+Extrapolating to reaches on a stream network, as opposed to points on the landscape, could improve the interpretability of the results.
+This would require a stream network with relevant attributes attached to each reach, similar to the GAAs we used.
+Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and builting a QRF model from that dataset.
+Even if the fish data was not done continously, estimates of capacity could be made directly from the QRF model across the entire stream network, without the need for an extrapolation model.
 
-<!-- Then open it up to next steps of extrapolation model.
-i.e., ideally one should collect spatially continuous habitat data.
-If done, we could eliminate the extrapolation model altogether.
-Transition to improving habitat data.
--->
+<!-- Do we need any more in this section? I commented out the paragraph below... -->
 
-While the Columbia River Basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging.
+<!-- While the Columbia River Basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging.
 On-the-ground habitat data collection is costly, time-consuming, and often not reproducible; thereby limiting the effectiveness of this type of extrapolation.
 Additionally, the extrapolation model relies on Global Available Attribute (GAA) data which ….
+-->
 

### 2020-02-05 analysis/R_scripts/fit_qrf_model.Rmd
added metadata for GAAs. Read in from QRFcapacity repo

@@
 
+# get some meta data for GAAs
+gaa_meta = readxl::read_excel('../../../QRFcapacity/data/raw/master_sample_pts/GAA_Metadata_20150506.xlsx') %>%
+              mutate(FieldName = recode(FieldName,
+                                        'ChanlType' = 'Channel_Type')) %>%
+              bind_rows(tibble(Covariate = "S2_02_11",
+                               Name = "NorWeST Avg. Aug. Temp",
+                               Units = "C",
+                               SpatialScale = "Site-300m"))
+
 # possible covariates from among the GAAs
@@
           "../data/derived_data/gaa_covars.rds")
+write_rds(gaa_meta,
+          "../data/derived_data/gaa_metadata.rds")
 write_rds(mod_data_weights,

### 2020-02-05 analysis/paper/paper.Rmd
used the gaa metadata (previously saved) for the GAA table, to avoid needing to read it in from another repo.

@@
 
-<!-- Need to flesh out this table with better names, and add units as a column -->
+```{r gaa-covars}
+gaa_meta = read_rds("../data/derived_data/gaa_metadata.rds")
 
-```{r gaa-covars}
 tibble(Covariate = gaa_covars) %>%
-  left_join(readxl::read_excel('../../../QRFcapacity/data/raw/master_sample_pts/GAA_Metadata_20150506.xlsx') %>%
-              mutate(FieldName = recode(FieldName,
-                                        'ChanlType' = 'Channel_Type')) %>%
+  left_join(gaa_meta %>%
               select(Covariate = FieldName,
-                     Name = DisplayName) %>%
-              bind_rows(tibble(Covariate = "S2_02_11",
-                               Name = "NorWeST Avg. Aug. Temp"))) %>%
-  mutate(Scale = if_else(Covariate %in% c('NatPrin1', 
-                                          'DistPrin1'),
-                         'Watershed (HUC12)',
-                         if_else(Covariate %in% c('WIDE_BF',
-                                                  'Channel_Type',
-                                                  'Elev_M'),
-                                 'Site (300 m)',
-                                 'Reach (2 km)'))) %>%
+                     Name = DisplayName,
+                     Units,
+                     SpatialScale)) %>%
+  mutate(Units = if_else(Units == "na",
+                         as.character(NA),
+                         Units),
+         Units = if_else(grepl('\\^2', Units),
+                         str_replace(Units, '\\^2', '$^2$'),
+                         Units)) %>%
   left_join(model_svy_df$mod_full[[1]] %>%
@@
         booktabs = T, 
+        escape = F,
         digits = 3)

### 2020-02-05 analysis/paper/paper.Rmd
Filled in a few missing citations. added to several discussion sections.

@@
 Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality.
-Recently released “de-listing” criteria (NOAA 2016 ***need reference details***) identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required.
+Recently released “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required.
 These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured.
@@
 Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU).
-The Upper Columbia Spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River Spring/Summer-run is listed as threatened (***need citation***).
+The Upper Columbia Spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River Spring/Summer-run is listed as threatened [@NOAA2016; @NOAA2016b].
 
@@
 For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity.
-After removing duplicate samples, our initial dataset contained `r n_distinct(fish_hab$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fh-samp-table)).
+After removing duplicate samples, our initial dataset contained `r n_distinct(qrf_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fish-hab-sites)).
 
@@
 To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
-The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from (***need citation***).
+The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from [@Larsen2016].
 Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)).
@@
 However, for visualization, predictions of areal capacity (fish/m$^2$) were used.
+
+<!-- Drop that last sentence unless we are going to include an example map of QRF capacity estimates -->
 
@@
 
+We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables.
 The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars).
-
-<!-- Probably need something more here -->
+From this, we have estimates of capacity at every master sample point in the CRB, which are located approximately one kilometer apart.
 
@@
 
-When it comes to estimating carrying capacity, spawner-recruit models are the gold standard (**citation?**).
+When it comes to estimating carrying capacity, spawner-recruit models are the gold standard (**citation??**).
 However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults.
@@
 
+Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity.
 Bioenergics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the same reach scale that our QRF model operates on [@Wall2016].
@@
 invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016].
-
-<!-- Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity.
-The QRF approach employed here allows for the analysis of multiple correlated habitat metrics, often with non-linear relationships to fish density, in order to assess fish habitat relationships for higher quantiles that are assumed to represent carrying capacity.
--->
+The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as function of habitat.
 
@@
 Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases.
-These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations?**).
+These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations??**).
 
@@
 <!-- Consider pulling back and thinking about whether these kinds of techniques could be applied to other systems (terrestial, ocean, etc.).
+One example might the breeding bird survey, if someone could gather some habitat metrics at all the various sampling routes (big GIS exercise).
 If not, we may want to target a more fisheries-centric journal -->
@@
 
-```{r }
+```{r fish-hab-sites}
 qrf_data %>%

### 2020-06-03 analysis/R_scripts/compile_wtsd_bndry.Rmd
re-did map of CHaMP watersheds, adding a north arrow and saving map in 4 different formats.

--- 2020-01-28 analysis/R_scripts/compile_wtsd_bndry.Rmd
@@
 library(ggrepel)
+library(ggspatial)
 
@@
                 #                           y = -5000)) +
+  annotation_north_arrow(location = 'tr',
+                         which_north = 'true',
+                         style = north_arrow_fancy_orienteering) +
   theme(axis.title = element_blank())
@@
 ```
+
+Now let's make a map that we can import into the manuscript
+
+```{r}
+champ_wtsds = read_rds('../data/derived_data/champ_wtds.rds')
+river_lines = QRFcapacity::get_flowlines(stream_order = 6,
+                                         champ_wtsds %>%
+                                           st_union() %>%
+                                           st_make_grid(n = 1)) %>%
+            filter(!grepl('^1701', reachcode),
+                   !grepl('^171', reachcode))
+
+# river_lines %>%
+#   filter(!gnis_name %in% c('Deschutes River',
+#                            'Crooked River',
+#                            'Payette River',
+#                            'North Fork Payette River')) %>%
+#   ggplot() +
+#   geom_sf(aes(color = gnis_name))
+
+wtsd_map = ggplot(data = champ_wtsds) +
+  geom_sf(data = river_lines %>%
+            filter(!gnis_name %in% c('Deschutes River',
+                                     'Crooked River',
+                                     'Payette River',
+                                     'North Fork Payette River')),
+          color = 'darkblue') +
+  geom_sf() +
+  geom_sf(data = champ_wtsds %>%
+            filter(fish_data),
+          fill = 'black') +
+  geom_sf_label(aes(label = id),
+                position = position_dodge(width = 200)) +
+  annotation_north_arrow(location = 'tr',
+                         which_north = 'true',
+                         style = north_arrow_fancy_orienteering) +
+  theme(axis.title = element_blank())
+
+# save a couple versions
+ggsave('../figures/ChampWtds.svg',
+       wtsd_map,
+       dpi = 600,
+       height = 6,
+       width = 6)
+ggsave('../figures/ChampWtds.png',
+       wtsd_map,
+       dpi = 600,
+       height = 6,
+       width = 6)
+ggsave('../figures/ChampWtds.tiff',
+       wtsd_map,
+       dpi = 600,
+       height = 6,
+       width = 6)
+ggsave('../figures/ChampWtds.jpeg',
+       wtsd_map,
+       dpi = 600,
+       height = 6,
+       width = 6)
+```

### 2020-06-03 analysis/paper/paper.Rmd
recoded Chiwawa R. to Chiwawa River for plots. Made some other text revisions. Added 2 photos as figures. re-knit as html and word

@@
 csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
+# csl: "../templates/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
 abstract: |
@@
 ```{r sp-rec-comp}
+# rename one population
+spawn_rec_params %<>%
+  mutate(Population = recode(Population,
+                             'Chiwawa R.' = 'Chiwawa River'))
+
 # pull out estimates of capacity
@@
 # which watersheds to plot?
-keep_wtsds = spawn_rec_params %>%
-  filter(form == 'QRF',
-         !is.na(est)) %>%
-  filter(!Population %in% c('Methow R.')) %>%
-  pull(Population)
+# keep_wtsds = spawn_rec_params %>%
+#   filter(form == 'QRF',
+#          !is.na(est)) %>%
+#   filter(!Population %in% c('Methow R.')) %>%
+#   pull(Population)
 
 keep_wtsds = c('Catherine Creek',
-               'Chiwawa R.',
+               'Chiwawa River',
                'Upper Grande Ronde River',
@@
 Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris.
-Habitat metrics measuring volume and area were scaled to the wetted area of each site.
+Habitat metrics measuring volume or area were scaled to the wetted area of each site.
 Within each category, metrics were ranked according to their MIC value (**Figure ??**).
@@
 Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
+
+***Worth putting the whole MIC habitat selection approach in an appendix or supplemental material, so we can include large figures and tables?***
 
@@
 We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
-The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010.
-We used the 90th quantile, rather than a higher quantile, to avoid using predictions that are aimed at the very upper tails of observed fish density, where the variability of predictions may be influenced by sample size issues.
+The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.
 
@@
 
-Using the QRF model, we predicted habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin.
+<!-- Using the QRF model, we predicted habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin.  -->
+
 To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
@@
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
-However, for visualization, predictions of areal capacity (fish/m$^2$) were used.
 
-<!-- Drop that last sentence unless we are going to include an example map of QRF capacity estimates -->
+<!-- However, for visualization, predictions of areal capacity (fish/m$^2$) were used.  -->
+<!-- The sentence above was dropped, unless we are going to include an example map of QRF capacity estimates -->
 
@@
 For the latter, estimates of parr were calculated using estimates of over-winter survival.
-A series of spawner-recruit functions were then fit to this data including Beverton-Hold, Ricker, and hockey stick.
-Estimates of capacity were made from each of these spawner-recruit curves and compared with QRF estimates of capacity (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)).
+A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick, using the *FSA* package [@R-FSA] in R.
+Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.
 
@@
 We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
-We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width:depth ratio, channel unit frequency, standard deviation of the wetted depth, braidedness, total fish cover, percent riparian canopy that was big trees, wetted width, observed discharge, percent fines less than 6mm, lower quantile of substrate size (D16), average august temperature and the frequency of large wood in the wetted channel (Table \@ref(tab:select-mets-table)).
+We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).
 
@@
 
-When it comes to estimating carrying capacity, spawner-recruit models are the gold standard (**citation??**).
+When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999].
 However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults.
@@
 The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations.
-Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases.
-These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations??**).
+Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is less fine sediment, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases.
+These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations??**) [@Mossop2006].
 
@@
 <!-- Try to include images of high capacity / low capacity sites  -->
+See Figures \@ref{fig:site-photo-high} and \@ref{fig:site-photo-low}.
 
@@
   filter(Population %in% keep_wtsds) %>%
+  mutate(Population = recode(Population,
+                             'Chiwawa R.' = 'Chiwawa River')) %>%
   # pander(caption = "(\\#tab:sr-table)Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
@@
 Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
-knitr::include_graphics('../figures/ChampWtds.svg')
-
-# for pdf
 # knitr::include_graphics('../figures/ChampWtds.png')
+knitr::include_graphics('../figures/ChampWtds.jpeg')
 ```
@@
 
-```{r sr-figure, fig.cap = "Spawner-recruit data from seven watersheds.
+```{r sr-figure, fig.cap = "Spawner-recruit data from nine watersheds.
 Solid lines are the spawner-recruit curve, dashed lines are the estimated capacity, and shaded polygons depict the 95% confidence intervals of capacity.
-Blue corresponds to Beverton-Holt models, green to Ricker models, yellow to hockey stick models, and red to QRF estimates.
+Red corresponds to Beverton-Holt models, purple to Ricker models, blue to hockey stick models, and green to QRF estimates.
 The QRF solid curve is a Beverton-Holt model with the capacity parameter fixed to the QRF estimate of capacity."}
@@
   theme(strip.text = element_text(size = 7))
+```
+
+```{r site-photo-high, fig.cap = 'Photo of CHaMP site, CBW05583-415923, with predicted high capacity.'}
+knitr::include_graphics('../figures/SitePhoto_CBW05583-415923.jpg')
+```
+
+```{r site-photo-low, fig.cap = 'Photo of CHaMP site, CBW05583-109994, with predicted low capacity.'}
+knitr::include_graphics('../figures/SitePhoto_CBW05583-109994.jpg')
 ```

### 2020-06-05 analysis/paper/paper.Rmd
added a paragraph to Conclusions section. updated several figures.

@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::pdf_document2:
-      fig_caption: yes
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
       - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::pdf_document2:
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
       - --lua-filter=../templates/pagebreak.lua
@@
   fig.path = "../figures/",
-  dpi = 300
+  dpi = 600
 )
@@
 Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is less fine sediment, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases.
-These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need citations??**) [@Mossop2006].
+These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need more citations??**) [@Mossop2006].
 
@@
 ## Conclusions and Next Steps
+
+If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment.
+This understanding could be of crucial importance in the realm of conservation, when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally.
+We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity.
+As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys.
+***other potential examples (coral reefs?
+turtles?)***
 
@@
               filter(is.na(MetricGroupName) | MetricGroupName != 'Tier 1 Summary')) %>%
-  mutate_at(vars(ShortName, Name),
+  mutate_at(vars(ShortName, Name, MetricCategory),
             list(~ fct_reorder(., .x = imp))) %>%
@@
              y = imp)) +
-  # geom_col(aes(fill = MetricCategory)) %>%
+  # geom_col(aes(fill = MetricCategory)) +
+  # scale_fill_brewer(palette = 'Set2') +
   geom_col(fill = 'gray40') +
@@
 
-```{r site-photo-high, fig.cap = 'Photo of CHaMP site, CBW05583-415923, with predicted high capacity.'}
+```{r site-photo-high, dpi = 300, fig.cap = 'Photo of CHaMP site, CBW05583-415923, with predicted high capacity.'}
 knitr::include_graphics('../figures/SitePhoto_CBW05583-415923.jpg')
@@
 
-```{r site-photo-low, fig.cap = 'Photo of CHaMP site, CBW05583-109994, with predicted low capacity.'}
+```{r site-photo-low, dpi = 300, fig.cap = 'Photo of CHaMP site, CBW05583-109994, with predicted low capacity.'}
 knitr::include_graphics('../figures/SitePhoto_CBW05583-109994.jpg')

### 2020-06-09 R/estimate_MIC.R
added estimate_MIC function from QRFcapacity repo

@@ -1,0 +1,50 @@
+#' @title Estimate MIC
+#'
+#' @description Estimate various MIC-related statistics
+#'
+#' @author Kevin See
+#'
+#' @param data data.frame containing response as one column, and various covariates as other columns
+#' @param covars vector of potential covariate names, which match column names in \code{data}
+#' @param response name of response, which matches one column name in \code{data}
+#'
+#' @import dplyr minerva
+#' @return NULL
+#' @export
+
+estimate_MIC = function(data,
+                        covars,
+                        response) {
+  
+  if (!requireNamespace("minerva", quietly = TRUE)) {
+    stop("Package \"minerva\" needed for this function to work. Please install it.",
+         call. = FALSE)
+  }
+  
+  res = minerva::mine(x = data %>%
+               dplyr::select(one_of(covars)),
+             y = data %>%
+               pull(response),
+             use = 'pairwise.complete.obs') %>%
+    map_df(.f = identity) %>%
+    mutate(Metric = data %>%
+             dplyr::select(one_of(covars)) %>%
+             names()) %>%
+    dplyr::select(Metric, everything()) %>%
+    left_join(data %>%
+                dplyr::select(one_of(covars)) %>%
+                gather(Metric, value) %>%
+                group_by(Metric) %>%
+                summarise(var = var(value, na.rm = T),
+                          obsCV = sd(value, na.rm = T) / abs(mean(value, na.rm = T)),
+                          non_NA = sum(!is.na(value)),
+                          is_NA = sum(is.na(value)),
+                          perc_NA = round(is_NA / (is_NA + non_NA), 3),
+                          non_0 = sum(value != 0, na.rm = T),
+                          perc_non0 = non_0 / n()),
+              by = 'Metric') %>%
+    ungroup() %>%
+    dplyr::select(Metric, var:perc_non0, everything())
+  
+  return(res)
+}

### 2020-06-09 analysis/paper/paper.Rmd
updated captions and cross references for photo figures
changed the name of habitat metrics for plotting. Added reference about master sample points.

@@
 champ_wtsds = read_rds("../data/derived_data/champ_wtds.rds")
+
+sel_hab_mets %<>%
+  left_join(hab_dict %>%
+              select(Metric = ShortName,
+                     Name) %>%
+              distinct()) %>%
+  mutate(Name = recode(Name,
+                       'Wetted Width To Depth Ratio Avg' = 'Width:Depth Ratio',
+                       'Wetted Depth SD' = 'Depth Complexity',
+                       'Wetted Channel Braidedness' = 'Braidedness',
+                       'Fish Cover: Total' = 'Fish Cover',
+                       'Riparian Cover: Some Canopy' = 'Riparian Canopy',
+                       'Wetted Width Integrated' = 'Wetted Width',
+                       'Substrate < 6mm' = 'Fines',
+                       # 'Substrate: D16',
+                       'Avg. August Temperature' = 'Avg. August Temp.',
+                       'Large Wood Volume: Wetted Slow Water' = 'Large Wood Freq. in Pools'))
+
 ```
@@
 To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
-The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from [@Larsen2016].
+The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
 Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)).
@@
 <!-- Try to include images of high capacity / low capacity sites  -->
-See Figures \@ref{fig:site-photo-high} and \@ref{fig:site-photo-low}.
+See Figures \@ref(fig:site-photo-high) and \@ref(fig:site-photo-low).
 
@@
 ## Conclusions and Next Steps
+
+<!-- The beginning of the following paragraph may need some word-smithing -->
 
@@
   select(-imp) %>%
-  left_join(hab_dict) %>%
+  left_join(hab_dict %>%
+              select(-Name) %>%
+              inner_join(sel_hab_mets %>%
+                           rename(ShortName = Metric))) %>%
   filter(MetricGroupName != 'Tier 1 Summary') %>%
@@
   rename(imp = IncNodePurity) %>%
+  mutate(imp = (imp / max(imp)) * 100) %>%
   arrange(desc(imp)) %>%
@@
               filter(is.na(MetricGroupName) | MetricGroupName != 'Tier 1 Summary')) %>%
+  select(-Name) %>%
+  left_join(sel_hab_mets %>%
+              select(ShortName = Metric,
+                     Name)) %>%
   mutate_at(vars(ShortName, Name, MetricCategory),
@@
                         data = qrf_data,
-                        data_dict = hab_dict,
+                        data_dict = hab_dict %>%
+                          select(-Name) %>%
+                          inner_join(sel_hab_mets %>%
+                                       select(ShortName = Metric,
+                                              Name)),
                         type = 'quantile',

### 2020-06-09 analysis/paper/appendix_s1.Rmd
added an appendix for the MIC habitat selection steps.

---
title: "Appendix S1"
subtitle: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@biomark.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@biomark.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richie.Carmichael@biomark.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@biomark.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@biomark.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      fig_height: 8.5
      fig_width: 6.5
      toc: no
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_appendixS.tex
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
---

<!-- Find a way to add the journal name (Ecological Applications) to the top  -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/appendix/",
  dpi = 600,
  appendix = TRUE
)
```

```{r packages}
# setwd('analysis/paper')

library(tidyverse)
library(minerva)
library(QRFpaper)
library(kableExtra)

# set theme for ggplot2
theme_set(theme_bw())

# options for table formating with "kable()"
options(knitr.kable.NA = '-')
options(knitr.table.format = "pandoc")

```

```{r load-data}
fish_hab = read_rds("../data/derived_data/fish_hab.rds")
hab_dict = read_rds("../data/derived_data/hab_dict.rds")
sel_hab_mets = read_rds("../data/derived_data/sel_hab_mets.rds")
hab_data = read_rds("../data/derived_data/hab_data.rds")
```

```{r}
# what are some possible habitat covariates?
poss_hab_mets = tibble(ShortName = names(fish_hab)) %>%
  distinct() %>%
  left_join(hab_dict %>%
              filter(! MetricGroupName %in% c('Tier 1 Summary', 'Tier 2 Summary')) %>%
              select(ShortName, MetricGroupName, Name, MetricCategory) %>%
              distinct()) %>%
  filter(is.na(MetricCategory) | MetricCategory != 'Categorical') %>%
  filter((!is.na(Name) |
            ShortName %in% c('Elev_M',
                             'CUMDRAINAG',
                             "DpthThlwg_Avg",
                             "SCSm_Area",
                             "SCSm_Freq",
                             "SCSm_Ct",
                             "SCSm_Vol",
                             "RipCovUstoryNone",
                             "RipCovGrndNone",
                             "SC_Area",
                             "SC_Area_Pct",
                             "ChnlUnitTotal_Ct",
                             "Discharge_fish",
                             "Temp",
                             "PercentIceCover",
                             "LWCount",
                             "SubEstBdrk",
                             "Ucut_Length",
                             "FishCovAll",
                             "SubEstCandBldr",
                             "UcutLgth",
                             "LWcnt_Wet"))) %>%
  mutate(Name = if_else(is.na(Name),
                        ShortName,
                        Name)) %>%
  filter(!ShortName %in% c('Tier1', 'Tier2')) %>%
  mutate(MetricCategory = if_else(grepl('SC', ShortName),
                                  'SideChannel',
                                  MetricCategory),
         MetricCategory = if_else(grepl('Sub', ShortName),
                                  'Substrate',
                                  MetricCategory),
         MetricCategory = if_else(grepl('^Rip', ShortName),
                                  'Riparian',
                                  MetricCategory),
         MetricCategory = if_else(grepl('^FishCov', ShortName) |
                                    grepl('Ucut', ShortName) |
                                    ShortName %in% c('PercentIceCover'),
                                  'Cover',
                                  MetricCategory),
         MetricCategory = if_else(grepl('^LW', ShortName),
                                  'Wood',
                                  MetricCategory),
         MetricCategory = if_else(grepl('Discharge', ShortName) |
                                    ShortName %in% c("DpthThlwg_Avg",
                                                     "Dpth_Max",
                                                     'DpthThlwgExit',
                                                     'DpthResid',
                                                     'TotalVol',
                                                     'CUMDRAINAG'),
                                  'Size',
                                  MetricCategory),
         MetricCategory = if_else(ShortName %in% c('Elev_M', 'Temp'),
                                  'Temperature',
                                  MetricCategory),
         MetricCategory = if_else(ShortName %in% c('ChnlUnitTotal_Ct'),
                                  'ChannelUnit',
                                  MetricCategory)) %>%
  select(-MetricGroupName)

# # what columns in fish_hab are not in poss_hab_mets?
# names(fish_hab)[!names(fish_hab) %in% poss_hab_mets$ShortName]
```

```{r mine-stats}
#-----------------------------------------------------------------
# generate MINE statistics
#-----------------------------------------------------------------
mine_res = tibble(species = unique(fish_hab$Species),
                  fh_data = list(fish_hab)) %>%
  mutate(metrics = map(fh_data,
                       .f = function(x) {
                         poss_hab_mets %>%
                           filter(ShortName %in% names(x)) %>%
                           pull(ShortName)
                       })) %>%
  mutate(mine_res = map2(fh_data,
                         metrics,
                         .f = function(x, y) {
                           if(sum(x$fish_dens == 0) == 0) {
                             try(x %>%
                                   mutate(fish_dens = log(fish_dens)) %>%
                                   estimate_MIC(covars = y,
                                                response = 'fish_dens'))
                           } else {
                             try(x %>%
                                   mutate(fish_dens = log(fish_dens + 0.005)) %>%
                                   estimate_MIC(covars = y,
                                                response = 'fish_dens'))
                           }
                         })) %>%
  select(species, mine_res) %>%
  unnest(cols = mine_res) %>%
  left_join(poss_hab_mets,
            by = c('Metric' = 'ShortName')) %>%
  select(species, MetricCategory, Metric, Name, everything()) %>%
  # put the metric names in descending order by MIC
  mutate_at(vars(Metric, Name),
            list(~ fct_reorder(., .x = MIC))) %>%
  arrange(species, MetricCategory, desc(MIC)) %>%
      # filter out some metrics with too many NAs or 0s
      # filter((perc_NA < 0.5 & non_0 > 100)) %>%
      # filter out metrics with very low variance
      # filter(var < 0.1) %>%
      # filter(obsCV < 0.1)
      # filter out area and volume metrics
      filter(!grepl('Area$', Metric),
             !grepl('Vol$', Metric),
             !Metric %in% c('Lgth_Wet', 
                            'Lgth_BfChnl',
                            'Lgth_WetChnl',
                            'Area_Wet', 
                            'Area_Bf', 
                            'WetVol', 
                            'BfVol')) %>%
  mutate(selected = if_else(Metric %in% sel_hab_mets$Metric,
                            T, F))

```

```{r calc-correlations}
library(corrr)

hab_corr = tibble(hab_data = list(hab_data)) %>%
  mutate(metrics = map(hab_data,
                       .f = function(x) {
                         mets = poss_hab_mets %>%
                           filter(ShortName %in% names(x)) %>%
                           pull(ShortName)
                         x %>%
                           select(one_of(mets)) %>%
                           gather(Metric, value) %>%
                           group_by(Metric) %>%
                           summarise(n_tot = n(),
                                     n_NA = sum(is.na(value)),
                                     non_0 = sum(value != 0, na.rm = T)) %>%
                           mutate(perc_NA = n_NA / n_tot,
                                  perc_non_0 = non_0 / n_tot) %>%
                           filter(perc_NA < 0.5,
                                  non_0 > 100) %>%
                           filter(!grepl('Area$', Metric),
                                  !grepl('Vol$', Metric),
                                  !Metric %in% c('Lgth_Wet', 
                                                 'Lgth_BfChnl',
                                                 'Lgth_WetChnl',
                                                 'Area_Wet', 
                                                 'Area_Bf', 
                                                 'WetVol', 
                                                 'BfVol')) %>%
                           pull(Metric)
                       }),
         cor_mat = map2(hab_data,
                        metrics,
                        .f = function(x, y) {
                          x %>%
                            select(one_of(y)) %>%
                            cor(use = 'pairwise',
                                method = 'spearman')
                        }),
         corr_mat = map(cor_mat,
                        .f = as_cordf),
         catg_cor = map2(hab_data,
                         metrics,
                         .f = function(x, y) {
                           catg_list = try(poss_hab_mets %>%
                                             filter(ShortName %in% y) %>%
                                             split(list(.$MetricCategory)) %>%
                                             map(.f = function(z) {
                                               if(nrow(z) == 1) {
                                                 return(NULL)
                                               } else {
                                                 x %>%
                                                   select(one_of(z$ShortName)) %>%
                                                   cor(use = 'pairwise',
                                                       method = 'spearman')
                                               }
                                               }))
                           return(catg_list)
                                             }))

```

# Choosing Habitat Covariates

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011]. The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.

Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring volume or area were scaled to the wetted area of each site. Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.). Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.

<!-- talk about how we chose not to use categorical variables like channle type or valley class, because we expected other metrics to capture the difference between those categories -->

<!-- The following line inserts a page break  -->
\newpage

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

\newpage

# Tables

```{r mic-tab}
mine_tab_df = mine_res %>%
  group_by(MetricCategory) %>%
  slice(1:10) %>%
  select(Category = MetricCategory,
         Name,
         Abbrv = Metric,
         MIC,
         selected)

mine_tab_df %>%
  select(-selected) %>%
  kable(digits = 3,
        format = 'html',
        caption = 'MIC statistic for top metrics in each habitat category. Metrics selected for the QRF model are in bold.') %>%
  kable_styling(fixed_thead = T) %>%
  row_spec(row = which(mine_tab_df$selected),
           bold = T)

```

# Figures

```{r mine-facet-fig, fig.cap = 'Barplots of MIC statistics, faceted by habitat category.'}
ggplot(data = mine_res,
       aes(x = Name,
           y = MIC)) +
  geom_col(fill = 'blue') +
  coord_flip() +
  facet_wrap(~ MetricCategory,
             scales = 'free_y',
             ncol = 3) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 5))
```

```{r mine-all-fig, fig.cap = 'Barplot of MIC statistics, colored by habitat category.'}
ggplot(data = mine_res,
       aes(x = Name,
           y = MIC,
           fill = MetricCategory)) +
  geom_col() +
  coord_flip() +
  # scale_fill_viridis_d() +
  scale_fill_brewer(palette = 'Set3',
                    guide = guide_legend(nrow = 2)) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 6)) +
  labs(fill = 'Category')
```

```{r corr-fig, results='hide', fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
  # map(.f = function(x) {
  #   rownames(x) <- tibble(Metric = rownames(x)) %>%
  #     left_join(hab_dict %>%
  #                 select(Metric = ShortName,
  #                        Name) %>%
  #                 distinct()) %>%
  #     pull(Name)
  #   
  #   # colnames(x) <- tibble(Metric = colnames(x)) %>%
  #   #   left_join(hab_dict %>%
  #   #               select(Metric = ShortName,
  #   #                      Name) %>%
  #   #               distinct()) %>%
  #   #   pull(Name)
  #   return(x)
  # }) %>%
    map(.f = corrplot::corrplot.mixed,
        upper = 'ellipse',
        tl.pos = 'lt',
        order = 'FPC')
```

```{r corr-fig2, results='hide', fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
  # map(.f = function(x) {
  #   # rownames(x) <- tibble(Metric = rownames(x)) %>%
  #   #   left_join(hab_dict %>%
  #   #               select(Metric = ShortName,
  #   #                      Name) %>%
  #   #               distinct()) %>%
  #   #   pull(Name)
  #   
  #   colnames(x) <- tibble(Metric = colnames(x)) %>%
  #     left_join(hab_dict %>%
  #                 select(Metric = ShortName,
  #                        Name) %>%
  #                 distinct()) %>%
  #     pull(Name)
  #   return(x)
  # }) %>%
  map(.f = ggcorrplot::ggcorrplot,
      method = 'circle',
      hc.order = T,
      type = 'lower',
      colors = c('blue4', 'white', 'red3'))
```

### 2020-06-09 analysis/paper/paper.Rmd
added needed line to load habitat data (hab_data)

@@
 ```{r load-data}
+hab_data = read_rds("../data/derived_data/hab_data.rds")
 fish_hab = read_rds("../data/derived_data/fish_hab.rds")

### 2020-06-09 analysis/paper/appendix_s1.Rmd
changed correlations to be done with Pearson correlation coefficients. Added some plots of selected metrics.

@@
       fig_caption: yes
+      fig_height: 8.5
+      fig_width: 6.5
       number_sections: FALSE
@@
                             cor(use = 'pairwise',
-                                method = 'spearman')
+                                method = 'pearson')
+                                # method = 'spearman')
                         }),
@@
                                                    cor(use = 'pairwise',
-                                                       method = 'spearman')
+                                                       method = 'pearson')
+                                                       # method = 'spearman')
                                                }
@@
 
-```{r corr-fig, results='hide', fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
+```{r corr-fig, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
 hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
@@
 
-```{r corr-fig2, results='hide', fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
-hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
-  # map(.f = function(x) {
-  #   # rownames(x) <- tibble(Metric = rownames(x)) %>%
-  #   #   left_join(hab_dict %>%
-  #   #               select(Metric = ShortName,
-  #   #                      Name) %>%
-  #   #               distinct()) %>%
-  #   #   pull(Name)
-  #   
-  #   colnames(x) <- tibble(Metric = colnames(x)) %>%
-  #     left_join(hab_dict %>%
-  #                 select(Metric = ShortName,
-  #                        Name) %>%
-  #                 distinct()) %>%
-  #     pull(Name)
-  #   return(x)
-  # }) %>%
-  map(.f = ggcorrplot::ggcorrplot,
+```{r corr-fig2, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
+cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
+                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
+  mutate(cor_plot = map2(cor_mat,
+                         Category,
+                         .f = function(x, y) {
+                           ggcorrplot::ggcorrplot(x,
                                                   method = 'circle',
@@
                                                   type = 'lower',
+                                                  title = y,
+                                                  colors = c('blue4', 'white', 'red3'))
+                         }))
+
+cor_plot_df$cor_plot
+```
+
+```{r selected-corr, fig.cap = 'Correlation plot of habitat metrics used in QRF model.'}
+hab_data %>% 
+  select(one_of(sel_hab_mets$Metric)) %>% 
+  cor(use = 'pairwise',
+      method = 'pearson') %>%
+      # method = 'spearman') %>% 
+  # as_cordf() %>%
+  ggcorrplot::ggcorrplot(method = 'circle',
+                         # hc.order = T,
+                         type = 'lower',
+                         # legend.title = "Correlation",
                          colors = c('blue4', 'white', 'red3'))
 ```
+
+```{r high-cor-mets, fig.cap = 'Pairs plot of habitat metrics used in QRF model with a correlation coefficient greater than 0.5.'}
+high_cor_mets = hab_data %>%
+  select(one_of(sel_hab_mets$Metric)) %>%
+  correlate(use = 'pairwise',
+            method = 'pearson') %>%
+            # method = 'spearman') %>%
+  stretch(na.rm = T,
+          remove.dups = T) %>%
+  arrange(desc(abs(r))) %>%
+  filter(abs(r) > 0.5) %>%
+  select(x, y) %>%
+  pivot_longer(cols = c(x, y)) %>%
+  pull(value) %>%
+  unique()
+
+hab_data %>%
+  select(one_of(high_cor_mets)) %>%
+  GGally::ggpairs()
+```

### 2020-06-09 analysis/paper/paper.Rmd
changed section numbering off for manuscript

@@
       fig_caption: yes
+      number_sections: FALSE
       pandoc_args:
@@
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
       pandoc_args:

### 2020-06-10 analysis/paper/paper.Rmd
test commit. Minor edits to biological expectations paragraph.

@@
 Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is less fine sediment, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases.
-These are all patterns that emerged from the fish-habitat data, and those fish-habitat relationships match what has been found qualitatively in many other studies (**need more citations??**) [@Mossop2006].
+These are all patterns that emerged from the fish-habitat data, and where available, those fish-habitat relationships match what has been found qualitatively in other studies (**need more citations??**) [@Mossop2006].
 

### 2020-06-10 analysis/paper/appendix_s1.Rmd
revised the language explaning our selection process. Dropped land classification metrics from MIC table. Added colophon to the bottom.

@@
 
-A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model.
+A key step in developing our QRF model to predict fish capacities was selecting the habitat covariates to include in the model.
 Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat).
 Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).
+Our goal was to select a group of covariates that captured as many different aspects of the stream habitat (e.g.
+substrate, flow, riparian condition, channel unit configuration, etc.) as possible, while still holding information about fish densities.
 
 To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site.
-To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities.
+Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, riparian areas, side channels, size, substrate, temperature, water quality, and woody debris.
+Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units).
+To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities.
 We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011].
+MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an expontential association the MIC value could be high, even when the standard correlation coefficient is low.
+We excluded categorical variables such as channel type (e.g.
+meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.
+
+Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)).
 The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
-
-Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris.
-Habitat metrics measuring volume or area were scaled to the wetted area of each site.
-Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)).
-Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.).
-Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
-
-<!-- talk about how we chose not to use categorical variables like channle type or valley class, because we expected other metrics to capture the difference between those categories -->
+We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior.
+For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed flow, so we chose to include that metric in our QRF model.
+We also tried to include covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
+Finally, we attempted to avoid metrics with too many missing values, or too many all zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].
 
@@
 mine_tab_df = mine_res %>%
+  filter(MetricCategory != 'Land Classification') %>%
   group_by(MetricCategory) %>%
@@
          MIC,
+         perc_non0,
          selected)
@@
 
-```{r corr-fig, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
+```{r corr-fig1, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
 hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
@@
 
-```{r corr-fig2, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
+```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
 cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
@@
 ```
+
+<!-- The following line inserts a page break  -->
+\newpage
+
+### Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+if ("devtools" %in% installed.packages()) devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? 
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+```

### 2020-06-10 analysis/paper/paper.Rmd
updated e-mails to @merck.com

@@
   - Kevin E. See:
-      email: Kevin.See@biomark.com
+      email: Kevin.See@merck.com
       institute: [biomark]
@@
   - Michael W. Ackerman:
-      email: Mike.Ackerman@biomark.com
+      email: Mike.Ackerman@merck.com
       institute: [biomark]
@@
   - Richard A. Carmichael:
-      email: Richie.Carmichael@biomark.com
+      email: Richie.Carmichael@merck.com
       institute: [biomark]
@@
   - Sarah L. Hoffmann:
-      email: Sarah.Hoffmann@biomark.com
+      email: Sarah.Hoffmann@merck.com
       institute: [biomark]
@@
   - Chris Beasley:
-      email: Chris.Beasley@biomark.com
+      email: Chris.Beasley@merck.com
       institute: [biomark]

### 2020-06-10 analysis/paper/paper.Rmd
edits to text in Biological Expectations from QRF Model section

@@
   - Richard A. Carmichael:
-      email: Richie.Carmichael@merck.com
+      email: Richard.Carmichael@merck.com
       institute: [biomark]
@@
 The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations.
-Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is less fine sediment, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for complexity) increases.
-These are all patterns that emerged from the fish-habitat data, and where available, those fish-habitat relationships match what has been found qualitatively in other studies (**need more citations??**) [@Mossop2006].
+Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is less fine sediment, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases.
+These are all patterns that emerged from the fish-habitat data, and where available, match those fish-habitat relationships identified qualitatively in other studies [@Mossop2006].
 
-The biggest driver of capacity is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response.
+The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response.
 In many ways, these metrics define habitat quantity.
-However, other metrics help define habitat quality, such as cooler temperatures in August, less pool-tail fine sendiment, and higher channel unit frequencies (another way of saying more pool-riffle interfaces).
+However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sendiment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.
 
@@
 See Figures \@ref(fig:site-photo-high) and \@ref(fig:site-photo-low).
+
+Moreover, predictions of available habitat capacity seemed to pass the "eye test" when comparing predictions to the actual physical habitat at CHaMP survey locations.
+That is, areas with higher predicted capacities had habitat that, in our opinion, looked like habitat that would support more juvenile Chinook salmon, whereas areas with lower predicted capacities looked like poorer rearing habitat.
+As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with an increases width-to-depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths.
+Alternatively, Figure \@ref(fig:site-photo-low) shows a site where the model predicted a much lower capacity, and visually, is much poorer rearing habitat for juvenile Chinook salmon e.g., lower wetted widths, discharge, width-to-depth ratio and standard deviation of depth.
+Pairing predictions of capacity from the QRF model with images and expert opinion of good versus bad habitat will be an important ongoing validation in addition to our comparisons with spawner-recruit models.
 

### 2020-06-10 analysis/paper/paper.Rmd
edited the habitat covariate selection section to refer to the appendix. Ensured that all methods that used the natural log of fish density said so. Added a paragraph to the extrapolation model discussion

--- 2020-06-10 analysis/paper/paper.Rmd
@@
 A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model.
-Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat).
-Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).
-
-To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site.
-To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities.
-We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011].
-The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
-
-Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris.
-Habitat metrics measuring volume or area were scaled to the wetted area of each site.
-Within each category, metrics were ranked according to their MIC value (**Figure ??**).
-Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.).
-Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
-
-***Worth putting the whole MIC habitat selection approach in an appendix or supplemental material, so we can include large figures and tables?***
+The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables.
+We sought to include a small set of covariates that were not overly redundant (i.e.
+not highly correlated), described many aspects of stream habitat (e.g.
+substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumeably because they contained information about what types of habitat fish sought or avoided.
+Full details of how covariates were selected can be found in Appendix S1.
 
@@
 
-Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months.
+Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months, using the natural log of fish densities as the response.
 After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model.
@@
 We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
-We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).
+We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).
 
@@
 
+One of the potential downsides to the extrapolation approach used here is that the globally available attributes generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat.
+While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model.
+This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale.
 <!-- Do we need any more in this section? I commented out the paragraph below... -->

### 2020-06-10 analysis/paper/appendix_s1.Rmd

@@
 
-A key step in developing our QRF model to predict fish capacities was selecting the habitat covariates to include in the model.
+One of the crucial steps in building this carrying capacity model was choosing which habitat covariates to include.
 Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat).
@@
 To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities.
-We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011].
+We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric [@Reshef2011].
 MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an expontential association the MIC value could be high, even when the standard correlation coefficient is low.

### 2020-06-10 analysis/paper/paper.Rmd

@@
   - Richard A. Carmichael:
-      email: Richie.Carmichael@biomark.com
+      email: Richard.Carmichael@merck.com
       institute: [biomark]

### 2020-06-10 analysis/paper/paper.Rmd
added cite keys to Biological Expectations section

@@
 The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations.
-Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler, when there is less fine sediment, when there is more fish cover, when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases.
+Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler [@Bjornn1991; @Raleigh1986; @Brett1952], when there is less fine sediment [@Allen2000; @Bjornn1991; @Hillman1987], when there is more fish cover [@Holecek2009; @Bjornn1991; @Hillman1987], when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases.
 These are all patterns that emerged from the fish-habitat data, and where available, match those fish-habitat relationships identified qualitatively in other studies [@Mossop2006].
@@
 However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sendiment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.
-
-<!-- Try to include images of high capacity / low capacity sites  -->
-See Figures \@ref(fig:site-photo-high) and \@ref(fig:site-photo-low).
 

### 2020-06-11 analysis/paper/paper.Rmd
Finished revisions through Methods

@@
 
-<!-- Need to define carrying capacity? -->
-
 Pacific salmon (*Oncorhynchus spp.*) species have experienced large declines in abundance throughout much of their range [@Good2005].
@@
 
-Most studies that have investigated fish habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity.
+Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity.
 Further, many of these studies focus on only one or two measures of habitat.
@@
 
-<!-- and it often fails to account for habitat covariates.
-This manuscript incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity.
-Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.
--->
+<!-- and it often fails to account for habitat covariates. -->
 
@@
 Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity.
-To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile regression forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics.
+To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics.
 QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001].
@@
 
-The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA.
-Within the interior CRB two major runs of Chinook salmon (*Oncorhynchus tshawytscha*; hereafter Chinook salmon) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics.
+The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River Basin, Pacific Northwest, USA.
+Within the interior Columbia River Basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics.
 Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts.
@@
 Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU).
-The Upper Columbia Spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River Spring/Summer-run is listed as threatened [@NOAA2016; @NOAA2016b].
+The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b].
+Hereafter, we refer to both ESUs simply as Chinook salmon.
 
@@
 1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
-1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measures (i.e., CHaMP sites within the Columbia River Basin),
+1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River Basin),
 1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
@@
 
-Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River Basin (CRB), Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)).
-The CRB covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
+Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River Basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)).
+The Columbia River Basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
 The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org.
@@
 
-CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin (CRB) selected based on a spatially balanced Generalized Random Tesselation Stratified (GRTS) sample selection algorithm [@Stevens1999; @Stevens2004].
+CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004].
 Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October.
@@
 
-Juvenile fish surveys were conducted for sp/sum Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol.
+Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol.
 Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, and single-pass electrofishing, as well as snorkeling.
-These data were used to estimate sp/sum Chinook salmon parr abundance at all CHaMP sites where fish survey data were available.
+These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available.
 Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013.
@@
 not highly correlated), described many aspects of stream habitat (e.g.
-substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumeably because they contained information about what types of habitat fish sought or avoided.
-Full details of how covariates were selected can be found in Appendix S1.
+substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided.
+Full details of how covariates used in the QRF model were selected can be found in Appendix S1.
 
@@
 
-Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months, using the natural log of fish densities as the response.
+Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for Chinook salmon parr, during summer months, using the natural log of fish densities as the response.
 After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model.
@@
 For CHaMP sites that have been sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions.
-In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of spring/summer Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.
+In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.
 
@@
 To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
-The GAA data used here was taken from the list of GRTS master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
+The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
 Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)).
@@
 We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else.
-We then made predictions of linear and areal capacity at all master sample sites throughout the interior CRB, generally spaced about one kilometer apart.
+We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River Basin, generally spaced about one kilometer apart.
 Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).
@@
 To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
-We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists.
+We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists.
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
@@
 
-Spawner-recruit data from several watersheds within the interior CRB were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity.
-Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts from rotary screw traps.
+Spawner-recruit data from several watersheds within the interior Columbia River Basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity.
+Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e.
+fall emigrants) from rotary screw traps.
 For the latter, estimates of parr were calculated using estimates of over-winter survival.
@@
 The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars).
-From this, we have estimates of capacity at every master sample point in the CRB, which are located approximately one kilometer apart.
+From this, we have estimates of capacity at every master sample point in the Columbia River Basin, which are located approximately one kilometer apart.
 
@@
 
-Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the CRB, there is a pressing need to develop new tools for habitat analyses.
+Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River Basin, there is a pressing need to develop new tools for habitat analyses.
 Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015].
@@
 As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys.
+The framework could also be applied to depleted, non-migratory, and isolated populations (e.g.
+desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed.
+Capacity estimates could also be used to potential translocation sites to determine if given sites can support an abundance considered viable before investing in translocation efforts.
 ***other potential examples (coral reefs?

### 2020-06-11 analysis/paper/paper.Rmd
completed revisions through Results

@@
 
-The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River Basin, Pacific Northwest, USA.
-Within the interior Columbia River Basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics.
+The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA.
+Within the interior Columbia River basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics.
 Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts.
@@
 1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
-1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River Basin),
+1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River basin),
 1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
@@
 
-Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River Basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)).
-The Columbia River Basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
+Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)).
+The Columbia River basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
 The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org.
@@
 
-CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004].
+CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004].
 Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October.
@@
 We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else.
-We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River Basin, generally spaced about one kilometer apart.
+We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.
 Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).
@@
 
-Spawner-recruit data from several watersheds within the interior Columbia River Basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity.
+Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity.
 Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e.
@@
 We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
-We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).
+We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average August temperature, wetted width:depth ratio, percent fines less than 6 mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).
 
@@
 After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)).
-Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots (PDP).
+Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots.
 These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)).
@@
 The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars).
-From this, we have estimates of capacity at every master sample point in the Columbia River Basin, which are located approximately one kilometer apart.
+From this, we have estimates of capacity at every master sample point in the Columbia River basin, which are located approximately one kilometer apart.
 
@@
 
-<!-- While the Columbia River Basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging.
+<!-- While the Columbia River basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging.
 On-the-ground habitat data collection is costly, time-consuming, and often not reproducible; thereby limiting the effectiveness of this type of extrapolation.
@@
 
-Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River Basin, there is a pressing need to develop new tools for habitat analyses.
+Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses.
 Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015].

### 2020-06-11 analysis/paper/appendix_s1.Rmd
minor text edits

@@
 To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site.
-Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, riparian areas, side channels, size, substrate, temperature, water quality, and woody debris.
+Habitat metrics were first grouped into broad categories that included channel unit configuration, complexity, fish cover, riparian areas, side channels, stream size, substrate, temperature, water quality, and woody debris.
 Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units).
@@
 We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric [@Reshef2011].
-MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an expontential association the MIC value could be high, even when the standard correlation coefficient is low.
+MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an quadratic association the MIC value could be high, even when the standard correlation coefficient is low.
 We excluded categorical variables such as channel type (e.g.
@@
 Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)).
-The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
+The MIC value of each measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
 We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior.
-For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed flow, so we chose to include that metric in our QRF model.
+For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model.
 We also tried to include covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
@@
 mine_tab_df = mine_res %>%
+  mutate(`Percent 0-value` = 1 - perc_non0) %>%
   filter(MetricCategory != 'Land Classification') %>%
@@
          MIC,
-         perc_non0,
+         `Percent Missing` = perc_NA,
+         `Percent 0-value`,
          selected)
@@
         format = 'html',
+        # format = 'latex',
+        booktabs = T,
         caption = 'MIC statistic for top metrics in each habitat category. Metrics selected for the QRF model are in bold.') %>%
@@
 ```{r mine-facet-fig, fig.cap = 'Barplots of MIC statistics, faceted by habitat category.'}
-ggplot(data = mine_res,
-       aes(x = Name,
+mine_res %>%
+  filter(MetricCategory != 'Land Classification') %>%
+  mutate(MetricCategory = recode(MetricCategory,
+                                 'ChannelUnit' = 'Channel\nUnit',
+                                 'SideChannel' = 'Side\nChannel',
+                                 'WaterQuality' = 'Water\nQuality')) %>%
+  ggplot(aes(x = Name,
              y = MIC)) +

### 2020-06-11 analysis/paper/paper.Rmd
added reference for North American Breeding Bird Survey

--- 2020-06-10 analysis/paper/paper.Rmd
@@
 We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity.
-As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys.
+As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey [@Sauer2017] to groundfish trawl surveys.
 ***other potential examples (coral reefs?

### 2020-06-11 analysis/paper/paper.Rmd
added reference for NOAA groundfish trawl survey

@@
 We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity.
-As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey [@Sauer2017] to groundfish trawl surveys.
+As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey [@Sauer2017] to groundfish trawl surveys [@Bradburn2011].
 ***other potential examples (coral reefs?

### 2020-06-11 analysis/paper/paper.Rmd

--- 2020-06-11 analysis/paper/paper.Rmd
@@
 We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity.
-As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys.
+As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey [@Sauer2017] to groundfish trawl surveys [@Bradburn2011].
 The framework could also be applied to depleted, non-migratory, and isolated populations (e.g.

### 2020-06-11 analysis/paper/paper.Rmd
revisions in Discussion of paper.Rmd

@@
 In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams.
-Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act (ESA) delisting.
-The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat restoration actions to recover depleted salmon populations.
-Moreover, the models presented here can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult).
-Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life stage specific habitat is limiting.
-For example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required for ESA delisting, but that juvenile capacity may not be sufficient to support those levels of adult spawning.
-In such a case, habitat restoration actions may be most cost-effectively and sustainably directed towards juvenile rearing habitat.
+Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting.
+The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations.
+Moreover, the models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult).
+Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting.
+As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support target abundance.
+In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat.
 Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.
+
+The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds goes a long way to justifying this approach.
+Although built from completely different data, when these mulitple lines of evidence converge it lends credence to the QRF capacity prediction results.
 
@@
 Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale.
-QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single season.
+QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single (or very few) season(s).
 Or the dataset could be built over time, sampling different locations in different years.
 If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.
-
-The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds goes a long way to justifying this approach.
-Although built from completely different data, when these mulitple lines of evidence converge it lends credence to the QRF results.
 
@@
 
-The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations.
-Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler [@Bjornn1991; @Raleigh1986; @Brett1952], when there is less fine sediment [@Allen2000; @Bjornn1991; @Hillman1987], when there is more fish cover [@Holecek2009; @Bjornn1991; @Hillman1987], when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases.
+The results of the QRF parr capacity model for Chinook salmon meet many biological expectations.
+Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width:depth ratio grow, when temperatures are cooler [@Bjornn1991; @Raleigh1986; @Brett1952], when there is less fine sediment [@Allen2000; @Bjornn1991; @Hillman1987], when there is more fish cover [@Holecek2009; @Bjornn1991; @Hillman1987], when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases.
 These are all patterns that emerged from the fish-habitat data, and where available, match those fish-habitat relationships identified qualitatively in other studies [@Mossop2006].
@@
 In many ways, these metrics define habitat quantity.
-However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sendiment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.
+However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.
 
@@
 That is, areas with higher predicted capacities had habitat that, in our opinion, looked like habitat that would support more juvenile Chinook salmon, whereas areas with lower predicted capacities looked like poorer rearing habitat.
-As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with an increases width-to-depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths.
-Alternatively, Figure \@ref(fig:site-photo-low) shows a site where the model predicted a much lower capacity, and visually, is much poorer rearing habitat for juvenile Chinook salmon e.g., lower wetted widths, discharge, width-to-depth ratio and standard deviation of depth.
+As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with an increases width:depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths.
+Alternatively, Figure \@ref(fig:site-photo-low) shows a site where the model predicted a much lower capacity, and visually, is much poorer rearing habitat for juvenile Chinook salmon e.g., lower wetted widths, discharge, width:depth ratio and standard deviation of depth.
 Pairing predictions of capacity from the QRF model with images and expert opinion of good versus bad habitat will be an important ongoing validation in addition to our comparisons with spawner-recruit models.
@@
 This would require a stream network with relevant attributes attached to each reach, similar to the GAAs we used.
-Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and builting a QRF model from that dataset.
-Even if the fish data was not done continously, estimates of capacity could be made directly from the QRF model across the entire stream network, without the need for an extrapolation model.
+Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and building a QRF model from that dataset.
+Even if the fish data was not done continuously, estimates of capacity could be made directly from the QRF model across the entire stream network, without the need for an extrapolation model.
 
@@
 This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale.
+
 <!-- Do we need any more in this section? I commented out the paragraph below... -->
@@
 UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling.
-Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].
+Advances in imaging techniques (e.g., multispectral imaging) and post-processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].
 
 Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly improve the broadscale application of QRF.
-Rapid advances drone technology further improves upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing.
+Rapid advances in drone technology further improves upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing.
 All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts.
-The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat restoration prioritization and planning.
+The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat rehabilitation prioritization and planning.
 
@@
 We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity.
-As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey [@Sauer2017] to groundfish trawl surveys [@Bradburn2011].
+As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys.
 The framework could also be applied to depleted, non-migratory, and isolated populations (e.g.
 desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed.
-Capacity estimates could also be used to potential translocation sites to determine if given sites can support an abundance considered viable before investing in translocation efforts.
-***other potential examples (coral reefs?
-turtles?)***
+Capacity estimates could also be used to evaluate potential translocation sites to determine if those sites could support an abundance considered viable before investing in translocation efforts.
 

### 2020-06-11 analysis/paper/paper.Rmd
minor text edits. Added shaded polygons in spawner-recruit figure for confidence intervals

@@
 output: 
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       number_sections: FALSE
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
+    bookdown::html_document2:
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       number_sections: FALSE
@@
 # define polygons for uncertainty shading
-plot_max = Inf
+# plot_max = Inf
 plot_max = 5e5
@@
          uprCI = est + se * qnorm(0.975)) %>%
+  rowwise() %>%
   mutate(lwrCI = if_else(lwrCI < 0, 0, lwrCI),
-         uprCI = if_else(uprCI > plot_max, plot_max, uprCI)) %>%
+         uprCI = if_else(uprCI > plot_max, max(est, plot_max), uprCI)) %>%
+  ungroup() %>%
   left_join(spawn_rec_params %>%
@@
              y = Parr)) +
-  # geom_polygon(data = cap_poly %>%
-  #                # filter out a couple populations with super poor fits
-  #                filter(Population %in% keep_wtsds),
-  #             aes(x = x,
-  #                 y = y,
-  #                 fill = form),
-  #             alpha = 0.2) +
+  geom_polygon(data = cap_poly %>%
+                 # filter out a couple populations with super poor fits
+                 filter(Population %in% keep_wtsds),
+              aes(x = x,
+                  y = y,
+                  fill = form),
+              alpha = 0.2) +
   geom_line(data = curve_fits %>%
@@
 
+```{r}
+cor_tab = cap_tbl %>%
+  select(-se, -cv) %>%
+  pivot_wider(names_from = "form",
+              values_from = "est") %>%
+  pivot_longer(cols = c(BevertonHolt, Ricker, Hockey),
+               names_to = "form",
+               values_to = "est") %>%
+  # filter(!is.na(est)) %>%
+  filter(Population %in% keep_wtsds) %>%
+  group_by(form) %>%
+  summarise(r = cor(x = QRF, 
+                    y = est, 
+                    use = 'pairwise'))
+```
+
 # Introduction
@@
 Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages.
-Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*Oncorhynchus mykiss*) in the Pacific Northwest, USA.
+Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*Oncorhynchus mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b].
 Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles.
@@
 
-<!-- and it often fails to account for habitat covariates. -->
-
-In fisheries it has long been recognized that that biotic and abiotic factors limit productivity within and across life-stages.
+In fisheries it has long been recognized that both biotic and abiotic factors limit productivity within and across life-stages.
 For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage.
@@
 1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
-1. Validate estimates of carrying capacity from QRF across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
+1. Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
 
@@
 
-Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)).
+Habitat data used in our study were collected from `r as.character(english(n_distinct(hab_data$Watershed)))` watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)).
 The Columbia River basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia.
@@
 Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`.
-Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
+Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds (see Table \@ref(tab:fish-hab-sites) and Figure \@ref(fig:wtsd-map)), at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
 
@@
 Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October.
-CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, size (depth, width, discharge), substrate, temperature, and water quality.
+CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, and water quality.
 
@@
 
-After model fitting, the QRF model was then be used to predict capacity using measurements of the habitat covariates that were used to fit the model.
+After model fitting, the QRF model was then used to predict capacity at sites with measurements of the habitat covariates that were used to fit the model.
 In our case, this includes all sites within CHaMP basins in the interior Columbia River basin.
-For CHaMP sites that have been sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions.
+For CHaMP sites that were sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions.
 In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.
@@
 Estimates of Chinook salmon parr capacity from the QRF and extrapolation models were comparable to independent estimates from spawner-recruit data (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)).
-QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the seven locations where comparisons were possible (Figure \@ref(fig:sr-figure)).
+QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the `r as.character(english(length(keep_wtsds)))` locations where comparisons were possible (Figure \@ref(fig:sr-figure)).
 Possible additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates.
-Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.905 (Beverton-Holt) to 0.965 (hockey stick).
+Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker).
 This favorable comparison provides strong validation, as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.
@@
 Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting.
-As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support target abundance.
+As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance.
 In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat.
@@
 That is, areas with higher predicted capacities had habitat that, in our opinion, looked like habitat that would support more juvenile Chinook salmon, whereas areas with lower predicted capacities looked like poorer rearing habitat.
-As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with an increases width:depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths.
+As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with high width:depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths.
 Alternatively, Figure \@ref(fig:site-photo-low) shows a site where the model predicted a much lower capacity, and visually, is much poorer rearing habitat for juvenile Chinook salmon e.g., lower wetted widths, discharge, width:depth ratio and standard deviation of depth.
@@
 Capacity estimates could also be used to evaluate potential translocation sites to determine if those sites could support an abundance considered viable before investing in translocation efforts.
-
-<!-- Consider pulling back and thinking about whether these kinds of techniques could be applied to other systems (terrestial, ocean, etc.).
-One example might the breeding bird survey, if someone could gather some habitat metrics at all the various sampling routes (big GIS exercise).
-If not, we may want to target a more fisheries-centric journal -->
 

### 2020-06-11 analysis/paper/appendix_s1.Rmd

@@
 Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).
-Our goal was to select a group of covariates that captured as many different aspects of the stream habitat (e.g.
-substrate, flow, riparian condition, channel unit configuration, etc.) as possible, while still holding information about fish densities.
+Our goal was to select a group of covariates that captured as many different aspects of the stream habitat (e.g., substrate, flow, riparian condition, channel unit configuration, etc.) as possible, while still holding information about fish densities.
 
@@
 We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric [@Reshef2011].
-MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an quadratic association the MIC value could be high, even when the standard correlation coefficient is low.
-We excluded categorical variables such as channel type (e.g.
-meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.
+MIC is a measure of correlation that incorporates potential non-linear associations; for example, if there is a quadratic association the MIC value could be high, even when the standard correlation coefficient is low.
+We excluded categorical variables such as channel type (e.g., meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.
 
 Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)).
-The MIC value of each measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
+The MIC value of each measured habitat characteristic and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
 We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior.
 For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model.
-We also tried to include covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.
-Finally, we attempted to avoid metrics with too many missing values, or too many all zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].
+We also tried to include covariates that can be directly influenced by rehabilitation actions or have been shown to impact salmonid juvenile density.
+Finally, we attempted to avoid metrics with too many missing values, or too many zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].
 
@@
         booktabs = T,
-        caption = 'MIC statistic for top metrics in each habitat category. Metrics selected for the QRF model are in bold.') %>%
+        caption = 'MIC statistic for top metrics within each habitat category, sorted by category and MIC value.
+The percent of records for which each habitat metric measurement was missing or zero is also shown.
+Metrics selected for the QRF model are in bold.') %>%
   kable_styling(fixed_thead = T) %>%
@@
 
-```{r corr-fig1, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
+```{r corr-fig1, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category."}
 hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
@@
 
-```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
+```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category."}
 cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
@@
 
-```{r selected-corr, fig.cap = 'Correlation plot of habitat metrics used in QRF model.'}
+```{r selected-corr, fig.cap = 'Correlation plot of habitat metrics used in the QRF model.'}
 hab_data %>% 

### 2020-06-11 analysis/paper/paper.Rmd
re-knit paper into html and word formats

@@
 output: 
-    bookdown::word_document2:
+    bookdown::html_document2:
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       number_sections: FALSE
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       number_sections: FALSE
@@
                values_to = "est") %>%
-  # filter(!is.na(est)) %>%
   filter(Population %in% keep_wtsds) %>%

### 2020-06-15 analysis/paper/paper.Rmd
incorporated Sarah's comments from a Word version

@@
 abstract: |
-  Needs more...
+  Might still need some refining...
 
-  The QRF model presented here provided estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale.
+  Establishing metrics of quality habitat is critical for the recovery of endangered Pacific salmonids.
+A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but rely on linear fish-habitat relationships and are limited in predictive power beyond specific watersheds.
+Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available.
+In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale.
+QRF models allow for the consideration of noisy data, correlated variables, non-linear relationships: common features in fish-habitat datasets.
+We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites.
+We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River Basin.
 Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models.
@@
 Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality.
-Recently released “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required.
+The most recent “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required.
 These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured.
@@
 Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles.
-However, estimating habitat carrying capacity, both historic and contemporary, for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge.
-Reliable methods to better understand fish-habitat relationships as well as to estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity, and further, to better direct tributary rehabilitation efforts.
+However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge.
+Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.
 
-@Fausch1988 conducted a thorough review of attempts to predict the standing crop of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal.
+@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal.
 Since that review, there has been progress in identifying some fish-habitat relationships for some salmonid species.
-@Nickelson1992 found that juvenile coho were found in higher densities within pool habitat on the Oregon coast.
+@Nickelson1992 demonstrated that juvenile coho were found in higher densities within pool habitat on the Oregon coast.
 Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001].
-@Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale, however their results demonstrated a negative relationship between abundance and the number of pools.
+@Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools.
 @Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools.
@@
 These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed.
-In addition, they all assumed some form of linear fish-habitat relationship, which often results in weak predictive power.
+In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.
 
@@
 The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003].
-Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
+Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
 @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead.
@@
 
-Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity.
+Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity.
 Further, many of these studies focus on only one or two measures of habitat.
 @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
-Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves.
-However, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].
+Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].
 
@@
 Ocean-type Chinook salmon enter freshwater later (e.g.
-fall or winter) spawn lower in the watershed, and the juveniles may only spend between several weeks and six months in freshwater before migrating to the ocean.
+fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean.
 Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow.
@@
 
-1. Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
-1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
-1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River basin),
-1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
-1. Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
+* Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
+* Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
+* Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River basin),
+* Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
+* Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
 
@@
 Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol.
-Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, and single-pass electrofishing, as well as snorkeling.
+Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, single-pass electrofishing, and snorkeling.
 These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available.
@@
 
-Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for Chinook salmon parr, during summer months, using the natural log of fish densities as the response.
+Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for Chinook salmon parr during summer months using the natural log of fish densities as the response.
 After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model.
-However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response.
+The individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response; therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response.
 There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012].
 We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)).
-The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.
+The 90th quantile of the predicted distribution was used as a proxy for carrying capacity following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.
 
@@
 
-<!-- Using the QRF model, we predicted habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin.  -->
+<!-- Drop the part about fitting an extrapolation model for areal density? -->
 
@@
 The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019].
-We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else.
+We fit two extrapolation models to each type of fish density, one that included watershed as covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else.
 We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.
@@
 
-To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
+To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
 We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists.
@@
 These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)).
-However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another, so care should be taken when interpretting them.
+However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another and care should be taken when interpreting them.
 After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.
@@
 Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker).
-This favorable comparison provides strong validation, as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.
+This favorable comparison provides strong validation as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.
 
@@
 
-## A Tool To Estimate Habitat Capacity
+## A Tool to Estimate Habitat Capacity
 
@@
 The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations.
-Moreover, the models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult).
+Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult).
 Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting.
@@
 The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response.
-In many ways, these metrics define habitat quantity.
-However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.
+In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.
 
@@
 
-Fish are mobile creatures, and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important.
+Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important.
 In the summer, for Chinook parr, there is clearly movement between multiple channel units (e.g.
 pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit.
-However, it is unlikely that they are moving up and down the entire watershed.
-Therefore, we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity.
-At the same time, we acknowledge that managers, life-cycle modelers and others are often interested in capacity estimates at larger spatial scales.
+However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity.
+At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales.
 While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population level scales.
@@
 Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and building a QRF model from that dataset.
-Even if the fish data was not done continuously, estimates of capacity could be made directly from the QRF model across the entire stream network, without the need for an extrapolation model.
+Even if the fish data were not collected continuously, estimates of capacity could be made directly from the QRF model across the entire stream network without the need for an extrapolation model.
 
@@
 
-Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly improve the broadscale application of QRF.
-Rapid advances in drone technology further improves upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing.
+Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly increase the broadscale application of QRF.
+Rapid advances in drone technology further improve upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing.
 All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts.
@@
 If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment.
-This understanding could be of crucial importance in the realm of conservation, when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally.
+This understanding could be of crucial importance in the realm of conservation when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally.
 We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity.

### 2020-06-16 analysis/paper/paper.Rmd
incorporated comments from Richie, and the new references he added.

@@
 
-  Establishing metrics of quality habitat is critical for the recovery of endangered Pacific salmonids.
+  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
 A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but rely on linear fish-habitat relationships and are limited in predictive power beyond specific watersheds.
@@
 In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale.
-QRF models allow for the consideration of noisy data, correlated variables, non-linear relationships: common features in fish-habitat datasets.
+QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets.
 We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites.
-We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River Basin.
+We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River basin.
 Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models.
@@
 @Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal.
-Since that review, there has been progress in identifying some fish-habitat relationships for some salmonid species.
+Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species.
 @Nickelson1992 demonstrated that juvenile coho were found in higher densities within pool habitat on the Oregon coast.
@@
 @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead.
-The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships.
+The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..
 
@@
 @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
-Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].
+Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which is usually unavailable [@Cramer2009].
 
-In fisheries it has long been recognized that both biotic and abiotic factors limit productivity within and across life-stages.
-For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage.
+<!-- Revise the first sentence of the following paragraph, it seems out of place with the following sentence and paragraph -->
+
+**In fisheries it has long been recognized that both biotic and abiotic factors limit productivity within and across life-stages.** For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage.
 We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality.
@@
 
-CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004].
+CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watershed within the interior Columbia River basin.
+The sites were selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004].
 Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October.
-CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, and water quality.
+CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.
 
@@
 
-To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
+To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear (parr / m) and areal (parr / m$^2$) capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
 The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
@@
 To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
-We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists.
+We first restricted the master sample points and lengths of streams to those within the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists.
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
@@
 
+All code and data for the analyses presented here can be found on a [GitHub repository](https://github.com/KevinSee/QRFpaper).
+
 # Results
@@
 
-We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m).
+We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate the Maximal Information Criteria (MIC) value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (parr/m) (See Appendix S1 for further details).
 We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average August temperature, wetted width:depth ratio, percent fines less than 6 mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).
@@
 The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars).
-From this, we have estimates of capacity at every master sample point in the Columbia River basin, which are located approximately one kilometer apart.
+From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, at roughly a one kilometer scale.
 
@@
 QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the `r as.character(english(length(keep_wtsds)))` locations where comparisons were possible (Figure \@ref(fig:sr-figure)).
-Possible additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates.
+Potential additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates.
 Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker).
@@
 In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat.
-Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.
+Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions and help guide not only the type of action, but also the location at which an action is performed.
 
-The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds goes a long way to justifying this approach.
-Although built from completely different data, when these mulitple lines of evidence converge it lends credence to the QRF capacity prediction results.
+The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds helps support and validate this approach.
+Although built from completely different data, when these multiple lines of evidence converge it lends credence to the QRF capacity prediction results.
 
@@
 However, there are some potential issues with how the food supply (i.e.
-invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016].
+invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Carmichael2020; @Wall2016].
 The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as function of habitat.
@@
 The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response.
-In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.
+In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and surrogate for the number of pool-riffle sequences or potential sheer areas providing feeding zones), and fish cover.
+
+<!-- Richie's comment on the paragraph below: I would say that this is an important part, but I question whether it is appropriate to include in the manuscript? I am fine either way. -->
 
@@
 At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales.
-While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population level scales.
+While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population levels.
 This is an efficient technique to leverage existing relationships for meaningful management decisions.
@@
 While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model.
-This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale.
+This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) maybe on the horizon [@Tonia2019] (Tonina et al.
+2019).
 
@@
 Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses.
-Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015].
+Unmanned Aerial Systems (UAS or drones, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015].
 UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling.
 Advances in imaging techniques (e.g., multispectral imaging) and post-processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].
+Further, developments in Light Detection and Ranging (LiDAR) technology have allowed for the characterization of watershed scale geomorphologic and hydraulic variables not previously possible [@Tonina2019; @McKean2008].
 
 Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly increase the broadscale application of QRF.
-Rapid advances in drone technology further improve upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing.
+Rapid advances in drone technology further improve upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated to automated data post-processing.
 All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts.
@@
 This understanding could be of crucial importance in the realm of conservation when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally.
-We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity.
+We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict a habitat's capacity.
 As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys.

### 2020-06-16 analysis/paper/paper.Rmd
dropped a questionable topic sentence (didn't really fit the rest of the paragraph)

@@
 
-<!-- Revise the first sentence of the following paragraph, it seems out of place with the following sentence and paragraph -->
-
-**In fisheries it has long been recognized that both biotic and abiotic factors limit productivity within and across life-stages.** For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage.
+For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage.
 We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality.

### 2020-06-16 analysis/paper/paper.Rmd
dropped any reference to an areal density extrapolation model. Added 2 references about hockey stick S-R model and natural/disturbance PCA. Improved formatting of some tables.

@@
 output: 
-    bookdown::html_document2:
+    bookdown::pdf_document2:
       fig_caption: yes
@@
       - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
@@
   Might still need some refining...
-
   Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
@@
 library(english)
+library(kableExtra)
 
@@
 
-panderOptions("big.mark", ',')
+# panderOptions("big.mark", ',')
 
@@
 
-<!-- Drop the part about fitting an extrapolation model for areal density? -->
-
-To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear (parr / m) and areal (parr / m$^2$) capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
+To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
 The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
-Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)).
-The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019].
-We fit two extrapolation models to each type of fish density, one that included watershed as covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else.
-We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.
-Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).
+Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)).
+The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019].
+We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else.
+We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.
 
@@
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
-
-<!-- However, for visualization, predictions of areal capacity (fish/m$^2$) were used.  -->
-<!-- The sentence above was dropped, unless we are going to include an example map of QRF capacity estimates -->
 
@@
 For the latter, estimates of parr were calculated using estimates of over-winter survival.
-A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick, using the *FSA* package [@R-FSA] in R.
+A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick [@Froese2008], using the *FSA* package [@R-FSA] in R.
 Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.
@@
 We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables.
-The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars).
+The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars) and the summary of the model fit is shown in Table \@ref(tab:extrap-r2-table).
 From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, at roughly a one kilometer scale.
@@
 pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit.
-However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity.
+However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 600 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity.
 At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales.
@@
 While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model.
-This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) maybe on the horizon [@Tonia2019] (Tonina et al.
-2019).
+This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) maybe on the horizon [@Tonina2019].
 
@@
          Percent = percent) %>%
-  # pander(caption = "(\\#tab:fh-samp-table)The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model.")
   kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model',
-        booktabs = T)
+        booktabs = T) %>%
+  kable_styling()
 
@@
          Description = DescriptiveText) %>%
-  # pander(caption = "(\\#tab:select-mets-table) Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
-Metrics are ranked in order of relative importance.")
   kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
@@
         booktabs = T,
-        format.args = list(width = c(2, 10, 10, 50)))
+        format.args = list(width = c(2, 10, 10, 50))) %>%
+  kable_styling()
 
@@
                          Units)) %>%
+  mutate(Name = if_else(Covariate == 'S2_02_11' & is.na(Name),
+                        'NorWeST Aug. Temperature',
+                        Name),
+         Units = if_else(Covariate == 'S2_02_11',
+                                'C',
+                                Units),
+         SpatialScale = if_else(Covariate == 'S2_02_11',
+                                'Reach-2km',
+                                SpatialScale)) %>%
   left_join(model_svy_df$mod_full[[1]] %>%
@@
               select(1:3)) %>%
-  # pander(caption = "(\\#tab:gaa-covars)Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.")
+  select(-Covariate) %>%
+  rename(`Spatial Scale` = SpatialScale,
+         Covariate = Name) %>%
   kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
@@
         escape = F,
-        digits = 3)
+        digits = 3) %>%
+  kable_styling()
 
@@
                            'chnk_per_m2' = 'fish/m$^2$')) %>%
+  filter(response == "fish/m") %>%
   select(Model = model,
@@
          `Adjusted $r^2$` = `Adjusted R2`) %>%
-  # pander(caption = "(\\#tab:extrap-r2-table)Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.")
-  kable(caption = "Summary of extrapolation model fits, split by each of the responses and whether the extrapolation model used CHaMP watershed as a covariate or not.",
+  kable(caption = "Summary of extrapolation model fits, split by whether the extrapolation model used CHaMP watershed as a covariate or not.",
         escape = F,
         booktabs = T,
-        digits = 3)
+        digits = 3) %>%
+  kable_styling()
 
@@
                              'Chiwawa R.' = 'Chiwawa River')) %>%
-  # pander(caption = "(\\#tab:sr-table)Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
-Numbers in parentheses are coefficients of variation.")
   kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
 Numbers in parentheses are coefficients of variation.",
-        booktabs = T)
+        booktabs = T) %>%
+  kable_styling()
 

### 2020-06-16 analysis/paper/paper.Rmd
deleted paragraph and figures referencing the "eye test" and pictures of a high capacity and low capacity site. Deemed unnecessary.

@@
 
-<!-- Richie's comment on the paragraph below: I would say that this is an important part, but I question whether it is appropriate to include in the manuscript? I am fine either way. -->
-
-Moreover, predictions of available habitat capacity seemed to pass the "eye test" when comparing predictions to the actual physical habitat at CHaMP survey locations.
-That is, areas with higher predicted capacities had habitat that, in our opinion, looked like habitat that would support more juvenile Chinook salmon, whereas areas with lower predicted capacities looked like poorer rearing habitat.
-As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with high width:depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths.
-Alternatively, Figure \@ref(fig:site-photo-low) shows a site where the model predicted a much lower capacity, and visually, is much poorer rearing habitat for juvenile Chinook salmon e.g., lower wetted widths, discharge, width:depth ratio and standard deviation of depth.
-Pairing predictions of capacity from the QRF model with images and expert opinion of good versus bad habitat will be an important ongoing validation in addition to our comparisons with spawner-recruit models.
-
 ## Extrapolation Model
@@
 
-```{r site-photo-high, dpi = 300, fig.cap = 'Photo of CHaMP site, CBW05583-415923, with predicted high capacity.'}
-knitr::include_graphics('../figures/SitePhoto_CBW05583-415923.jpg')
-```
-
-```{r site-photo-low, dpi = 300, fig.cap = 'Photo of CHaMP site, CBW05583-109994, with predicted low capacity.'}
-knitr::include_graphics('../figures/SitePhoto_CBW05583-109994.jpg')
-```
-
 <!-- The following line inserts a page break  -->

### 2020-06-18 analysis/R_scripts/fit_qrf_model.Rmd
updated spawner recruit parameters in QRFcapacity repo, so updated the results here.

@@
 
-data("spawn_rec_data")
+data("spawn_recr_data")
 # fit Beverton-Holt curve with capacity fixed at QRF capacity prediction
@@
 spawn_rec_params %<>%
-  bind_rows(qrf_params)
+  bind_rows(qrf_params %>%
+              select(one_of(names(spawn_rec_params)))) %>%
+  arrange(Population, form)
 

### 2020-06-18 analysis/paper/appendix/appendix_s1.Rmd
moved appendix to it's own folder. added a _bookdown.yml file to make the figure and table numbering correct for the appendix.

--- 2020-06-11 analysis/paper/appendix_s1.Rmd
@@
       pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
+      - --lua-filter=../../templates/scholarly-metadata.lua
+      - --lua-filter=../../templates/author-info-blocks.lua
+      - --lua-filter=../../templates/pagebreak.lua
     bookdown::word_document2:
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
+      - --lua-filter=../../templates/scholarly-metadata.lua
+      - --lua-filter=../../templates/author-info-blocks.lua
+      - --lua-filter=../../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
+      - --lua-filter=../../templates/scholarly-metadata.lua
+      - --lua-filter=../../templates/author-info-blocks2.lua
+      - --lua-filter=../../templates/pagebreak.lua
       includes:
-        in_header: ../templates/header_appendixS.tex
+        in_header: ../../templates/header_appendixS.tex
 bibliography:
-- references.bib
-- packages.bib
-csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
+- ../references.bib
+- ../packages.bib
+csl: "../../templates/ecological-applications.csl" # Insert path for the bib-style
 ---
@@
   comment = "#>",
-  fig.path = "../figures/appendix/",
+  fig.path = "../../figures/appendix/",
   dpi = 600,
@@
 options(knitr.kable.NA = '-')
-options(knitr.table.format = "pandoc")
+# options(knitr.table.format = "pandoc")
 
@@
 ```{r load-data}
-fish_hab = read_rds("../data/derived_data/fish_hab.rds")
-hab_dict = read_rds("../data/derived_data/hab_dict.rds")
-sel_hab_mets = read_rds("../data/derived_data/sel_hab_mets.rds")
-hab_data = read_rds("../data/derived_data/hab_data.rds")
+fish_hab = read_rds("../../data/derived_data/fish_hab.rds")
+hab_dict = read_rds("../../data/derived_data/hab_dict.rds")
+sel_hab_mets = read_rds("../../data/derived_data/sel_hab_mets.rds")
+hab_data = read_rds("../../data/derived_data/hab_data.rds")
 ```
@@
   select(-MetricGroupName)
+
+# remove Land Classification metrics
+poss_hab_mets <- poss_hab_mets %>%
+  filter(MetricCategory != 'Land Classification')
 
@@
   mutate(`Percent 0-value` = 1 - perc_non0) %>%
-  filter(MetricCategory != 'Land Classification') %>%
   group_by(MetricCategory) %>%
@@
   kable(digits = 3,
-        format = 'html',
+        # format = 'html',
         # format = 'latex',
         booktabs = T,
+        longtable = T,
         caption = 'MIC statistic for top metrics within each habitat category, sorted by category and MIC value.
@@
 Metrics selected for the QRF model are in bold.') %>%
-  kable_styling(fixed_thead = T) %>%
+  kable_styling(fixed_thead = T,
+                latex_options = c('repeat_header',
+                                  'scale_down',
+                                  'HOLD_position')) %>%
   row_spec(row = which(mine_tab_df$selected),
-           bold = T)
+           bold = T) %>%
+  column_spec(2, width = "2in") %>%
+  column_spec(3, width = "1in") %>%
+  column_spec(c(4:6), width = "0.5in")
 
@@
 mine_res %>%
-  filter(MetricCategory != 'Land Classification') %>%
   mutate(MetricCategory = recode(MetricCategory,
@@
              scales = 'free_y',
-             ncol = 3) +
+             ncol = 2) +
   theme(legend.position = 'bottom',
@@
 
-```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category."}
+```{r corr-fig, eval = T, results='hide', out.width = "90%", fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category.
+Larger circles correspond to larger correlation coefficients.
+The colors correspond to the correlation coefficient as well, and range from blue (-1) to white (0) to red (1)."}
 cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
@@
                                                   colors = c('blue4', 'white', 'red3'))
+                         }))
+
+cor_plot_df$cor_plot
+
+```
+
+```{r corr-fig2, eval = F, out.width = "50%", fig.cap = "Correlation plots of metrics, faceted by habitat category.
+Larger circles correspond to larger correlation coefficients.
+The colors correspond to the correlation coefficient as well, and range from blue (-1) to white (0) to red (1)."}
+cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
+                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
+  mutate(cor_plot = map2(cor_mat,
+                         Category,
+                         .f = function(x, y) {
+                           corrplot::corrplot.mixed(x,
+                                                    upper = 'ellipse',
+                                                    tl.pos = 'lt',
+                                                    order = 'FPC')
                          }))

### 2020-06-18 analysis/paper/appendix_s1.Rmd
moved appendix to it's own folder. added a _bookdown.yml file to make the figure and table numbering correct for the appendix.

@@
 output: 
-    bookdown::html_document2:
-      fig_caption: yes
-      fig_height: 8.5
-      fig_width: 6.5
-      number_sections: FALSE
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_appendixS.tex
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      fig_height: 8.5
+      fig_width: 6.5
+      number_sections: FALSE
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 bibliography:
@@
 
-```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category."}
+```{r corr-fig, eval = T, results='hide', out.width = "50%", fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category.
+Larger circles correspond to larger correlation coefficients.
+The colors correspond to the correlation coefficient as well, and range from blue (-1) to white (0) to red (1)."}
 cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
@@
                                                   colors = c('blue4', 'white', 'red3'))
+                         }))
+
+# for(i in 1:nrow(cor_plot_df)) {
+#   print(cor_plot_df$cor_plot[i])
+# }
+cor_plot_df$cor_plot
+
+# cowplot::plot_grid(plotlist = cor_plot_df$cor_plot,
+#                    ncol = 2)
+
+```
+
+```{r corr-fig2, eval = F, out.width = "50%", fig.cap = "Correlation plots of metrics, faceted by habitat category.
+Larger circles correspond to larger correlation coefficients.
+The colors correspond to the correlation coefficient as well, and range from blue (-1) to white (0) to red (1)."}
+cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
+                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
+  mutate(cor_plot = map2(cor_mat,
+                         Category,
+                         .f = function(x, y) {
+                           corrplot::corrplot.mixed(x,
+                                                    upper = 'ellipse',
+                                                    tl.pos = 'lt',
+                                                    order = 'FPC')
                          }))

### 2020-06-18 analysis/paper/appendix/appendix_s1.Rmd
changed emails in the Appendix to be merck, not biomark.

@@
 output: 
-    bookdown::html_document2:
-      fig_caption: yes
-      fig_height: 8.5
-      fig_width: 6.5
-      number_sections: FALSE
-      pandoc_args:
-      - --lua-filter=../../templates/scholarly-metadata.lua
-      - --lua-filter=../../templates/author-info-blocks.lua
-      - --lua-filter=../../templates/pagebreak.lua
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../../templates/scholarly-metadata.lua
-      - --lua-filter=../../templates/author-info-blocks.lua
-      - --lua-filter=../../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../../templates/header_appendixS.tex
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
+      pandoc_args:
+      - --lua-filter=../../templates/scholarly-metadata.lua
+      - --lua-filter=../../templates/author-info-blocks.lua
+      - --lua-filter=../../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      fig_height: 8.5
+      fig_width: 6.5
+      number_sections: FALSE
+      pandoc_args:
+      - --lua-filter=../../templates/scholarly-metadata.lua
+      - --lua-filter=../../templates/author-info-blocks.lua
+      - --lua-filter=../../templates/pagebreak.lua
 bibliography:

### 2020-06-18 analysis/paper/paper.Rmd
fixed some formating of tables

@@
 output:
-    bookdown::pdf_document2:
+    bookdown::html_document2:
       fig_caption: yes
       number_sections: FALSE
+      toc: no
       pandoc_args:
       - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
@@
       number_sections: FALSE
+      toc: no
       pandoc_args:
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
+    bookdown::pdf_document2:
       fig_caption: yes
       number_sections: FALSE
+      toc: no
       pandoc_args:
       - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
       - --lua-filter=../templates/pagebreak.lua
@@
 abstract: |
-  Might still need some refining...
+  **Might still need some refining...**
   Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
@@
 options(knitr.kable.NA = '-')
-options(knitr.table.format = "pandoc")
+# options(knitr.table.format = "pandoc")
+# options(knitr.table.format = "latex")
 
@@
 champ_wtsds = read_rds("../data/derived_data/champ_wtds.rds")
+gaa_meta = read_rds("../data/derived_data/gaa_metadata.rds")
 
@@
         format.args = list(width = c(2, 10, 10, 50))) %>%
-  kable_styling()
+  kable_styling() %>%
+  column_spec(4, width = '3in')
 
@@
 ```{r gaa-covars}
-gaa_meta = read_rds("../data/derived_data/gaa_metadata.rds")
-
 tibble(Covariate = gaa_covars) %>%
@@
         booktabs = T) %>%
-  kable_styling()
+  kable_styling() %>%
+  # kable_styling(latex_options = c('scale_down')) %>%
+  column_spec(1, width = "1in") %>%
+  column_spec(c(5:8), width = "0.5in")
 

### 2020-06-19 analysis/paper/paper.Rmd
updated the template to match ecological applications suggestions. Made a few minor revisions to text and figures, put each table and figure on it's own page, and re-knit into Word, PDF and html.

@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::word_document2:
+      fig_caption: yes
+      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 bibliography:
@@
 abstract: |
-  **Might still need some refining...**
   Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
-A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but rely on linear fish-habitat relationships and are limited in predictive power beyond specific watersheds.
+A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods often fail the capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds.
 Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available.
@@
 library(tidyverse)
-# library(QRFcapacity) # for the moment, until we know exactly what data to copy over to this repo
 library(QRFpaper)
@@
 
-# panderOptions("big.mark", ',')
+options(knitr.kable.NA = '-')
 
-options(knitr.kable.NA = '-')
+# when knitting to Word, use this
 # options(knitr.table.format = "pandoc")
-# options(knitr.table.format = "latex")
 
@@
 
+# change some of the names of metrics for plotting purposes
 sel_hab_mets %<>%
@@
                        # 'Substrate: D16',
+                       'Channel Unit Frequency' = 'Channel Unit\nFrequency',
                        'Avg. August Temperature' = 'Avg. August Temp.',
-                       'Large Wood Volume: Wetted Slow Water' = 'Large Wood Freq. in Pools'))
+                       'Large Wood Volume: Wetted Slow Water' = 'Large Wood\nFreq. in Pools'))
 
@@
 
+# scale the parr numbers for the plot
+scale_parr = 100000
+
 spawn_rec_comp_p = spawn_rec_params %>%
@@
   distinct() %>%
+  mutate_at(vars(Parr),
+            list(~ . / scale_parr)) %>%
   ggplot(aes(x = Spawners,
@@
   geom_polygon(data = cap_poly %>%
+                 mutate_at(vars(y),
+                           list(~ . / scale_parr)) %>%
                  # filter out a couple populations with super poor fits
@@
   geom_line(data = curve_fits %>%
+              mutate_at(vars(Parr),
+                        list(~ . / scale_parr)) %>%
               # filter out a couple populations with super poor fits
@@
   geom_hline(data = cap_tbl %>%
+               mutate_at(vars(est),
+                         list(~ . / scale_parr)) %>%
                # filter out a couple populations with super poor fits
@@
                     name = 'Model') +
-  labs(shape = 'Spawner\nType') +
+  labs(shape = 'Spawner\nType',
+       y = 'Parr (x100,000)') +
+  scale_y_continuous(breaks = scales::breaks_pretty(n = 3)) +
   theme(legend.position = 'bottom')
@@
 
-```{r select-mets-table}
+\newpage
+
+```{r select-mets-table, eval = T}
 qrf_mod$importance %>%
@@
               inner_join(sel_hab_mets %>%
-                           rename(ShortName = Metric))) %>%
+                           rename(ShortName = Metric)) %>%
+              mutate(Name = str_replace(Name, '\n', ' '))) %>%
   filter(MetricGroupName != 'Tier 1 Summary') %>%
@@
 
-```{r gaa-covars}
+\newpage
+
+```{r gaa-covars, eval = T}
 tibble(Covariate = gaa_covars) %>%
@@
 
-```{r extrap-r2-table}
+\newpage
+
+```{r extrap-r2-table, eval = T}
 model_svy_df %>%
@@
 
-```{r sr-table}
+\newpage
+
+```{r sr-table, eval = T}
 spawn_rec_params %>%
@@
 Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
-# knitr::include_graphics('../figures/ChampWtds.png')
-knitr::include_graphics('../figures/ChampWtds.jpeg')
+knitr::include_graphics('../figures/ChampWtds.png')
+# knitr::include_graphics('../figures/ChampWtds.jpeg')
 ```
+
+\newpage
 
@@
 
+\newpage
+
 ```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer Chinook salmon parr capacity quantile regression forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
@@
 
-```{r sr-figure, fig.cap = "Spawner-recruit data from nine watersheds.
+\newpage
+
+```{r sr-figure, fig.height = 5, fig.cap = "Spawner-recruit data from nine watersheds.
 Solid lines are the spawner-recruit curve, dashed lines are the estimated capacity, and shaded polygons depict the 95% confidence intervals of capacity.
 Red corresponds to Beverton-Holt models, purple to Ricker models, blue to hockey stick models, and green to QRF estimates.
-The QRF solid curve is a Beverton-Holt model with the capacity parameter fixed to the QRF estimate of capacity."}
+The QRF solid curve is a Beverton-Holt model with the capacity parameter fixed to the QRF estimate of capacity.
+A few curves with high capacity estimates were not plotted to improve readability."}
 
@@
 
-```{r colophon, cache = FALSE}
+```{r colophon, cache = FALSE, eval = T}
 # which R packages and versions?
@@
 
-```{r}
+```{r, eval = T}
 # what commit is this file at? 

### 2020-06-26 analysis/paper/paper.Rmd

@@
   Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
-A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods often fail the capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds.
+A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods often fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds.
 Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available.
@@
 We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites.
-We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River basin.
+We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin.
 Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models.
@@
 Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing.
-For perhaps the first time, the necessity of tributary habitat restoration can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”
+For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”
 
-Pacific salmon (*Oncorhynchus spp.*) species have experienced large declines in abundance throughout much of their range [@Good2005].
+Pacific salmon species have experienced large declines in abundance throughout much of their range [@Good2005].
 Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages.
-Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*Oncorhynchus mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b].
+Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O.
+mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b].
 Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles.
@@
 Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species.
-@Nickelson1992 demonstrated that juvenile coho were found in higher densities within pool habitat on the Oregon coast.
+@Nickelson1992 demonstrated that juvenile coho salmon (*O.
+kisutch*) were found in higher densities within pool habitat on the Oregon coast.
 Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001].
 @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools.
-@Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools.
+@Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O.
+nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools.
 Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002].
-@Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient.
+@Mossop2006 examined juvenile Chinook salmon (*O.
+tshawytscha*) in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient.
 These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed.
@@
 
-A number of studies have utilized other modeling approaches to elicit non-salmonid fish habitat relationships.
-@Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density.
-The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003].
+A number of studies have used other modeling approaches to elicit non-salmonid fish habitat relationships.
+@Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O.
+clarkii*) densities and the width:depth ratio of a stream for the upper quantiles of trout density.
+The same approach was also used to map the potential extent of sole (*Solea solea*) in the English Channel and southern North Sea [@Eastwood2003].
 Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
@@
 Further, many of these studies focus on only one or two measures of habitat.
-@Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
-Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which is usually unavailable [@Cramer2009].
+@Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
+Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are usually unavailable [@Cramer2009].
 
@@
 Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity.
-To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics.
+To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance and density and habitat characteristics.
 QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001].
@@
 The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA.
-Within the interior Columbia River basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics.
-Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts.
-Ocean-type Chinook salmon enter freshwater later (e.g.
-fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean.
-Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow.
+Within the interior Columbia River basin two major runs of Chinook salmon occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics.
+Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts.
+Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings.
+Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage.
 Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU).
@@
 
-In this study, we developed a QRF model to
+In this study, we developed a QRF model to:
 
@@
 
-This manuscript incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity.
+Our study incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity.
 Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.
@@
 Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`.
-Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds (see Table \@ref(tab:fish-hab-sites) and Figure \@ref(fig:wtsd-map)), at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
+Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds (see Table \@ref(tab:fish-hab-sites) and Figure \@ref(fig:wtsd-map)), at CHaMP survey reaches and were graciously provided by a number of agencies and projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].
 
@@
 CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watershed within the interior Columbia River basin.
-The sites were selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004].
+The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004].
 Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October.
@@
 The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables.
-We sought to include a small set of covariates that were not overly redundant (i.e.
-not highly correlated), described many aspects of stream habitat (e.g.
-substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided.
+We sought to include a small set of covariates that were not overly redundant (i.e., not highly correlated), described many aspects of stream habitat (e.g., substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided.
 Full details of how covariates used in the QRF model were selected can be found in Appendix S1.
@@
 For CHaMP sites that were sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions.
-In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.
+In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork Salmon.
 
@@
 To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
-The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
+The GAA data used here was taken from the list of generalized random tesselation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
 Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)).
@@
 To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
-We first restricted the master sample points and lengths of streams to those within the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists.
+We first restricted the master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists.
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
@@
 
-Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity.
-Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e.
-fall emigrants) from rotary screw traps.
+Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of Chinook salmon parr capacity.
+Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e., fall emigrants) from rotary screw traps.
 For the latter, estimates of parr were calculated using estimates of over-winter survival.
@@
 
-All code and data for the analyses presented here can be found on a [GitHub repository](https://github.com/KevinSee/QRFpaper).
+All code and data for the analyses presented here can be found in a [GitHub repository](https://github.com/KevinSee/QRFpaper).
 
@@
 We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables.
-The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars) and the summary of the model fit is shown in Table \@ref(tab:extrap-r2-table).
-From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, at roughly a one kilometer scale.
+The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars) and the summary of the model fit in Table \@ref(tab:extrap-r2-table).
+From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, each representing roughly one kilometer of stream length.
 
@@
 
-In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams.
+In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months and in wadeable streams.
 Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting.
@@
 
-The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds helps support and validate this approach.
+The favorable comparison between QRF estimates of carrying capacity and the spawner-recruit based estimates in select watersheds helps support and validate this approach.
 Although built from completely different data, when these multiple lines of evidence converge it lends credence to the QRF capacity prediction results.
@@
 invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Carmichael2020; @Wall2016].
-The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as function of habitat.
+The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as a function of habitat.
 
@@
 Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important.
-In the summer, for Chinook parr, there is clearly movement between multiple channel units (e.g.
-pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit.
+In the summer, for Chinook salmon parr, there is clearly movement between multiple channel units (e.g., pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit.
 However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 600 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity.
@@
 
-One of the potential downsides to the extrapolation approach used here is that the globally available attributes generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat.
+One of the potential downsides to the extrapolation approach used here is that the GAAs generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat.
 While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model.
@@
 As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys.
-The framework could also be applied to depleted, non-migratory, and isolated populations (e.g.
-desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed.
+The framework could also be applied to depleted, non-migratory, and isolated populations (e.g., desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed.
 Capacity estimates could also be used to evaluate potential translocation sites to determine if those sites could support an abundance considered viable before investing in translocation efforts.
@@
          Percent = percent) %>%
-  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer Chinook salmon parr capacity QRF model',
+  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer-run Chinook salmon parr capacity QRF model',
         booktabs = T) %>%
@@
          Description = DescriptiveText) %>%
-  kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer Chinook salmon parr capacity (during summer months).
+  kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer-run Chinook salmon parr capacity.
 Metrics are ranked in order of relative importance.",
@@
          Covariate = Name) %>%
-  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
+  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
         booktabs = T, 
@@
 
-```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile regression forest (QRF) model to predict habitat capacity, during summer months, for spring/summer Chinook salmon parr"}
+```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile regression forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr"}
 qrf_mod$importance %>%
@@
 
-```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer Chinook salmon parr capacity quantile regression forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
+```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile regression forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
 Tick marks along the X-axis depict observed values."}

### 2020-06-26 analysis/paper/paper.Rmd
added a header for pdf to get line numbering and doublespacing. Added some YAML for font family and font size. Re-knitted all versions.

@@
 output:
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::pdf_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::pdf_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      geometry: margin=1in
+      latex_engine: "xelatex"
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
+      includes:
+        in_header: ../templates/header_manuscript.tex
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+fontsize: 12pt
+mainfont: Times New Roman
 bibliography:
@@
 # when knitting to Word, use this
-# options(knitr.table.format = "pandoc")
+options(knitr.table.format = "pandoc")
 

### 2020-06-29 analysis/paper/paper.Rmd
Changed "References" to "Literature Cited"

@@
 
-# References 
+# Literature Cited 
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

### 2020-07-13 analysis/paper/paper.Rmd
changed the title for submitting to Ecosphere, made it appeal to a slightly more general audience. Added an if statement for knitting to Word

@@
 ---
-title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
+title: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
 author:
@@
 # when knitting to Word, use this
+# what kind of document is being created?
+doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to')
+
+if(doc.type == 'docx') {
   options(knitr.table.format = "pandoc")
+}
 

### 2020-08-07 analysis/paper/appendix/appendix_s1.Rmd
made some formatting changes and minor edits for submission to Ecosphere.

@@
 title: "Appendix S1"
-subtitle: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
+subtitle: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
 author:
@@
 output: 
+    bookdown::word_document2:
+      fig_caption: yes
+      toc: no
+      always_allow_html: true
+      number_sections: FALSE
+      fig_height: 7
+      fig_width: 6
+      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
+      pandoc_args:
+      - --lua-filter=../../templates/scholarly-metadata.lua
+      - --lua-filter=../../templates/author-info-blocks.lua
+      - --lua-filter=../../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       toc: no
+      number_sections: FALSE
+      geometry: margin=1in
+      latex_engine: "xelatex"
       pandoc_args:
@@
         in_header: ../../templates/header_appendixS.tex
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../../templates/scholarly-metadata.lua
-      - --lua-filter=../../templates/author-info-blocks.lua
-      - --lua-filter=../../templates/pagebreak.lua
     bookdown::html_document2:
@@
       - --lua-filter=../../templates/pagebreak.lua
+fontsize: 12pt
+mainfont: Times New Roman
 bibliography:
@@
 ```{r packages}
-# setwd('analysis/paper')
+# setwd('analysis/paper/appendix')
 
@@
 options(knitr.kable.NA = '-')
-# options(knitr.table.format = "pandoc")
 
+# what kind of document is being created?
+doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to')
+# when knitting to Word, use this
+if(doc.type == 'docx') {
+  options(knitr.table.format = "pandoc")
+}
+
+# cat(doc.type)
 ```
@@
                            filter(ShortName %in% names(x)) %>%
+                           filter(ShortName != "ChnlUnitTotal_Ct") %>%
                            pull(ShortName)
@@
 
+# create data.frame with correlation coefficients and habitat categories
+corrr_df = hab_corr$catg_cor[[1]] %>%
+  map(.f = as_cordf) %>%
+  map(.f = stretch,
+      remove.dups = T,
+      na.rm = T) %>%
+  map_df(.id = 'Category',
+         .f = identity) %>%
+  left_join(poss_hab_mets %>%
+              select(x = ShortName,
+                     `Metric 1` = Name)) %>%
+  left_join(poss_hab_mets %>%
+              select(y = ShortName,
+                     `Metric 2` = Name)) %>%
+  select(Category,
+         `Metric 1`,
+         `Metric 2`,
+         r) %>%
+  mutate_at(vars(Category),
+            list(str_replace_all),
+            pattern = "([:lower:])([:upper:])",
+            replacement = "\\1 \\2") %>%
+  arrange(Category,
+          desc(abs(r)))
+
 ```
@@
 
-Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)).
+Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figures \@ref(fig:mine-facet-fig) and \@ref(fig:mine-all-fig)).
 The MIC value of each measured habitat characteristic and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.
-We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior.
+We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Table \@ref(tab:corr-tab)), while focusing on covariates we thought could influence fish behavior.
 For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model.
@@
 
+# Results
+
+We chose `r nrow(sel_hab_mets)` metrics, highlighted in bold in Table \@ref(tab:mic-tab).
+Their correlations with each other are displayed in Figure \@ref(fig:selected-corr).
+Those with a high correlation coefficient ($\geq$ 0.5 or $\leq$ -0.5) are plotted against each other in Figure \@ref(fig:high-cor-mets) to show the variety in values even for pairs of metrics with relatively high correlations.
+
 <!-- The following line inserts a page break  -->
@@
 
+if (doc.type != "docx") {
 mine_tab_df %>%
@@
   kable(digits = 3,
-        # format = 'html',
-        # format = 'latex',
         booktabs = T,
@@
   column_spec(c(4:6), width = "0.5in")
+}
+
+if(doc.type == 'docx') {
+  library(pander)
+  mine_tab_df %>%
+    select(-selected) %>%
+    pander(digits = 3,
+           caption = '(\\#tab:mic-tab)MIC statistic for top metrics within each habitat category, sorted by category and MIC value.
+The percent of records for which each habitat metric measurement was missing or zero is also shown.
+Metrics selected for the QRF model are in bold.',
+           emphasize.strong.rows = which(mine_tab_df$selected),
+           split.tables = Inf)
+}
+
+```
+
+```{r corr-tab}
+new_catg_rows = corrr_df %>%
+  mutate(lead_catg = lead(Category),
+         new_catg = if_else(Category != lead_catg, 
+                            T, F)) %>%
+  pull(new_catg) %>%
+  which()
+
+if (doc.type == "latex") {
+  corrr_df %>%
+    kable(digits = 2,
+          booktabs = T,
+          longtable = T,
+          format = "latex",
+          caption = "Pearson correlation coefficient between each variable within a habitat category.") %>%
+    kable_styling(fixed_thead = T,
+                  latex_options = c('repeat_header',
+                                    'scale_down',
+                                    'HOLD_position')) %>%
+    row_spec(new_catg_rows,
+             hline_after = T) %>%
+    column_spec(1, width = "1in") %>%
+    column_spec(c(2, 3), width = "2in")
+}
+
+if (doc.type == "html") {
+  corrr_df %>%
+    kable(digits = 2,
+          booktabs = T,
+          # longtable = T,
+          caption = "Pearson correlation coefficient between each variable within a habitat category.") %>%
+    kable_styling(fixed_thead = T,
+                  latex_options = c('repeat_header',
+                                    'scale_down',
+                                    'HOLD_position')) %>%
+    collapse_rows(columns = 1,
+                  latex_hline = "major",
+                  valign = "top")
+}
+
+if(doc.type == 'docx') {
+  library(pander)
+
+  corrr_df %>%
+    pander(digits = 2,
+           caption = "(\\#tab:corr-tab)Pearson correlation coefficient between each variable within a habitat category.",
+           split.tables = Inf)
+
+}
 
@@
 ```{r mine-all-fig, fig.cap = 'Barplot of MIC statistics, colored by habitat category.'}
-ggplot(data = mine_res,
-       aes(x = Name,
+mine_res %>%
+    mutate_at(vars(MetricCategory),
+            list(str_replace_all),
+            pattern = "([:lower:])([:upper:])",
+            replacement = "\\1 \\2") %>%
+
+ggplot(aes(x = Name,
            y = MIC,
@@
   scale_fill_brewer(palette = 'Set3',
-                    guide = guide_legend(nrow = 2)) +
+                    guide = guide_legend(ncol = 3)) +
   theme(legend.position = 'bottom',
@@
 
-```{r corr-fig, eval = T, results='hide', out.width = "90%", fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category.
+```{r corr-fig, eval = F, results='hide', out.width = "90%", fig.show = "hold", fig.cap = "Correlation plots of metrics, divided by habitat category.
 Larger circles correspond to larger correlation coefficients.
@@
 
-```{r high-cor-mets, fig.cap = 'Pairs plot of habitat metrics used in QRF model with a correlation coefficient greater than 0.5.'}
+```{r high-cor-mets, fig.cap = 'Pairs plot of habitat metrics used in QRF model with a correlation coefficient greater than 0.5 or less than -0.5.'}
 high_cor_mets = hab_data %>%
@@
 ```
-
-<!-- The following line inserts a page break  -->
-\newpage
-
-### Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
-
-The current Git commit details are:
-
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```

### 2020-09-15 analysis/paper/paper.Rmd
made some formatting changes and minor edits for submission to Ecosphere.
made some revisions based on reviewer #2's comments

@@
 
-A number of studies have used other modeling approaches to elicit non-salmonid fish habitat relationships.
+A number of studies have used other modeling approaches to elicit fish habitat relationships.
 @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O.
@@
 Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
-@Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead.
 The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..
@@
 @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
-Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are usually unavailable [@Cramer2009].
+Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are unavailable in most watersheds [@Cramer2009].
 
@@
 We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality.
-Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity.
+Furthermore, we assert that observed fish density is a poor proxy of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity.
 To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance and density and habitat characteristics.
@@
 Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010].
-Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006].
-Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].
+Quantile random forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006].
+In addition, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].
 
@@
 
-CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watershed within the interior Columbia River basin.
+CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watersheds within the interior Columbia River basin.
 The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004].
@@
 The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019].
+The design weights are generated from how much of the watershed each site is meant to represent.
+Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients [@Nahorniak2015].
 We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else.
@@
 To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
-We first restricted the master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists.
+We first restricted the upstream limit of master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists, and the downstream limit by when streams were no longer wadeable (often by some combination of estimated bankfull width and cumulative drainage area).
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
@@
 Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important.
-In the summer, for Chinook salmon parr, there is clearly movement between multiple channel units (e.g., pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit.
+In the summer, for Chinook salmon parr, our fish data clearly shows movement between multiple channel units (e.g., pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit.
 However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 600 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity.
@@
 While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model.
-This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) maybe on the horizon [@Tonina2019].
+This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) may be on the horizon [@Tonina2019].
 
@@
          Covariate = Name) %>%
-  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile regression forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
+  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile random forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
         booktabs = T, 
@@
                              'Chiwawa R.' = 'Chiwawa River')) %>%
-  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile regression forest (QRF) model.
+  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile random forest (QRF) model.
 Numbers in parentheses are coefficients of variation.",
@@
 
-```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile regression forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr"}
+```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile random forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr"}
 qrf_mod$importance %>%
@@
 
-```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile regression forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
+```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile random forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
 Tick marks along the X-axis depict observed values."}

### 2020-09-21 analysis/R_scripts/fit_qrf_model.Rmd
edited code to deal with changes to snapPointsToLines output.
added code (but commented it out) for using fish / m2 as the response.

@@
   mutate(RipCovCanSome = 100 - RipCovCanNone)
+
+# # to use fish / m2
+# fish_hab %<>%
+#   mutate(fish_dens = N / FishWettedArea) %>%
+#   filter(!is.na(fish_dens),
+#          FishWettedArea > 0)
 
@@
 chnk_sites = chnk_samps %>%
+  bind_cols(chnk_samps %>%
               as_Spatial() %>%
@@
                                             as_Spatial(),
-                              maxDist = max_snap_dist,
-                              withAttrs = TRUE,
+                                          withAttrs = F,
                                           idField = 'id') %>%
               as('sf') %>%
-  select(-nearest_line_id, -snap_dist)
+              st_drop_geometry() %>%
+              as_tibble() %>%
+              select(snap_dist)) %>%
+  filter(snap_dist <= max_snap_dist)
 
@@
   unique()
-
-# poss_hab_mets = c(poss_hab_mets, 'aug_temp', 'avg_aug_temp') %>%
-#   unique()
 
@@
 
+# # if using fish / m2 as response
+# # predict at all CHaMP sites
+# pred_hab_sites = hab_impute %>%
+#   mutate(chnk_per_m2 = predict(qrf_mod,
+#                               newdata = select(., one_of(covars)),
+#                               what = pred_quant),
+#          chnk_per_m2 = exp(chnk_per_m2) - dens_offset,
+#          chnk_per_m = chnk_per_m2 * Area_Wet / Lgth_Wet)
+
 # filter out sites outside the Chinook domain
 data("chnk_domain")
-snap_dist = 1000
+my_snap_dist = 1000
 
 pred_hab_sites_chnk = pred_hab_sites %>%
+  filter(!is.na(LON_DD)) %>%
+  bind_cols(pred_hab_sites %>%
               filter(!is.na(LON_DD)) %>%
@@
                                             as_Spatial(),
-                              maxDist = snap_dist,
-                              withAttrs = T,
                                           idField = 'id') %>%
               as('sf') %>%
-  as_tibble() %>%
+              as_tibble()) %>%
+  filter(snap_dist <= my_snap_dist) %>%
   select(-nearest_line_id, -snap_dist, -geometry)

### 2020-09-21 analysis/paper/paper.Rmd
made several edits based on Sarah's comments and response to reviewers.

@@
                    "quantregForest", 
+                   "randomForest",
                    "missForest",
@@
 Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality.
-The most recent “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required.
-These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured.
-Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing.
-For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”
-
-Pacific salmon species have experienced large declines in abundance throughout much of their range [@Good2005].
-Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages.
+Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2018; @Hinrichsen2020; @NOAA2020].
 Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O.
@@
 Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles.
-However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge.
+However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2018; @Hinrichsen2020; @NOAA2020].
 Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.
 
+When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999].
+However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults.
+Such data are unavailable in most watersheds [@Cramer2009], and they do not necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale.
+Bioenergetics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the 200 - 600 m reach scale [@Wall2016].
+However, there are some potential issues with how the food supply (i.e.
+invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Wall2016; @Carmichael2020].
+Taking a different approach, @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
+
+Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity.
+Further, many of these studies focus on only one or two measures of habitat.
 @Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal.
 Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species.
-@Nickelson1992 demonstrated that juvenile coho salmon (*O.
+@Nickelson1992 demonstrated that juvenile Coho salmon (*O.
 kisutch*) were found in higher densities within pool habitat on the Oregon coast.
-Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001].
-@Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools.
+Similarly, pool and pond densities were good predictors of Coho smolt densities in western Washington [@Sharma2001].
+@Bryant2009 found that juvenile Coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools.
 @Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O.
 nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools.
-Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002].
+Densities of adult spawning Coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002].
 @Mossop2006 examined juvenile Chinook salmon (*O.
@@
 The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..
-
-Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity.
-Further, many of these studies focus on only one or two measures of habitat.
-@Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
-Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are unavailable in most watersheds [@Cramer2009].
 
@@
 Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.
+Additionally, this approach for estimating life-stage specific habitat-based carrying capacity can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing.
+Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon.
+For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other "H's."
 
@@
 In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork Salmon.
+CHaMP sampled between 1 and 28% of the Chinook domain within these watersheds, with an average of 11%.
 
@@
 A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity.
-After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)).
+After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)), quantified by the average decrease in residual sum of squares for splits on that variable amidst the trees in the random forest, implemented by the *importance* function from the *randomForest* package [@randomForest2002].
 Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots.
@@
 
-## What Carrying Capacity Models Are Currently Available?
-
-When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999].
-However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults.
-Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale.
-QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single (or very few) season(s).
-Or the dataset could be built over time, sampling different locations in different years.
-If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.
-
-Historically, fish-habitat relationship models have focused on species distribution or average abundance/density rather than directly addressing carrying capacity.
-Bioenergics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the same reach scale that our QRF model operates on [@Wall2016].
-However, there are some potential issues with how the food supply (i.e.
-invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Carmichael2020; @Wall2016].
-The QRF approach described here allows us to analyze multiple (often correlated) habitat metrics and assess the potentially non-linear fish habitat relationships that define capacity as a function of habitat.
+There are two aspects that make this approach "data hungry", meaning a large dataset is needed to fit a QRF model like this.
+First, random forest models generally require more data that parametric models, due to the lack of parametric distribution assumptions and the lack of an assumed form of the relationship between dependent and independent variables.
+Second, it takes larger data sets to accurately predict the lower and higher quantiles in a quantile regression framework.
+For example, if a data set consisted of thousands, rather than hundreds, of data points, a researcher might feel comfortable using the 95th or the 98th quantile as a proxy for capacity, rather than the 90th.
+Our data set consisted of `r n_distinct(qrf_data$Site)` sites, across a variety of habitats and years, providing contrast in all the habitat covariates and presumeably satisfying the data hungriness of a QRF model, based on our validation with spawner-recruit data.
 

### 2020-09-21 analysis/paper/paper.Rmd
made a few more edits to the manuscript, increased the font size in some figures, added a model framework figure, dropped the colophon

@@
 output:
-    bookdown::word_document2:
+    bookdown::html_document2:
       fig_caption: yes
-      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
+      number_sections: false
       toc: no
-      always_allow_html: true
+      fig_width: 7
       pandoc_args:
@@
       fig_caption: yes
-      number_sections: FALSE
+      number_sections: false
       toc: no
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
-      number_sections: FALSE
+      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: false
       toc: no
+      always_allow_html: true
+      fig_width: 6
       pandoc_args:
@@
 csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
-# csl: "../templates/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
 abstract: |
@@
 invertebrate drift) is measured with these methods that could lead to biases in capacity estimates [@Dodrill2016] as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach [@Wall2016; @Carmichael2020].
-Taking a different approach, @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
+In addition, those authors did not take the step of scaling the capacity predictions at the reach scale to entire watersheds.
+In contrast, @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr at the watershed scale, using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
+Estimates of salmonid carrying capacity that leverage fish-habitat relationships are lacking for the watershed scale in the Pacific Northwest.
+
+<!-- How's that transition sentence at the end of the paragraph above? -->
 
@@
 
-A number of studies have used other modeling approaches to elicit fish habitat relationships.
+A number of studies have used other modeling approaches to elicit fish-habitat relationships.
 @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O.
@@
 Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
-The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..
+The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish-habitat relationships and provide motivation for furthering research in this realm..
 
@@
 * Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
-* Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
-* Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River basin),
-* Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
+* Elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density, using paired fish and habitat measurements,
+* Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured,
+* Extrapolate capacity predictions at measured habitat sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
 * Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
@@
 Additionally, this approach for estimating life-stage specific habitat-based carrying capacity can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing.
-Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon.
+Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish-habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon.
 For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other "H's."
@@
 For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity.
-After removing duplicate samples, our initial dataset contained `r n_distinct(qrf_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fish-hab-sites)).
+After removing duplicate sites, our initial dataset contained `r n_distinct(qrf_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fish-hab-sites)).
+We did explore using areal fish densities (parr/m$^2$) as the response, but found very similar results so in the interest of brevity we only present results based on linear fish densities.
 
@@
 We sought to include a small set of covariates that were not overly redundant (i.e., not highly correlated), described many aspects of stream habitat (e.g., substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided.
-Full details of how covariates used in the QRF model were selected can be found in Appendix S1.
+Full details of how the `r as.character(english(n_distinct(sel_hab_mets$Metric)))` covariates used in the QRF model were selected can be found in Appendix S1.
 
@@
 We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.
+These points do not represent specific segments of streams, however, so we needed to do some spatial averaging of capacity predictions to generate larger scale capacity estimates.
 
 To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary.
-We first restricted the upstream limit of master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists, and the downstream limit by when streams were no longer wadeable (often by some combination of estimated bankfull width and cumulative drainage area).
+We first restricted the upstream limit of master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists, and the downstream limit by when streams were no longer wadeable (often determined by some combination of estimated bankfull width and cumulative drainage area).
 The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.
+A conceptual diagram showing the data and modeling framework of the QRF and extrapolation models is shown in Figure \@ref(fig:model-diag).
 
@@
 Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of Chinook salmon parr capacity.
-Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e., fall emigrants) from rotary screw traps.
-For the latter, estimates of parr were calculated using estimates of over-winter survival.
+Some watersheds had direct estimates of parr, while some had estimates of pre-smolts and smolts (i.e., fall and spring emigrants) from rotary screw traps.
+For the latter, estimates of parr were calculated using estimates of over-winter survival to back-calculate parr from smolt estimates, and then adding that to pre-smolt estimates.
 A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick [@Froese2008], using the *FSA* package [@R-FSA] in R.
@@
 In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and surrogate for the number of pool-riffle sequences or potential sheer areas providing feeding zones), and fish cover.
+Metrics that describe habitat quantity set some bounds around possible capacity estimates, while metrics describing habitat quality refine those estimates to better match conditions at that site.
 
@@
         booktabs = T) %>%
-  kable_styling()
+  kable_styling(latex_options = c("hold_position"))
 
@@
         format.args = list(width = c(2, 10, 10, 50))) %>%
-  kable_styling() %>%
+  kable_styling(latex_options = c("hold_position")) %>%
   column_spec(4, width = '3in')
@@
         digits = 3) %>%
-  kable_styling()
+  kable_styling(latex_options = c("hold_position"))
 
@@
         booktabs = T) %>%
-  kable_styling() %>%
+  kable_styling(latex_options = c("hold_position")) %>%
   # kable_styling(latex_options = c('scale_down')) %>%
@@
 
-```{r wtsd-map, fig.cap = "Watersheds with CHaMP habitat data.
+```{r wtsd-map, out.width = "100%", fig.cap = "Watersheds with CHaMP habitat data.
 Watersheds in black also contain paired fish data.
@@
 knitr::include_graphics('../figures/ChampWtds.png')
-# knitr::include_graphics('../figures/ChampWtds.jpeg')
+```
+
+\newpage
+
+```{r model-diag, out.width = "100%", fig.cap = "Conceptual diagram showing input data sources, modeling decisions, model outputs etc. for the QRF and extrapolation models."}
+knitr::include_graphics("../figures/QRF_flowchart.png")
 ```
@@
   theme(legend.position = 'none',
-        strip.text = element_text(size = 6))
+        strip.text = element_text(size = 10))
 ```
@@
 spawn_rec_comp_p +
-  theme(strip.text = element_text(size = 7))
+  theme(strip.text = element_text(size = 10))
 ```
-
-<!-- The following line inserts a page break  -->
-\newpage
-
-### Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE, eval = T}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
-
-The current Git commit details are:
-
-```{r, eval = T}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```

### 2020-09-22 analysis/paper/paper.Rmd
updated Coho to coho

@@
 Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species.
-@Nickelson1992 demonstrated that juvenile Coho salmon (*O.
+@Nickelson1992 demonstrated that juvenile coho salmon (*O.
 kisutch*) were found in higher densities within pool habitat on the Oregon coast.
-Similarly, pool and pond densities were good predictors of Coho smolt densities in western Washington [@Sharma2001].
-@Bryant2009 found that juvenile Coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools.
+Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001].
+@Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools.
 @Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O.
 nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools.
-Densities of adult spawning Coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002].
+Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002].
 @Mossop2006 examined juvenile Chinook salmon (*O.

### 2020-09-22 R/plot_partial_dependence.R
removed theme_bw() from plot_partial_dependence function

@@ -4,14 +4,14 @@
 #'
 #' @author Kevin See
 #'
-#' @param rf_mod the \code{quantregForest} model 
+#' @param rf_mod the \code{quantregForest} model
 #' @param data the data.frame used to fit the \code{rf_mod}
 #' @param plot_covars character vector of covariates to plot
 #' @param data_dict data.frame containing columns \code{}
 #' @param type whether to predict a particular quantile, or the mean
 #' @param pred_quantile if \code{type} is "quantile", which quantile to predict, between 0 and 1? The default value is 0.9
-#' @param n_pts how many points to use in predictions? 
-#' @param log_transform was the response log transformed? 
+#' @param n_pts how many points to use in predictions?
+#' @param log_transform was the response log transformed?
 #' @param log_offset if an offset was used before log transforming, enter it here.
 #' @param scales ggplot facet_wrap argument \code{scales} for facetting
 #' @param ... other arguements to be passed to \code{ggplot}
@@ -30,14 +30,14 @@ plot_partial_dependence = function(rf_mod,
                                    log_transform = T,
                                    log_offset = 0.005,
                                    scales = 'free') {
-  
+
   if(is.null(data_dict)) {
     data(hab_dict_2017)
     data_dict = hab_dict_2017
   }
-  
+
   type = match.arg(type)
-  
+
   # relative importance of covariates
   rel_imp = as_tibble(rf_mod$importance,
                       rownames = 'Metric') %>%
@@ -45,30 +45,30 @@ plot_partial_dependence = function(rf_mod,
     mutate_at(vars(Metric),
               list(~ fct_reorder(., relImp))) %>%
     arrange(desc(Metric))
-  
+
   # names of covariates
   covars = rel_imp$Metric
   if(is.null(plot_covars)) plot_covars = covars
-  
+
   # get means and ranges of all
   covar_range = data %>%
     select(one_of(as.character(plot_covars))) %>%
     gather(Metric, value) %>%
     group_by(Metric) %>%
     summarise_at(vars(value),
-                 list(mean = mean, 
-                      median = median, 
-                      min = min, 
+                 list(mean = mean,
+                      median = median,
+                      min = min,
                       max = max),
                  na.rm = T) %>%
     ungroup()
-  
+
   # create data.frame with sequence of values across each covariate, keeping other covariates at their average value
   pdp_df = covar_range %>%
     split(.$Metric) %>%
     map_df(.id = 'Metric',
[…]

### 2020-09-23 analysis/paper/paper.Rmd
a few more small revisions, based on an internal review. Removed comments that were no longer relevant.

@@
 output:
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: false
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: false
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
   Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
-A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods often fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds.
+A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds.
 Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available.
@@
 keywords: |
-  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
+  carrying capacity; quantile regression; random forests; Chinook; quantile random forest
 highlights: |
@@
 ```{r packages}
+# run this if running chunks directly
+# setwd('analysis/paper')
+
+# what packages are necessary
 library(survey)
@@
 library(kableExtra)
+library(ggpubr)
 
-theme_set(theme_bw())
+theme_set(theme_pubr())
 
@@
 
-# run this if running chunks directly
-# setwd('analysis/paper')
+if(doc.type %in% c("docx", "html")) {
+  theme_set(theme_pubr(base_family = "Times New Roman"))
+}
+
 ```
@@
 The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend.
-These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation.
+These actions are often categorized into four H’s: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation.
 Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality.
@@
 When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999].
-However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults.
+However, such models require a long time-series of accurate estimates of abundance for adults and juveniles, with variation in the number of adults.
 Such data are unavailable in most watersheds [@Cramer2009], and they do not necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale.
@@
 In contrast, @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr at the watershed scale, using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area.
-Estimates of salmonid carrying capacity that leverage fish-habitat relationships are lacking for the watershed scale in the Pacific Northwest.
-
-<!-- How's that transition sentence at the end of the paragraph above? -->
+Estimates of salmonid carrying capacity that leverage fish-habitat relationships are lacking at the watershed scale in the Pacific Northwest.
 
@@
 Further, many of these studies focus on only one or two measures of habitat.
-@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal.
+@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985 and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal.
 Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species.
@@
 @Mossop2006 examined juvenile Chinook salmon (*O.
-tshawytscha*) in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient.
+tshawytscha*) in the Yukon River and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient.
 These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed.
@@
 The same approach was also used to map the potential extent of sole (*Solea solea*) in the English Channel and southern North Sea [@Eastwood2003].
-Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012].
-The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish-habitat relationships and provide motivation for furthering research in this realm..
+Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; @Compton2012].
+The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish-habitat relationships and provide motivation for furthering research in this realm.
 
@@
 
-* Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
-* Elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density, using paired fish and habitat measurements,
-* Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured,
-* Extrapolate capacity predictions at measured habitat sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
-* Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
+* identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
+* elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density, using paired fish and habitat measurements,
+* predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured,
+* extrapolate capacity predictions at measured habitat sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
+* validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
 
@@
 Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish-habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon.
-For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other "H's."
+For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated, and the magnitude of required change can be placed in context with the other "H's."
 
@@
 
-CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watersheds within the interior Columbia River basin.
+CHaMP sites are 200 m to 600 m reaches in wadeable streams across select watersheds within the interior Columbia River basin.
 The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004].
@@
 Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions.
-We included a random effect for fish crew / watershed.
+We included a random effect for fish crew/watershed.
 We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.
@@
 After removing duplicate sites, our initial dataset contained `r n_distinct(qrf_data$Site)` unique sites with paired fish-habitat data (Table \@ref(tab:fish-hab-sites)).
-We did explore using areal fish densities (parr/m$^2$) as the response, but found very similar results so in the interest of brevity we only present results based on linear fish densities.
+We did explore using areal fish densities (parr/m$^2$) as the response but found very similar results so in the interest of brevity we only present results based on linear fish densities.
 
@@
 To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population.
-The GAA data used here was taken from the list of generalized random tesselation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
-Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)).
+The GAA data used here was taken from the list of generalized random tessellation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016].
+Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classification [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)).
 The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019].
@@
 
-All code and data for the analyses presented here can be found in a [GitHub repository](https://github.com/KevinSee/QRFpaper).
+All code and data for the analyses presented here can be found in a GitHub repository ([(https://github.com/KevinSee/QRFpaper)](https://github.com/KevinSee/QRFpaper)).
 
@@
 In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months and in wadeable streams.
-Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting.
+Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation that may be necessary to support Endangered Species Act delisting.
 The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations.
 Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult).
-Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting.
+Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stages and/or specific habitat patches may be limiting.
 As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance.
@@
 There are two aspects that make this approach "data hungry", meaning a large dataset is needed to fit a QRF model like this.
-First, random forest models generally require more data that parametric models, due to the lack of parametric distribution assumptions and the lack of an assumed form of the relationship between dependent and independent variables.
+First, random forest models generally require more data than parametric models, due to the lack of parametric distribution assumptions and the lack of an assumed form of the relationship between dependent and independent variables.
 Second, it takes larger data sets to accurately predict the lower and higher quantiles in a quantile regression framework.
 For example, if a data set consisted of thousands, rather than hundreds, of data points, a researcher might feel comfortable using the 95th or the 98th quantile as a proxy for capacity, rather than the 90th.
-Our data set consisted of `r n_distinct(qrf_data$Site)` sites, across a variety of habitats and years, providing contrast in all the habitat covariates and presumeably satisfying the data hungriness of a QRF model, based on our validation with spawner-recruit data.
+Our data set consisted of `r n_distinct(qrf_data$Site)` sites, across a variety of habitats and years, providing contrast in all the habitat covariates and presumably satisfying the data hungriness of a QRF model, based on our validation with spawner-recruit data.
 
@@
 
-<!-- Do we need any more in this section? I commented out the paragraph below... -->
-
-<!-- While the Columbia River basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging.
-On-the-ground habitat data collection is costly, time-consuming, and often not reproducible; thereby limiting the effectiveness of this type of extrapolation.
-Additionally, the extrapolation model relies on Global Available Attribute (GAA) data which ….
--->
-
-<!-- Add here why GAA are messy. And then anything else which may be a consideration for using the extrapolation model moving forward.  -->
-
 ## The Future: Improving Habitat Data
-
-<!-- Discussion of improving habitat data to possibly avoid extrapolation. Collect cheaper/better spatially continuous habitat data. -->
-
-<!-- I have no idea if this is what you are looking for so feel free to delete if I am way off base.  -->
 
@@
 
-<!-- The beginning of the following paragraph may need some word-smithing -->
-
 If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment.
@@
 
-Model development efforts have been funded by the Bonneville Power Administration through projects 2003-017-00 and 2011-006-00 and by the Bureau of Reclamation and Idaho Office of Species Conservation through contract BOR002 16.
-Fish sampling work in the Lemhi River was also funded through the Idaho Office of Species Conservation through the Pacific Coast Salmon Recovery Fund.
+Model development efforts have been funded by the Bonneville Power Administration through projects 2003-017-00 and 2011-006-00 and by the Bureau of Reclamation and Idaho Governor's Office of Species Conservation through contract BOR002 16.
+Fish sampling work in the Lemhi River was also funded through the Idaho Governor's Office of Species Conservation through the Pacific Coast Salmon Recovery Fund.
 Special thanks to staff from the following agencies for providing data to calculate Chinook salmon parr abundance and density estimates: Columbia River Inter-Tribal Fish Commission, Oregon Department of Fish and Wildlife, and U.S.
@@
 
-```{r wtsd-map, out.width = "100%", fig.cap = "Watersheds with CHaMP habitat data.
+```{r wtsd-map, out.width = "90%", fig.cap = "Watersheds with CHaMP habitat data.
 Watersheds in black also contain paired fish data.
@@
 
-```{r rel-imp-figure, fig.cap = "Relative importance of each habitat covariate included in the quantile random forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr"}
+```{r rel-imp-figure, fig.height=5, fig.cap = "Relative importance of each habitat covariate included in the quantile random forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr"}
 qrf_mod$importance %>%
@@
 
-```{r pdp-figure, fig.cap = "Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile random forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
+```{r pdp-figure, fig.height = 6, fig.cap = "Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile random forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
 Tick marks along the X-axis depict observed values."}
@@
   theme(legend.position = 'none',
-        strip.text = element_text(size = 10))
+        strip.text = element_text(size = 10),
+        axis.text = element_text(size = 8))
 ```
@@
 
-```{r sr-figure, fig.height = 5, fig.cap = "Spawner-recruit data from nine watersheds.
+```{r sr-figure, fig.height = 6, fig.cap = "Spawner-recruit data from nine watersheds.
 Solid lines are the spawner-recruit curve, dashed lines are the estimated capacity, and shaded polygons depict the 95% confidence intervals of capacity.
@@
 spawn_rec_comp_p +
-  theme(strip.text = element_text(size = 10))
+  theme(strip.text = element_text(size = 10),
+        axis.text = element_text(size = 8))
 ```

### 2020-12-01 analysis/paper/paper.Rmd
for final submission, kept the .tex file, changed format of saved figures. Also added linesep = "" to all tables, and added list of figures

@@
       latex_engine: "xelatex"
+      keep_tex: true
       pandoc_args:
@@
   fig.path = "../figures/",
-  dpi = 600
+  dpi = 400,
+  dev = c("pdf", 'tiff')
 )
@@
 
-theme_set(theme_pubr())
+theme_set(theme_pubr(base_family = "Helvetica"))
 
@@
   kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer-run Chinook salmon parr capacity QRF model',
-        booktabs = T) %>%
+        booktabs = T,
+        linesep = "") %>%
   kable_styling(latex_options = c("hold_position"))
@@
         booktabs = T,
+        linesep = "",
         format.args = list(width = c(2, 10, 10, 50))) %>%
@@
         booktabs = T,
+        linesep = "", 
         escape = F,
@@
         booktabs = T,
+        linesep = "",
         digits = 3) %>%
@@
 Numbers in parentheses are coefficients of variation.",
-        booktabs = T) %>%
+        booktabs = T,
+        linesep = "") %>%
   kable_styling(latex_options = c("hold_position")) %>%
@@
 ```
+
+\newpage
+
+\listoffigures
 

### 2020-12-01 analysis/paper/paper.tex
for final submission, kept the .tex file, changed format of saved figures. Also added linesep = "" to all tables, and added list of figures

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models},
  pdfauthor={Kevin E. See1,, Michael W. Ackerman1, Richard A. Carmichael1, Sarah L. Hoffmann1, and Chris Beasley1},
  pdfkeywords={carrying capacity; quantile regression; random forests; Chinook; quantile random forest},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% double spacing
\usepackage{setspace}
\doublespacing
% to double space figure captions
\usepackage{caption}
\captionsetup{font = doublespacing}

% line numbers
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models}
\author{Kevin E. See\textsuperscript{1,*}, Michael W. Ackerman\textsuperscript{1}, Richard A. Carmichael\textsuperscript{1}, Sarah L. Hoffmann\textsuperscript{1}, and Chris Beasley\textsuperscript{1}}
\date{December 01, 2020}

\begin{document}
\maketitle
\begin{abstract}
Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
\end{abstract}

\textsuperscript{1} Biomark, Inc.~705 South 8th St., Boise, Idaho, 83702, USA

\textsuperscript{*} Correspondence: \href{mailto:Kevin.See@merck.com}{Kevin E. See \textless{}\href{mailto:Kevin.See@merck.com}{\nolinkurl{Kevin.See@merck.com}}\textgreater{}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The decline of anadromous Pacific salmonids (\emph{Oncorhynchus spp.}) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H's: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids (Bond et al. 2018, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020). Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (\emph{O. mykiss}) in the Pacific Northwest, USA (NOAA Fisheries 2016a, 2016b). Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge (Bond et al. 2018, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020). Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard (Moussalli and Hilborn 1986, Myers et al. 1999). However, such models require a long time-series of accurate estimates of abundance for adults and juveniles, with variation in the number of adults. Such data are unavailable in most watersheds (Cramer and Ackerman 2009), and they do not necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. Bioenergetics approaches, such as the net rate of energy intake (NREI) have been applied to salmonids to estimate capacity on the 200 - 600 m reach scale (Wall et al. 2016). However, there are some potential issues with how the food supply (i.e.~invertebrate drift) is measured with these methods that could lead to biases in capacity estimates (Dodrill and Yackulic 2016) as well as difficulty in properly constraining drift depletion and inter-species competition, and computational and spatial limitation of this modelling approach (Wall et al. 2016, Carmichael et al. 2020). In addition, those authors did not take the step of scaling the capacity predictions at the reach scale to entire watersheds. In contrast, Sweka and Mackey (2010) estimated carrying capacity of Atlantic salmon (\emph{Salmo salar}) parr at the watershed scale, using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Estimates of salmonid carrying capacity that leverage fish-habitat relationships are lacking at the watershed scale in the Pacific Northwest.

Most studies that have investigated fish-habitat relationships focus on predicting a species' distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. Fausch et al. (1988) conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985 and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. Nickelson et al. (1992) demonstrated that juvenile coho salmon (\emph{O. kisutch}) were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington (Sharma and Hilborn 2001). Bryant and Woodsmith (2009) found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. Braun and Reynolds (2011) similarly found positive associations between spawner densities of sockeye salmon (\emph{O. nerka}) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed (Pess et al. 2002). Mossop and Bradford (2006) examined juvenile Chinook salmon (\emph{O. tshawytscha}) in the Yukon River and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have used other modeling approaches to elicit fish-habitat relationships. Dunham et al. (2002) used a quantile regression approach to show a negative relationship between cutthroat trout (\emph{O. clarkii}) densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole (\emph{Solea solea}) in the English Channel and southern North Sea (Eastwood et al. 2003). Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species (Pittman et al. 2009, Knudby et al. 2010, Compton et al. 2012). The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish-habitat relationships and provide motivation for furthering research in this realm.

For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor proxy of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) (Meinshausen 2006) models using measurements of fish abundance and density and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling (Koenker and Bassett Jr 1978, Cade and Noon 2003) with the flexibility and framework of random forest models (Breiman 2001). They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts (Knudby et al. 2010). Quantile random forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers (Prasad et al. 2006). In addition, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors (Terrell et al. 1996, Cade and Noon 2003).

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts. Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings. Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened (NOAA Fisheries 2016a, 2016b). Hereafter, we refer to both ESUs simply as Chinook salmon.

In this study, we developed a QRF model to:

\begin{itemize}
\tightlist
\item
  identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
\item
  elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density, using paired fish and habitat measurements,
\item
  predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured,
\item
  extrapolate capacity predictions at measured habitat sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
\item
  validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).
\end{itemize}

Our study incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations. Additionally, this approach for estimating life-stage specific habitat-based carrying capacity can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. Given the multitude of (often correlated) habitat metrics and the potentially non-linear fish-habitat relationships that define capacity as a function of habitat, we explore the application of QRF modeling to habitat capacity estimation, validated using data from Columbia River Chinook salmon. For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated, and the magnitude of required change can be placed in context with the other ``H's.''

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{study-site}{%
\subsection{Study Site}\label{study-site}}

Habitat data used in our study were collected from eleven watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \ref{fig:wtsd-map}). The Columbia River basin covers more than 668,000 km\(^2\) draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) (Volk et al. 2017) and were downloaded from \url{https://www.champmonitoring.org}. Data from the following eleven CHaMP watersheds were used in this study: Asotin, Entiat, John Day, Lemhi, Methow, Minam, South Fork Salmon, Tucannon, Upper Grande Ronde, Wenatchee and Yankee Fork. Juvenile density and abundance data were collected in a subset of seven watersheds (see Table \ref{tab:fish-hab-sites} and Figure \ref{fig:wtsd-map}), at CHaMP survey reaches and were graciously provided by a number of agencies and projects, including the Integrated Status and Effectiveness Monitoring Project (Volk et al. 2017).

\hypertarget{data}{%
\subsection{Data}\label{data}}

CHaMP sites are 200 m to 600 m reaches in wadeable streams across select watersheds within the interior Columbia River basin. The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm (Stevens Jr and Olsen 1999, 2004). Habitat data within CHaMP sites were collected using the CHaMP protocol (CHaMP 2016) which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, single-pass electrofishing, and snorkeling. These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator (Carle and Strub 1978), following advice from Hedger et al. (2013). Two-pass removal estimates used the estimator described by Seber (2002). Mark-recapture estimates used Chapman's modified Lincoln-Peterson estimator (Chapman 1951) and were deemed valid if they met the criteria described in Robson and Regier (1964). These estimates were made using the \emph{removal} function from the \emph{FSA} package (Ogle et al. 2020) or the \emph{closedp.bc} function from the \emph{Rcapture} package (Rivest and Baillargeon 2019) in R software (R Core Team 2019). Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew/watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

Abundance estimates at all sites were then translated into linear (parr/m) fish densities which were paired with the associated CHaMP habitat data. For sites that were sampled in multiple years, only the fish and habitat data from the year with the highest observed fish density was retained to avoid possible pseudo-replication, while remaining consistent with our goal of estimating carrying capacity. After removing duplicate sites, our initial dataset contained 327 unique sites with paired fish-habitat data (Table \ref{tab:fish-hab-sites}). We did explore using areal fish densities (parr/m\(^2\)) as the response but found very similar results so in the interest of brevity we only present results based on linear fish densities.

\hypertarget{habitat-covariate-selection}{%
\subsection{Habitat Covariate Selection}\label{habitat-covariate-selection}}

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables. We sought to include a small set of covariates that were not overly redundant (i.e., not highly correlated), described many aspects of stream habitat (e.g., substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided. Full details of how the twelve covariates used in the QRF model were selected can be found in Appendix S1.

\hypertarget{qrf-model-fit}{%
\subsection{QRF Model Fit}\label{qrf-model-fit}}

Using the selected habitat covariates (Table \ref{tab:select-mets-table}), we fit a QRF model to predict habitat rearing capacity for Chinook salmon parr during summer months using the natural log of fish densities as the response. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. The individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response; therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the \emph{missForest} R package (Stekhoven and Bühlmann 2012, Stekhoven 2013). We fit the QRF models using the \emph{quantregForest} function from the \emph{quantregForest} package (Meinshausen 2017) in R software (R Core Team 2019), incorporating data from 327 records (paired fish-habitat data) and twelve habitat covariates (27.2 data points per covariate) (Table \ref{tab:select-mets-table}). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity following the suggestion of Sweka and Mackey (2010), and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

After model fitting, the QRF model was then used to predict capacity at sites with measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that were sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated 589 predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork Salmon. CHaMP sampled between 1 and 28\% of the Chinook domain within these watersheds, with an average of 11\%.

\hypertarget{extrapolating-to-other-sites}{%
\subsection{Extrapolating to Other Sites}\label{extrapolating-to-other-sites}}

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tessellation stratified master sample sites that the CHaMP sites were originally selected from (Larsen et al. 2008, 2016). Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classification (Whittier et al. 2011), the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \ref{tab:gaa-covars}). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the \emph{svyglm} function from the \emph{survey} package (Lumley 2020) in R software (R Core Team 2019). The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients (Nahorniak et al. 2015). We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. These points do not represent specific segments of streams, however, so we needed to do some spatial averaging of capacity predictions to generate larger scale capacity estimates.

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the upstream limit of master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet (\url{http://www.streamnet.org}) or using expert opinion from local biologists, and the downstream limit by when streams were no longer wadeable (often determined by some combination of estimated bankfull width and cumulative drainage area). The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale. A conceptual diagram showing the data and modeling framework of the QRF and extrapolation models is shown in Figure \ref{fig:model-diag}.

\hypertarget{model-validation}{%
\subsection{Model Validation}\label{model-validation}}

Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of pre-smolts and smolts (i.e., fall and spring emigrants) from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival to back-calculate parr from smolt estimates, and then adding that to pre-smolt estimates. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick (Froese 2008), using the \emph{FSA} package (Ogle et al. 2020) in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

All code and data for the analyses presented here can be found in a GitHub repository (\href{https://github.com/KevinSee/QRFpaper}{(https://github.com/KevinSee/QRFpaper)}).

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{habitat-covariate-selection-1}{%
\subsection{Habitat Covariate Selection}\label{habitat-covariate-selection-1}}

We categorized 165 habitat measurements collected using the CHaMP habitat protocol (CHaMP 2016) into eleven habitat categories, and for each habitat covariate the Maximal Information Criteria (MIC) value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (parr/m) (See Appendix S1 for further details). We chose the following twelve CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average August temperature, wetted width:depth ratio, percent fines less than 6 mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \ref{tab:select-mets-table}).

\hypertarget{qrf-model}{%
\subsection{QRF Model}\label{qrf-model}}

A QRF model was fit using those metrics and the \emph{quantregForest} package (Meinshausen 2006) in R (R Core Team 2019) and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \ref{fig:rel-imp-figure}), quantified by the average decrease in residual sum of squares for splits on that variable amidst the trees in the random forest, implemented by the \emph{importance} function from the \emph{randomForest} package (Liaw and Wiener 2002). Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots. These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \ref{fig:pdp-figure}). However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another and care should be taken when interpreting them. After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

\hypertarget{extrapolating-to-other-sites-1}{%
\subsection{Extrapolating to Other Sites}\label{extrapolating-to-other-sites-1}}

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \ref{tab:gaa-covars} and the summary of the model fit in Table \ref{tab:extrap-r2-table}. From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, each representing roughly one kilometer of stream length.

\hypertarget{model-validation-1}{%
\subsection{Model Validation}\label{model-validation-1}}

Estimates of Chinook salmon parr capacity from the QRF and extrapolation models were comparable to independent estimates from spawner-recruit data (Table \ref{tab:sr-table}, Figure \ref{fig:sr-figure}). QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the nine locations where comparisons were possible (Figure \ref{fig:sr-figure}). Potential additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates. Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker). This favorable comparison provides strong validation as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{a-tool-to-estimate-habitat-capacity}{%
\subsection{A Tool to Estimate Habitat Capacity}\label{a-tool-to-estimate-habitat-capacity}}

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months and in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation that may be necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stages and/or specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions and help guide not only the type of action, but also the location at which an action is performed.

The favorable comparison between QRF estimates of carrying capacity and the spawner-recruit based estimates in select watersheds helps support and validate this approach. Although built from completely different data, when these multiple lines of evidence converge it lends credence to the QRF capacity prediction results.

There are two aspects that make this approach ``data hungry'', meaning a large dataset is needed to fit a QRF model like this. First, random forest models generally require more data than parametric models, due to the lack of parametric distribution assumptions and the lack of an assumed form of the relationship between dependent and independent variables. Second, it takes larger data sets to accurately predict the lower and higher quantiles in a quantile regression framework. For example, if a data set consisted of thousands, rather than hundreds, of data points, a researcher might feel comfortable using the 95th or the 98th quantile as a proxy for capacity, rather than the 90th. Our data set consisted of 327 sites, across a variety of habitats and years, providing contrast in all the habitat covariates and presumably satisfying the data hungriness of a QRF model, based on our validation with spawner-recruit data.

\hypertarget{biological-expectations-from-qrf-model}{%
\subsection{Biological Expectations from QRF Model}\label{biological-expectations-from-qrf-model}}

The results of the QRF parr capacity model for Chinook salmon meet many biological expectations. Focusing on the partial dependence plots (Figure \ref{fig:pdp-figure}), the QRF model predicts capacity to increase when the wetted width, discharge and the width:depth ratio grow, when temperatures are cooler (Brett 1952, Raleigh et al. 1986, Bjornn and Reiser 1991), when there is less fine sediment (Hillman et al. 1987, Bjornn and Reiser 1991, Allen 2000), when there is more fish cover (Hillman et al. 1987, Bjornn and Reiser 1991, Holecek et al. 2009), when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases. These are all patterns that emerged from the fish-habitat data, and where available, match those fish-habitat relationships identified qualitatively in other studies (Mossop and Bradford 2006).

The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response. In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and surrogate for the number of pool-riffle sequences or potential sheer areas providing feeding zones), and fish cover. Metrics that describe habitat quantity set some bounds around possible capacity estimates, while metrics describing habitat quality refine those estimates to better match conditions at that site.

\hypertarget{extrapolation-model}{%
\subsection{Extrapolation Model}\label{extrapolation-model}}

Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important. In the summer, for Chinook salmon parr, our fish data clearly shows movement between multiple channel units (e.g., pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit. However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 600 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity. At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales. While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population levels. This is an efficient technique to leverage existing relationships for meaningful management decisions.

Our extrapolation model was focused on extrapolating to other master sample points, because that is the dataset available to us, but the methodology could be improved. Extrapolating to reaches on a stream network, as opposed to points on the landscape, could improve the interpretability of the results. This would require a stream network with relevant attributes attached to each reach, similar to the GAAs we used. Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and building a QRF model from that dataset. Even if the fish data were not collected continuously, estimates of capacity could be made directly from the QRF model across the entire stream network without the need for an extrapolation model.

One of the potential downsides to the extrapolation approach used here is that the GAAs generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale, but with improvements in remote sensing, spatially continuous data (modeled or measured) may be on the horizon (Tonina et al. 2019).

\hypertarget{the-future-improving-habitat-data}{%
\subsection{The Future: Improving Habitat Data}\label{the-future-improving-habitat-data}}

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses. Unmanned Aerial Systems (UAS or drones, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency (Jones IV et al. 2006, Chabot and Bird 2015). UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post-processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection (Whitehead and Hugenholtz 2014, LeCun et al. 2015, Weinstein 2018). Further, developments in Light Detection and Ranging (LiDAR) technology have allowed for the characterization of watershed scale geomorphologic and hydraulic variables not previously possible (McKean et al. 2008, Tonina et al. 2019).

Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly increase the broadscale application of QRF. Rapid advances in drone technology further improve upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated to automated data post-processing. All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts. The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat rehabilitation prioritization and planning.

\hypertarget{conclusions-and-next-steps}{%
\subsection{Conclusions and Next Steps}\label{conclusions-and-next-steps}}

If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment. This understanding could be of crucial importance in the realm of conservation when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally. We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict a habitat's capacity. As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys. The framework could also be applied to depleted, non-migratory, and isolated populations (e.g., desert pupfish \emph{Cyprinodon macularius}) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed. Capacity estimates could also be used to evaluate potential translocation sites to determine if those sites could support an abundance considered viable before investing in translocation efforts.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

Model development efforts have been funded by the Bonneville Power Administration through projects 2003-017-00 and 2011-006-00 and by the Bureau of Reclamation and Idaho Governor's Office of Species Conservation through contract BOR002 16. Fish sampling work in the Lemhi River was also funded through the Idaho Governor's Office of Species Conservation through the Pacific Coast Salmon Recovery Fund. Special thanks to staff from the following agencies for providing data to calculate Chinook salmon parr abundance and density estimates: Columbia River Inter-Tribal Fish Commission, Oregon Department of Fish and Wildlife, and U.S. Fish and Wildlife Service. The models in this study were improved by conversations with Eric Buhle.

\newpage

\hypertarget{literature-cited}{%
\section{Literature Cited}\label{literature-cited}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Allen2000}{}%
Allen, M. A. 2000. Seasonal microhabitat use by juvenile spring chinook salmon in the yakima river basin, washington. Rivers 7:314--332.

\leavevmode\hypertarget{ref-Bjornn1991}{}%
Bjornn, T., and D. Reiser. 1991. Habitat requirements of salmonids in streams. American Fisheries Society Special Publication 19:138.

\leavevmode\hypertarget{ref-Bond2018}{}%
Bond, M. H., T. G. Nodine, T. J. Beechie, and R. W. Zabel. 2018. Estimating the benefits of widespread floodplain reconnection for columbia river chinook salmon. Canadian Journal of Fisheries and Aquatic Sciences:1--15.

\leavevmode\hypertarget{ref-Braun2011}{}%
Braun, D. C., and J. D. Reynolds. 2011. Relationships between habitat characteristics and breeding population densities in sockeye salmon (oncorhynchus nerka). Canadian Journal of Fisheries and Aquatic Sciences 68:758--767.

\leavevmode\hypertarget{ref-Breiman2001}{}%
Breiman, L. 2001. Random forests. Machine Learning 45:5--32.

\leavevmode\hypertarget{ref-Brett1952}{}%
Brett, J. R. 1952. Temperature tolerance in young pacific salmon, genus oncorhynchus. Journal of the Fisheries Board of Canada 9:265--323.

\leavevmode\hypertarget{ref-Bryant2009}{}%
Bryant, M., and R. Woodsmith. 2009. The response of salmon populations to geomorphic measurements at three scales. North American Journal of Fisheries Management 29:549--559.

\leavevmode\hypertarget{ref-Cade2003}{}%
Cade, B. S., and B. R. Noon. 2003. A gentle introduction to quantile regression for ecologists. Frontiers in Ecology and the Environment 1:412--420.

\leavevmode\hypertarget{ref-Carle1978}{}%
Carle, F. L., and M. R. Strub. 1978. A new method for estimating population size from removal data. Biometrics:621--630.

\leavevmode\hypertarget{ref-Carmichael2020}{}%
Carmichael, R. A., D. Tonina, E. R. Keeley, R. M. Benjankar, and K. E. See. 2020. Some like it slow: A bioenergetic evaluation of habitat quality for juvenile chinook salmon in the lemhi river, idaho. Canadian Journal of Fisheries and Aquatic Sciences.

\leavevmode\hypertarget{ref-Chabot2015}{}%
Chabot, D., and D. M. Bird. 2015. Wildlife research and management methods in the 21st century: Where do unmanned aircraft fit in? Journal of Unmanned Vehicle Systems 3:137--155.

\leavevmode\hypertarget{ref-CHaMP2016}{}%
CHaMP. 2016. Scientific protocol for salmonid habitat surveys within the columbia habitat monitoring program. Bonneville Power Administration Portland, Oregon, USA.

\leavevmode\hypertarget{ref-Chapman1951}{}%
Chapman, D. G. 1951. Some properties of the hypergeometric distribution with applications to zoological sample censuses. University of California Press.

\leavevmode\hypertarget{ref-Compton2012}{}%
Compton, T. J., M. A. Morrison, J. R. Leathwick, and G. D. Carbines. 2012. Ontogenetic habitat associations of a demersal fish species, pagrus auratus, identified using boosted regression trees. Marine Ecology Progress Series 462:219--230.

\leavevmode\hypertarget{ref-Cramer2009}{}%
Cramer, S. P., and N. K. Ackerman. 2009. Linking stream carrying capacity for salmonids to habitat features. Pages 225--254 \emph{in} American fisheries society symposium.

\leavevmode\hypertarget{ref-Dodrill2016}{}%
Dodrill, M. J., and C. B. Yackulic. 2016. Nonlinear relationships can lead to bias in biomass calculations and drift-foraging models when using summaries of invertebrate drift data. Environmental biology of fishes 99:659--670.

\leavevmode\hypertarget{ref-Dunham2002}{}%
Dunham, J. B., B. S. Cade, and J. W. Terrell. 2002. Influences of spatial and temporal variation on fish-habitat relationships defined by regression quantiles. Transactions of the American Fisheries Society 131:86--98.

\leavevmode\hypertarget{ref-Eastwood2003}{}%
Eastwood, P. D., G. J. Meaden, A. Carpentier, and S. I. Rogers. 2003. Estimating limits to the spatial extent and suitability of sole (solea solea) nursery grounds in the dover strait. Journal of Sea Research 50:151--165.

\leavevmode\hypertarget{ref-Fausch1988}{}%
Fausch, K. D., C. L. Hawkes, and M. G. Parsons. 1988. Models that predict standing crop of stream fish from habitat variables: 1950-85. Notes.

\leavevmode\hypertarget{ref-Froese2008}{}%
Froese, R. 2008. The continuous smooth hockey stick: A newly proposed spawner-recruitment model. Journal of Applied Ichthyology 24:703--704.

\leavevmode\hypertarget{ref-Hedger2013}{}%
Hedger, R. D., E. De Eyto, M. Dillane, O. H. Diserud, K. Hindar, P. McGinnity, R. Poole, and G. Rogan. 2013. Improving abundance estimates from electrofishing removal sampling. Fisheries Research 137:104--115.

\leavevmode\hypertarget{ref-Hillman1987}{}%
Hillman, T. W., J. Griffith, and W. Platts. 1987. Summer and winter habitat selection by juvenile chinook salmon in a highly sedimented idaho stream. Transactions of the American Fisheries Society 116:185--195.

\leavevmode\hypertarget{ref-Hinrichsen2020}{}%
Hinrichsen, R. A., and C. M. Paulsen. 2020. Low carrying capacity a risk for threatened Chinook salmon. Ecological Modelling 432:109223.

\leavevmode\hypertarget{ref-Holecek2009}{}%
Holecek, D. E., K. J. Cromwell, and B. P. Kennedy. 2009. Juvenile chinook salmon summer microhabitat availability, use, and selection in a central idaho wilderness stream. Transactions of the American Fisheries Society 138:633--644.

\leavevmode\hypertarget{ref-Jones2006}{}%
Jones IV, G. P., L. G. Pearlstine, and H. F. Percival. 2006. An assessment of small unmanned aerial vehicles for wildlife research. Wildlife society bulletin 34:750--758.

\leavevmode\hypertarget{ref-Knudby2010}{}%
Knudby, A., A. Brenning, and E. LeDrew. 2010. New approaches to modelling fish-habitat relationships. Ecological Modelling 221:503--511.

\leavevmode\hypertarget{ref-Koenker1978}{}%
Koenker, R., and G. Bassett Jr. 1978. Regression quantiles. Econometrica: journal of the Econometric Society:33--50.

\leavevmode\hypertarget{ref-Larsen2008}{}%
Larsen, D. P., A. R. Olsen, and D. L. Stevens. 2008. Using a master sample to integrate stream monitoring programs. Journal of Agricultural, Biological, and Environmental Statistics 13:243.

\leavevmode\hypertarget{ref-Larsen2016}{}%
Larsen, D. P., C. J. Volk, D. L. Stevens Jr, A. R. Olsen, and C. E. Jordan. 2016. An overview of the Columbia Habitat Monitoring Program's (CHaMP) spatial-temporal design framework. South Fork Research.

\leavevmode\hypertarget{ref-LeCun2015}{}%
LeCun, Y., Y. Bengio, and G. Hinton. 2015. Deep learning. Nature 521:436--444.

\leavevmode\hypertarget{ref-randomForest2002}{}%
Liaw, A., and M. Wiener. 2002. Classification and regression by randomForest. R News 2:18--22.

\leavevmode\hypertarget{ref-R-survey}{}%
Lumley, T. 2020. Survey: Analysis of complex survey samples.

\leavevmode\hypertarget{ref-McKean2008}{}%
McKean, J. A., D. J. Isaak, and C. W. Wright. 2008. Geomorphic controls on salmon nesting patterns described by a new, narrow-beam terrestrial--aquatic lidar. Frontiers in Ecology and the Environment 6:125--130.

\leavevmode\hypertarget{ref-R-quantregForest}{}%
Meinshausen, N. 2017. QuantregForest: Quantile regression forests.

\leavevmode\hypertarget{ref-Meinshausen2006}{}%
Meinshausen, N. 2006. Quantile regression forests. Journal of Machine Learning Research 7:983--999.

\leavevmode\hypertarget{ref-Mossop2006}{}%
Mossop, B., and M. J. Bradford. 2006. Using thalweg profiling to assess and monitor juvenile salmon (oncorhynchus spp.) habitat in small streams. Canadian Journal of Fisheries and Aquatic Sciences 63:1515--1525.

\leavevmode\hypertarget{ref-Moussalli1986}{}%
Moussalli, E., and R. Hilborn. 1986. Optimal stock size and harvest rate in multistage life history models. Canadian Journal of Fisheries and Aquatic Sciences 43:135--141.

\leavevmode\hypertarget{ref-Myers1999}{}%
Myers, R. A., K. G. Bowen, and N. J. Barrowman. 1999. Maximum reproductive rate of fish at low population sizes. Canadian Journal of Fisheries and Aquatic Sciences 56:2404--2419.

\leavevmode\hypertarget{ref-Nahorniak2015}{}%
Nahorniak, M., D. P. Larsen, C. Volk, and C. E. Jordan. 2015. Using inverse probability bootstrap sampling to eliminate sample induced bias in model based analysis of unequal probability samples. Plos one 10:e0131765.

\leavevmode\hypertarget{ref-Nickelson1992}{}%
Nickelson, T. E., J. D. Rodgers, S. L. Johnson, and M. F. Solazzi. 1992. Seasonal changes in habitat use by juvenile coho salmon (oncorhynchus kisutch) in oregon coastal streams. Canadian Journal of Fisheries and Aquatic Sciences 49:783--789.

\leavevmode\hypertarget{ref-NOAA2016}{}%
NOAA Fisheries. 2016a. 2016 5-year review: Summary \& evaluation of Upper Columbia River steelhead and Upper Columbia River spring-run Chinook salmon. NOAA.

\leavevmode\hypertarget{ref-NOAA2016b}{}%
NOAA Fisheries. 2016b. 2016 5-year review: Summary \& evaluation of Snake River Sockeye, Snake River spring-summer Chinook, Snake River fall-run Chinook and Snake River steelhead. NOAA.

\leavevmode\hypertarget{ref-NOAA2020}{}%
NOAA Fisheries. 2020. Endangered Species Act Section 7(a)(2) Biological opinion and Magnuson-Stevens fishery conservation and management act essential fish habitat response for the continued operation and maintenance of the Columbia River system. NMFS Consultation, NOAA National Marine Fisheries Service.

\leavevmode\hypertarget{ref-R-FSA}{}%
Ogle, D., P. Wheeler, and A. Dinno. 2020. FSA: Simple fisheries stock assessment methods.

\leavevmode\hypertarget{ref-Pess2002}{}%
Pess, G. R., D. R. Montgomery, E. A. Steel, R. E. Bilby, B. E. Feist, and H. M. Greenberg. 2002. Landscape characteristics, land use, and coho salmon (oncorhynchus kisutch) abundance, snohomish river, wash., usa. Canadian Journal of Fisheries and Aquatic Sciences 59:613--623.

\leavevmode\hypertarget{ref-Pittman2009}{}%
Pittman, S. J., B. M. Costa, and T. A. Battista. 2009. Using lidar bathymetry and boosted regression trees to predict the diversity and abundance of fish and corals. Journal of Coastal Research 53:27--38.

\leavevmode\hypertarget{ref-Prasad2006}{}%
Prasad, A., L. Iverson, and A. Liaw. 2006. Newer classification and regression tree techniques: Bagging and random forests for ecological prediction. Ecosystems 9:181--199.

\leavevmode\hypertarget{ref-Raleigh1986}{}%
Raleigh, R. F., W. J. Miller, and P. C. Nelson. 1986. Habitat suitability index models and instream flow suitability curves: Chinook salmon. National Ecology Center.

\leavevmode\hypertarget{ref-Rsoftware2019}{}%
R Core Team. 2019. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\hypertarget{ref-R-Rcapture}{}%
Rivest, L.-P., and S. Baillargeon. 2019. Rcapture: Loglinear models for capture-recapture experiments.

\leavevmode\hypertarget{ref-Robson1964}{}%
Robson, D., and H. Regier. 1964. Sample size in petersen mark--recapture experiments. Transactions of the American Fisheries Society 93:215--226.

\leavevmode\hypertarget{ref-Seber2002}{}%
Seber, G. A. F. 2002. The estimation of animal abundance and related parameters. Blackburn Press Caldwell, New Jersey.

\leavevmode\hypertarget{ref-Sharma2001}{}%
Sharma, R., and R. Hilborn. 2001. Empirical relationships between watershed characteristics and coho salmon (oncorhynchus kisutch) smolt abundance in 14 western washington streams. Canadian Journal of Fisheries and Aquatic Sciences 58:1453--1463.

\leavevmode\hypertarget{ref-R-missForest}{}%
Stekhoven, D. J. 2013. MissForest: Nonparametric missing value imputation using random forest.

\leavevmode\hypertarget{ref-Stekhoven2012}{}%
Stekhoven, D. J., and P. Bühlmann. 2012. MissForest---non-parametric missing value imputation for mixed-type data. Bioinformatics 28:112--118.

\leavevmode\hypertarget{ref-Stevens1999}{}%
Stevens Jr, D. L., and A. R. Olsen. 1999. Spatially restricted surveys over time for aquatic resources. Journal of Agricultural, Biological, and Environmental Statistics:415--428.

\leavevmode\hypertarget{ref-Stevens2004}{}%
Stevens Jr, D. L., and A. R. Olsen. 2004. Spatially balanced sampling of natural resources. Journal of the American Statistical Association 99:262--278.

\leavevmode\hypertarget{ref-Sweka2010}{}%
Sweka, J. A., and G. Mackey. 2010. A functional relationship between watershed size and atlantic salmon parr density. Journal of Fish and Wildlife Management 1:3--10.

\leavevmode\hypertarget{ref-Terrell1996}{}%
Terrell, J. W., B. S. Cade, J. Carpenter, and J. M. Thompson. 1996. Modeling stream fish habitat limitations from wedge-shaped patterns of variation in standing stock. Transactions of the American Fisheries Society 125:104--117.

\leavevmode\hypertarget{ref-Tonina2019}{}%
Tonina, D., J. A. McKean, R. M. Benjankar, C. W. Wright, J. R. Goode, Q. Chen, W. J. Reeder, R. A. Carmichael, and M. R. Edmondson. 2019. Mapping river bathymetries: Evaluating topobathymetric lidar survey. Earth Surface Processes and Landforms 44:507--520.

\leavevmode\hypertarget{ref-Volk2017}{}%
Volk, C., N. Bouwes, C. Jordan, J. Wheaton, P. Nelle, C. Beasley, S. Walker, B. Bouwes, M. Nahorniak, A. Hill, J. Heitke, K. Whitehead, S. Bangen, and E. Ward. 2017. Integrated status and effectiveness monitoring program (ISEMP) and columbia habitat monitoring program (CHaMP) 2016 annual combined technical report. Bonneville Power Administration.

\leavevmode\hypertarget{ref-Wall2016}{}%
Wall, C. E., N. Bouwes, J. M. Wheaton, W. C. Saunders, and S. N. Bennett. 2016. Net rate of energy intake predicts reach-level steelhead (oncorhynchus mykiss) densities in diverse basins from a large monitoring program. Canadian Journal of Fisheries and Aquatic Sciences 73:1081--1091.

\leavevmode\hypertarget{ref-Weinstein2018}{}%
Weinstein, B. G. 2018. A computer vision for animal ecology. Journal of Animal Ecology 87:533--545.

\leavevmode\hypertarget{ref-Whitehead2014}{}%
Whitehead, K., and C. H. Hugenholtz. 2014. Remote sensing of the environment with small unmanned aircraft systems (uass), part 1: A review of progress and challenges. Journal of Unmanned Vehicle Systems 2:69--85.

\leavevmode\hypertarget{ref-Whittier2011}{}%
Whittier, T. R., A. T. Herlihy, C. E. Jordan, and C. Volk. 2011. Landscape classification of pacific northwest hydrologic units based on natural features and human disturbance to support salmonid research. NOAA, National Marine Fisheries Service.

\newpage

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\hypertarget{tables}{}

\begin{table}[!h]

\caption{\label{tab:fish-hab-sites}The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer-run Chinook salmon parr capacity QRF model}
\centering
\begin{tabular}[t]{lrl}
\toprule
Watershed & n Sites & Percent\\
\midrule
Entiat & 61 & 18.7\%\\
John Day & 75 & 22.9\%\\
Lemhi & 33 & 10.1\%\\
Minam & 20 & 6.1\%\\
South Fork Salmon & 30 & 9.2\%\\
Upper Grande Ronde & 86 & 26.3\%\\
Wenatchee & 22 & 6.7\%\\
Total & 327 & 100.0\%\\
\bottomrule
\end{tabular}
\end{table}

\newpage

\begin{table}[!h]

\caption{\label{tab:select-mets-table}Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer-run Chinook salmon parr capacity. Metrics are ranked in order of relative importance.}
\centering
\begin{tabular}[t]{rll>{\raggedright\arraybackslash}p{3in}}
\toprule
Rank & Metric & Metric Category & Description\\
\midrule
1 & Wetted Width & Size & Average width of the wetted polygon for a site.\\
2 & Discharge & Size & The sum of station discharge across all  stations.  Station discharge is calculated as depth x velocity x station increment for all stations except first and last.  Station discharge for first and last station is 0.5 x station width x depth x velocity.\\
3 & Avg. August Temp. & Temperature & Average predicted daily August temperature from NorWest, averaged across the years 2002-2011.\\
4 & Width:Depth Ratio & Complexity & Average width to depth ratio of the wetted channel measured from cross-sections. Depths represent an average of depths along each cross-section.\\
5 & Fines & Substrate & Average percentage of pool tail substrates comprised of sediment <6 mm.\\
6 & Fish Cover & Cover & Percent of wetted area with the following types of cover: aquatic vegetation, artificial, woody debris, and terrestrial vegetation.\\
7 & Channel Unit Frequency & ChannelUnit & Number of channel units per 100 meters.\\
8 & Depth Complexity & Complexity & Standard Deviation of water depths within the wetted channel.\\
9 & Large Wood Freq. in Pools & Wood & Total volume of  large wood pieces  within the wetted channel and Slow Water/Pool channel units, scaled by site length.\\
10 & Riparian Canopy & Riparian & Percent of riparian canopy with some vegetation.\\
11 & Substrate: D16 & Substrate & Diameter of the 16th percentile particle derived from pebble counts.\\
12 & Braidedness & Complexity & Ratio of the total length of the wetted mainstem channel plus side channels and the length of the mainstem channel.\\
\bottomrule
\end{tabular}
\end{table}

\newpage

\begin{table}[!h]

\caption{\label{tab:gaa-covars}Globally available attribute (GAA) habitat covariates used to extrapolate quantile random forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.}
\centering
\begin{tabular}[t]{lllrr}
\toprule
Covariate & Units & Spatial Scale & Estimate & Std. Error\\
\midrule
Temeprature Range & C & Reach-2km & -0.044 & 0.081\\
Elevation & m & Site-300m & -0.243 & 0.155\\
CHaMP Watershed & - & Region & - & -\\
Natural Class PCA 1 & - & Watershed-HUC12 & -0.092 & 0.070\\
Disturbance Class PCA 1 & - & Watershed-HUC12 & -0.064 & 0.064\\
Drainage Area (sqrt) & km$^2$ (square root) & Reach-2km & -0.141 & 0.077\\
Stream Power & - & Reach-2km & 0.049 & 0.033\\
Slope & m/m & Reach-2km & -0.513 & 0.100\\
Channel Type & - & Site-300m & - & -\\
Bankfull Width - modeled & m & Site-300m & 0.216 & 0.099\\
NorWeST Aug. Temperature & C & Reach-2km & -0.149 & 0.119\\
\bottomrule
\end{tabular}
\end{table}

\newpage

\begin{table}[!h]

\caption{\label{tab:extrap-r2-table}Summary of extrapolation model fits, split by whether the extrapolation model used CHaMP watershed as a covariate or not.}
\centering
\begin{tabular}[t]{llrr}
\toprule
Model & Response & $r^2$ & Adjusted $r^2$\\
\midrule
CHaMP & fish/m & 0.481 & 0.454\\
non-CHaMP & fish/m & 0.360 & 0.339\\
\bottomrule
\end{tabular}
\end{table}

\newpage

\begin{table}[!h]

\caption{\label{tab:sr-table}Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile random forest (QRF) model. Numbers in parentheses are coefficients of variation.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1in}rll>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}}
\toprule
Population & n Yrs & Adult Data & Parr Data & Beverton Holt & Ricker & Hockey Stick & QRF\\
\midrule
Catherine Creek & 20 & Spawners & RST & 135,387 (0.269) & 103,021 (0.141) & 99,921 (0.21) & 190,857 (0.162)\\
Chiwawa River & 20 & Spawners & Parr Surveys & 248,586 (0.24) & 166,139 (0.148) & 174,216 (0.184) & 216,451 (0.363)\\
Hayden Creek & 7 & Spawners & RST & 58,394 (0.244) & 65,958 (0.195) & 48,351 (0.174) & 121,676 (0.202)\\
Lostine River & 17 & Redds & RST & 196,259 (0.24) & 146,982 (0.159) & 144,415 (0.201) & 152,493 (0.316)\\
Minam River & 14 & Spawners & RST & 1,309,223 (2.18) & 484,810 (1.444) & 662,802 (1.726) & 365,338 (0.261)\\
South Fork Salmon River & 17 & Redds & RST & 87,260 (0.407) & 62,456 (0.265) & 64,654 (0.317) & 221,362 (0.142)\\
Tucannon River & 27 & Redds & RST & 4,791,131 (13.016) & 1,234,653 (8.566) & 1,922,692 (10.082) & 529,223 (0.196)\\
Upper Grande Ronde River & 8 & Spawners & RST & 171,607 (0.388) & 168,137 (0.298) & 127,052 (0.317) & 200,228 (0.23)\\
Upper Lemhi & 7 & Spawners & RST & 333,229 (0.322) & 229,635 (0.212) & 242,637 (0.252) & 269,626 (0.217)\\
\bottomrule
\end{tabular}
\end{table}

\newpage

\listoffigures

\newpage

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\hypertarget{figures}{}

\begin{figure}
\includegraphics[width=0.9\linewidth]{../figures/ChampWtds} \caption{Watersheds with CHaMP habitat data. Watersheds in black also contain paired fish data. Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork.}\label{fig:wtsd-map}
\end{figure}

\newpage

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/QRF_flowchart} \caption{Conceptual diagram showing input data sources, modeling decisions, model outputs etc. for the QRF and extrapolation models.}\label{fig:model-diag}
\end{figure}

\newpage

\begin{figure}
\centering
\includegraphics{figures/rel-imp-figure-1.pdf}
\caption{\label{fig:rel-imp-figure}Relative importance of each habitat covariate included in the quantile random forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr}
\end{figure}

\newpage

\begin{figure}
\centering
\includegraphics{figures/pdp-figure-1.pdf}
\caption{\label{fig:pdp-figure}Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile random forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values. Tick marks along the X-axis depict observed values.}
\end{figure}

\newpage

\begin{figure}
\centering
\includegraphics{figures/sr-figure-1.pdf}
\caption{\label{fig:sr-figure}Spawner-recruit data from nine watersheds. Solid lines are the spawner-recruit curve, dashed lines are the estimated capacity, and shaded polygons depict the 95\% confidence intervals of capacity. Red corresponds to Beverton-Holt models, purple to Ricker models, blue to hockey stick models, and green to QRF estimates. The QRF solid curve is a Beverton-Holt model with the capacity parameter fixed to the QRF estimate of capacity. A few curves with high capacity estimates were not plotted to improve readability.}
\end{figure}

\end{document}

### 2020-12-01 README.Rmd
updated README to show that the article is published in Ecosphere
updated the README to have the correct title

@@
 # Please put your title here to include it in the file below.
-Title <- "Estimating Life-Stage Specific Habitat Capacity for Anadromous Salmonids using Quantile Random Forest Models"
+Title <- "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
 Authors <- "See, Kevin E., Michael W. Ackerman, Richard A. Carmichael, Sarah Hoffman, and Chris Beasley"
 yr <- 2020
-Journal <- "Ecological Modelling"
+Journal <- "Ecosphere"
 ```

### 2020-12-01 R/estimate_MIC.R
copied the most up-to-date estimate_MIC function from QRFcapacity repo

@@ -26,7 +26,11 @@ estimate_MIC = function(data,
              y = data %>%
                pull(response),
              use = 'pairwise.complete.obs') %>%
-    map_df(.f = identity) %>%
+    map_df(.f = function(x) {
+      x %>%
+        as_tibble() %>%
+        pull(1)
+    }) %>%
     mutate(Metric = data %>%
              dplyr::select(one_of(covars)) %>%
              names()) %>%

### 2020-12-01 analysis/paper/appendix/appendix_s1.Rmd
ensured that the Figures section starts on a new page, and the first figure is on that page as well.

@@
 output: 
-    bookdown::word_document2:
-      fig_caption: yes
-      toc: no
-      always_allow_html: true
-      number_sections: FALSE
-      fig_height: 7
-      fig_width: 6
-      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../../templates/scholarly-metadata.lua
-      - --lua-filter=../../templates/author-info-blocks.lua
-      - --lua-filter=../../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../../templates/header_appendixS.tex
+    bookdown::word_document2:
+      fig_caption: yes
+      toc: no
+      always_allow_html: true
+      number_sections: FALSE
+      fig_height: 7
+      fig_width: 6
+      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
+      pandoc_args:
+      - --lua-filter=../../templates/scholarly-metadata.lua
+      - --lua-filter=../../templates/author-info-blocks.lua
+      - --lua-filter=../../templates/pagebreak.lua
     bookdown::html_document2:
@@
   fig.path = "../../figures/appendix/",
-  dpi = 600,
+  dpi = 400,
   appendix = TRUE
@@
 
+\newpage
 # Figures
 
-```{r mine-facet-fig, fig.cap = 'Barplots of MIC statistics, faceted by habitat category.'}
+```{r mine-facet-fig, fig.height = 6.7, fig.cap = 'Barplots of MIC statistics, faceted by habitat category.'}
 mine_res %>%

### 2020-12-01 analysis/paper/appendix/appendix_s1.Rmd
updated the appendix to keep a copy of the .tex file. Replaced the formatting of author names in that .tex file (to avoid overrunning a line), and re-compiled the pdf

@@
 ---
-title: "Appendix S1"
-subtitle: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
+title: Appendix S1 - _Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models_
+subtitle: "Ecospheres"
 author:

### 2020-12-01 analysis/paper/appendix/appendix_s1.tex
updated the appendix to keep a copy of the .tex file. Replaced the formatting of author names in that .tex file (to avoid overrunning a line), and re-compiled the pdf

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Appendix S1 - Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models},
  pdfauthor={Kevin E. See1,, Michael W. Ackerman1, Richard A. Carmichael1, Sarah L. Hoffmann1, and Chris Beasley1},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% set figure and table numbering
\usepackage{longtable}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

% double spacing
\usepackage{setspace}
\doublespacing
% to double space figure captions
\usepackage{caption}
\captionsetup{font = doublespacing}

% % line numbers
% \usepackage{lineno}
% \linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Appendix S1 - \emph{Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Ecospheres}
\author{Kevin E. See\textsuperscript{1,*}, 
    \and Michael W. Ackerman\textsuperscript{1}, 
    \and Richard A. Carmichael\textsuperscript{1},
    \and Sarah L. Hoffmann\textsuperscript{1}, 
    \and Chris Beasley\textsuperscript{1}}
\date{December 01, 2020}

\begin{document}
\maketitle

\textsuperscript{1} Biomark, Inc.~705 South 8th St., Boise, Idaho, 83702, USA

\textsuperscript{*} Correspondence: \href{mailto:Kevin.See@merck.com}{Kevin E. See \textless{}\href{mailto:Kevin.See@merck.com}{\nolinkurl{Kevin.See@merck.com}}\textgreater{}}

\hypertarget{choosing-habitat-covariates}{%
\section{Choosing Habitat Covariates}\label{choosing-habitat-covariates}}

One of the crucial steps in building this carrying capacity model was choosing which habitat covariates to include. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points). Our goal was to select a group of covariates that captured as many different aspects of the stream habitat (e.g., substrate, flow, riparian condition, channel unit configuration, etc.) as possible, while still holding information about fish densities.

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. Habitat metrics were first grouped into broad categories that included channel unit configuration, complexity, fish cover, riparian areas, side channels, stream size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units). To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics (Reshef et al. 2011) to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities. We calculated the maximal information coefficient (MIC), using the R package \emph{minerva} (Filosi et al. 2019), to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric (Reshef et al. 2011). MIC is a measure of correlation that incorporates potential non-linear associations; for example, if there is a quadratic association the MIC value could be high, even when the standard correlation coefficient is low. We excluded categorical variables such as channel type (e.g., meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.

Within each category, metrics were ranked according to their MIC value (Table \ref{tab:mic-tab} and Figures \ref{fig:mine-facet-fig} and \ref{fig:mine-all-fig}). The MIC value of each measured habitat characteristic and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Table \ref{tab:corr-tab}), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by rehabilitation actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many zero values, in the data set, as well as metrics that may have too much observer error (Rosgen et al. 2018).

\hypertarget{results}{%
\section{Results}\label{results}}

We chose 12 metrics, highlighted in bold in Table \ref{tab:mic-tab}. Their correlations with each other are displayed in Figure \ref{fig:selected-corr}. Those with a high correlation coefficient (\(\geq\) 0.5 or \(\leq\) -0.5) are plotted against each other in Figure \ref{fig:high-cor-mets} to show the variety in values even for pairs of metrics with relatively high correlations.

\newpage

\hypertarget{literature-cited}{%
\section{Literature Cited}\label{literature-cited}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-R-minerva}{}%
Filosi, M., R. Visintainer, and D. Albanese. 2019. Minerva: Maximal information-based nonparametric exploration for variable analysis.

\leavevmode\hypertarget{ref-Reshef2011}{}%
Reshef, D. N., Y. A. Reshef, H. K. Finucane, S. R. Grossman, G. McVean, P. J. Turnbaugh, E. S. Lander, M. Mitzenmacher, and P. C. Sabeti. 2011. Detecting novel associations in large data sets. Science 334:1518--1524.

\leavevmode\hypertarget{ref-Rosgen2018}{}%
Rosgen, D., A. Taillacq, B. Rosgen, and D. Geenen. 2018. A technical review of the Columbia Habitat Monitoring Program's protocol, data quality.

\newpage

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\begin{longtable}[t]{l>{\raggedright\arraybackslash}p{2in}>{\raggedright\arraybackslash}p{1in}>{\raggedleft\arraybackslash}p{0.5in}>{\raggedleft\arraybackslash}p{0.5in}>{\raggedleft\arraybackslash}p{0.5in}}
\caption{\label{tab:mic-tab}MIC statistic for top metrics within each habitat category, sorted by category and MIC value. The percent of records for which each habitat metric measurement was missing or zero is also shown. Metrics selected for the QRF model are in bold.}\\
\toprule
Category & Name & Abbrv & MIC & Percent Missing & Percent 0-value\\
\midrule
\endfirsthead
\caption[]{\label{tab:mic-tab}MIC statistic for top metrics within each habitat category, sorted by category and MIC value. The percent of records for which each habitat metric measurement was missing or zero is also shown. Metrics selected for the QRF model are in bold. \textit{(continued)}}\\
\toprule
Category & Name & Abbrv & MIC & Percent Missing & Percent 0-value\\
\midrule
\endhead
\
\endfoot
\bottomrule
\endlastfoot
\textbf{ChannelUnit} & \textbf{Channel Unit Frequency} & \textbf{CU\_Freq} & \textbf{0.241} & \textbf{0.021} & \textbf{0.021}\\
ChannelUnit & Fast Turbulent Frequency & FstTurb\_Freq & 0.230 & 0.021 & 0.082\\
ChannelUnit & Fast NonTurbulent Frequency & FstNT\_Freq & 0.209 & 0.021 & 0.308\\
ChannelUnit & Slow Water Frequency & SlowWater\_Freq & 0.208 & 0.021 & 0.073\\
ChannelUnit & Fast Turbulent Percent & FstTurb\_Pct & 0.195 & 0.021 & 0.082\\
\addlinespace
ChannelUnit & Channel Unit Count & CU\_Ct & 0.189 & 0.021 & 0.021\\
ChannelUnit & Fast Turbulent Count & FstTurb\_Ct & 0.178 & 0.021 & 0.082\\
ChannelUnit & Slow Water Percent & SlowWater\_Pct & 0.177 & 0.021 & 0.073\\
ChannelUnit & Fast NonTurbulent Percent & FstNT\_Pct & 0.169 & 0.021 & 0.308\\
ChannelUnit & Fast NonTurbulent Count & FstNT\_Ct & 0.166 & 0.021 & 0.308\\
\addlinespace
\textbf{Complexity} & \textbf{Wetted Width To Depth Ratio Avg} & \textbf{WetWDRat\_Avg} & \textbf{0.247} & \textbf{0.003} & \textbf{0.003}\\
Complexity & Bankfull Width To Depth Ratio Avg & BfWDRat\_Avg & 0.245 & 0.003 & 0.003\\
\textbf{Complexity} & \textbf{Wetted Depth SD} & \textbf{DpthWet\_SD} & \textbf{0.232} & \textbf{0.003} & \textbf{0.003}\\
\textbf{Complexity} & \textbf{Wetted Channel Braidedness} & \textbf{WetBraid} & \textbf{0.212} & \textbf{0.003} & \textbf{0.003}\\
Complexity & Bankfull Channel Braidedness & BfBraid & 0.211 & 0.003 & 0.003\\
\addlinespace
Complexity & Wetted Channel Qualifying Island Count & Wet\_QIsland\_Ct & 0.209 & 0.003 & 0.835\\
Complexity & Bankfull Width CV & BfWdth\_CV & 0.209 & 0.003 & 0.003\\
Complexity & Bankfull Width To Depth Ratio CV & BfWDRat\_CV & 0.202 & 0.003 & 0.003\\
Complexity & Detrended Elevation SD & DetrendElev\_SD & 0.196 & 0.003 & 0.003\\
Complexity & Bankfull Channel Qualifying Island Count & Bf\_QIsland\_Ct & 0.193 & 0.003 & 0.780\\
\addlinespace
\textbf{Cover} & \textbf{Fish Cover: Total} & \textbf{FishCovTotal} & \textbf{0.225} & \textbf{0.021} & \textbf{0.030}\\
Cover & Fish Cover: None & FishCovNone & 0.224 & 0.021 & 0.021\\
Cover & Fish Cover: LW & FishCovLW & 0.213 & 0.021 & 0.155\\
Cover & Fish Cover: Terrestrial Vegetation & FishCovTVeg & 0.204 & 0.021 & 0.052\\
Cover & Percent Undercut by Length & UcutLgth\_Pct & 0.185 & 0.000 & 0.476\\
\addlinespace
Cover & Percent Undercut by Area & UcutArea\_Pct & 0.184 & 0.000 & 0.476\\
Cover & Fish Cover: Aquatic Vegetation & FishCovAqVeg & 0.166 & 0.296 & 0.631\\
Cover & Fish Cover: Artificial & FishCovArt & 0.136 & 0.021 & 0.851\\
Riparian & Riparian Cover: Understory & RipCovUstory & 0.206 & 0.000 & 0.000\\
Riparian & RipCovUstoryNone & RipCovUstoryNone & 0.206 & 0.000 & 0.000\\
\addlinespace
Riparian & Riparian Cover: No Canopy & RipCovCanNone & 0.194 & 0.000 & 0.000\\
\textbf{Riparian} & \textbf{Riparian Cover: Some Canopy} & \textbf{RipCovCanSome} & \textbf{0.194} & \textbf{0.000} & \textbf{0.095}\\
Riparian & Riparian Cover: Big Tree & RipCovBigTree & 0.184 & 0.000 & 0.183\\
Riparian & Riparian Cover: Ground & RipCovGrnd & 0.182 & 0.000 & 0.000\\
Riparian & RipCovGrndNone & RipCovGrndNone & 0.170 & 0.000 & 0.003\\
\addlinespace
Riparian & Riparian Cover: Woody & RipCovWood & 0.168 & 0.000 & 0.000\\
Riparian & Riparian Cover: Non-Woody & RipCovNonWood & 0.166 & 0.000 & 0.000\\
Riparian & Riparian Cover: Coniferous & RipCovConif & 0.164 & 0.009 & 0.192\\
SideChannel & Bankfull Side Channel Width & BfSCWdth & 0.223 & 0.796 & 0.796\\
SideChannel & Wetted Side Channel Width & WetSCWdth & 0.213 & 0.832 & 0.832\\
\addlinespace
SideChannel & Wetted Side Channel Percent By Area & WetSC\_Pct & 0.209 & 0.021 & 0.820\\
SideChannel & SCSm\_Freq & SCSm\_Freq & 0.153 & 0.021 & 0.921\\
SideChannel & SCSm\_Ct & SCSm\_Ct & 0.153 & 0.021 & 0.921\\
SideChannel & SC\_Area\_Pct & SC\_Area\_Pct & 0.153 & 0.021 & 0.921\\
Size & Mean Annual Flow & MeanU & 0.346 & 0.476 & 0.476\\
\addlinespace
\textbf{Size} & \textbf{Wetted Width Integrated} & \textbf{WetWdth\_Int} & \textbf{0.332} & \textbf{0.003} & \textbf{0.003}\\
Size & Bankfull Width Integrated & BfWdthInt & 0.324 & 0.003 & 0.003\\
Size & Wetted Width Avg & WetWdth\_Avg & 0.324 & 0.003 & 0.003\\
Size & Drainage Area (Flowline) & CUMDRAINAG & 0.302 & 0.341 & 0.341\\
Size & Bankfull Width Avg & BfWdth\_Avg & 0.298 & 0.003 & 0.003\\
\addlinespace
Size & DpthThlwg\_Avg & DpthThlwg\_Avg & 0.280 & 0.003 & 0.003\\
\textbf{Size} & \textbf{Discharge} & \textbf{Q} & \textbf{0.259} & \textbf{0.037} & \textbf{0.037}\\
Size & Bankfull Depth Avg & DpthBf\_Avg & 0.245 & 0.018 & 0.018\\
Size & Bankfull Depth Max & DpthBf\_Max & 0.240 & 0.018 & 0.018\\
\textbf{Substrate} & \textbf{Substrate < 6mm} & \textbf{SubLT6} & \textbf{0.237} & \textbf{0.049} & \textbf{0.055}\\
\addlinespace
Substrate & Substrate < 2mm & SubLT2 & 0.227 & 0.049 & 0.082\\
\textbf{Substrate} & \textbf{Substrate: D16} & \textbf{SubD16} & \textbf{0.219} & \textbf{0.012} & \textbf{0.012}\\
Substrate & Substrate: Embeddedness Avg & SubEmbed\_Avg & 0.204 & 0.293 & 0.317\\
Substrate & Substrate: D50 & SubD50 & 0.197 & 0.012 & 0.012\\
Substrate & Substrate Est: Sand and Fines & SubEstSandFines & 0.190 & 0.021 & 0.030\\
\addlinespace
Substrate & Substrate Est: Cobbles & SubEstCbl & 0.185 & 0.021 & 0.027\\
Substrate & Substrate: D84 & SubD84 & 0.185 & 0.012 & 0.012\\
Substrate & Substrate Est: Boulders & SubEstBldr & 0.183 & 0.021 & 0.149\\
Substrate & Substrate: Embeddedness SD & SubEmbed\_SD & 0.181 & 0.302 & 0.320\\
\textbf{Temperature} & \textbf{Avg. August Temperature} & \textbf{avg\_aug\_temp} & \textbf{0.272} & \textbf{0.000} & \textbf{0.000}\\
\addlinespace
Temperature & Elev\_M & Elev\_M & 0.262 & 0.363 & 0.363\\
Temperature & August Temperature & aug\_temp & 0.188 & 0.155 & 0.155\\
Temperature & Solar Access: Summer Avg & SolarSummr\_Avg & 0.186 & 0.070 & 0.070\\
WaterQuality & Conductivity & Cond & 0.254 & 0.024 & 0.027\\
WaterQuality & Alkalinity & Alk & 0.225 & 0.009 & 0.027\\
\addlinespace
WaterQuality & Drift Biomass & DriftBioMass & 0.000 & 0.277 & 0.384\\
Wood & Large Wood Volume: Bankfull Slow Water & LWVol\_BfSlow & 0.213 & 0.003 & 0.232\\
\textbf{Wood} & \textbf{Large Wood Volume: Wetted Slow Water} & \textbf{LWVol\_WetSlow} & \textbf{0.207} & \textbf{0.003} & \textbf{0.290}\\
Wood & Large Wood Frequency: Wetted & LWFreq\_Wet & 0.199 & 0.003 & 0.125\\
Wood & Large Wood Volume: Bankfull & LWVol\_Bf & 0.189 & 0.003 & 0.085\\
\addlinespace
Wood & Large Wood Volume: Wetted Fast Turbulent & LWVol\_WetFstTurb & 0.187 & 0.003 & 0.274\\
Wood & Large Wood Frequency: Bankfull & LWFreq\_Bf & 0.178 & 0.003 & 0.085\\
Wood & Large Wood Volume: Bankfull Fast NonTurbulent & LWVol\_BfFstNT & 0.175 & 0.003 & 0.521\\
Wood & Large Wood Volume: Wetted & LWVol\_Wet & 0.166 & 0.003 & 0.125\\
Wood & Large Wood Volume: Wetted Fast NonTurbulent & LWVol\_WetFstNT & 0.159 & 0.003 & 0.595\\*
\end{longtable}

\begin{longtable}[t]{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{2in}>{\raggedright\arraybackslash}p{2in}r}
\caption{\label{tab:corr-tab}Pearson correlation coefficient between each variable within a habitat category.}\\
\toprule
Category & Metric 1 & Metric 2 & r\\
\midrule
\endfirsthead
\caption[]{\label{tab:corr-tab}Pearson correlation coefficient between each variable within a habitat category. \textit{(continued)}}\\
\toprule
Category & Metric 1 & Metric 2 & r\\
\midrule
\endhead
\
\endfoot
\bottomrule
\endlastfoot
Channel Unit & Slow Water Frequency & Channel Unit Frequency & 0.88\\
Channel Unit & Slow Water Count & Channel Unit Count & 0.87\\
Channel Unit & Fast Turbulent Frequency & Channel Unit Frequency & 0.84\\
Channel Unit & Fast NonTurbulent Count & Fast NonTurbulent Frequency & 0.84\\
Channel Unit & Fast Turbulent Count & Fast Turbulent Frequency & 0.80\\
\addlinespace
Channel Unit & Slow Water Count & Slow Water Frequency & 0.77\\
Channel Unit & Slow Water Percent & Fast Turbulent Percent & -0.74\\
Channel Unit & Channel Unit Count & Channel Unit Frequency & 0.73\\
Channel Unit & Fast Turbulent Count & Channel Unit Count & 0.73\\
Channel Unit & Slow Water Frequency & Channel Unit Count & 0.69\\
\addlinespace
Channel Unit & Fast NonTurbulent Count & Fast NonTurbulent Percent & 0.65\\
Channel Unit & Fast NonTurbulent Frequency & Channel Unit Frequency & 0.63\\
Channel Unit & Fast Turbulent Count & Channel Unit Frequency & 0.60\\
Channel Unit & Slow Water Count & Channel Unit Frequency & 0.59\\
Channel Unit & Slow Water Frequency & Fast Turbulent Frequency & 0.59\\
\addlinespace
Channel Unit & Fast Turbulent Frequency & Channel Unit Count & 0.56\\
Channel Unit & Fast NonTurbulent Count & Channel Unit Count & 0.56\\
Channel Unit & Fast Turbulent Percent & Fast NonTurbulent Percent & -0.55\\
Channel Unit & Slow Water Count & Slow Water Percent & 0.54\\
Channel Unit & Fast NonTurbulent Percent & Fast NonTurbulent Frequency & 0.52\\
\addlinespace
Channel Unit & Fast NonTurbulent Frequency & Channel Unit Count & 0.46\\
Channel Unit & Slow Water Percent & Slow Water Frequency & 0.46\\
Channel Unit & Fast Turbulent Percent & Fast NonTurbulent Count & -0.45\\
Channel Unit & Slow Water Count & Fast Turbulent Count & 0.41\\
Channel Unit & Slow Water Count & Fast Turbulent Percent & -0.40\\
\addlinespace
Channel Unit & Fast NonTurbulent Count & Channel Unit Frequency & 0.39\\
Channel Unit & Fast Turbulent Count & Slow Water Frequency & 0.39\\
Channel Unit & Slow Water Frequency & Fast NonTurbulent Frequency & 0.38\\
Channel Unit & Fast Turbulent Percent & Fast NonTurbulent Frequency & -0.37\\
Channel Unit & Fast Turbulent Frequency & Fast NonTurbulent Frequency & 0.36\\
\addlinespace
Channel Unit & Fast Turbulent Percent & Slow Water Frequency & -0.31\\
Channel Unit & Slow Water Count & Fast Turbulent Frequency & 0.30\\
Channel Unit & Slow Water Count & Fast NonTurbulent Count & 0.29\\
Channel Unit & Fast Turbulent Percent & Channel Unit Count & -0.26\\
Channel Unit & Fast Turbulent Count & Fast Turbulent Percent & 0.26\\
\addlinespace
Channel Unit & Slow Water Percent & Channel Unit Count & 0.26\\
Channel Unit & Slow Water Percent & Channel Unit Frequency & 0.21\\
Channel Unit & Slow Water Count & Fast NonTurbulent Frequency & 0.21\\
Channel Unit & Fast NonTurbulent Count & Slow Water Frequency & 0.20\\
Channel Unit & Fast Turbulent Count & Fast NonTurbulent Frequency & 0.19\\
\addlinespace
Channel Unit & Fast Turbulent Count & Fast NonTurbulent Count & 0.19\\
Channel Unit & Fast Turbulent Percent & Channel Unit Frequency & -0.19\\
Channel Unit & Fast Turbulent Count & Fast NonTurbulent Percent & -0.19\\
Channel Unit & Fast Turbulent Percent & Fast Turbulent Frequency & 0.18\\
Channel Unit & Slow Water Percent & Fast Turbulent Count & -0.17\\
\addlinespace
Channel Unit & Fast NonTurbulent Percent & Fast Turbulent Frequency & -0.16\\
Channel Unit & Slow Water Percent & Fast NonTurbulent Percent & -0.14\\
Channel Unit & Fast NonTurbulent Count & Fast Turbulent Frequency & 0.13\\
Channel Unit & Fast NonTurbulent Percent & Slow Water Frequency & -0.11\\
Channel Unit & Slow Water Count & Fast NonTurbulent Percent & -0.09\\
\addlinespace
Channel Unit & Slow Water Percent & Fast Turbulent Frequency & -0.09\\
Channel Unit & Fast NonTurbulent Percent & Channel Unit Count & 0.06\\
Channel Unit & Slow Water Percent & Fast NonTurbulent Frequency & 0.03\\
Channel Unit & Fast NonTurbulent Percent & Channel Unit Frequency & 0.02\\
Channel Unit & Slow Water Percent & Fast NonTurbulent Count & 0.01\\
\hline
\addlinespace
Complexity & Bankfull Width CV & Bankfull Width To Depth Ratio CV & 0.84\\
Complexity & Wetted Channel Braidedness & Bankfull Channel Braidedness & 0.83\\
Complexity & Bankfull Width To Depth Ratio Avg & Wetted Width To Depth Ratio Avg & 0.78\\
Complexity & Wetted Channel Braidedness & Wetted Channel Qualifying Island Count & 0.78\\
Complexity & Bankfull Channel Braidedness & Bankfull Channel Qualifying Island Count & 0.78\\
\addlinespace
Complexity & Bankfull Width CV & Wetted Width CV & 0.76\\
Complexity & Wetted Channel Qualifying Island Count & Bankfull Channel Qualifying Island Count & 0.76\\
Complexity & Wetted Channel Braidedness & Bankfull Channel Qualifying Island Count & 0.66\\
Complexity & Bankfull Channel Braidedness & Wetted Channel Qualifying Island Count & 0.61\\
Complexity & Bankfull Width To Depth Ratio CV & Wetted Width CV & 0.57\\
\addlinespace
Complexity & Wetted Width CV & Wetted Width To Depth Ratio CV & 0.57\\
Complexity & Thalweg Depth CV & Wetted Width To Depth Ratio CV & 0.51\\
Complexity & Bankfull Width To Depth Ratio CV & Wetted Width To Depth Ratio CV & 0.50\\
Complexity & Bankfull Width CV & Wetted Width To Depth Ratio CV & 0.46\\
Complexity & Thalweg Depth CV & Wetted Width CV & 0.44\\
\addlinespace
Complexity & Wetted Depth SD & Bankfull Width To Depth Ratio Avg & 0.35\\
Complexity & Thalweg Depth CV & Bankfull Width To Depth Ratio CV & 0.30\\
Complexity & Thalweg Depth CV & Bankfull Width CV & 0.29\\
Complexity & Wetted Channel Qualifying Island Count & Bankfull Width CV & 0.25\\
Complexity & Bankfull Channel Qualifying Island Count & Bankfull Width CV & 0.25\\
\addlinespace
Complexity & Sinuosity & Wetted Depth SD & 0.24\\
Complexity & Detrended Elevation SD & Wetted Depth SD & 0.23\\
Complexity & Bankfull Width CV & Wetted Width To Depth Ratio Avg & -0.23\\
Complexity & Bankfull Channel Qualifying Island Count & Bankfull Width To Depth Ratio CV & 0.23\\
Complexity & Wetted Channel Qualifying Island Count & Bankfull Width To Depth Ratio CV & 0.22\\
\addlinespace
Complexity & Wetted Channel Braidedness & Bankfull Width CV & 0.21\\
Complexity & Bankfull Channel Braidedness & Bankfull Width CV & 0.21\\
Complexity & Wetted Depth SD & Wetted Width To Depth Ratio Avg & 0.21\\
Complexity & Sinuosity & Wetted Width To Depth Ratio CV & 0.20\\
Complexity & Bankfull Width To Depth Ratio CV & Wetted Width To Depth Ratio Avg & -0.20\\
\addlinespace
Complexity & Bankfull Channel Qualifying Island Count & Wetted Width CV & 0.20\\
Complexity & Sinuosity & Bankfull Width To Depth Ratio CV & 0.20\\
Complexity & Bankfull Channel Braidedness & Bankfull Width To Depth Ratio CV & 0.19\\
Complexity & Wetted Channel Qualifying Island Count & Wetted Width CV & 0.19\\
Complexity & Detrended Elevation SD & Bankfull Width To Depth Ratio CV & -0.19\\
\addlinespace
Complexity & Sinuosity & Thalweg Depth CV & 0.19\\
Complexity & Wetted Channel Braidedness & Bankfull Width To Depth Ratio CV & 0.18\\
Complexity & Bankfull Channel Braidedness & Wetted Width CV & 0.17\\
Complexity & Wetted Channel Braidedness & Bankfull Width To Depth Ratio Avg & 0.17\\
Complexity & Bankfull Channel Braidedness & Bankfull Width To Depth Ratio Avg & 0.16\\
\addlinespace
Complexity & Detrended Elevation SD & Bankfull Width CV & -0.16\\
Complexity & Wetted Channel Braidedness & Wetted Width CV & 0.16\\
Complexity & Sinuosity & Wetted Width CV & 0.14\\
Complexity & Sinuosity & Bankfull Width CV & 0.14\\
Complexity & Detrended Elevation SD & Wetted Width CV & -0.13\\
\addlinespace
Complexity & Detrended Elevation SD & Wetted Width To Depth Ratio CV & -0.13\\
Complexity & Sinuosity & Wetted Width To Depth Ratio Avg & -0.12\\
Complexity & Wetted Channel Qualifying Island Count & Bankfull Width To Depth Ratio Avg & 0.12\\
Complexity & Bankfull Channel Qualifying Island Count & Wetted Width To Depth Ratio CV & 0.11\\
Complexity & Bankfull Channel Qualifying Island Count & Bankfull Width To Depth Ratio Avg & 0.11\\
\addlinespace
Complexity & Detrended Elevation SD & Thalweg Depth CV & -0.11\\
Complexity & Bankfull Channel Braidedness & Wetted Width To Depth Ratio CV & 0.10\\
Complexity & Sinuosity & Detrended Elevation SD & -0.10\\
Complexity & Bankfull Width To Depth Ratio Avg & Wetted Width CV & 0.10\\
Complexity & Detrended Elevation SD & Bankfull Channel Qualifying Island Count & -0.09\\
\addlinespace
Complexity & Detrended Elevation SD & Wetted Channel Qualifying Island Count & -0.09\\
Complexity & Wetted Channel Qualifying Island Count & Wetted Width To Depth Ratio CV & 0.08\\
Complexity & Wetted Width CV & Wetted Width To Depth Ratio Avg & -0.08\\
Complexity & Bankfull Channel Braidedness & Wetted Depth SD & 0.08\\
Complexity & Wetted Depth SD & Wetted Width To Depth Ratio CV & 0.08\\
\addlinespace
Complexity & Detrended Elevation SD & Wetted Width To Depth Ratio Avg & 0.08\\
Complexity & Detrended Elevation SD & Wetted Channel Braidedness & -0.07\\
Complexity & Thalweg Depth CV & Bankfull Channel Qualifying Island Count & 0.07\\
Complexity & Detrended Elevation SD & Bankfull Channel Braidedness & -0.07\\
Complexity & Wetted Depth SD & Bankfull Width To Depth Ratio CV & 0.07\\
\addlinespace
Complexity & Bankfull Width To Depth Ratio CV & Bankfull Width To Depth Ratio Avg & 0.07\\
Complexity & Bankfull Channel Braidedness & Wetted Width To Depth Ratio Avg & 0.07\\
Complexity & Wetted Channel Braidedness & Wetted Depth SD & 0.06\\
Complexity & Thalweg Depth CV & Wetted Depth SD & 0.06\\
Complexity & Thalweg Depth CV & Bankfull Channel Braidedness & 0.06\\
\addlinespace
Complexity & Sinuosity & Bankfull Channel Qualifying Island Count & 0.06\\
Complexity & Wetted Channel Braidedness & Wetted Width To Depth Ratio CV & 0.06\\
Complexity & Sinuosity & Wetted Channel Qualifying Island Count & 0.06\\
Complexity & Wetted Depth SD & Wetted Width CV & 0.05\\
Complexity & Wetted Channel Braidedness & Wetted Width To Depth Ratio Avg & 0.04\\
\addlinespace
Complexity & Bankfull Width To Depth Ratio Avg & Wetted Width To Depth Ratio CV & 0.04\\
Complexity & Thalweg Depth CV & Wetted Width To Depth Ratio Avg & -0.04\\
Complexity & Sinuosity & Bankfull Channel Braidedness & 0.03\\
Complexity & Thalweg Depth CV & Bankfull Width To Depth Ratio Avg & -0.03\\
Complexity & Bankfull Channel Qualifying Island Count & Wetted Depth SD & 0.03\\
\addlinespace
Complexity & Wetted Width To Depth Ratio CV & Wetted Width To Depth Ratio Avg & 0.03\\
Complexity & Thalweg Depth CV & Wetted Channel Qualifying Island Count & 0.02\\
Complexity & Wetted Channel Qualifying Island Count & Wetted Width To Depth Ratio Avg & -0.02\\
Complexity & Sinuosity & Bankfull Width To Depth Ratio Avg & 0.02\\
Complexity & Wetted Depth SD & Bankfull Width CV & 0.02\\
\addlinespace
Complexity & Sinuosity & Wetted Channel Braidedness & 0.02\\
Complexity & Bankfull Width CV & Bankfull Width To Depth Ratio Avg & 0.01\\
Complexity & Detrended Elevation SD & Bankfull Width To Depth Ratio Avg & -0.01\\
Complexity & Wetted Channel Qualifying Island Count & Wetted Depth SD & 0.01\\
Complexity & Thalweg Depth CV & Wetted Channel Braidedness & 0.01\\
\addlinespace
Complexity & Bankfull Channel Qualifying Island Count & Wetted Width To Depth Ratio Avg & -0.01\\
\hline
Cover & Fish Cover: None & Fish Cover: Total & -0.94\\
Cover & Percent Undercut by Length & Percent Undercut by Area & 0.80\\
Cover & Fish Cover: Aquatic Vegetation & Fish Cover: Total & 0.70\\
Cover & Fish Cover: Terrestrial Vegetation & Fish Cover: Total & 0.68\\
\addlinespace
Cover & Fish Cover: None & Fish Cover: Aquatic Vegetation & -0.67\\
Cover & Fish Cover: Terrestrial Vegetation & Fish Cover: None & -0.65\\
Cover & Fish Cover: LW & Fish Cover: Total & 0.54\\
Cover & Fish Cover: LW & Fish Cover: None & -0.50\\
Cover & Fish Cover: LW & Fish Cover: Terrestrial Vegetation & 0.35\\
\addlinespace
Cover & Fish Cover: Total & Percent Undercut by Area & 0.29\\
Cover & Fish Cover: Aquatic Vegetation & Percent Undercut by Length & 0.28\\
Cover & Fish Cover: None & Percent Undercut by Area & -0.27\\
Cover & Fish Cover: Total & Percent Undercut by Length & 0.25\\
Cover & Fish Cover: None & Percent Undercut by Length & -0.22\\
\addlinespace
Cover & Fish Cover: LW & Percent Undercut by Area & 0.20\\
Cover & Fish Cover: Terrestrial Vegetation & Percent Undercut by Area & 0.20\\
Cover & Fish Cover: Aquatic Vegetation & Percent Undercut by Area & 0.18\\
Cover & Fish Cover: Artificial & Fish Cover: None & -0.12\\
Cover & Fish Cover: Artificial & Fish Cover: Total & 0.11\\
\addlinespace
Cover & Fish Cover: LW & Percent Undercut by Length & 0.11\\
Cover & Fish Cover: Artificial & Fish Cover: Aquatic Vegetation & 0.07\\
Cover & Fish Cover: Terrestrial Vegetation & Percent Undercut by Length & 0.06\\
Cover & Fish Cover: Artificial & Percent Undercut by Length & -0.05\\
Cover & Fish Cover: Artificial & Percent Undercut by Area & -0.03\\
\addlinespace
Cover & Fish Cover: LW & Fish Cover: Aquatic Vegetation & -0.02\\
Cover & Fish Cover: Terrestrial Vegetation & Fish Cover: Aquatic Vegetation & 0.02\\
Cover & Fish Cover: LW & Fish Cover: Artificial & 0.02\\
Cover & Fish Cover: Terrestrial Vegetation & Fish Cover: Artificial & -0.01\\
\hline
Riparian & Riparian Cover: Understory & RipCovUstoryNone & -1.00\\
\addlinespace
Riparian & Riparian Cover: No Canopy & Riparian Cover: Some Canopy & -1.00\\
Riparian & Riparian Cover: Ground & RipCovGrndNone & -0.99\\
Riparian & Riparian Cover: Understory & Riparian Cover: Woody & 0.85\\
Riparian & Riparian Cover: Woody & RipCovUstoryNone & -0.85\\
Riparian & Riparian Cover: Ground & Riparian Cover: Non-Woody & 0.83\\
\addlinespace
Riparian & Riparian Cover: Big Tree & Riparian Cover: No Canopy & -0.83\\
Riparian & Riparian Cover: Big Tree & Riparian Cover: Some Canopy & 0.83\\
Riparian & Riparian Cover: Non-Woody & RipCovGrndNone & -0.83\\
Riparian & Riparian Cover: Woody & Riparian Cover: No Canopy & -0.73\\
Riparian & Riparian Cover: Woody & Riparian Cover: Some Canopy & 0.73\\
\addlinespace
Riparian & Riparian Cover: Big Tree & Riparian Cover: Coniferous & 0.59\\
Riparian & Riparian Cover: Big Tree & Riparian Cover: Woody & 0.59\\
Riparian & Riparian Cover: Coniferous & Riparian Cover: No Canopy & -0.52\\
Riparian & Riparian Cover: Coniferous & Riparian Cover: Some Canopy & 0.52\\
Riparian & Riparian Cover: Understory & Riparian Cover: No Canopy & -0.49\\
\addlinespace
Riparian & Riparian Cover: Understory & Riparian Cover: Some Canopy & 0.49\\
Riparian & Riparian Cover: No Canopy & RipCovUstoryNone & 0.49\\
Riparian & RipCovUstoryNone & Riparian Cover: Some Canopy & -0.49\\
Riparian & Riparian Cover: Coniferous & Riparian Cover: Woody & 0.43\\
Riparian & Riparian Cover: Big Tree & Riparian Cover: Understory & 0.35\\
\addlinespace
Riparian & Riparian Cover: Big Tree & RipCovUstoryNone & -0.35\\
Riparian & Riparian Cover: Coniferous & Riparian Cover: Understory & 0.24\\
Riparian & Riparian Cover: Coniferous & RipCovUstoryNone & -0.24\\
Riparian & Riparian Cover: Non-Woody & Riparian Cover: Woody & -0.21\\
Riparian & Riparian Cover: Big Tree & Riparian Cover: Non-Woody & -0.20\\
\addlinespace
Riparian & Riparian Cover: Non-Woody & Riparian Cover: No Canopy & 0.17\\
Riparian & Riparian Cover: Non-Woody & Riparian Cover: Some Canopy & -0.17\\
Riparian & RipCovUstoryNone & RipCovGrndNone & 0.17\\
Riparian & Riparian Cover: Understory & RipCovGrndNone & -0.17\\
Riparian & Riparian Cover: Ground & Riparian Cover: Understory & 0.16\\
\addlinespace
Riparian & Riparian Cover: Ground & RipCovUstoryNone & -0.16\\
Riparian & Riparian Cover: Woody & RipCovGrndNone & -0.16\\
Riparian & Riparian Cover: Ground & Riparian Cover: Woody & 0.15\\
Riparian & Riparian Cover: Coniferous & Riparian Cover: Non-Woody & -0.13\\
Riparian & Riparian Cover: Big Tree & Riparian Cover: Ground & -0.12\\
\addlinespace
Riparian & Riparian Cover: Big Tree & RipCovGrndNone & 0.11\\
Riparian & Riparian Cover: Ground & Riparian Cover: No Canopy & 0.09\\
Riparian & Riparian Cover: Ground & Riparian Cover: Some Canopy & -0.09\\
Riparian & Riparian Cover: No Canopy & RipCovGrndNone & -0.08\\
Riparian & RipCovGrndNone & Riparian Cover: Some Canopy & 0.08\\
\addlinespace
Riparian & Riparian Cover: Coniferous & RipCovGrndNone & -0.01\\
Riparian & Riparian Cover: Non-Woody & Riparian Cover: Understory & -0.01\\
Riparian & Riparian Cover: Non-Woody & RipCovUstoryNone & 0.01\\
Riparian & Riparian Cover: Coniferous & Riparian Cover: Ground & 0.00\\
\hline
Side Channel & SCSm\_Freq & SCSm\_Ct & 0.90\\
\addlinespace
Side Channel & SCSm\_Freq & SC\_Area\_Pct & 0.49\\
Side Channel & SCSm\_Ct & SC\_Area\_Pct & 0.49\\
Side Channel & Wetted Side Channel Percent By Area & SCSm\_Ct & 0.44\\
Side Channel & Wetted Side Channel Percent By Area & SCSm\_Freq & 0.40\\
Side Channel & Wetted Side Channel Percent By Area & SC\_Area\_Pct & 0.30\\
\hline
\addlinespace
Size & Wetted Width Integrated & Wetted Width Avg & 0.98\\
Size & Bankfull Width Integrated & Bankfull Width Avg & 0.97\\
Size & Bankfull Width Avg & Wetted Width Avg & 0.95\\
Size & Wetted Width Integrated & Bankfull Width Avg & 0.95\\
Size & Wetted Width Integrated & Bankfull Width Integrated & 0.92\\
\addlinespace
Size & Bankfull Width Integrated & Wetted Width Avg & 0.91\\
Size & Bankfull Depth Avg & DpthThlwg\_Avg & 0.85\\
Size & Bankfull Depth Max & DpthThlwg\_Avg & 0.84\\
Size & Bankfull Width Avg & Bankfull Depth Avg & 0.83\\
Size & Wetted Width Avg & Bankfull Depth Avg & 0.82\\
\addlinespace
Size & Wetted Width Avg & Discharge & 0.82\\
Size & Bankfull Depth Max & Bankfull Depth Avg & 0.82\\
Size & Bankfull Depth Max & Residual Pool Depth & 0.81\\
Size & Wetted Width Integrated & Discharge & 0.79\\
Size & Bankfull Width Integrated & Bankfull Depth Avg & 0.79\\
\addlinespace
Size & Wetted Width Integrated & Bankfull Depth Avg & 0.79\\
Size & Wetted Width Avg & DpthThlwg\_Avg & 0.79\\
Size & Wetted Width Integrated & DpthThlwg\_Avg & 0.78\\
Size & Bankfull Width Avg & DpthThlwg\_Avg & 0.77\\
Size & Bankfull Width Avg & Discharge & 0.77\\
\addlinespace
Size & DpthThlwg\_Avg & Residual Pool Depth & 0.76\\
Size & Bankfull Width Integrated & DpthThlwg\_Avg & 0.75\\
Size & Bankfull Width Integrated & Discharge & 0.74\\
Size & Bankfull Width Avg & Bankfull Depth Max & 0.74\\
Size & DpthThlwg\_Avg & Discharge & 0.74\\
\addlinespace
Size & Drainage Area (Flowline) & Wetted Width Avg & 0.73\\
Size & Bankfull Width Integrated & Bankfull Depth Max & 0.73\\
Size & Drainage Area (Flowline) & Wetted Width Integrated & 0.72\\
Size & Drainage Area (Flowline) & Bankfull Width Avg & 0.72\\
Size & Drainage Area (Flowline) & Discharge & 0.70\\
\addlinespace
Size & Wetted Width Avg & Bankfull Depth Max & 0.70\\
Size & Bankfull Width Avg & Residual Pool Depth & 0.69\\
Size & Wetted Width Integrated & Bankfull Depth Max & 0.69\\
Size & Drainage Area (Flowline) & Bankfull Width Integrated & 0.68\\
Size & Bankfull Depth Avg & Residual Pool Depth & 0.68\\
\addlinespace
Size & Bankfull Depth Avg & Discharge & 0.68\\
Size & Wetted Width Avg & Residual Pool Depth & 0.67\\
Size & Wetted Width Integrated & Residual Pool Depth & 0.67\\
Size & Bankfull Width Integrated & Residual Pool Depth & 0.67\\
Size & Residual Pool Depth & Discharge & 0.60\\
\addlinespace
Size & Bankfull Depth Max & Discharge & 0.59\\
Size & Drainage Area (Flowline) & Bankfull Depth Avg & 0.56\\
Size & Drainage Area (Flowline) & Residual Pool Depth & 0.54\\
Size & Drainage Area (Flowline) & DpthThlwg\_Avg & 0.52\\
Size & Drainage Area (Flowline) & Bankfull Depth Max & 0.46\\
\addlinespace
Size & Gradient & Bankfull Width Avg & -0.44\\
Size & Gradient & Bankfull Width Integrated & -0.43\\
Size & Gradient & Wetted Width Avg & -0.42\\
Size & Gradient & Wetted Width Integrated & -0.42\\
Size & Gradient & DpthThlwg\_Avg & -0.39\\
\addlinespace
Size & Gradient & Residual Pool Depth & -0.33\\
Size & Gradient & Bankfull Depth Avg & -0.33\\
Size & Drainage Area (Flowline) & Gradient & -0.31\\
Size & Gradient & Bankfull Depth Max & -0.31\\
Size & Gradient & Discharge & -0.30\\
\addlinespace
Size & Precipitation & Drainage Area (Flowline) & -0.15\\
Size & Precipitation & Bankfull Depth Avg & 0.11\\
Size & Precipitation & Bankfull Depth Max & 0.10\\
Size & Precipitation & DpthThlwg\_Avg & 0.09\\
Size & Precipitation & Discharge & -0.05\\
\addlinespace
Size & Precipitation & Gradient & 0.03\\
Size & Precipitation & Wetted Width Integrated & -0.02\\
Size & Precipitation & Bankfull Width Integrated & -0.02\\
Size & Precipitation & Wetted Width Avg & -0.02\\
Size & Precipitation & Bankfull Width Avg & -0.01\\
\addlinespace
Size & Precipitation & Residual Pool Depth & -0.01\\
\hline
Substrate & Substrate < 2mm & Substrate < 6mm & 0.95\\
Substrate & Substrate: D50 & Substrate: D84 & 0.87\\
Substrate & Substrate: Embeddedness Avg & Substrate: Embeddedness SD & 0.86\\
Substrate & Substrate: D84 & Substrate Est: Boulders & 0.80\\
\addlinespace
Substrate & Substrate: D50 & Substrate Est: Boulders & 0.71\\
Substrate & Substrate < 2mm & Substrate Est: Sand and Fines & 0.68\\
Substrate & Substrate < 6mm & Substrate Est: Sand and Fines & 0.67\\
Substrate & Substrate Est: Sand and Fines & Substrate Est: Cobbles & -0.67\\
Substrate & Substrate: D16 & Substrate: D50 & 0.61\\
\addlinespace
Substrate & Substrate Est: Coarse and Fine Gravel & Substrate Est: Boulders & -0.59\\
Substrate & Substrate: D84 & Substrate Est: Coarse and Fine Gravel & -0.52\\
Substrate & Substrate < 6mm & Substrate Est: Cobbles & -0.46\\
Substrate & Substrate: D50 & Substrate Est: Sand and Fines & -0.45\\
Substrate & Substrate: D16 & Substrate Est: Sand and Fines & -0.43\\
\addlinespace
Substrate & Substrate < 2mm & Substrate Est: Cobbles & -0.43\\
Substrate & Substrate: D50 & Substrate Est: Coarse and Fine Gravel & -0.43\\
Substrate & Substrate: D16 & Substrate Est: Cobbles & 0.42\\
Substrate & Substrate: Embeddedness Avg & Substrate < 2mm & 0.42\\
Substrate & Substrate: D16 & Substrate < 6mm & -0.42\\
\addlinespace
Substrate & Substrate: Embeddedness Avg & Substrate Est: Sand and Fines & 0.40\\
Substrate & Substrate: D50 & Substrate Est: Cobbles & 0.40\\
Substrate & Substrate: D84 & Substrate Est: Sand and Fines & -0.39\\
Substrate & Substrate: D16 & Substrate < 2mm & -0.39\\
Substrate & Substrate: D50 & Substrate < 6mm & -0.38\\
\addlinespace
Substrate & Substrate: Embeddedness Avg & Substrate < 6mm & 0.38\\
Substrate & Substrate: D16 & Substrate: D84 & 0.38\\
Substrate & Substrate Est: Sand and Fines & Substrate Est: Boulders & -0.36\\
Substrate & Substrate: D50 & Substrate < 2mm & -0.35\\
Substrate & Substrate: Embeddedness SD & Substrate < 2mm & 0.34\\
\addlinespace
Substrate & Substrate: D84 & Substrate Est: Cobbles & 0.33\\
Substrate & Substrate: Embeddedness SD & Substrate Est: Sand and Fines & 0.31\\
Substrate & Substrate: Embeddedness SD & Substrate < 6mm & 0.31\\
Substrate & Substrate: Embeddedness Avg & Substrate: D16 & -0.30\\
Substrate & Substrate Est: Coarse and Fine Gravel & Substrate Est: Cobbles & -0.30\\
\addlinespace
Substrate & Substrate: D16 & Substrate Est: Boulders & 0.29\\
Substrate & Substrate: D84 & Substrate < 6mm & -0.28\\
Substrate & Substrate: Embeddedness SD & Substrate: D16 & -0.27\\
Substrate & Substrate: Embeddedness Avg & Substrate Est: Cobbles & -0.25\\
Substrate & Substrate: D84 & Substrate < 2mm & -0.25\\
\addlinespace
Substrate & Substrate Est: Coarse and Fine Gravel & Substrate Est: Sand and Fines & -0.23\\
Substrate & Substrate < 2mm & Substrate Est: Coarse and Fine Gravel & -0.22\\
Substrate & Substrate Est: Boulders & Substrate Est: Cobbles & 0.22\\
Substrate & Substrate < 6mm & Substrate Est: Boulders & -0.19\\
Substrate & Substrate: Embeddedness SD & Substrate Est: Cobbles & -0.19\\
\addlinespace
Substrate & Substrate < 2mm & Substrate Est: Boulders & -0.19\\
Substrate & Substrate: Embeddedness Avg & Substrate Est: Coarse and Fine Gravel & -0.18\\
Substrate & Substrate < 6mm & Substrate Est: Coarse and Fine Gravel & -0.18\\
Substrate & Substrate: Embeddedness Avg & Substrate: D50 & -0.17\\
Substrate & Substrate: Embeddedness SD & Substrate Est: Coarse and Fine Gravel & -0.15\\
\addlinespace
Substrate & Substrate: Embeddedness SD & Substrate: D50 & -0.13\\
Substrate & Substrate: D16 & Substrate Est: Coarse and Fine Gravel & -0.12\\
Substrate & Substrate: Embeddedness Avg & Substrate: D84 & -0.08\\
Substrate & Substrate: Embeddedness SD & Substrate: D84 & -0.05\\
Substrate & Substrate: Embeddedness Avg & Substrate Est: Boulders & -0.03\\
\addlinespace
Substrate & Substrate: Embeddedness SD & Substrate Est: Boulders & -0.02\\
\hline
Temperature & Avg. August Temperature & August Temperature & 0.98\\
Temperature & Elev\_M & Avg. August Temperature & -0.61\\
Temperature & Elev\_M & August Temperature & -0.61\\
Temperature & Solar Access: Summer Avg & August Temperature & 0.19\\
\addlinespace
Temperature & Solar Access: Summer Avg & Avg. August Temperature & 0.13\\
Temperature & Elev\_M & Solar Access: Summer Avg & 0.12\\
\hline
Water Quality & Conductivity & Alkalinity & 0.78\\
Water Quality & Alkalinity & Drift Biomass & 0.16\\
Water Quality & Conductivity & Drift Biomass & 0.16\\
\hline
\addlinespace
Wood & Large Wood Volume: Wetted Slow Water & Large Wood Volume: Bankfull Slow Water & 0.91\\
Wood & Large Wood Frequency: Wetted & Large Wood Frequency: Bankfull & 0.91\\
Wood & Large Wood Volume: Wetted & Large Wood Volume: Bankfull & 0.90\\
Wood & Large Wood Volume: Wetted & Large Wood Volume: Wetted Slow Water & 0.88\\
Wood & Large Wood Volume: Bankfull & Large Wood Volume: Bankfull Slow Water & 0.84\\
\addlinespace
Wood & Large Wood Volume: Wetted & Large Wood Volume: Bankfull Slow Water & 0.81\\
Wood & Large Wood Volume: Wetted Fast NonTurbulent & Large Wood Volume: Bankfull Fast NonTurbulent & 0.79\\
Wood & Large Wood Volume: Bankfull & Large Wood Volume: Wetted Slow Water & 0.78\\
Wood & Large Wood Volume: Wetted & Large Wood Volume: Wetted Fast NonTurbulent & 0.60\\
Wood & Large Wood Volume: Bankfull & Large Wood Volume: Bankfull Fast NonTurbulent & 0.56\\
\addlinespace
Wood & Large Wood Volume: Bankfull & Large Wood Frequency: Bankfull & 0.56\\
Wood & Large Wood Volume: Wetted & Large Wood Volume: Wetted Fast Turbulent & 0.53\\
Wood & Large Wood Volume: Wetted & Large Wood Volume: Bankfull Fast NonTurbulent & 0.52\\
Wood & Large Wood Volume: Wetted & Large Wood Frequency: Wetted & 0.51\\
Wood & Large Wood Volume: Wetted & Large Wood Frequency: Bankfull & 0.51\\
\addlinespace
Wood & Large Wood Volume: Bankfull & Large Wood Volume: Wetted Fast NonTurbulent & 0.50\\
Wood & Large Wood Volume: Bankfull & Large Wood Volume: Wetted Fast Turbulent & 0.50\\
Wood & Large Wood Volume: Bankfull & Large Wood Frequency: Wetted & 0.45\\
Wood & Large Wood Volume: Wetted Slow Water & Large Wood Volume: Wetted Fast NonTurbulent & 0.45\\
Wood & Large Wood Volume: Bankfull Slow Water & Large Wood Frequency: Bankfull & 0.43\\
\addlinespace
Wood & Large Wood Volume: Wetted Slow Water & Large Wood Frequency: Wetted & 0.42\\
Wood & Large Wood Volume: Wetted Slow Water & Large Wood Volume: Bankfull Fast NonTurbulent & 0.40\\
Wood & Large Wood Volume: Wetted Slow Water & Large Wood Frequency: Bankfull & 0.40\\
Wood & Large Wood Volume: Bankfull Slow Water & Large Wood Volume: Wetted Fast NonTurbulent & 0.39\\
Wood & Large Wood Volume: Bankfull Slow Water & Large Wood Frequency: Wetted & 0.39\\
\addlinespace
Wood & Large Wood Volume: Bankfull Slow Water & Large Wood Volume: Bankfull Fast NonTurbulent & 0.38\\
Wood & Large Wood Volume: Wetted Fast Turbulent & Large Wood Frequency: Bankfull & 0.37\\
Wood & Large Wood Volume: Wetted Fast Turbulent & Large Wood Frequency: Wetted & 0.35\\
Wood & Large Wood Frequency: Bankfull & Large Wood Volume: Bankfull Fast NonTurbulent & 0.27\\
Wood & Large Wood Volume: Wetted Fast NonTurbulent & Large Wood Frequency: Wetted & 0.24\\
\addlinespace
Wood & Large Wood Frequency: Wetted & Large Wood Volume: Bankfull Fast NonTurbulent & 0.22\\
Wood & Large Wood Volume: Wetted Fast NonTurbulent & Large Wood Frequency: Bankfull & 0.22\\
Wood & Large Wood Volume: Wetted Slow Water & Large Wood Volume: Wetted Fast Turbulent & 0.15\\
Wood & Large Wood Volume: Bankfull Slow Water & Large Wood Volume: Wetted Fast Turbulent & 0.14\\
Wood & Large Wood Volume: Wetted Fast Turbulent & Large Wood Volume: Wetted Fast NonTurbulent & 0.12\\
\addlinespace
Wood & Large Wood Volume: Wetted Fast Turbulent & Large Wood Volume: Bankfull Fast NonTurbulent & 0.12\\*
\end{longtable}

\newpage

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\begin{figure}
\centering
\includegraphics{../../figures/appendix/mine-facet-fig-1.pdf}
\caption{\label{fig:mine-facet-fig}Barplots of MIC statistics, faceted by habitat category.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../../figures/appendix/mine-all-fig-1.pdf}
\caption{\label{fig:mine-all-fig}Barplot of MIC statistics, colored by habitat category.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../../figures/appendix/selected-corr-1.pdf}
\caption{\label{fig:selected-corr}Correlation plot of habitat metrics used in the QRF model.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../../figures/appendix/high-cor-mets-1.pdf}
\caption{\label{fig:high-cor-mets}Pairs plot of habitat metrics used in QRF model with a correlation coefficient greater than 0.5 or less than -0.5.}
\end{figure}

\end{document}

### 2020-12-01 analysis/paper/paper.Rmd
updated the link to the GitHub repo with the DOI link.

@@
 
-All code and data for the analyses presented here can be found in a GitHub repository ([(https://github.com/KevinSee/QRFpaper)](https://github.com/KevinSee/QRFpaper)).
+All code and data for the analyses presented here can be found in a GitHub repository ([(https://doi.org/10.5281/zenodo.4300942)](https://doi.org/10.5281/zenodo.4300942)).
+
+<!-- ([(https://github.com/KevinSee/QRFpaper)](https://github.com/KevinSee/QRFpaper)) -->
 

### 2020-12-02 analysis/R_scripts/compile_wtsd_bndry.Rmd
updated R notebook for making watershed map. Changed what file types to save, deleted some obsolete versions.

@@
 # set default theme for ggplot
-theme_set(theme_bw())
+theme_set(theme_bw(base_family = "Helvetica"))
 # setwd('analysis/R_scripts')
@@
                                            st_make_grid(n = 1)) %>%
-            filter(!grepl('^1701', reachcode),
-                   !grepl('^171', reachcode))
+            filter(!grepl('^1701', REACHCODE),
+                   !grepl('^171', REACHCODE))
 
@@
 # save a couple versions
-ggsave('../figures/ChampWtds.svg',
+for(file_ext in c("png", 'pdf', 'tiff')) {
+  ggsave(paste0('../figures/ChampWtds.', file_ext),
        wtsd_map,
-       dpi = 600,
+       dpi = 400,
        height = 6,
        width = 6)
-ggsave('../figures/ChampWtds.png',
-       wtsd_map,
-       dpi = 600,
-       height = 6,
-       width = 6)
-ggsave('../figures/ChampWtds.tiff',
-       wtsd_map,
-       dpi = 600,
-       height = 6,
-       width = 6)
-ggsave('../figures/ChampWtds.jpeg',
-       wtsd_map,
-       dpi = 600,
-       height = 6,
-       width = 6)
+}
+
 ```

### 2020-12-02 analysis/paper/paper.Rmd
finalized how figures are saved, changed one thing with the GitHub DOI link, and re-knit to all 3 formats

@@
 output:
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: false
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       fig_width: 6
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: false
-      toc: no
-      fig_width: 7
       pandoc_args:
@@
   fig.path = "../figures/",
-  dpi = 400,
-  dev = c("pdf", 'tiff')
+  dev = c("pdf", "png", 'tiff'),
+  dpi = 400
 )
@@
 
-if(doc.type %in% c("docx", "html")) {
-  theme_set(theme_pubr(base_family = "Times New Roman"))
-}
+# if(doc.type %in% c("docx", "html")) {
+#   theme_set(theme_pubr(base_family = "Times New Roman"))
+# }
 
@@
 
-All code and data for the analyses presented here can be found in a GitHub repository ([(https://doi.org/10.5281/zenodo.4300942)](https://doi.org/10.5281/zenodo.4300942)).
+All code and data for the analyses presented here can be found in a GitHub repository ([https://doi.org/10.5281/zenodo.4300942](https://doi.org/10.5281/zenodo.4300942)).
 

### 2020-12-02 analysis/paper/paper.tex
finalized how figures are saved, changed one thing with the GitHub DOI link, and re-knit to all 3 formats

@@
 
-All code and data for the analyses presented here can be found in a GitHub repository (\href{https://github.com/KevinSee/QRFpaper}{(https://github.com/KevinSee/QRFpaper)}).
+All code and data for the analyses presented here can be found in a GitHub repository (\url{https://doi.org/10.5281/zenodo.4300942}).
 
@@
 \centering
-\includegraphics{figures/rel-imp-figure-1.pdf}
+\includegraphics{../figures/rel-imp-figure-1.pdf}
 \caption{\label{fig:rel-imp-figure}Relative importance of each habitat covariate included in the quantile random forest (QRF) model to predict habitat capacity, during summer months, for spring/summer-run Chinook salmon parr}
@@
 \centering
-\includegraphics{figures/pdp-figure-1.pdf}
+\includegraphics{../figures/pdp-figure-1.pdf}
 \caption{\label{fig:pdp-figure}Partial dependence plots for the spring/summer-run Chinook salmon parr capacity quantile random forest (QRF) model, depicting how parr capacity shifts as each habitat metric changes, assuming all other habitat metrics remain at their mean values.
@@
 \centering
-\includegraphics{figures/sr-figure-1.pdf}
+\includegraphics{../figures/sr-figure-1.pdf}
 \caption{\label{fig:sr-figure}Spawner-recruit data from nine watersheds.

### 2020-12-02 analysis/paper/paper.Rmd
added Chris Jordan to acknowledgements (Sorry Chris!)

@@
         in_header: ../templates/header_manuscript.tex
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: false
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
 }
-
-# if(doc.type %in% c("docx", "html")) {
-#   theme_set(theme_pubr(base_family = "Times New Roman"))
-# }
 
@@
 Fish and Wildlife Service.
-The models in this study were improved by conversations with Eric Buhle.
+The models in this study were improved by conversations with Chris Jordan and Eric Buhle.
 
@@
 
-```{r wtsd-map, out.width = "90%", fig.cap = "Watersheds with CHaMP habitat data.
+```{r wtsd-map, out.width = "85%", fig.cap = "Watersheds with CHaMP habitat data.
 Watersheds in black also contain paired fish data.

### 2020-12-02 analysis/paper/paper.tex
added Chris Jordan to acknowledgements (Sorry Chris!)

@@
 \begin{figure}
-\includegraphics[width=0.9\linewidth]{../figures/ChampWtds} \caption{Watersheds with CHaMP habitat data.
+\includegraphics[width=0.85\linewidth]{../figures/ChampWtds} \caption{Watersheds with CHaMP habitat data.
 Watersheds in black also contain paired fish data.

### 2020-12-02 analysis/paper/appendix/appendix_s1.Rmd
made the font size 10 in all tables in main document and appendix

@@
 
-if (doc.type != "docx") {
+if(doc.type != "docx") {
   mine_tab_df %>%
     select(-selected) %>%
+    mutate(Category = recode(Category,
+                             "ChannelUnit" = "Channel Unit",
+                             "SideChannel" = "Side Channel")) %>%
   kable(digits = 3,
@@
     kable_styling(fixed_thead = T,
+                  font_size = 10,
                   latex_options = c('repeat_header',
@@
   library(pander)
+  panderOptions("graph.fontsize", 10)
   mine_tab_df %>%
@@
   library(pander)
-
+  panderOptions("graph.fontsize", 10)
   corrr_df %>%

### 2020-12-02 analysis/paper/appendix/appendix_s1.tex
made the font size 10 in all tables in main document and appendix

@@
 \subtitle{Ecospheres}
-\author{Kevin E. See\textsuperscript{1,*}, 
-    \and Michael W. Ackerman\textsuperscript{1}, 
-    \and Richard A. Carmichael\textsuperscript{1},
-    \and Sarah L. Hoffmann\textsuperscript{1}, 
-    \and Chris Beasley\textsuperscript{1}}
-\date{December 01, 2020}
+\author{Kevin E. See\textsuperscript{1,*}, Michael W. Ackerman\textsuperscript{1}, Richard A. Carmichael\textsuperscript{1}, Sarah L. Hoffmann\textsuperscript{1}, and Chris Beasley\textsuperscript{1}}
+\date{December 02, 2020}
 
@@
 
+\begingroup\fontsize{10}{12}\selectfont
+
 \begin{longtable}[t]{l>{\raggedright\arraybackslash}p{2in}>{\raggedright\arraybackslash}p{1in}>{\raggedleft\arraybackslash}p{0.5in}>{\raggedleft\arraybackslash}p{0.5in}>{\raggedleft\arraybackslash}p{0.5in}}
@@
 \endlastfoot
-\textbf{ChannelUnit} & \textbf{Channel Unit Frequency} & \textbf{CU\_Freq} & \textbf{0.241} & \textbf{0.021} & \textbf{0.021}\\
-ChannelUnit & Fast Turbulent Frequency & FstTurb\_Freq & 0.230 & 0.021 & 0.082\\
-ChannelUnit & Fast NonTurbulent Frequency & FstNT\_Freq & 0.209 & 0.021 & 0.308\\
-ChannelUnit & Slow Water Frequency & SlowWater\_Freq & 0.208 & 0.021 & 0.073\\
-ChannelUnit & Fast Turbulent Percent & FstTurb\_Pct & 0.195 & 0.021 & 0.082\\
+\textbf{Channel Unit} & \textbf{Channel Unit Frequency} & \textbf{CU\_Freq} & \textbf{0.241} & \textbf{0.021} & \textbf{0.021}\\
+Channel Unit & Fast Turbulent Frequency & FstTurb\_Freq & 0.230 & 0.021 & 0.082\\
+Channel Unit & Fast NonTurbulent Frequency & FstNT\_Freq & 0.209 & 0.021 & 0.308\\
+Channel Unit & Slow Water Frequency & SlowWater\_Freq & 0.208 & 0.021 & 0.073\\
+Channel Unit & Fast Turbulent Percent & FstTurb\_Pct & 0.195 & 0.021 & 0.082\\
 \addlinespace
-ChannelUnit & Channel Unit Count & CU\_Ct & 0.189 & 0.021 & 0.021\\
-ChannelUnit & Fast Turbulent Count & FstTurb\_Ct & 0.178 & 0.021 & 0.082\\
-ChannelUnit & Slow Water Percent & SlowWater\_Pct & 0.177 & 0.021 & 0.073\\
-ChannelUnit & Fast NonTurbulent Percent & FstNT\_Pct & 0.169 & 0.021 & 0.308\\
-ChannelUnit & Fast NonTurbulent Count & FstNT\_Ct & 0.166 & 0.021 & 0.308\\
+Channel Unit & Channel Unit Count & CU\_Ct & 0.189 & 0.021 & 0.021\\
+Channel Unit & Fast Turbulent Count & FstTurb\_Ct & 0.178 & 0.021 & 0.082\\
+Channel Unit & Slow Water Percent & SlowWater\_Pct & 0.177 & 0.021 & 0.073\\
+Channel Unit & Fast NonTurbulent Percent & FstNT\_Pct & 0.169 & 0.021 & 0.308\\
+Channel Unit & Fast NonTurbulent Count & FstNT\_Ct & 0.166 & 0.021 & 0.308\\
 \addlinespace
@@
 Riparian & Riparian Cover: Coniferous & RipCovConif & 0.164 & 0.009 & 0.192\\
-SideChannel & Bankfull Side Channel Width & BfSCWdth & 0.223 & 0.796 & 0.796\\
-SideChannel & Wetted Side Channel Width & WetSCWdth & 0.213 & 0.832 & 0.832\\
+Side Channel & Bankfull Side Channel Width & BfSCWdth & 0.223 & 0.796 & 0.796\\
+Side Channel & Wetted Side Channel Width & WetSCWdth & 0.213 & 0.832 & 0.832\\
 \addlinespace
-SideChannel & Wetted Side Channel Percent By Area & WetSC\_Pct & 0.209 & 0.021 & 0.820\\
-SideChannel & SCSm\_Freq & SCSm\_Freq & 0.153 & 0.021 & 0.921\\
-SideChannel & SCSm\_Ct & SCSm\_Ct & 0.153 & 0.021 & 0.921\\
-SideChannel & SC\_Area\_Pct & SC\_Area\_Pct & 0.153 & 0.021 & 0.921\\
+Side Channel & Wetted Side Channel Percent By Area & WetSC\_Pct & 0.209 & 0.021 & 0.820\\
+Side Channel & SCSm\_Freq & SCSm\_Freq & 0.153 & 0.021 & 0.921\\
+Side Channel & SCSm\_Ct & SCSm\_Ct & 0.153 & 0.021 & 0.921\\
+Side Channel & SC\_Area\_Pct & SC\_Area\_Pct & 0.153 & 0.021 & 0.921\\
 Size & Mean Annual Flow & MeanU & 0.346 & 0.476 & 0.476\\
@@
 \end{longtable}
+\endgroup{}
+
+\begingroup\fontsize{10}{12}\selectfont
 
@@
 \end{longtable}
+\endgroup{}
 

### 2020-12-02 analysis/paper/paper.Rmd
made the font size 10 in all tables in main document and appendix

@@
 output:
+    bookdown::word_document2:
+      fig_caption: yes
+      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: false
+      toc: no
+      always_allow_html: true
+      fig_width: 6
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: false
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: false
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
-      fig_caption: yes
-      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: false
-      toc: no
-      always_allow_html: true
-      fig_width: 6
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
 keywords: |
-  carrying capacity; quantile regression; random forests; Chinook; quantile random forest
+  carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
 highlights: |
@@
   fig.path = "../figures/",
-  dev = c("pdf", "png", 'tiff'),
+  dev = c("pdf", "png", "tiff"),
   dpi = 400
@@
 
-Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity.
+Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence/absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity.
 Further, many of these studies focus on only one or two measures of habitat.
@@
         linesep = "") %>%
-  kable_styling(latex_options = c("hold_position"))
+  kable_styling(font_size = 10,
+                latex_options = c("hold_position"))
 
@@
         format.args = list(width = c(2, 10, 10, 50))) %>%
-  kable_styling(latex_options = c("hold_position")) %>%
+  kable_styling(font_size = 10,
+                latex_options = c("hold_position")) %>%
   column_spec(4, width = '3in')
@@
         digits = 3) %>%
-  kable_styling(latex_options = c("hold_position"))
+  kable_styling(font_size = 10,
+                latex_options = c("hold_position"))
 
@@
         linesep = "") %>%
-  kable_styling(latex_options = c("hold_position")) %>%
+  kable_styling(font_size = 10,
+                latex_options = c("hold_position")) %>%
   # kable_styling(latex_options = c('scale_down')) %>%
@@
   theme(legend.position = 'none',
-        strip.text = element_text(size = 10),
+        strip.text = element_text(size = 9),
         axis.text = element_text(size = 8))
@@
 spawn_rec_comp_p +
-  theme(strip.text = element_text(size = 10),
+  theme(strip.text = element_text(size = 9),
         axis.text = element_text(size = 8))

### 2020-12-02 analysis/paper/paper.tex
made the font size 10 in all tables in main document and appendix

@@
   pdfauthor={Kevin E. See1,, Michael W. Ackerman1, Richard A. Carmichael1, Sarah L. Hoffmann1, and Chris Beasley1},
-  pdfkeywords={carrying capacity; quantile regression; random forests; Chinook; quantile random forest},
+  pdfkeywords={carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests},
   hidelinks,
@@
 \author{Kevin E. See\textsuperscript{1,*}, Michael W. Ackerman\textsuperscript{1}, Richard A. Carmichael\textsuperscript{1}, Sarah L. Hoffmann\textsuperscript{1}, and Chris Beasley\textsuperscript{1}}
-\date{December 01, 2020}
+\date{December 02, 2020}
 
@@
 
-Most studies that have investigated fish-habitat relationships focus on predicting a species' distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity.
+Most studies that have investigated fish-habitat relationships focus on predicting a species' distribution (presence/absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity.
 Further, many of these studies focus on only one or two measures of habitat.
@@
 \centering
+\fontsize{10}{12}\selectfont
 \begin{tabular}[t]{lrl}
@@
 \centering
+\fontsize{10}{12}\selectfont
 \begin{tabular}[t]{rll>{\raggedright\arraybackslash}p{3in}}
@@
 \centering
+\fontsize{10}{12}\selectfont
 \begin{tabular}[t]{lllrr}
@@
 \centering
+\fontsize{10}{12}\selectfont
 \begin{tabular}[t]{llrr}
@@
 \centering
+\fontsize{10}{12}\selectfont
 \begin{tabular}[t]{>{\raggedright\arraybackslash}p{1in}rll>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}}

### 2020-12-04 analysis/paper/appendix/appendix_s1.Rmd
Implemented revisions suggested by Mike, Sarah and Richie before final submission

@@
 ---
-title: Appendix S1 - _Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models_
-subtitle: "Ecospheres"
+title: Appendix S1 - _Estimating carrying capacity for juvenile salmon using quantile random forest models_
+subtitle: "Ecosphere"
 author:
@@
 
-To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site.
+To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol [@CHaMP2016] describing the quantity and quality of fish habitat for each survey site.
 Habitat metrics were first grouped into broad categories that included channel unit configuration, complexity, fish cover, riparian areas, side channels, stream size, substrate, temperature, water quality, and woody debris.

### 2020-12-04 analysis/paper/appendix/appendix_s1.tex
updated the author formatting in the appendix so they don't overrun a single line
Implemented revisions suggested by Mike, Sarah and Richie before final submission

@@
 \hypersetup{
-  pdftitle={Appendix S1 - Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models},
+  pdftitle={Appendix S1 - Estimating carrying capacity for juvenile salmon using quantile random forest models},
   pdfauthor={Kevin E. See1,, Michael W. Ackerman1, Richard A. Carmichael1, Sarah L. Hoffmann1, and Chris Beasley1},
@@
 
-\title{Appendix S1 - \emph{Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models}}
+\title{Appendix S1 - \emph{Estimating carrying capacity for juvenile salmon using quantile random forest models}}
 \usepackage{etoolbox}
@@
 \makeatother
-\subtitle{Ecospheres}
-\author{Kevin E. See\textsuperscript{1,*}, Michael W. Ackerman\textsuperscript{1}, Richard A. Carmichael\textsuperscript{1}, Sarah L. Hoffmann\textsuperscript{1}, and Chris Beasley\textsuperscript{1}}
-\date{December 02, 2020}
+\subtitle{Ecosphere}
+\author{Kevin E. See\textsuperscript{1,*}, 
+    \and Michael W. Ackerman\textsuperscript{1}, 
+    \and Richard A. Carmichael\textsuperscript{1},
+    \and Sarah L. Hoffmann\textsuperscript{1}, 
+    \and Chris Beasley\textsuperscript{1}}
+\date{December 03, 2020}
 
@@
 
-To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site.
+To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol (CHaMP 2016) describing the quantity and quality of fish habitat for each survey site.
 Habitat metrics were first grouped into broad categories that included channel unit configuration, complexity, fish cover, riparian areas, side channels, stream size, substrate, temperature, water quality, and woody debris.
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-CHaMP2016}{}%
+CHaMP. 2016. Scientific protocol for salmonid habitat surveys within the Columbia habitat monitoring program. Bonneville Power Administration Portland, Oregon, USA.
+
 \leavevmode\hypertarget{ref-R-minerva}{}%

### 2020-12-04 analysis/paper/paper.Rmd
Implemented revisions suggested by Mike, Sarah and Richie before final submission

@@
 ---
-title: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
+title: Estimating carrying capacity for juvenile salmon using quantile random forest models
 author:
@@
 abstract: |
-  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
+
+  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids (*Oncorhynchus* spp.).
 A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds.
 Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available.
-In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale.
-QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets.
-We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites.
+In this study, we developed a quantile random forest model to  provide estimates of habitat carrying capacity for Chinook salmon (*O.
+tshawytscha*) parr during the summer months, at both the site and watershed scale.
+Quantile random forest models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets.
+We leveraged Columbia Habitat Monitoring Program data to select habitat co-variates and predict capacity at those sites.
 We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin.
 Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models.
-Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
+Carrying capacity estimates based on quantile random forest models, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
+  **Key words:** carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
 keywords: |
@@
 
+# modify name of one watershed
+hab_data %<>%
+  mutate(Watershed = recode(Watershed,
+                            "Yankee Fork" = "Yankee Fork Salmon"))
+
 # change some of the names of metrics for plotting purposes
@@
   mutate(Population = recode(Population,
-                             'Chiwawa R.' = 'Chiwawa River'))
+                             'Chiwawa R.' = 'Chiwawa River',
+                             "Upper Lemhi" = "Upper Lemhi River"))
 
@@
                'Tucannon River',
-               'Upper Lemhi',
+               'Upper Lemhi River',
                'Lostine River',
@@
 
-The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend.
+The decline of anadromous Pacific salmon (*Oncorhynchus* spp.) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend.
 These actions are often categorized into four H’s: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation.
 Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality.
-Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2018; @Hinrichsen2020; @NOAA2020].
+Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2019; @Hinrichsen2020; @NOAA2020].
 Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O.
@@
 Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles.
-However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2018; @Hinrichsen2020; @NOAA2020].
+However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2019; @Hinrichsen2020; @NOAA2020].
 Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.
@@
 Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts.
-Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings.
+Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter), spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings.
 Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage.
@@
 
-## Study Site
+## Study site
 
@@
 
-## Habitat Covariate Selection
+## Habitat covariate selection
 
@@
 
-## QRF Model Fit
+## QRF model fit
 
@@
 
-## Extrapolating to Other Sites
+## Extrapolating to other Sites
 
@@
 We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.
-These points do not represent specific segments of streams, however, so we needed to do some spatial averaging of capacity predictions to generate larger scale capacity estimates.
+These points do not represent specific segments of streams, however, so some form of spatial averaging of capacity predictions to generate larger scale capacity estimates is necessary.
 
@@
 
-## Model Validation
+## Model validation
 
@@
 
-## QRF Model
+## QRF model
 
@@
 
-## Extrapolating to Other Sites
+## Extrapolating to other sites
 
@@
 
-## A Tool to Estimate Habitat Capacity
+## A tool to estimate habitat capacity
 
@@
 
-## Biological Expectations from QRF Model
+## Biological expectations from QRF model
 
@@
 
-## Extrapolation Model
+## Extrapolation model
 
@@
 
-## The Future: Improving Habitat Data
+## The future: improving habitat data
 
@@
 
-## Conclusions and Next Steps
+## Conclusions and next steps
 
@@
                 latex_options = c("hold_position"))
-
 ```
@@
                 latex_options = c("hold_position")) %>%
-  # kable_styling(latex_options = c('scale_down')) %>%
-  column_spec(1, width = "1in") %>%
+  column_spec(1, width = "1.5in") %>%
   column_spec(c(5:8), width = "0.5in")
@@
 Watersheds in black also contain paired fish data.
-Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
+Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork Salmon."}
 knitr::include_graphics('../figures/ChampWtds.png')

### 2020-12-04 analysis/paper/paper.tex
Implemented revisions suggested by Mike, Sarah and Richie before final submission

@@
 \hypersetup{
-  pdftitle={Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models},
+  pdftitle={Estimating carrying capacity for juvenile salmon using quantile random forest models},
   pdfauthor={Kevin E. See1,, Michael W. Ackerman1, Richard A. Carmichael1, Sarah L. Hoffmann1, and Chris Beasley1},
@@
 
-\title{Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models}
+\title{Estimating carrying capacity for juvenile salmon using quantile random forest models}
 \author{Kevin E. See\textsuperscript{1,*}, Michael W. Ackerman\textsuperscript{1}, Richard A. Carmichael\textsuperscript{1}, Sarah L. Hoffmann\textsuperscript{1}, and Chris Beasley\textsuperscript{1}}
-\date{December 02, 2020}
+\date{December 03, 2020}
 
@@
 \begin{abstract}
-Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids.
+Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids (\emph{Oncorhynchus} spp.).
 A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds.
 Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available.
-In this study, we developed a quantile random forest (QRF) model to provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale.
-QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets.
-We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites.
+In this study, we developed a quantile random forest model to provide estimates of habitat carrying capacity for Chinook salmon (\emph{O.
+tshawytscha}) parr during the summer months, at both the site and watershed scale.
+Quantile random forest models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets.
+We leveraged Columbia Habitat Monitoring Program data to select habitat co-variates and predict capacity at those sites.
 We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin.
 Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models.
-Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
+Carrying capacity estimates based on quantile random forest models, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
+\textbf{Key words:} carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
 \end{abstract}
@@
 
-The decline of anadromous Pacific salmonids (\emph{Oncorhynchus spp.}) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend.
+The decline of anadromous Pacific salmon (\emph{Oncorhynchus} spp.) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend.
 These actions are often categorized into four H's: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation.
@@
 Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids (Bond et al.
-2018, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020).
+2019, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020).
 Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (\emph{O.
@@
 However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge (Bond et al.
-2018, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020).
+2019, Hinrichsen and Paulsen 2020, NOAA Fisheries 2020).
 Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.
@@
 To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) (Meinshausen 2006) models using measurements of fish abundance and density and habitat characteristics.
-QRF models combine the theory and justification of quantile regression modeling (Koenker and Bassett Jr 1978, Cade and Noon 2003) with the flexibility and framework of random forest models (Breiman 2001).
+QRF models combine the theory and justification of quantile regression modeling (Koenker and Bassett Jr.
+1978, Cade and Noon 2003) with the flexibility and framework of random forest models (Breiman 2001).
 They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density.
@@
 \hypertarget{study-site}{%
-\subsection{Study Site}\label{study-site}}
+\subsection{Study site}\label{study-site}}
 
@@
 2017) and were downloaded from \url{https://www.champmonitoring.org}.
-Data from the following eleven CHaMP watersheds were used in this study: Asotin, Entiat, John Day, Lemhi, Methow, Minam, South Fork Salmon, Tucannon, Upper Grande Ronde, Wenatchee and Yankee Fork.
+Data from the following eleven CHaMP watersheds were used in this study: Asotin, Entiat, John Day, Lemhi, Methow, Minam, South Fork Salmon, Tucannon, Upper Grande Ronde, Wenatchee and Yankee Fork Salmon.
 Juvenile density and abundance data were collected in a subset of seven watersheds (see Table \ref{tab:fish-hab-sites} and Figure \ref{fig:wtsd-map}), at CHaMP survey reaches and were graciously provided by a number of agencies and projects, including the Integrated Status and Effectiveness Monitoring Project (Volk et al.
@@
 \hypertarget{habitat-covariate-selection}{%
-\subsection{Habitat Covariate Selection}\label{habitat-covariate-selection}}
+\subsection{Habitat covariate selection}\label{habitat-covariate-selection}}
 
@@
 \hypertarget{qrf-model-fit}{%
-\subsection{QRF Model Fit}\label{qrf-model-fit}}
+\subsection{QRF model fit}\label{qrf-model-fit}}
 
@@
 \hypertarget{extrapolating-to-other-sites}{%
-\subsection{Extrapolating to Other Sites}\label{extrapolating-to-other-sites}}
+\subsection{Extrapolating to other Sites}\label{extrapolating-to-other-sites}}
 
@@
 \hypertarget{model-validation}{%
-\subsection{Model Validation}\label{model-validation}}
+\subsection{Model validation}\label{model-validation}}
 
@@
 \hypertarget{habitat-covariate-selection-1}{%
-\subsection{Habitat Covariate Selection}\label{habitat-covariate-selection-1}}
+\subsection{Habitat covariate selection}\label{habitat-covariate-selection-1}}
 
@@
 \hypertarget{qrf-model}{%
-\subsection{QRF Model}\label{qrf-model}}
+\subsection{QRF model}\label{qrf-model}}
 
@@
 \hypertarget{extrapolating-to-other-sites-1}{%
-\subsection{Extrapolating to Other Sites}\label{extrapolating-to-other-sites-1}}
+\subsection{Extrapolating to other sites}\label{extrapolating-to-other-sites-1}}
 
@@
 \hypertarget{model-validation-1}{%
-\subsection{Model Validation}\label{model-validation-1}}
+\subsection{Model validation}\label{model-validation-1}}
 
@@
 \hypertarget{a-tool-to-estimate-habitat-capacity}{%
-\subsection{A Tool to Estimate Habitat Capacity}\label{a-tool-to-estimate-habitat-capacity}}
+\subsection{A tool to estimate habitat capacity}\label{a-tool-to-estimate-habitat-capacity}}
 
@@
 \hypertarget{biological-expectations-from-qrf-model}{%
-\subsection{Biological Expectations from QRF Model}\label{biological-expectations-from-qrf-model}}
+\subsection{Biological expectations from QRF model}\label{biological-expectations-from-qrf-model}}
 
@@
 \hypertarget{extrapolation-model}{%
-\subsection{Extrapolation Model}\label{extrapolation-model}}
+\subsection{Extrapolation model}\label{extrapolation-model}}
 
@@
 \hypertarget{the-future-improving-habitat-data}{%
-\subsection{The Future: Improving Habitat Data}\label{the-future-improving-habitat-data}}
+\subsection{The future: improving habitat data}\label{the-future-improving-habitat-data}}
 
@@
 \hypertarget{conclusions-and-next-steps}{%
-\subsection{Conclusions and Next Steps}\label{conclusions-and-next-steps}}
+\subsection{Conclusions and next steps}\label{conclusions-and-next-steps}}
 
@@
 \leavevmode\hypertarget{ref-Allen2000}{}%
-Allen, M. A. 2000. Seasonal microhabitat use by juvenile spring chinook salmon in the yakima river basin, washington. Rivers 7:314--332.
+Allen, M. A. 2000. Seasonal microhabitat use by juvenile spring Chinook salmon in the Yakima River basin, Washington. Rivers 7:314--332.
 
@@
 
-\leavevmode\hypertarget{ref-Bond2018}{}%
+\leavevmode\hypertarget{ref-Bond2019}{}%
 Bond, M.
@@
 Zabel.
-2018.
-Estimating the benefits of widespread floodplain reconnection for columbia river chinook salmon.
-Canadian Journal of Fisheries and Aquatic Sciences:1--15.
+2019.
+Estimating the benefits of widespread floodplain reconnection for Columbia River Chinook salmon.
+Canadian Journal of Fisheries and Aquatic Sciences 76:1212--1226.
 
@@
 2011.
-Relationships between habitat characteristics and breeding population densities in sockeye salmon (oncorhynchus nerka).
+Relationships between habitat characteristics and breeding population densities in sockeye salmon (\emph{Oncorhynchus nerka}).
 Canadian Journal of Fisheries and Aquatic Sciences 68:758--767.
@@
 \leavevmode\hypertarget{ref-Brett1952}{}%
-Brett, J. R. 1952. Temperature tolerance in young pacific salmon, genus oncorhynchus. Journal of the Fisheries Board of Canada 9:265--323.
+Brett, J. R. 1952. Temperature tolerance in young pacific salmon, genus \emph{oncorhynchus}. Journal of the Fisheries Board of Canada 9:265--323.
 
@@
 \leavevmode\hypertarget{ref-Carle1978}{}%
-Carle, F. L., and M. R. Strub. 1978. A new method for estimating population size from removal data. Biometrics:621--630.
+Carle, F. L., and M. R. Strub. 1978. A new method for estimating population size from removal data. Biometrics 34:621--630.
 
@@
 2020.
-Some like it slow: A bioenergetic evaluation of habitat quality for juvenile chinook salmon in the lemhi river, idaho.
-Canadian Journal of Fisheries and Aquatic Sciences.
+Some like it slow: A bioenergetic evaluation of habitat quality for juvenile Chinook salmon in the Lemhi River, Idaho.
+Canadian Journal of Fisheries and Aquatic Sciences 77:1221--1232.
 
@@
 \leavevmode\hypertarget{ref-CHaMP2016}{}%
-CHaMP. 2016. Scientific protocol for salmonid habitat surveys within the columbia habitat monitoring program. Bonneville Power Administration Portland, Oregon, USA.
+CHaMP. 2016. Scientific protocol for salmonid habitat surveys within the Columbia habitat monitoring program. Bonneville Power Administration Portland, Oregon, USA.
 
 \leavevmode\hypertarget{ref-Chapman1951}{}%
-Chapman, D. G. 1951. Some properties of the hypergeometric distribution with applications to zoological sample censuses. University of California Press.
+Chapman, D. G. 1951. Some properties of the hypergeometric distribution with applications to zoological sample censuses. University of California Press, Berkeley, California, USA.
 
@@
 2012.
-Ontogenetic habitat associations of a demersal fish species, pagrus auratus, identified using boosted regression trees.
+Ontogenetic habitat associations of a demersal fish species, \emph{Pagrus auratus}, identified using boosted regression trees.
 Marine Ecology Progress Series 462:219--230.
@@
 \leavevmode\hypertarget{ref-Cramer2009}{}%
-Cramer, S. P., and N. K. Ackerman. 2009. Linking stream carrying capacity for salmonids to habitat features. Pages 225--254 \emph{in} American fisheries society symposium.
+Cramer, S. P., and N. K. Ackerman. 2009. Linking stream carrying capacity for salmonids to habitat features. Pages 225--254 \emph{in} American Fisheries Society symposium.
 
@@
 Nonlinear relationships can lead to bias in biomass calculations and drift-foraging models when using summaries of invertebrate drift data.
-Environmental biology of fishes 99:659--670.
+Environmental Biology of Fishes 99:659--670.
 
@@
 2003.
-Estimating limits to the spatial extent and suitability of sole (solea solea) nursery grounds in the dover strait.
+Estimating limits to the spatial extent and suitability of sole (Solea solea) nursery grounds in the dover strait.
 Journal of Sea Research 50:151--165.
@@
 \leavevmode\hypertarget{ref-Fausch1988}{}%
-Fausch, K. D., C. L. Hawkes, and M. G. Parsons. 1988. Models that predict standing crop of stream fish from habitat variables: 1950-85. Notes.
+Fausch, K.
+D., C.
+L.
+Hawkes, and M.
+G.
+Parsons.
+1988.
+Models that predict standing crop of stream fish from habitat variables: 1950-85.
+Notes.
+General Technical Report, United States Department of Agriculture, Forest Service, Portland, Oregon, USA.
 
@@
 1987.
-Summer and winter habitat selection by juvenile chinook salmon in a highly sedimented idaho stream.
+Summer and winter habitat selection by juvenile Chinook salmon in a highly sedimented Idaho stream.
 Transactions of the American Fisheries Society 116:185--195.
@@
 2009.
-Juvenile chinook salmon summer microhabitat availability, use, and selection in a central idaho wilderness stream.
+Juvenile Chinook salmon summer microhabitat availability, use, and selection in a central Idaho wilderness stream.
 Transactions of the American Fisheries Society 138:633--644.
@@
 \leavevmode\hypertarget{ref-Jones2006}{}%
-Jones IV, G. P., L. G. Pearlstine, and H. F. Percival. 2006. An assessment of small unmanned aerial vehicles for wildlife research. Wildlife society bulletin 34:750--758.
+Jones IV, G. P., L. G. Pearlstine, and H. F. Percival. 2006. An assessment of small unmanned aerial vehicles for wildlife research. Wildlife Society Bulletin 34:750--758.
 
@@
 \leavevmode\hypertarget{ref-Koenker1978}{}%
-Koenker, R., and G. Bassett Jr. 1978. Regression quantiles. Econometrica: journal of the Econometric Society:33--50.
+Koenker, R., and G. Bassett Jr. 1978. Regression quantiles. Econometrica: Journal of the Econometric Society 46:33--50.
 
@@
 An overview of the Columbia Habitat Monitoring Program's (CHaMP) spatial-temporal design framework.
-South Fork Research.
+South Fork Research, North Bend, Washington, USA.
 
@@
 \leavevmode\hypertarget{ref-R-survey}{}%
-Lumley, T. 2020. Survey: Analysis of complex survey samples.
+Lumley, T. 2020. survey: Analysis of complex survey samples.
 
@@
 \leavevmode\hypertarget{ref-R-quantregForest}{}%
-Meinshausen, N. 2017. QuantregForest: Quantile regression forests.
+Meinshausen, N. 2017. quantregForest: Quantile regression forests.
 
@@
 2006.
-Using thalweg profiling to assess and monitor juvenile salmon (oncorhynchus spp.) habitat in small streams.
+Using thalweg profiling to assess and monitor juvenile salmon (\emph{Oncorhynchus} spp.) habitat in small streams.
 Canadian Journal of Fisheries and Aquatic Sciences 63:1515--1525.
@@
 Using inverse probability bootstrap sampling to eliminate sample induced bias in model based analysis of unequal probability samples.
-Plos one 10:e0131765.
+PLoS ONE 10:e0131765.
 
@@
 1992.
-Seasonal changes in habitat use by juvenile coho salmon (oncorhynchus kisutch) in oregon coastal streams.
+Seasonal changes in habitat use by juvenile coho salmon (\emph{Oncorhynchus kisutch}) in Oregon coastal streams.
 Canadian Journal of Fisheries and Aquatic Sciences 49:783--789.
@@
 \leavevmode\hypertarget{ref-NOAA2016}{}%
-NOAA Fisheries. 2016a. 2016 5-year review: Summary \& evaluation of Upper Columbia River steelhead and Upper Columbia River spring-run Chinook salmon. NOAA.
+NOAA Fisheries.
+2016a.
+2016 5-year review: Summary \& evaluation of Upper Columbia River steelhead and Upper Columbia River spring-run Chinook salmon.
+U.S.
+Department of Commerce, National Oceanic; Atmospheric Administration, National Marine Fisheries Service, West Coast Region, Portland, Oregon, USA.
 
 \leavevmode\hypertarget{ref-NOAA2016b}{}%
-NOAA Fisheries. 2016b. 2016 5-year review: Summary \& evaluation of Snake River Sockeye, Snake River spring-summer Chinook, Snake River fall-run Chinook and Snake River steelhead. NOAA.
+NOAA Fisheries.
+2016b.
+2016 5-year review: Summary \& evaluation of Snake River Sockeye, Snake River spring-summer Chinook, Snake River fall-run Chinook and Snake River steelhead.
+U.S.
+Department of Commerce, National Oceanic; Atmospheric Administration, National Marine Fisheries Service, West Coast Region, Portland, Oregon, USA.
 
@@
 Endangered Species Act Section 7(a)(2) Biological opinion and Magnuson-Stevens fishery conservation and management act essential fish habitat response for the continued operation and maintenance of the Columbia River system.
-NMFS Consultation, NOAA National Marine Fisheries Service.
+NMFS Consultation, U.S.
+Department of Commerce, National Oceanic; Atmospheric Administration, National Marine Fisheries Service, West Coast Region, Portland, Oregon, USA.
 
@@
 2002.
-Landscape characteristics, land use, and coho salmon (oncorhynchus kisutch) abundance, snohomish river, wash., usa.
+Landscape characteristics, land use, and coho salmon (\emph{Oncorhynchus kisutch}) abundance, Snohomish River, Washington, USA.
 Canadian Journal of Fisheries and Aquatic Sciences 59:613--623.
@@
 \leavevmode\hypertarget{ref-Raleigh1986}{}%
-Raleigh, R. F., W. J. Miller, and P. C. Nelson. 1986. Habitat suitability index models and instream flow suitability curves: Chinook salmon. National Ecology Center.
+Raleigh, R.
+F., W.
+J.
+Miller, and P.
+C.
+Nelson.
+1986.
+Habitat suitability index models and instream flow suitability curves: Chinook salmon.
+Biological Report, US Department of the Interior, Fish; Wildlife Service; National Ecology Center, Fort Collins, Colorado, USA.
 
@@
 \leavevmode\hypertarget{ref-Robson1964}{}%
-Robson, D., and H. Regier. 1964. Sample size in petersen mark--recapture experiments. Transactions of the American Fisheries Society 93:215--226.
+Robson, D., and H. Regier. 1964. Sample size in Petersen mark--recapture experiments. Transactions of the American Fisheries Society 93:215--226.
 
 \leavevmode\hypertarget{ref-Seber2002}{}%
-Seber, G. A. F. 2002. The estimation of animal abundance and related parameters. Blackburn Press Caldwell, New Jersey.
+Seber, G. A. F. 2002. The estimation of animal abundance and related parameters. Blackburn Press Caldwell, New Jersey, USA.
 
@@
 2001.
-Empirical relationships between watershed characteristics and coho salmon (oncorhynchus kisutch) smolt abundance in 14 western washington streams.
+Empirical relationships between watershed characteristics and coho salmon (\emph{Oncorhynchus kisutch}) smolt abundance in 14 western Washington streams.
 Canadian Journal of Fisheries and Aquatic Sciences 58:1453--1463.
@@
 \leavevmode\hypertarget{ref-R-missForest}{}%
-Stekhoven, D. J. 2013. MissForest: Nonparametric missing value imputation using random forest.
+Stekhoven, D. J. 2013. missForest: Nonparametric missing value imputation using random forest.
 
@@
 \leavevmode\hypertarget{ref-Stevens1999}{}%
-Stevens Jr, D. L., and A. R. Olsen. 1999. Spatially restricted surveys over time for aquatic resources. Journal of Agricultural, Biological, and Environmental Statistics:415--428.
+Stevens Jr, D. L., and A. R. Olsen. 1999. Spatially restricted surveys over time for aquatic resources. Journal of Agricultural, Biological, and Environmental Statistics 4:415--428.
 
@@
 \leavevmode\hypertarget{ref-Sweka2010}{}%
-Sweka, J. A., and G. Mackey. 2010. A functional relationship between watershed size and atlantic salmon parr density. Journal of Fish and Wildlife Management 1:3--10.
+Sweka, J. A., and G. Mackey. 2010. A functional relationship between watershed size and Atlantic salmon parr density. Journal of Fish and Wildlife Management 1:3--10.
 
@@
 2017.
-Integrated status and effectiveness monitoring program (ISEMP) and columbia habitat monitoring program (CHaMP) 2016 annual combined technical report.
-Bonneville Power Administration.
+Integrated status and effectiveness monitoring program (ISEMP) and Columbia habitat monitoring program (CHaMP).
+Annual Combined Technical Report, Bonneville Power Administration, Portland, Oregon, USA.
 
@@
 \fontsize{10}{12}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1in}rll>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}}
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1.5in}rll>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}>{\raggedright\arraybackslash}p{0.5in}}
 \toprule
@@
 Upper Grande Ronde River & 8 & Spawners & RST & 171,607 (0.388) & 168,137 (0.298) & 127,052 (0.317) & 200,228 (0.23)\\
-Upper Lemhi & 7 & Spawners & RST & 333,229 (0.322) & 229,635 (0.212) & 242,637 (0.252) & 269,626 (0.217)\\
+Upper Lemhi River & 7 & Spawners & RST & 333,229 (0.322) & 229,635 (0.212) & 242,637 (0.252) & 269,626 (0.217)\\
 \bottomrule
@@
 Watersheds in black also contain paired fish data.
-Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork.}\label{fig:wtsd-map}
+Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork Salmon.}\label{fig:wtsd-map}
 \end{figure}

### 2020-12-04 analysis/paper/paper.Rmd
added "Freshwater Ecology" as subtitle, to indicate what section it's being submitted under
removed some white space

@@
 title: Estimating carrying capacity for juvenile salmon using quantile random forest models
+subtitle: Freshwater Ecology
 author:
@@
 abstract: |
-
   Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids (*Oncorhynchus* spp.).

### 2021-03-09 README.Rmd
added the DOI to the README
updated some references.
Added DOI info to the README page, now that the paper is published.

@@
 Authors <- "See, Kevin E., Michael W. Ackerman, Richard A. Carmichael, Sarah Hoffman, and Chris Beasley"
-yr <- 2020
+yr <- 2021
 Journal <- "Ecosphere"
@@
 
+[![DOI](https://zenodo.org/badge/225946308.svg)](https://zenodo.org/badge/latestdoi/225946308)
+
 [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/KevinSee/QRFpaper/master?urlpath=rstudio)
@@
 
-> `r Authors` (`r yr`). _`r Title`_. `r Journal` <https://doi.org/xxx/xxx>
-
-Our pre-print is online here:
-
-> `r Authors` (`r yr`). _`r Title`_. `r Journal`, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> `r Authors` (`r yr`). _`r Title`_. `r Journal` <http://doi.org/10.1002/ecs2.3404 >
 
@@
 
-> `r Authors`, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> `r Authors`, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/10.5281/zenodo.4300942>
 

### 2025-02-20 analysis/paper/paper.Rmd
updated function to create partial dependence plots. Added script to create an example of a subset of PDPs for a capacity review paper.

@@
 library(tidyverse)
-library(QRFpaper)
+# library(QRFpaper)
+devtools::load_all()
 library(janitor)

### 2025-02-20 R/plot_partial_dependence.R

@@ -52,7 +52,7 @@ plot_partial_dependence = function(rf_mod,
 
   # get means and ranges of all
   covar_range = data %>%
-    select(one_of(as.character(plot_covars))) %>%
+    select(one_of(as.character(covars))) %>%
     gather(Metric, value) %>%
     group_by(Metric) %>%
     summarise_at(vars(value),
@@ -65,6 +65,7 @@ plot_partial_dependence = function(rf_mod,
 
   # create data.frame with sequence of values across each covariate, keeping other covariates at their average value
   pdp_df = covar_range %>%
+    filter(Metric %in% plot_covars) %>%
     split(.$Metric) %>%
     map_df(.id = 'Metric',
            .f = function(x) {
@@ -124,7 +125,8 @@ plot_partial_dependence = function(rf_mod,
   # get covariate labels
   pdp_df = pdp_df %>%
     left_join(data_dict %>%
-                select(Metric = ShortName, covar_label = Name)) %>%
+                select(Metric = ShortName, covar_label = Name) %>%
+                distinct()) %>%
     # put covariates in order by relative importance
     left_join(rel_imp %>%
                 select(Metric, relImp)) %>%

### 2025-02-20 analysis/R_scripts/extra_pdp_figure.R

@@ -1,0 +1,68 @@
+# load packages
+library(tidyverse)
+library(here)
+library(ggpubr)
+devtools::load_all()
+
+# set theme for plotting
+theme_set(theme_pubr(base_family = "Helvetica"))
+
+# load some data
+qrf_data = read_rds(here("analysis/data/derived_data/qrf_data.rds"))
+dens_offset = read_rds(here("analysis/data/derived_data/dens_offset.rds"))
+sel_hab_mets = read_rds(here("analysis/data/derived_data/sel_hab_mets.rds"))
+qrf_mod = read_rds(here("analysis/data/derived_data/qrf_mod.rds"))
+hab_dict = read_rds(here("analysis/data/derived_data/hab_dict.rds"))
+
+# change some of the names of metrics for plotting purposes
+sel_hab_mets <-
+  sel_hab_mets %>%
+  left_join(hab_dict %>%
+              select(Metric = ShortName,
+                     Name) %>%
+              distinct()) %>%
+  mutate(Name = recode(Name,
+                       'Wetted Width To Depth Ratio Avg' = 'Width:Depth Ratio',
+                       'Wetted Depth SD' = 'Depth Complexity',
+                       'Wetted Channel Braidedness' = 'Braidedness',
+                       'Fish Cover: Total' = 'Fish Cover',
+                       'Riparian Cover: Some Canopy' = 'Riparian Canopy',
+                       'Wetted Width Integrated' = 'Wetted Width',
+                       'Substrate < 6mm' = 'Fines',
+                       # 'Substrate: D16',
+                       'Channel Unit Frequency' = 'Channel Unit\nFrequency',
+                       'Avg. August Temperature' = 'Avg. August Temp.',
+                       'Large Wood Volume: Wetted Slow Water' = 'Large Wood\nFreq. in Pools'))
+
+
+# which habitat metrics should be plotted?
+plot_metrics <-
+  c("avg_aug_temp",
+    "FishCovTotal",
+    "CU_Freq",
+    "RipCovCanSome")
+
+# plot partial dependence plots
+plot_partial_dependence(rf_mod = qrf_mod,
+                        data = qrf_data,
+                        plot_covars = plot_metrics,
+                        data_dict = hab_dict %>%
+                          select(-Name) %>%
+                          inner_join(sel_hab_mets %>%
+                                       select(ShortName = Metric,
+                                              Name)) |>
+                          filter(ShortName %in% plot_metrics),
[…]
